---
title: "Project Title (Placeholder)"
author: "Camden Bergquist"
date: "Last compiled on `r format(Sys.time(), '%B %d, %Y')`"
number-sections: true
format:
  jasa-pdf:
    toc: true
    lof: true
    lot: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{url}
        \usepackage{xurl}
        \usepackage{booktabs}
        \usepackage{tabularx}
        \usepackage{xltabular}
        \usepackage{array}
        \usepackage{float}
        \keepXColumns
        \usepackage{ragged2e}
        \newcolumntype{T}[1]{>{\RaggedRight\ttfamily\arraybackslash}p{#1}}
        \newcommand{\us}{\_\allowbreak}
bibliography: references.bib
execute:
  echo: FALSE
  warning: FALSE
  message: FALSE
---


<!--
Note: Find way to center-align figures in YAML header
-->

```{r Libraries, output = FALSE}
# Need I explain these?
library(tidyverse) 
library(scales)

library(arrow) # For reading parquet files
library(DBI) # To open DuckDB connection
library(duckdb) # To manipulate data on disk instead of in environment, saving on RAM
library(ggthemes) # Provides ready-made ggplot themes

```

```{r Setup, output = FALSE}
# Data Paths
base_dir <- "../data/derived/full_datasets"
paths <- list(
  pm = file.path(base_dir, "player_month.parquet"),
  pm_tc = file.path(base_dir, "player_month_tc.parquet"),
  pom_tc = file.path(base_dir, "player_opening_month_tc.parquet"),
  ptm_tc = file.path(base_dir, "player_time_month_tc.parquet"),
  mom_tc = file.path(base_dir, "meta_opening_month_tc.parquet")
)

# Open database connection
con <- dbConnect(duckdb(), dbdir=":memory:")

# Helper for safe `read_parquet()` path quoting
qpath <- function(p) paste0("'", gsub("'", "''", normalizePath(p, winslash="/", mustWork=FALSE)), "'")

```



Make sure every code chunk has echo=FALSE in its preamble unless you need to highlight the code for some reason in the paper.

## Abstract
Complete this after all other sections have been completed.

\newpage

## Introduction

Competitive games have become a major source of large-scale behavioral data, enabling researchers to study strategic decision-making, skill development, and population-level trends using observational logs at scales unheard of compared to controlled studies [@edelmann2020css; @watts2014css]. Chess is one of the most famous competitive games in the world, and the relatively recent growth of online platforms has created datasets with millions to billions of games played under standardized rules and well-defined competitive contexts (time controls, ratings, game outcomes, etc.) [@watts2014css; @lichessDatabase]. One of the most widely used sources is the Lichess Open Database, which provides public monthly exports of rated games played on lichess.org in PGN format [@lichessDatabase; @lichessDatabaseGitHub]. These exports make it possible to study how strategy and performance vary across skill levels and time controls in a modern, online environment.

This project intends to use Lichess game data to investigate contemporary metagame behavior and player development. Here, 'metagame' refers to the population-level distribution of strategic choices and styles observed in practice – such as opening families, risk preferences, and game outcomes – within a defined time window and player population. In addition to describing how strategic behavior differs by rating and time control, the project will attempt to connect these strategic signatures to measures of player progression over time. Because chess is a repeatedly played, fully competitive setting with a mature rating infrastructure, it offers a natural testbed for quantifying skill and comparing behavior across different players [@lichessRatingSystems; @glickmanGlicko2Desc].

A central component of modern chess analytics is the rating system used to summarize player strength. The ELO rating system – a system originally developed for chess – models player performance probabilistically, and remains a foundational reference point for interpreting skill differences and expected outcomes in competitive settings [@elo1978rating; @fideElo2025]. Lichess uses a modified version of this system called 'Glicko-2', which, in addition to rating, tracks uncertainty via rating deviation and volatility [@lichessRatingSystems; @glickmanGlicko2Desc]. Ratings are especially useful for observational studies because they provide a consistent scale by which to stratify players and construct longitudinal measures of improvement [@glickmanGlicko2Desc].

Overall, this project attempts to marry data science, computational social science, and strategic behavior: It treats online chess as a real-world environment where strategic choices and performance outcomes can be measured at scale and subsequently analyzed using statistical learning methods [@edelmann2020css; @watts2014css].

### Background {#sec-background}

A growing research literature uses large online chess datasets to study human performance, decision-making, and strategic patterns. For example, Chowdhary et al. leverage an online chess dataset at the scale of over 100 million games to quantify performance dynamics and career-like trajectories, illustrating how large game archives support population-level behavioral measurement across skill levels [@chowdhary2023performance].

One important line of work focuses on time usage and decision quality under time constraints. Online chess platforms record both move sequences and clock information, which allows researchers to study how players allocate thinking time and how time pressure shapes decisions. Russek et al. analyze millions of Lichess games to relate move times to the estimated value of additional computation, and show that time allocation patterns vary systematically with playing strength and time control [@russek2025thinking]. These findings support the idea that online chess is a rich setting for studying bounded rationality and performance under constraints, especially when comparing bullet, blitz, rapid, and classical formats [@russek2025thinking].

Chess opening choice is also an active research topic. Because openings are discrete, labelable strategic choices that occur early in every game, they are well suited to population-level analysis. De Marzo and Servedio use online chess community data to quantify opening similarity and complexity, including representing openings in network form and studying how opening-related structure can be used to forecast future opening adoption [@demarzo2023openings]. This motivates analyzing metagame development in terms of shifting opening popularity and success rates across time, player skill, and time controls [@demarzo2023openings; @chowdhary2023performance].

Finally, there is an emerging interest in identifying player styles or archetypes from game logs. While traditional chess analysis often focuses on engine evaluation of positions, a complementary approach is to extract behavioral features from move sequences and outcomes (e.g., game length, opening selection, and other style proxies) and use statistical learning to characterize player types. For instance, McIlroy-Young et al. demonstrate that models trained on move sequences can capture individual-specific signatures (“style”) in chess decisions, providing evidence that player behavior is meaningfully distinguishable from game records alone [@mcilroyyoung2021stylometry]. This creates a bridge between strategic analysis (what players choose to do) and longitudinal outcomes (how players improve or persist over time), which aligns closely with the goals of this project.

Taken together, existing work suggests that large-scale online chess datasets support three relevant kinds of analysis: (1) descriptive metagame analysis of strategic choices across populations, (2) modeling how constraints like time pressure affect behavior and outcomes, and (3) data-driven characterization of player types and their evolution [@chowdhary2023performance; @russek2025thinking; @demarzo2023openings; @mcilroyyoung2021stylometry]. This project intends to build on those themes by using Lichess data to analyze modern opening/style patterns and connect them to measures of player performance and development over time.


\newpage

## The Data {#sec-data}

### Source {#sec-source}
Describe where this data comes from, why it was collected, who collected it, and how it was collected. A Citation may be useful here.

Lichess is the largest open-source chess server is the world, and releases free, monthly exports of their game data under the Creative Commons CCO license [@lichessDatabase]. The site requires a log of the game data in order to function while a game is being played, but holds onto it after the fact to allow users to analyze their games after playing them. Though Lichess offers a powerful API, I elected to download data directly from their website in `.pgn.zst` format. The PGN (portable game notation) format is a plaintext format for chess games, containing moves, headers like players, date, and event, and even optional annotation. The `.zst` file extension (sometimes also written as `zstd`) represents Zstandard compression, which is a modern compressor similar to gzip or xz that's commonly used because it's fast and compresses well; think of it as a variation on a ZIP or RAR file. Zstandard compression can be expected to shrink PGN files by 5 times on its faster, less-efficient settings, and even more on it's slower ones.

For the purposes of this project, I downloaded 24 months of data spanning from January 2024 to December 2025, amounting to 677 gigabytes of raw data when compressed. Obviously, nearly 700 gigabytes of compressed data – 3~4 terabytes of uncompressed PGN data – is an insane amount to try to work with, capstone project or not. I have a powerful home computer, but that kind of scale requires something closer to a high-end server than a personal computer; it won't fly. Instead of tackling the problem head-on, I constructed a hybrid R/Python pipeline that downloads the data, deterministically samples and parses it into small chunks (in the form of parquet files), and then builds a selection of smaller, ready-for-analysis datasets from those chunks. Individually, no one dataset has the 20+ variables required in the project outline, but since I've built multiple datasets for multiple different types of analysis, they collectively meet that benchmark 

The download and parsing section of the pipeline handles the following:

- Reads a config file `pipeline.yml` that controls the scope and parameters of the pipeline's operation.
- Downloads monthly Lichess 'standard rated' PGN archives for each month specified in the YAML.
- Decompresses each Zstandard file as a stream, scanning games one-by-one so it doesn't run into RAM bottlenecks.
- Parses game headers first to decide whether a game is relevant before doing any heavy work (not all games are actually getting sampled, so it skips over the ones not being used).
  - Extracts core metadata (players, ratings, time control, result, termination, opening, timestamps).
  - Buckets each game into blitz/rapid/classical using the Lichess time-control convention.
- Applies filtering rules from the YAML, which in this case are:
  - Only include blitz/rapid/classical time controls.
  - Only include games from players in the 800–2600 rating range.
  - Drops unwanted terminations (aborted games).
- Uses deterministic hash-based sampling to create reproducible subsets without needing to store the whole population:
  - A meta sample keyed by `game_id` for population/metagame summaries, at a rate of 5/500 games.
  - A player cohort keyed by `username` for longitudinal/player tracking at a rate of 3/500 players.
  - A moves sample keyed by `game_id` for clock/time-per-move data at a rate of 5/10000 moves.
- Deterministically anonymizes usernames based on a stable, salted hash.
  - Cohort selection is computed on the real username before anonymization, so it remains reproducible.
- Writes month-partitioned parquet outputs for game-level data.

The dataset-building portion of the pipeline takes those outputs and, using DuckDB to read and write parquet files on the disk, builds seven datasets from them:

- Player-level longitudinal summary tables from `games_cohort`:
  - `player_month` (`pm`): player * month, 249,761 observations.
  - `player_month_tc` (`pm_tc`): player * month * time control, 320,869 observations.
- Player-opening 'repertoire' table from `games_cohort`:
  - `player_opening_month_tc` (`pom_tc`): player * month * time control * opening, 7,167,580 observations.
- Metagame tables from `games_meta`:
  - `meta_month_tc` (`mm_tc`): month * time control, 72 observations.
  - `meta_opening_month_tc` (`mom_tc`): month * time control * opening, 130,672 observations.
- Time-usage tables from `moves_meta`:
  - `moves_month_tc` (`mvm_tc`) month * time control, 72 observations.
  - `player_time_month_tc` (`ptm_tc`): player * month * time control, 1,038,050 observations.

The pipeline logs session and version info for both R and Python and records checksums for downloaded raw files to help with reproducibility.

### Variables
This project is working with multiple related-yet-distinct datasets. As such, it's very difficult to properly introduce the data without an according quantity of tables, which results in several pages worth of figures. My personal recommendation is to read through table 1, featured below, before moving straight to @sec-observations.^[I'm considering consolidating the Variables and Observations sections into one, as I feel they're somewhat redundant, or else perhaps moving the Variable tables entirely into the appendix. I need to consult you for permission on that first, though.] Most variable names are self-explanatory, so this section should largely be treated as a reference to skim through if a variable introduced later is confusing or unclear. Additionally, I elected to exclude tables `mm_tc` and `mvm_tc` from this section, as they're summary-level datasets that won't be explicitly modeled. Please refer to the <appendix placeholder; update when written> for information on them.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.8cm} X}
\caption{Core identifiers and categorical variables used across derived datasets.}\label{tab:variables-core}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & categorical & Deterministic pseudonymous user ID (e.g., \texttt{u\us ...}). Appears in player-level tables (\texttt{pm}, \texttt{pm\us tc}, \texttt{pom\us tc}, \texttt{ptm\us tc}). \\
\texttt{month} & categorical & Month of play as \texttt{YYYY-MM}. Primary time index for metagame and longitudinal analysis. \\
\texttt{tc\us bucket} & categorical & Time control bucket: \texttt{blitz}, \texttt{rapid}, or \texttt{classical}. Used to stratify behavior by constraint level. \\
\texttt{eco} & categorical & ECO opening code (e.g., \texttt{C41}). Standardized opening taxonomy useful for grouping openings into families. \\
\texttt{opening} & categorical & Human-readable opening family name (e.g., \texttt{Philidor Defense}). Used in repertoire and metagame summaries. \\
\end{xltabular}




```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player-month outcome and covariate variables (tables: \texttt{pm}, \texttt{pm\us tc}).}\label{tab:variables-player-month}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of games played in the month (and time-control bucket for \texttt{pm\us tc}). Also a reliability/weighting signal. \\
\texttt{wins} & numeric (count) & Count of wins in the period. \\
\texttt{losses} & numeric (count) & Count of losses in the period. \\
\texttt{draws} & numeric (count) & Count of draws in the period. \\
\texttt{win\us rate} & numeric (proportion) & Key performance metric: wins divided by games. Natural response variable for performance models. \\
\texttt{draw\us rate} & numeric (proportion) & Draws divided by games. Useful because draw propensity varies by skill, time control, and strategic choices. \\
\texttt{avg\us rating} & numeric & Primary skill proxy: average of the player’s pre-game rating across games in the period. Used for stratification and longitudinal change. \\
\texttt{avg\us opp\us rating} & numeric & Average opponent rating. Contextualizes performance and approximates strength of schedule. \\
\texttt{avg\us rating\us diff} & numeric & Average (player rating minus opponent rating). Summarizes typical matchup difficulty. \\
\texttt{avg\us moves} & numeric & Average game length (moves). Proxy for style/complexity. \\
\texttt{opening\us diversity} & numeric (count) & Number of distinct openings used in the period. Proxy for repertoire breadth. \\
\texttt{top\us opening} & categorical & Most frequently played opening during the period (ties broken deterministically). \\
\end{xltabular}



```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player-opening table variables (table: \texttt{pom\us tc}).}\label{tab:variables-player-opening}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of games the player played with the given opening in the month/time-control bucket. Used as a frequency weight. \\
\texttt{win\us rate} & numeric (proportion) & Opening-specific win rate for the player and period. Used to study effectiveness conditional on opening. \\
\texttt{draw\us rate} & numeric (proportion) & Opening-specific draw rate. Helps characterize risk profile and style. \\
\texttt{avg\us rating} & numeric & Average player rating in games where the opening was played. Enables comparison of opening choice across skill. \\
\texttt{avg\us rating\us diff} & numeric & Average rating difference in games where the opening was played. Captures whether openings are used in tougher/easier matchups. \\
\end{xltabular}




```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player time-use table variables (table: \texttt{ptm\us tc}).}\label{tab:variables-player-time}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{move\us rows} & numeric (count) & Number of move observations contributing to the summary (after sampling and filtering). Reliability signal. \\
\texttt{games} & numeric (count) & Number of distinct games contributing move records for the player-month-time-control group. \\
\texttt{avg\us time\us spent} & numeric (seconds) & Mean time spent per move. Captures typical deliberation time. \\
\texttt{median\us time\us spent} & numeric (seconds) & Median time spent per move; robust to outliers and occasional long thinks. \\
\texttt{p90\us time\us spent} & numeric (seconds) & 90th percentile of time spent per move. Measures tendency for occasional deep computation. \\
\texttt{avg\us clock\us after} & numeric (seconds) & Mean remaining clock time after moves. Proxy for time management and pacing. \\
\texttt{low\us time\us rate} & numeric (proportion) & Fraction of moves under a low-time threshold. Proxy for time trouble frequency. \\
\texttt{avg\us rating} & numeric & Average player rating across move records in the group; allows stratification by skill. \\
\texttt{avg\us rating\us diff} & numeric & Average rating difference across the move sample; controls for opponent strength effects. \\
\texttt{avg\us time\us opening} & numeric (seconds) & Average time spent per move during the opening phase (if phase segmentation is available; game phases are defined by chess-engine analysis). \\
\texttt{avg\us time\us middlegame} & numeric (seconds) & Average time spent per move during middlegame phase (may be missing for very short games). \\
\texttt{avg\us time\us endgame} & numeric (seconds) & Average time spent per move during endgame phase (often missing because many games never reach endgame). \\
\end{xltabular}



```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Metagame opening table variables (table: \texttt{mom\us tc}).}\label{tab:variables-metagame-opening}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of sampled games for the opening within month and time control. Precision/support indicator. \\
\texttt{share} & numeric (proportion) & Core metagame measure: fraction of games using the opening within a month/time-control bucket. Tracks adoption shifts. \\
\texttt{white\us win\us rate} & numeric (proportion) & Fraction of games where White wins. Used for opening-level success patterns and color asymmetries. \\
\texttt{black\us win\us rate} & numeric (proportion) & Fraction of games where Black wins. Complements White win rate. \\
\texttt{draw\us rate} & numeric (proportion) & Fraction of games drawn. Draw propensity varies by opening and time control. \\
\texttt{upset\us rate} & numeric (proportion) & Fraction of games where the lower-rated player wins (pipeline definition). Proxy for volatility/competitiveness. \\
\texttt{avg\us abs\us rating\us diff} & numeric & Mean absolute rating gap between players in the opening/month/tc group. Context for outcome rates. \\
\texttt{avg\us moves} & numeric & Mean game length for the opening/month/tc group. Proxy for complexity and style at population level. \\
\end{xltabular}


```

### Observations {#sec-observations}
Each derived dataset is organized at a different unit of analysis (e.g., player-by-month or opening-by-month). In all cases, the raw unit of data is a rated chess game from the Lichess public database, but the pipeline aggregates games (and, for time-use, moves) into analysis-ready summaries. Because many variables overlap across tables, the observation structure can be summarized with a small set of representative row types. The examples below demonstrate the core units: (1) player-period summaries (with optional time-use extensions), (2) player opening-level summaries, and (3) population-level metagame summaries for openings.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observation for player-period outcome summaries (tables: \texttt{pm}, \texttt{pm\us tc}).}\label{tab:obs-player-period-outcomes}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & \texttt{u\us 0001cf80083d0517} (stable pseudonymous identifier) \\
\texttt{month} & \texttt{2024-12} (calendar month) \\
\texttt{tc\us bucket} & \texttt{rapid} (present in \texttt{pm\us tc}; omitted in \texttt{pm}) \\
\texttt{games} & 5 (total games in this player-month-(tc) group) \\
\texttt{wins} & 4 \\
\texttt{losses} & 0 \\
\texttt{draws} & 1 \\
\texttt{win\us rate} & 0.80 (wins / games) \\
\texttt{draw\us rate} & 0.20 (draws / games) \\
\texttt{avg\us rating} & 1544.0 (mean pre-game rating across games) \\
\texttt{avg\us opp\us rating} & 1503.2 (mean opponent rating across games) \\
\texttt{avg\us rating\us diff} & 40.8 (mean rating difference: player minus opponent) \\
\texttt{avg\us moves} & 40.4 (mean game length in moves) \\
\texttt{opening\us diversity} & 4 (distinct openings played in the group) \\
\texttt{top\us opening} & \texttt{Ponziani Opening: Jaenisch Counterattack} (most frequent opening) \\
\end{xltabular}


```


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observation for player-period time-use summaries (table: \texttt{ptm\us tc}).}\label{tab:obs-player-timeuse}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & \texttt{u\us 00004578077e9fa4} (stable pseudonymous identifier) \\
\texttt{month} & \texttt{2024-08} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{move\us rows} & 70 (number of move records contributing to this summary) \\
\texttt{games} & 1 (distinct games represented among the sampled moves) \\
\texttt{avg\us time\us spent} & 1.66 seconds (mean time spent per move) \\
\texttt{median\us time\us spent} & 1.0 seconds (median time spent per move) \\
\texttt{p90\us time\us spent} & 3.1 seconds (90th percentile time spent per move) \\
\texttt{avg\us clock\us after} & 122.5 seconds (mean remaining clock after moves) \\
\texttt{low\us time\us rate} & 0.00 (share of moves flagged as low-time) \\
\texttt{avg\us rating} & 1677.0 (mean pre-move rating across sampled moves) \\
\texttt{avg\us rating\us diff} & 11.0 (mean rating difference across sampled moves) \\
\texttt{avg\us time\us opening} & 0.9 seconds (mean time spent in opening phase) \\
\texttt{avg\us time\us middlegame} & 2.2 seconds (mean time spent in middlegame phase) \\
\texttt{avg\us time\us endgame} & 1.64 seconds (mean time spent in endgame phase; may be missing) \\
\end{xltabular}


```


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observations for opening-level summaries: player repertoire entries (\texttt{pom\us tc}) and population metagame cells (\texttt{mom\us tc}).}\label{tab:obs-opening-level}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\multicolumn{2}{l}{\textbf{Player opening summary (table: \texttt{pom\us tc})}} \\
\texttt{player} & \texttt{u\us 0001cf80083d0517} \\
\texttt{month} & \texttt{2024-11} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{eco} & \texttt{B13} \\
\texttt{opening} & \texttt{Caro-Kann Defense: Exchange Variation} \\
\texttt{games} & 4 (games where this player used the opening in this month/tc) \\
\texttt{win\us rate} & 0.50 (opening-conditional win rate for the player) \\
\texttt{draw\us rate} & 0.00 (opening-conditional draw rate for the player) \\
\texttt{avg\us rating} & 1965.3 (mean rating in games using this opening) \\
\texttt{avg\us rating\us diff} & 12.0 (mean rating difference in those games) \\

\midrule
\multicolumn{2}{l}{\textbf{Population opening summary (table: \texttt{mom\us tc})}} \\
\texttt{month} & \texttt{2024-01} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{eco} & \texttt{C41} \\
\texttt{opening} & \texttt{Philidor Defense} \\
\texttt{games} & 8{,}825 (sampled games using this opening in the month/tc group) \\
\texttt{share} & 0.0189 (opening share of the month/tc metagame) \\
\texttt{white\us win\us rate} & 0.512 (fraction of games White wins) \\
\texttt{black\us win\us rate} & 0.442 (fraction of games Black wins) \\
\texttt{draw\us rate} & 0.0466 (fraction drawn) \\
\texttt{upset\us rate} & 0.436 (fraction where lower-rated player wins) \\
\texttt{avg\us abs\us rating\us diff} & 50.2 (mean rating gap magnitude) \\
\texttt{avg\us moves} & 35.5 (mean game length) \\
\end{xltabular}


```

Together, these observation types support longitudinal player performance modeling (`pm`, `pm_tc`), repertoire/style measurement from opening choice (`pom_tc`), metagame dynamics via opening adoption and outcomes (`mom_tc`), and bounded-rationality/time-management analyses using move-time features (`ptm_tc`).


### Cleaning

A large amount of cleaning was performed during the pipeline's parsing steps. The filtering and deterministic sampling was already touched upon in #sec-source, but additional standardization and feature construction include:

- `TimeControl` tag parsed into:
  - `initial` (seconds).
  - `increment` (seconds).
  - `tc_bucket` (< 180 = bullet, < 600 = blitz, < 1800 = rapid, else classical),
- Converted PGN result strings into:
  - `winner` (white/black/draw).
  - `score_white` (1/0.5/0).
- Computed upset indicators using ratings:
  - `rating_diff` = `white_rating` - `black rating`.
  - `abs_rating_diff`
  - `is_upset`
- Game-length variables:
  - Computed `piles` and `moves` from move sequences (used to construct final datasets).

The information was then restructured and aggregated into a tidy format.

## Exploratory Data Analysis {#sec-eda}

@fig-coverage, the first and most important of the EDA plots, shows the number of observations the parser sampled per player-month, colored by time control.^[Throughout, 'rows' refer to aggregated observation units (e.g., player-month-time-control or opening-month-time-control records), whereas 'games' refer to the underlying sampled matches contributing to those aggregates. A change in row counts therefore reflects changes in the breadth of observed groups, not necessarily a change in total game volume.] It highlights a fundamental flaw in my parsing method— one that I failed to consider until seeing it visualized. See, while the parser filters the raw data down to a smaller subset of time controls and rating ranges, it does nothing to ensure that there's an equal distribution of games for given time controls or players of given ratings. Rather, it simply samples in a pseudo-random manner. This means that the distribution of certain variables in the cleaned dataset will reflect the true distribution of those variables among all games stored in the Lichess database. In more straightforward terms, popular time controls like blitz are played more often than less-popular ones like classical, and that relationship is mirrored in the data, as evidenced in @fig-coverage.

This finding isn't the end of the world, but it *is* extremely important to control for. It means that some 'missingness' will be structural rather than truly random, and that there will be a general imbalance across time controls, ratings, and months. It also justifies restricting certain analyses to blitz/rapid (or else collapsing them entirely), adding minimum-support thresholds to ensure there's enough data, and treating month as a potential confound only when coverage is stable enough; buckets with adequate support will be emphasized, and low-support buckets should be treated as exploratory only.


```{r Coverage Plot}
#| label: fig-coverage
#| fig-cap: "Proportion of observations per time control is skewed due to the sampling method used when parsing. Any estimate that compares time controls will have unequal precision, especially where classical is concerned. "
#| fig-align: center

cov_pm_tc <- dbGetQuery(con, sprintf("
  SELECT month, tc_bucket, COUNT(*) AS n_rows
  FROM read_parquet(%s)
  GROUP BY month, tc_bucket
  ORDER BY month, tc_bucket
", qpath(paths$pm_tc)))

ggplot(cov_pm_tc, aes(x = month, y = n_rows, group = tc_bucket, color = tc_bucket)) +
  geom_line(linewidth = 1) +
  coord_cartesian(clip = "off") +
  labs(x = "Month", y = "Observations (Player-Month-TC)", title = "Coverage of Player-Month-TC Table", color = "Time Control") +
  scale_color_few(labels = c("Blitz", "Classical", "Rapid")) +
  theme_few() + 
    theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")
```

@fig-reliability and @fig-panel-depth display two different kinds of unreliability inherent to player-month: the former being the number of games played in a given month, and the latter the number of months each player is present for in the data (total; not necessarily consecutive). For @fig-reliability, many player-months contain few games, making outcomes like win rate or average rating noisy at the unit level, justifying a minimum-games filter for player-month analyses, weighting player-month observations by games in aggregated models, and preferring aggregated measures such as bins over raw point clouds; game count should be treated as a precision/weighting signal.



```{r Reliability Plot}
#| label: fig-reliability
#| fig-cap: "Player-month observations have highly unequal reliability. The dataset has a long, right-skewed tail of heavy-activity players, which risk dominating aggregates if not uncontrolled for."
#| fig-align: center

gdist <- dbGetQuery(con, sprintf("
  SELECT games
  FROM read_parquet(%s)
  WHERE games <= 250
", qpath(paths$pm)))

ggplot(gdist, aes(x=games)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "steelblue") +
  coord_cartesian(xlim = c(0, 250)) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Games per Player in a Given Month", y = "Rows", title = "Player-Month Reliability Varies Widely") +
  theme_few()
```
For @fig-panel-depth, a majority of players only appear for a few months, which is to be expected for a hobby like chess. There's a notable spike at the 24-month column, but that makes sense— that column reflects long-time players who have likely played for much longer than two-odd years; if this data were being visualized without a two-year window, that spike would reflect a '24+' bin rather than players who have played for exactly 24 months. This attrition justifies the reporting of effective sample size for any next- or k-month analysis, using consecutive-month joins rather than simple lags on sparse panels, and potentially restricting to 'stable' players or months when clean identification is important.

```{r Panel Depth Plot}
#| label: fig-panel-depth
#| fig-cap: "The panel is unbalanced: Many players appear briefly; fewer are observed consistently. Lag-based outcomes are only defined for a subset of observations."
#| fig-align: center

panel_len <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, COUNT(DISTINCT month) AS n_months
    FROM read_parquet(%s)
    GROUP BY player
  )
  SELECT n_months, COUNT(*) AS n_players
  FROM x
  GROUP BY n_months
  ORDER BY n_months
", qpath(paths$pm)))

ggplot(panel_len, aes(x = n_months, y = n_players)) +
  geom_col(fill = "steelblue", color = "steelblue") +
  scale_y_continuous() +
  labs(x = "Months Observed per Player", y = "Players",
       title = "Panel Depth: How Long Players Remain Observable") +
  theme_few()
```

Pictured in @fig-delta-rating is without contest the single most perfect Gaussian distribution I have ever seen in a real-world dataset. This is not a parsing mistake, nor the result of some sort of transformation, but a natural consequence of the Glicko-2 rating system employed by Lichess. Both Glicko-2 and its predecessor Glicko assume that player skill – both for individuals *and* for a given population – follows a normal distribution. Because of that, implementations of Glicko systems almost always result in normally-distributed skill ratings, provided the player count is high enough.^[It's important to note that Glicko-based systems don't 'force' ratings to conform to a normal distribution; it's possible for them to fall into any form, but the way the system is designed tends to converge towards normality in most cases.] Practically, this means that $\Delta$ rating should be treated as a noisy outcome requiring careful aggregation, preferring binned summaries over unconstrained smoothing. It also encourages caution when interpreting statistically-significant-yet-tiny effects, as small, downright impractical or meaningless outcomes can still read as significant when dealing with a distribution like this.

```{r Rating Changes Plot}
#| label: fig-delta-rating
#| fig-cap: "Rating change is centered near zero with roughly symmetric dispersion. Most month-to-month changes are small, so effect sizes will likely be modest."
#| fig-align: center

chg <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT (avg_rating - prev_rating) AS d_rating
  FROM y
  WHERE prev_rating IS NOT NULL AND d_rating >= -300 AND d_rating <= 300
", qpath(paths$pm)))

ggplot(chg, aes(x=d_rating)) +
  geom_histogram(bins=80, fill = "steelblue", color = "steelblue") +
  labs(x = "Month-to-Month Change in Average Rating", y = "Count", 
       title = "Distribution of Within-Player Rating Change") +
  theme_few()
```
Unsurprisingly, @fig-volatility shows that the variance in players' month-to-moth ratings decreases with the number of games played in a given month. Importantly, the figure is log-transformed; though the plot visualizes a linear relationship, the first hundred games or so are comparatively more impactful than the next thousand. This is a relationship that I would expect to see with any rating system – law of large numbers and all – but was something I was expecting from Glicko-2 in particular. On top of rating, Glicko systems also measure uncertainty, which is a Bayesian-style measure of how confident it is in a player's rating. Because uncertainty always reduces as more games are played, it makes sense that this relationship is so clearly present in the data. This figure helps reinforce the need for minimum-games filters (which will reduce heteroskedastic noise) and shows that weighting by games is defensible as precision weighting. It also supports longer-horizon $\Delta$ (over multiple months instead of just one) as a way to increase robustness because it averages noise.

```{r Volatility Plot}
#| label: fig-volatility
#| fig-cap: "Observed rating movement behaves like a measurement-error problem; the number of games played is inversely proportional to rating-change volatility. This is expected behavior for a rating system."
#| fig-align: center

vol <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, games, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      games,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT
    games,
    ABS(avg_rating - prev_rating) AS abs_d_rating
  FROM y
  WHERE prev_rating IS NOT NULL
", qpath(paths$pm)))

ggplot(vol, aes(x = games, y = abs_d_rating)) +
  geom_point(alpha = 0.15) +
  scale_x_continuous(trans = "log10") +
  labs(x = "Games per Month (Log-Transformed)", y = "Absolute Change in Rating (|Δ|)",
       title = "Monthly Rating Volatility Shrinks as Game Volume Rises",
       subtitle = "Log-Transformed for Legibility") +
  theme_few()
```

@fig-low-time shows that low-time rate is massively zero-inflated, to the degree that it's difficult to identify a meaningful relationship when modeled continuously; binning is practically a necessity. Not shown directly in the figure is the fact that rapid and classical time controls don't have enough data to be meaningfully modeled on their own; due to how low-time rate is defined in the parser, a much smaller proportion of moves classify as being performed under time pressure.^[Low-time definition: A move is flagged as 'low time' if the remaining clock time after the move satisfies $\text{clock\_after} \le \max\!\left(10,\; 2 \times \text{increment}\right)$ seconds, where $\text{increment}$ is the per-move increment of the game's time control.] This justifies collapsing across time controls via move-weighted aggregations, near-exclusive use of binned metrics (including a 0% bin) instead of smoothers, and a combination of minimum move-row thresholds and n-per-bin reports for defensibility.

```{r LTR}
#| label: fig-low-time
#| fig-cap: "Low-time rate is zero-inflated and noisy. Time-control-specific curves are underpowered (with the exception of blitz), and pooling is needed for stable inference. Any relationshop between time trouble and improvement is likely both small and nonlinear, which is why loess smoothing wasn't applied to this figure."
#| fig-align: center

min_games_pm   <- 5     # require at least this many games in the month (across all TCs)
min_moves_ptm  <- 25    # require at least this many move rows in the month (across all TCs)

ltr_scatter_all <- dbGetQuery(con, sprintf("
  WITH pm_tc AS (
    SELECT
      player,
      month,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  pm AS (
    -- collapse across time controls using games-weighted avg_rating
    SELECT
      player,
      month,
      SUM(games) AS games_total,
      SUM(avg_rating * games) / SUM(games) AS avg_rating_all
    FROM pm_tc
    GROUP BY player, month
    HAVING SUM(games) >= %d
  ),
  pm_next AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      LEAD(avg_rating_all) OVER (PARTITION BY player ORDER BY month) AS next_rating_all
    FROM pm
  ),
  delta AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      (next_rating_all - avg_rating_all) AS d_rating_next
    FROM pm_next
    WHERE next_rating_all IS NOT NULL
  ),
  ptm_tc AS (
    SELECT
      player,
      month,
      low_time_rate,
      move_rows
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND low_time_rate IS NOT NULL
      AND move_rows IS NOT NULL
  ),
  ptm AS (
    -- collapse across time controls using move_rows-weighted low_time_rate
    SELECT
      player,
      month,
      SUM(move_rows) AS move_rows_total,
      SUM(low_time_rate * move_rows) / SUM(move_rows) AS low_time_rate_all
    FROM ptm_tc
    GROUP BY player, month
    HAVING SUM(move_rows) >= %d
  )
  SELECT
    d.player,
    d.month,
    d.avg_rating_all,
    d.games_total,
    d.d_rating_next,
    p.low_time_rate_all,
    p.move_rows_total
  FROM delta d
  JOIN ptm p USING (player, month)
", qpath(paths$pm_tc), min_games_pm,
   qpath(paths$ptm_tc), min_moves_ptm))

ggplot(ltr_scatter_all, aes(x = low_time_rate_all, y = d_rating_next)) +
  geom_point(alpha = 0.25) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  ylim(-225, 225) +
  labs(
    x = "Low-Time Rate",
    y = "Next-Month Δ Rating",
    title = "Time Trouble vs Next-Month Rating Change",
    subtitle = "Collapsed Across Time Controls"
  ) +
  theme_few()

```

Collectively, these diagnostics show that the dataset is (1) imbalanced across time controls (and ratings, though not visualized), (2) highly variable in within-unit support, and (3) noisy in short-horizon outcomes. Therefore, methods should adopt minimum-support thresholds, use precision weights, and rely on binned or aggregated summaries for stable inference.

### Opening Share

Opening trends over time are one of the main ways this project aims to study metagame development. The results, featured in @fig-opening and @fig-opening-rating, are rather disappointing. In particular, @fig-opening displays an amount of stability that I never would have expected from a dataset spanning as many months as this. It should be noted that similar figures for rapid and classical time controls show similar patterns, but with more variance. I suspect this difference in variance is a result of the fewer number of rows associated with each format, as classical stands out the most. Even if variance differs, however, the pattern remains regardless of time control— there's very little drift in terms of opening popularity, at least so far as the most popular choices are concerned. A better way of exploring this dynamic would have used select months over several years, but doing so would have required a completely different parsing pipeline, so this will have to suffice. 

```{r Opening Trends}
#| label: fig-opening
#| fig-cap: "Opening preference remains remarkably stable in blitz games over the project's two-year period."
#| fig-align: center
#| fig-height: 4

top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz' AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT 8
", qpath(paths$mom_tc)))

open_list <- paste0("('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')")

trend <- dbGetQuery(con, sprintf("
  WITH filtered AS (
    SELECT month, tc_bucket, opening, games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz'
      AND opening IN %s
      AND opening IS NOT NULL
  ),
  opening_level AS (
    SELECT
      month,
      tc_bucket,
      opening,
      SUM(games) AS games
    FROM filtered
    GROUP BY month, tc_bucket, opening
  )
  SELECT
    month,
    tc_bucket,
    opening,
    games,
    CAST(games AS DOUBLE) / SUM(games) OVER (PARTITION BY month, tc_bucket) AS share
  FROM opening_level
  ORDER BY month, opening
", qpath(paths$mom_tc), open_list))


ggplot(trend, aes(x=month, y=share, color=opening, group=opening)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels=percent_format(accuracy=0.1)) +
  labs(x="Month", y="Share of Games", color = "", 
  title="Opening Share Over Time (Blitz)", subtitle = "Top Eight Openings in Terms of Game Share") +
  scale_color_few() +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle=60, hjust=1)
  ) + 
  guides(color = guide_legend(nrow = 4, byrow = TRUE))
```

@fig-opening-rating, while still more stable than expected, is more interesting. This figure shows clear changes in preferences as player rating increases, particularly in the cases of the Pirc and Philidor defenses— the former becoming more popular and the latter less. Even where the more stable openings are concerned, there's a more pronounced trend present than anything in @fig-opening. Even if the relationship is minor, this is evidence enough to warrant attempts at modeling opening popularity as a function of player skill.

```{r Opening by Rating}
#| label: fig-opening-rating
#| fig-cap: "Opening trends by rating are more interesting. The Pirc and Philidor defenses show particularly interesting movement as player rating increases."
#| fig-align: center
#| fig-height: 4

# settings
tc_use <- "blitz"
bin_width <- 200
top_k <- 8             # number of openings to plot
min_bin_total <- 2000  # min player-games in a rating bin

# pick top openings within tc_bucket (weighted by games)
top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT %d
", qpath(paths$pom_tc), tc_use, top_k))

open_list <- paste0(
  "('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')"
)

# compute opening share within rating bins
trend_rating <- dbGetQuery(con, sprintf("
  WITH base AS (
    SELECT
      tc_bucket,
      opening,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IN %s
      AND opening IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  binned AS (
    SELECT
      opening,
      CAST(FLOOR(avg_rating / %d) * %d AS INTEGER) AS rating_bin_lo,
      SUM(games) AS player_games
    FROM base
    GROUP BY opening, rating_bin_lo
  ),
  totals AS (
    SELECT
      rating_bin_lo,
      SUM(player_games) AS bin_total
    FROM binned
    GROUP BY rating_bin_lo
  )
  SELECT
    b.rating_bin_lo,
    (b.rating_bin_lo + %d - 1) AS rating_bin_hi,
    b.opening,
    b.player_games,
    t.bin_total,
    CAST(b.player_games AS DOUBLE) / t.bin_total AS share
  FROM binned b
  JOIN totals t USING (rating_bin_lo)
  WHERE t.bin_total >= %d
  ORDER BY b.rating_bin_lo, b.opening
", qpath(paths$pom_tc), tc_use, open_list,
   bin_width, bin_width, bin_width, min_bin_total))


trend_rating <- trend_rating |>
  mutate(
    rating_bin = paste0(rating_bin_lo, "\u2013", rating_bin_hi),
    rating_bin = factor(rating_bin, levels = unique(rating_bin))
  )

ggplot(trend_rating, aes(x = rating_bin, y = share, color = opening, group = opening)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, alpha = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = "Rating",
    y = "Share of Player-Games",
    title = paste0("Opening Share by Rating (Blitz)"),
    color = ""
  ) +
  scale_color_few() +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle=60, hjust=1)
  ) + 
  guides(color = guide_legend(nrow = 4, byrow = TRUE))
```

### Time Usage

@fig-time and @fig-time-allocation are much more in line with what I was expecting than the opening-popularity figures. @fig-time shows a clear difference in thinking time, both between time controls and player ratings. Blitz and classical time controls are mirrors of one another: As players get better, they spend less time per move in blitz formats and more per move in classical formats. Rapid, in contrast, shows an interesting curve, where time usage dips for a little and then increases. Overall, @fig-time is enough on its own to justify time control as a factor when modeling with move time as an outcome.

```{r Time Use}
#| label: fig-time
#| fig-cap: "All three time controls have distinct time-usage patterns as player rating increases."
#| fig-align: center

time_bin <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    floor(avg_rating/100)*100 AS rating_bin,
    AVG(avg_time_spent) AS mean_time_spent,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 20000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

ggplot(time_bin, aes(x=rating_bin, y=mean_time_spent)) +
  geom_line() +
    facet_wrap(~factor(tc_bucket, levels = c("blitz", "rapid", "classical")), scales="free_y", labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))) +
  labs(x="Player Rating", y="Average Time per Move (Seconds)", title="Thinking Time Varies by Skill and Time Control") +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

@fig-time-allocation is in my opinion one of the most interesting. It breaks down the time-per-move from @fig-time into the amount of time allocated in opening, middlegame, and endgame phases.^[Game phases are defined by the Stockfish chess engine rather than a consistent standard.] Regardless of skill or time control, the consistent throughline is that players spend the most time thinking in the middlegame and the least in the opening, with the endgame in the middle. This is about what I'd expect. The opening has the fewest possible permutations, and is often studied and memorized, and while the end- and middlegames are similar, the former tends to be defined when fewer pieces remain on the board, which reduces complexity and thus thinking time. It sugggests factoring by gamestate may also be helpful when modeling move time as an outcome.

```{r Allocation Profile by Rating}
#| label: fig-time-allocation
#| fig-cap: "All three time controls have distinct time-usage patterns when game phases are additionally factored in."
#| fig-align: center

phase <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    FLOOR(avg_rating/200)*200 AS rating_bin,
    AVG(avg_time_opening) AS t_open,
    AVG(avg_time_middlegame) AS t_mid,
    AVG(avg_time_endgame) AS t_end,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 10000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

phase_long <- phase |>
  pivot_longer(cols=c(t_open, t_mid, t_end),
                      names_to="phase", values_to="sec")

ggplot(phase_long, aes(x=rating_bin, y=sec, color=phase, group=phase)) +
  geom_line() +
  facet_wrap(~factor(tc_bucket, levels = c("blitz", "rapid", "classical")), scales="free_y", labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))) +
  labs(x="Rating", y="Seconds per Move", color = "Phase",
       title="Phase Time Allocation by Rating") +
  scale_color_few(labels = c("Endgame", "Middlegame", "Opening")) +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


### Opening Diversity

@fig-strata is the strangest of the figures I've chosen to share here. It's also the most complex, faceting by rating bin and coloring the plot's lines by number of games played. For the most part, players' month-to-month rating seems to remain mostly stable regardless of opening diversity. A few patterns remain consistent, however: (1) At lower ratings, opening diversity has a minor-yet-positive relationship with rating change, indicating slight improvement on average for players with a larger repertoire. This relationship seems to falls off once a player hits a rating of around 1800. And (2) the left tail of each plot tends to skew negative, regardless of where the plot is centered. I'm not entirely sure how useful any of this information will be for modeling, but I felt it was an interesting enough visualization to share.

```{r Div Strata}
#| label: fig-strata
#| fig-cap: "Line color indicates number of games played, while the x-axis show the number of unique openings a player used within a given month."
#| fig-align: center
#| fig-height: 6
#| fig-width: 8

div_strata <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, opening_diversity, games, avg_rating
    FROM read_parquet(%s)
    WHERE opening_diversity IS NOT NULL
      AND games IS NOT NULL
      AND avg_rating IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating
    FROM x
  ),
  z AS (
    SELECT
      CAST(FLOOR(avg_rating/200)*200 AS INTEGER) AS rating_bin,
      CASE
        WHEN games < 10 THEN '<10'
        WHEN games < 25 THEN '10–24'
        WHEN games < 60 THEN '25–59'
        ELSE '60+'
      END AS games_bin,
      opening_diversity,
      (next_rating - avg_rating) AS d_rating_next
    FROM y
    WHERE next_rating IS NOT NULL
  )
  SELECT
    rating_bin,
    games_bin,
    opening_diversity,
    AVG(d_rating_next) AS mean_d,
    COUNT(*) AS n
  FROM z
  GROUP BY rating_bin, games_bin, opening_diversity
  HAVING n >= 5
  ORDER BY rating_bin, games_bin, opening_diversity
", qpath(paths$pm)))

div_strata <- div_strata |>
  mutate(games_bin = factor(games_bin, levels = c("60+", "25–59", "10–24", "<10")))

ggplot(div_strata, aes(x=opening_diversity, y=mean_d, color=games_bin)) +
  geom_line(alpha = 1) +
  facet_wrap(~rating_bin, scales="free_y") +
  labs(x="Opening Diversity (Within-Month)", y="Mean Next-Month Δ Rating",
       color="Games",
       title="Opening Diversity vs Improvement",
       subtitle = "Stratified by Games and Rating") +
  scale_color_few() +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


\newpage


## Methodology
<!--
Describe the methods used to analyze the data. Do not discuss the analysis itself. Make sure to discuss any train/test data splitting here.
-->

At the core of this project are three themes, as listed in the @sec-background: (1) descriptive metagame dynamics, (2) how time constraints relate to behavior and performance, and (3) characterization of player types or styles. The analysis is strictly non-causal; models are used for description, association, and prediction under out-of-sample evaluation.

When splitting the data into train and test sets, all predictive modeling uses forward-in-time splits to avoid data leakage. This means that the training set contains earlier months in the sample window, while the validation/test set are a contiguous block of later months that were held out from the training data. This split is applied at the month level so that predictors from any given month are never trained using information from future months. For player progression tasks, where the target is a future rating change, using time-forward splits is essential for keeping evaluation realistic.

When appropriate, models are additionally checked for robustness under time-control and rating-bin stratification, to ensure performance isn't driven by a single skill or time-control band.

### Data Tranformations
<!--
Describe any transformation you did to the data to assist in the analyzing of the data such as creating dummy variables.
-->

Transformations serve to stabilize noisy estimates, make quantities comparable across players and contexts, and construct interpretable features aligned with all three research topics. Different transformations are applied for different purposes.

#### Aggregation-Level Alignment {#sec-aggregation}

Rather than a single, monolithic dataset, this project handles several datasets aligned to differing observational units. Strictly speaking, this alignment occurs during the data pipeline itself rather than a formal, pre-modeling cleaning step but it's important to note nonetheless. Broadly, the observational units for each table can be separated into four categories: metagame (population-level games), player outcomes (player-level games), time usage, and opening repertoire. This, however, is an oversimplification; please refer to @sec-data for more detail.

#### Binning and Standardization

To support nonparametric EDA and stable estimation, rating is often binned for plots and stratified summaries— several of which have already been featured in @sec-eda. For clustering (player type/style discovery), features are standardized using z-scores so that variables measured on different scales (e.g., seconds, proportions, counts) contribute in a comparable manner.

#### Exposure Weighting and Time-Control-Collapsed Quantities

As displayed by @fig-coverage in @sec-eda, in some contexts it's justifiable to collapse observations across time controls, particularly where classical is concerned. When a model needs 'overall' player-month measures rather than those specific to a given time control, overall rating is constructed as a games-weighted average of `avg_rating` across time controls. Similarly, overall time-trouble rate becomes an average of `low_time_rate` across time controls, weighted by move rows. This prevents the volume disparity between time controls from disproportionately influencing aggregated quantities.

#### Reliability Thresholds

Many variables are rates estimated from finite samples. Often, these sample sizes are extremely small, increasing volatility. In order to reduce such noise, minimum thresholds for observation count may be applied; player-month observations may require a minimum number of games per player, time-use observations a number of move-rows, and opening-level estimates a number of games per group. This filtering may still be applied even if the model or models being fit don't explicitly require it to function (or as part of a fundamental model assumption) to ensure that modeled signals are not dominated by small-sample volatility.

#### Derived Features

A set of interpetable style features is constructed from player-opening and time-use tables for use in type discovery and player-progression models. Similarly to @sec-aggregation, some of these features were created during the data pipeline itself, while others are derived directly from them. These include:

- Repertoire breath, measured by `opening_diversity`.
- Repertoire concentration, computed as a share of games in the most-played opening from `player_opening_month`.
- Opening switching, indicated by whether `top_opening` differs from the previous month.
- Repertoire churn, measured jointly as:
  - The distance between consecutive-month opening-share vectors.
  - The share of games in brand new openings relative to the month prior.
- Time-pressure behavior, based on a combination of `low_time_rate` plus summaries like `avg_time_spent`, `p90_time_spent`, and `avg_time_opening`.

For more information on pre-existing features, see @sec-data.

#### Missingness

Finally, phase-specific time metrics are often missing for structural reasons. Almost always, the phase missing is endgame, as many games never reach a computer-defined endgame due to resignation or blundered, early losses. Models that use phase variables either include explicit missingness handling or analyze phases separately with appropriate inclusion rules.

### Types of Models
Because this project aims to explore a variety of questions by modeling on multiple datasets, strictly categorizing models by family makes little sense.^[Here, I define "model family" as sets of  models operating under similar principles and fundamental assumptions (e.g., general linear models, tree-based models, etc.), whereas "model category" refers to the five groupings of models I specifically define for this paper.] Were there a single theme and associated dataset, doing so would make it easy to illustrate exactly how each type of model contributes to the project's singular goal. In this case, however, I think it more appropriate to categorize models by their goal; to define them by the questions they intend to answer and the methods they'll use in seeking them.^[It should be noted that this form of model categorization allows for slight overlap between model families; some families may be used by multiple categories of model.] 


#### Metagame Dynamics {#sec-metagame-dynamics}
The goal of metagame models are to quantify how aggregate metagame metrics vary across factors such as time and format. The primary observational unit is `meta_opening_month_tc`, while response variables include `share` and outcome rates (e.g., `white_win_rate`). The model families which fall under this category include generalized additive models (GAM) and weighted linear models (WLM), as well as multinomial models. 

GAM and WLM models are useful because they produce simple, easy-to-read trend estimates. The incorporation of weights let high-support opening-month cells impact the fit more than small, noisy cells, and GAM-style smoothing allows for nonlinear time trends without forcing a specific form. Both are strong at tracking a small set of common openings over time (which, given the EDA, is very relevant #ref) while remaining easy to justify and visualize. Importantly, because opening shares are bounded and compositional, these models can only ever approximate the true structure of opening share, meaning that results will necessarily be biased or even potentially spurious.

Multinomial models exist to solve the problem of compositional variables. By modeling on log-odds and log-ratios, they avoid predicting impossible shares ($\notin 0 < x < 1$); in essence, they're capable of implicitly understanding that if one opening goes up, others must go down. This provides interpretable effects for opening adoption relative to a specific baseline, meaning they're best when relative substitution between openings is of import. Multinomial models tend to work better when modeling a large portion of a distribution (rather than just the top k), but as a consequence are more complicated to both implement and interpret.

Combined, these models attempt to ask questions about the state of the metagame – the Lichess 'ecosystem' –  and its evolution rather than just fitting adoption curves. 

#### Opening Popularity
The goal of opening-popularity models is to describe how opening preferences differ by player skill and time control. You'll notice that this goal shares substantial overlap with that of metagame models, and indeed, opening popularity is a subset of metagame dynamics. The core difference between the two is that opening-popularity models exclusively target the adoption rate of openings at the expense of all other metagame descriptors; they care not for the diversity of the opening pool, or how outcome rates, game length, and rating gaps shift, for instance. While both use `meta_opening_month_tc` as an observational unit, popularity models prefer to treat each opening separately (top-k) or as a compositional set focused on `share`. Models in this category include multinomial models, count models, and weighted regression.

Contrary to the metagame category, multinomial models are front and center for opening popularity because the target is explicitly compositional. Metagame models seek to observe large-scale patterns more than anything, so in their case it's possible to get away with using multinomials as a backup for increased accuracy, but when it comes to opening popularity there's no real excuse. Multinomial choice models build player-game as a categorical choice, producing probabilities of choosing each opening as a function of things like rating and time control. They're computationally heavy, but are worth the price if they have access to enough data.

Count models such as poisson and negative binomial are a simpler answer to the same problem, treating observed opening use as counts out of a total number of games. By applying an offset (e.g., taking the log of total games), they can effectively model rates while remaining in count space. Negative binomial can handle overdispersion, which is common in count data. Their main advantage over multinomial models is their ability to produce interpretable, multiplicative effects, or rate ratios, which allows for another angle of analysis.

Finally, weighted regression on `share` takes the 'simple is best' philosophy to an extreme. Just as in the prior section, they don't enforce composition, meaning they're liable to misbehave at the boundaries if used for extrapolation; they can't replace multinomial or count models and only serve as an approximation. They remain for much the same reasons as before: they're easy to implement, easy to explain, and good when the only goal is a half-decent descriptive model focused on top openings. Weighted regression certainly won't be the star of the show, but it's such a simple addition that it costs nothing to include.

Together, these models seek to hone in on opening popularity with a higher degree of specificity than the metagame models, each through varying levels of complexity.

#### Time-Use
Time use models attempt to characterize the way time allocation varies across skill and time control. Observational units are `player_time_month_tc` for player-level behavior and `moves_month_tc` for macro behavior. Response variables include time spent, time management, pressure, and phase allocation, while predictors primarily consist of `avg_rating` and `tc_bucket`. Three related pairs of models populate this category: additive- and interaction regression, robust- and quantile regression, and beta- and logistic-style modeling.

Additive regression and interaction regression provide a clean separation between skill effects and constraint effects. Additive regression assumes each predictor contributes its own effect, and that those effects simply add up, best for when predictors are believed to act independently of one another. They also tend to be a substitute for interaction regression when there's not enough data. Interaction regression, as the name implies, allows the effect of one predictor to depend on the level of another predictor. Both are relatively easy to interpret.

Robust regression and quantile regression are useful in situations when ordinary least squares (OLS) is a poor fit. Notably, both solve different problems. Robust regression still estimates the conditional mean like OLS, but makes it less sensitive to outliers, whereas quantile regression breaks the mold by estimating a conditional quantile (e.g., median, 90th percentile, etc.) rather than the mean. As a consequence, they must also be interpreted slightly differently, as they ask how predictors shift a percentile rather than the mean. They're valuable for modeling time use because time-related variables often present heavy tails, which these models are better equipped to handle.

Beta and logistic modeling are along for the ride because of a specific problem. Namely, `low_time_rate` is bounded and almost always piles up at 0. Both beta and logit-style approaches keep predictions bounded in [0, 1], handling proportion structure better than OLS, the core difference between the two being that logistic modeling uses a binary response (or a count of successes out of a total number of trials), while beta modeling uses a continuous proportion between 0 and 1. By binning low-time rate, it's possible to produce stable summaries that avoid smoothing failures, making for a more-easily-defensible model

Together, these models thoroughly explore several different methods of time-use analysis. The reason this section contains three pairs of complimentary models rather than just the 'better' of each pair is that the EDA did not produce conclusive enough evidence to justify jumping straight to a smaller subset of model options; the data is complex enough that a simple visual inspection was insufficient for ruling out one choice or the other, and so the most reliable solution is to test them all.

#### Player-Progression Models
Player-progression models are straightforward: They wish to relate a player's current-month behavior to future rating change. The purpose of this category is simple, asking if it's possible to isolate certain factors that prove to have a marked effect on a player's growth— positive *or* negative. The baseline observational unit for these models is `player_month`, which can be stratified by time control (`player_month_tc`) or joined to other tables for additional information (e.g., `player_time_month_tc` or `player_opening_month_tc`). The response variables are derived from `avg_rating` and all represent $\Delta rating$ in some form. Horizon spans may differ (i.e., rating change over spans greater than a single month), and constructing a stability model using $|\Delta|$ is also possible. Predictors include the number of games and moves played during a given timespan, style proxies, and time-use features, controlled by baseline rating and other fixed effects. Comprising this category are model families such as regularized linear models, mixed-/fixed effects models, and tree-based regression.

Regularized linear models (ridge, lasso, elastic net) help account for the fact that this category's predictors are correlated. Regularization stabilizes these estimates. Ridge regression is good for prediction under multicollinearity, and lasso, though perhaps less relevant here is good for feature selection. Elastic net is a hybrid of the two. As a whole, they should provide a strong baseline predictive model without sacrificing interpretability.

Fixed effects models and mixed (random) effects models are good at handling longitudinal data, though they differ in how they handle time-invariant predictors. The former controls for time-invariant differences without attempting to predict their effects, while the latter tries to estimate said effects. Fixed effects push towards within-player identification, while mixed effects will pool players to improve stability. Crucially, mixed effects models assume that unit intercepts are uncorrelated to the regressors, which can't be reasonably assumed with this data. Fixed effects models are likely to be the better performing of the two (if non-fixed models even pass muster at all), though the implementation of something like a correlated random effects (Mundlak) model might help bridge the gap.

Tree-based regression both naturally captures nonlinearities and thresholds in the data and automatically handles interactions without explicit specification. They're best for exploratory prediction and detecting complex feature relationships. As with all tree-based models, this comes at the cost of interpretability; they require interpretation tools such as feature importance, partial dependence plots, individual conditional expectation plots, et cetera, and even then struggle to match the interpretability of more-straightforward models.

Together, these models account for the varying facets of longitudinal player-level modeling in order to identify the best angle to attack the problem from.

#### Player Categorization
As the name implies, player-categorization models want to categorize players. More specifically, they seek to identify clusters using behavioral features, stratifying by rating and format, and observing for stability. This is by far the most nebulous of the model categories because they're primarily exploratory in nature; rather than beginning with pre-defined player archetypes and asking how their performance differs, these models ask which style clusters (if any) can be identified in the first place. The most straightforward observational unit is a join between `player_month_tc`, `player_time_month_tc`, and repertoire-derived features, and inputs include performance, outcomes, repertoire breadth, time use, and general context. The core model families in this category are dimensionality reduction and clustering, though additional modeling becomes possible after playstyle has been categorized.

Dimensionality reduction methods such as principal component analysis (PCA) and uniform manifold approximation and projection (UMAP) provide low-dimensional representation. PCA yields interpretable axes and helps with noise reduction, whereas UMAP is useful for visual separations and nonlinear manifolds. PCA is a linear method which tries to preserve large-scale structure, while UMAP is a nonlinear method prefers preserving cluster separation. Dimensionality reduction is good at for summarizing multi-feature space *before* clustering or visualization, which should prove helpful in the case of a very broad 'style space'.

Clustering methods (e.g., k-means, hierarchical, GMM) are potentially capable of turning continuous player-behavior into discrete, describable archetypes. K-means clustering is fast, but requires tuning towards an unknown k. Hierarchical clustering is tree-based and doesn't require a known k, but requires a dendrogram for interpretation, and Gaussian mixture models (GMM) creates clusters of varying sizes with associated membership probabilities, which is useful if player archetypes overlap. 

Though perhaps not within the scope of this project, dimensionality reduction and clustering are notable as a potential method of creating discrete types for downstream modeling. Once each player-month has a type, it becomes possible to study stability and movement between types or perform supervised follow-up using type labels (e.g., as predictors of progression).

Together, these models explore the bounds of player-archetype/playstyle categorization.

### Model Selection
Because this project does not intend to settle on a single, final, 'best' model, I have not outlined an exact process for model selection. Rather, the sheer quantity and breadth of models tested naturally pushes toward a broader set of more-general rules to follow. The rules I have defined are as follows:

- Start with defensible baselines.
  - Make the simplest possible model (or set of models) in a given category that fulfill reasonable standards for transparency, methodology, and fulfilled model assumptions.
  - Compare increasingly-complex models to their baseline counterparts when assessing their worth.
- Prefer simpler models when performance is comparable.
  - When two models perform similarly, the one with fewer assumptions and clearer interpretation is preferable for reporting.
- Check stability to thresholds and stratification.
  - Key results must be checked for sensitivity to minimum thresholds and to whether analyses are stratified by time control and/or rating bin.
  - If one model performs worse than another but the better-performing model fails to pass for stability, then the 'worse', more-stable model should be preferred.
- For clustering specifically, select solutions by stability and interpretability.
  - Candidate clusterings should be preferred when feature profiles are distinct and assignments are reasonably stable across adjacent months or within rating strata.

Because no single framework is sufficient for selecting between all models tested, a large portion of the model-selection process will instead be covered in @sec-results.^[As a note, I heavily suspect I'll be reworking this section before the final draft of this paper.]

\newpage


## Results {#sec-results}
Describe the results that you received from the methods described earlier.

### Model 1 (Change to appropriate name)
Show the results for model type one. At least one fully written out model should be displayed for each model type.

### Model 2 (Change to appropriate name)
Show the results for model type two.

### etc

### Selected model
Display the table that shows the model selection. If using more than 10 models only show the top 10 performing in the table, the full table should be in the appendix. The final model should be fully displayed.


\newpage



## Discussion

### Final model interpolation  
Discuss what information the selected model provides that may be new and interesting

### Use of Model
Describe how this model might be used by others.

### Comparison
Compare the things learned from this selected model to the previous information.


\newpage

## Furture Work
Discuss what work could be done with this model to make it better.





\newpage
## References
List the reference used for your paper. This should include anything cited in your background section and the source of your data.


\newpage
## Appendix
### Appendix A Graphs (Use appropriate name)
```{r appendix A, echo=FALSE}
# Use this code to display useful graphs that do not belong in the body of the paper.
# For example use this area for model validation graphs like residual versus fit.
```

#### A.1 (Use appropriate name)
If showing multiple graphs make sure to index them by 0.1 each


\newpage

### Appendix B Tables (Use appropriate name)
```{r appendix B, echo=FALSE}
# Use this area to show results that are in table format
```

#### Appendix B.1 (Use appropriate name)
If showing multiple tables make sure to index them by 0.1 each


### Code
```{r cleaning,echo=FALSE,eval=FALSE}
# Write your cleaning code here
```


```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
```

## Cleanup (Header To-Be Removed for Final Draft)

```{r Cleanup, output = FALSE}
if (exists("con") && DBI::dbIsValid(con)) {
  DBI::dbDisconnect(con, shutdown = TRUE)
}
```





