---
title: "Project Title (Placeholder)"
author: "Camden Bergquist"
date: "Last compiled on `r format(Sys.time(), '%B %d, %Y')`"
abstract: >
  Lorem Ipsum.
number-sections: true
format:
  jasa-pdf:
    toc: true
    lof: true
    lot: true
    keep-tex: true
    include-in-header:
      text: |
        \usepackage{url}
        \usepackage{xurl}
        \usepackage{booktabs}
        \usepackage{tabularx}
        \usepackage{xltabular}
        \usepackage{array}
        \usepackage{float}
        \keepXColumns
        \usepackage{ragged2e}
        \newcolumntype{T}[1]{>{\RaggedRight\ttfamily\arraybackslash}p{#1}}
        \newcommand{\us}{\_\allowbreak}
        \setlength{\parindent}{2em}
bibliography: references.bib
execute:
  echo: FALSE
  warning: FALSE
  message: FALSE
---


<!--
Note: Find way to center-align figures in YAML header
-->

```{r Libraries, output = FALSE}
# Need I explain these?
library(tidyverse) 
library(scales)

library(arrow) # For reading parquet files
library(DBI) # To open DuckDB connection
library(duckdb) # To manipulate data on disk instead of in environment, saving on RAM
library(ggthemes) # Provides ready-made ggplot themes

```

```{r Setup, output = FALSE}
# Data Paths
base_dir <- "../data/derived/full_datasets"
paths <- list(
  pm = file.path(base_dir, "player_month.parquet"),
  pm_tc = file.path(base_dir, "player_month_tc.parquet"),
  pom_tc = file.path(base_dir, "player_opening_month_tc.parquet"),
  ptm_tc = file.path(base_dir, "player_time_month_tc.parquet"),
  mom_tc = file.path(base_dir, "meta_opening_month_tc.parquet")
)

# Open database connection
con <- dbConnect(duckdb(), dbdir=":memory:")

# Helper for safe `read_parquet()` path quoting
qpath <- function(p) paste0("'", gsub("'", "''", normalizePath(p, winslash="/", mustWork=FALSE)), "'")

```

<!--

Make sure every code chunk has echo=FALSE in its preamble unless you need to highlight the code for some reason in the paper.

## Abstract
Complete this after all other sections have been completed.

-->

\newpage

## Introduction

Competitive games have become a major source of large-scale behavioral data, enabling researchers to study strategic decision-making, skill development, and population-level trends using observational logs at scales unheard of compared to controlled studies [@edelmann2020css; @watts2014css]. Chess is one of the most famous competitive games in the world, and the relatively recent growth of online platforms has created datasets with millions to billions of games played under standardized rules and well-defined competitive contexts (time controls, ratings, game outcomes, etc.) [@watts2014css; @lichessDatabase]. One of the most widely used sources is the Lichess Open Database, which provides public monthly exports of rated games played on lichess.org in PGN format [@lichessDatabase; @lichessDatabaseGitHub]. These exports make it possible to study how strategy and performance vary across skill levels and time controls in a modern, online environment.

This project intends to use Lichess game data to investigate contemporary metagame behavior and player development. Here, 'metagame' refers to the population-level distribution of strategic choices and styles observed in practice – such as opening families, risk preferences, and game outcomes – within a defined time window and player population. In addition to describing how strategic behavior differs by rating and time control, the project will attempt to connect these strategic signatures to measures of player progression over time. Because chess is a repeatedly played, fully competitive setting with a mature rating infrastructure, it offers a natural testbed for quantifying skill and comparing behavior across different players [@lichessRatingSystems; @glickmanGlicko2Desc].

A central component of modern chess analytics is the rating system used to summarize player strength. The ELO rating system – a system originally developed for chess – models player performance probabilistically, and remains a foundational reference point for interpreting skill differences and expected outcomes in competitive settings [@elo1978rating; @fideElo2025]. Lichess uses a modified version of this system called 'Glicko-2', which, in addition to rating, tracks uncertainty via rating deviation and volatility [@lichessRatingSystems; @glickmanGlicko2Desc]. Ratings are especially useful for observational studies because they provide a consistent scale by which to stratify players and construct longitudinal measures of improvement [@glickmanGlicko2Desc].

Overall, this project attempts to marry data science, computational social science, and strategic behavior: It treats online chess as a real-world environment where strategic choices and performance outcomes can be measured at scale and subsequently analyzed using statistical learning methods [@edelmann2020css; @watts2014css].

### Background {#sec-background}

A growing research literature uses large online chess datasets to study human performance, decision-making, and strategic patterns. For example, Chowdhary et al. leverage an online chess dataset at the scale of over 100 million games to quantify performance dynamics and career-like trajectories, illustrating how large game archives support population-level behavioral measurement across skill levels [@chowdhary2023performance].

One important line of work focuses on time usage and decision quality under time constraints. Online chess platforms record both move sequences and clock information, which allows researchers to study how players allocate thinking time and how time pressure shapes decisions. Russek et al. analyze millions of Lichess games to relate move times to the estimated value of additional computation, and show that time allocation patterns vary systematically with playing strength and time control [@russek2025thinking]. These findings support the idea that online chess is a rich setting for studying bounded rationality and performance under constraints, especially when comparing bullet, blitz, rapid, and classical formats [@russek2025thinking].

Chess opening choice is also an active research topic. Because openings are discrete, labelable strategic choices that occur early in every game, they are well suited to population-level analysis. De Marzo and Servedio use online chess community data to quantify opening similarity and complexity, including representing openings in network form and studying how opening-related structure can be used to forecast future opening adoption [@demarzo2023openings]. This motivates analyzing metagame development in terms of shifting opening popularity and success rates across time, player skill, and time controls [@demarzo2023openings; @chowdhary2023performance].

Finally, there is an emerging interest in identifying player styles or archetypes from game logs. While traditional chess analysis often focuses on engine evaluation of positions, a complementary approach is to extract behavioral features from move sequences and outcomes (e.g., game length, opening selection, and other style proxies) and use statistical learning to characterize player types. For instance, McIlroy-Young et al. demonstrate that models trained on move sequences can capture individual-specific signatures (“style”) in chess decisions, providing evidence that player behavior is meaningfully distinguishable from game records alone [@mcilroyyoung2021stylometry]. This creates a bridge between strategic analysis (what players choose to do) and longitudinal outcomes (how players improve or persist over time), which aligns closely with the goals of this project.

Taken together, existing work suggests that large-scale online chess datasets support three relevant kinds of analysis: (1) descriptive metagame analysis of strategic choices across populations, (2) modeling how constraints like time pressure affect behavior and outcomes, and (3) data-driven characterization of player types and their evolution [@chowdhary2023performance; @russek2025thinking; @demarzo2023openings; @mcilroyyoung2021stylometry]. This project intends to build on those themes by using Lichess data to analyze modern opening/style patterns and connect them to measures of player performance and development over time.

<!--
\newpage
-->

## The Data {#sec-data}

### Source {#sec-source}
<!--
Describe where this data comes from, why it was collected, who collected it, and how it was collected. A Citation may be useful here.
-->
Lichess is the largest open-source chess server is the world, and releases free, monthly exports of their game data under the Creative Commons CCO license [@lichessDatabase]. The site requires a log of the game data in order to function while a game is being played, but holds onto it after the fact to allow users to analyze their games after playing them. Though Lichess offers a powerful API, I elected to download data directly from their website in `.pgn.zst` format. The PGN (portable game notation) format is a plaintext format for chess games, containing moves, headers like players, date, and event, and even optional annotation. The `.zst` file extension (sometimes also written as `zstd`) represents Zstandard compression, which is a modern compressor similar to gzip or xz that's commonly used because it's fast and compresses well; think of it as a variation on a ZIP or RAR file. Zstandard compression can be expected to shrink PGN files by 5 times on its faster, less-efficient settings, and even more on it's slower ones.

For the purposes of this project, I downloaded 24 months of data spanning from January 2024 to December 2025, amounting to 677 gigabytes of raw data when compressed. Obviously, nearly 700 gigabytes of compressed data – 3~4 terabytes of uncompressed PGN data – is an insane amount to try to work with, capstone project or not. I have a powerful home computer, but that kind of scale requires something closer to a high-end server than a personal computer; it won't fly. Instead of tackling the problem head-on, I constructed a hybrid R/Python pipeline that downloads the data, deterministically samples and parses it into small chunks (in the form of parquet files), and then builds a selection of smaller, ready-for-analysis datasets from those chunks. Individually, no one dataset has the 20+ variables required in the project outline, but since I've built multiple datasets for multiple different types of analysis, they collectively meet that benchmark 

The download and parsing section of the pipeline handles the following:

- Reads a config file `pipeline.yml` that controls the scope and parameters of the pipeline's operation.
- Downloads monthly Lichess 'standard rated' PGN archives for each month specified in the YAML.
- Decompresses each Zstandard file as a stream, scanning games one-by-one so it doesn't run into RAM bottlenecks.
- Parses game headers first to decide whether a game is relevant before doing any heavy work (not all games are actually getting sampled, so it skips over the ones not being used).
  - Extracts core metadata (players, ratings, time control, result, termination, opening, timestamps).
  - Buckets each game into blitz/rapid/classical using the Lichess time-control convention.
- Applies filtering rules from the YAML, which in this case are:
  - Only include blitz/rapid/classical time controls.
  - Only include games from players in the 800–2600 rating range.
  - Drops unwanted terminations (aborted games).
- Uses deterministic hash-based sampling to create reproducible subsets without needing to store the whole population:
  - A meta sample keyed by `game_id` for population/metagame summaries, at a rate of 5/500 games.
  - A player cohort keyed by `username` for longitudinal/player tracking at a rate of 3/500 players.
  - A moves sample keyed by `game_id` for clock/time-per-move data at a rate of 5/10000 moves.
- Deterministically anonymizes usernames based on a stable, salted hash.
  - Cohort selection is computed on the real username before anonymization, so it remains reproducible.
- Writes month-partitioned parquet outputs for game-level data.

The dataset-building portion of the pipeline takes those outputs and, using DuckDB to read and write parquet files on the disk, builds seven datasets from them:

- Player-level longitudinal summary tables from `games_cohort`:
  - `player_month` (`pm`): player * month, 249,761 observations.
  - `player_month_tc` (`pm_tc`): player * month * time control, 320,869 observations.
- Player-opening 'repertoire' table from `games_cohort`:
  - `player_opening_month_tc` (`pom_tc`): player * month * time control * opening, 7,167,580 observations.
- Metagame tables from `games_meta`:
  - `meta_month_tc` (`mm_tc`): month * time control, 72 observations.
  - `meta_opening_month_tc` (`mom_tc`): month * time control * opening, 130,672 observations.
- Time-usage tables from `moves_meta`:
  - `moves_month_tc` (`mvm_tc`) month * time control, 72 observations.
  - `player_time_month_tc` (`ptm_tc`): player * month * time control, 1,038,050 observations.

The pipeline logs session and version info for both R and Python and records checksums for downloaded raw files to help with reproducibility.

### Variables
This project is working with multiple related-yet-distinct datasets. As such, it's very difficult to properly introduce the data without an according quantity of tables, which results in several pages worth of figures. My personal recommendation is to read through table 1, featured below, before moving straight to @sec-observations.^[I'm considering consolidating the Variables and Observations sections into one, as I feel they're somewhat redundant, or else perhaps moving the Variable tables entirely into the appendix. I need to consult you for permission on that first, though.] Most variable names are self-explanatory, so this section should largely be treated as a reference to skim through if a variable introduced later is confusing or unclear. Additionally, I elected to exclude tables `mm_tc` and `mvm_tc` from this section, as they're summary-level datasets that won't be explicitly modeled. Please refer to the <appendix placeholder; update when written> for information on them.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.8cm} X}
\caption{Core identifiers and categorical variables used across derived datasets.}\label{tab:variables-core}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & categorical & Deterministic pseudonymous user ID (e.g., \texttt{u\us ...}). Appears in player-level tables (\texttt{pm}, \texttt{pm\us tc}, \texttt{pom\us tc}, \texttt{ptm\us tc}). \\
\texttt{month} & categorical & Month of play as \texttt{YYYY-MM}. Primary time index for metagame and longitudinal analysis. \\
\texttt{tc\us bucket} & categorical & Time control bucket: \texttt{blitz}, \texttt{rapid}, or \texttt{classical}. Used to stratify behavior by constraint level. \\
\texttt{eco} & categorical & ECO opening code (e.g., \texttt{C41}). Standardized opening taxonomy useful for grouping openings into families. \\
\texttt{opening} & categorical & Human-readable opening family name (e.g., \texttt{Philidor Defense}). Used in repertoire and metagame summaries. \\
\end{xltabular}




```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player-month outcome and covariate variables (tables: \texttt{pm}, \texttt{pm\us tc}).}\label{tab:variables-player-month}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of games played in the month (and time-control bucket for \texttt{pm\us tc}). Also a reliability/weighting signal. \\
\texttt{wins} & numeric (count) & Count of wins in the period. \\
\texttt{losses} & numeric (count) & Count of losses in the period. \\
\texttt{draws} & numeric (count) & Count of draws in the period. \\
\texttt{win\us rate} & numeric (proportion) & Key performance metric: wins divided by games. Natural response variable for performance models. \\
\texttt{draw\us rate} & numeric (proportion) & Draws divided by games. Useful because draw propensity varies by skill, time control, and strategic choices. \\
\texttt{avg\us rating} & numeric & Primary skill proxy: average of the player’s pre-game rating across games in the period. Used for stratification and longitudinal change. \\
\texttt{avg\us opp\us rating} & numeric & Average opponent rating. Contextualizes performance and approximates strength of schedule. \\
\texttt{avg\us rating\us diff} & numeric & Average (player rating minus opponent rating). Summarizes typical matchup difficulty. \\
\texttt{avg\us moves} & numeric & Average game length (moves). Proxy for style/complexity. \\
\texttt{opening\us diversity} & numeric (count) & Number of distinct openings used in the period. Proxy for repertoire breadth. \\
\texttt{top\us opening} & categorical & Most frequently played opening during the period (ties broken deterministically). \\
\end{xltabular}



```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player-opening table variables (table: \texttt{pom\us tc}).}\label{tab:variables-player-opening}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of games the player played with the given opening in the month/time-control bucket. Used as a frequency weight. \\
\texttt{win\us rate} & numeric (proportion) & Opening-specific win rate for the player and period. Used to study effectiveness conditional on opening. \\
\texttt{draw\us rate} & numeric (proportion) & Opening-specific draw rate. Helps characterize risk profile and style. \\
\texttt{avg\us rating} & numeric & Average player rating in games where the opening was played. Enables comparison of opening choice across skill. \\
\texttt{avg\us rating\us diff} & numeric & Average rating difference in games where the opening was played. Captures whether openings are used in tougher/easier matchups. \\
\end{xltabular}




```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Player time-use table variables (table: \texttt{ptm\us tc}).}\label{tab:variables-player-time}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{move\us rows} & numeric (count) & Number of move observations contributing to the summary (after sampling and filtering). Reliability signal. \\
\texttt{games} & numeric (count) & Number of distinct games contributing move records for the player-month-time-control group. \\
\texttt{avg\us time\us spent} & numeric (seconds) & Mean time spent per move. Captures typical deliberation time. \\
\texttt{median\us time\us spent} & numeric (seconds) & Median time spent per move; robust to outliers and occasional long thinks. \\
\texttt{p90\us time\us spent} & numeric (seconds) & 90th percentile of time spent per move. Measures tendency for occasional deep computation. \\
\texttt{avg\us clock\us after} & numeric (seconds) & Mean remaining clock time after moves. Proxy for time management and pacing. \\
\texttt{low\us time\us rate} & numeric (proportion) & Fraction of moves under a low-time threshold. Proxy for time trouble frequency. \\
\texttt{avg\us rating} & numeric & Average player rating across move records in the group; allows stratification by skill. \\
\texttt{avg\us rating\us diff} & numeric & Average rating difference across the move sample; controls for opponent strength effects. \\
\texttt{avg\us time\us opening} & numeric (seconds) & Average time spent per move during the opening phase (if phase segmentation is available; game phases are defined by chess-engine analysis). \\
\texttt{avg\us time\us middlegame} & numeric (seconds) & Average time spent per move during middlegame phase (may be missing for very short games). \\
\texttt{avg\us time\us endgame} & numeric (seconds) & Average time spent per move during endgame phase (often missing because many games never reach endgame). \\
\end{xltabular}



```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.6cm} X}
\caption{Metagame opening table variables (table: \texttt{mom\us tc}).}\label{tab:variables-metagame-opening}\\
\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endfirsthead

\toprule
\textbf{Variable} & \textbf{Type} & \textbf{Description} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{games} & numeric (count) & Number of sampled games for the opening within month and time control. Precision/support indicator. \\
\texttt{share} & numeric (proportion) & Core metagame measure: fraction of games using the opening within a month/time-control bucket. Tracks adoption shifts. \\
\texttt{white\us win\us rate} & numeric (proportion) & Fraction of games where White wins. Used for opening-level success patterns and color asymmetries. \\
\texttt{black\us win\us rate} & numeric (proportion) & Fraction of games where Black wins. Complements White win rate. \\
\texttt{draw\us rate} & numeric (proportion) & Fraction of games drawn. Draw propensity varies by opening and time control. \\
\texttt{upset\us rate} & numeric (proportion) & Fraction of games where the lower-rated player wins (pipeline definition). Proxy for volatility/competitiveness. \\
\texttt{avg\us abs\us rating\us diff} & numeric & Mean absolute rating gap between players in the opening/month/tc group. Context for outcome rates. \\
\texttt{avg\us moves} & numeric & Mean game length for the opening/month/tc group. Proxy for complexity and style at population level. \\
\end{xltabular}


```

### Observations {#sec-observations}
Each derived dataset is organized at a different unit of analysis (e.g., player-by-month or opening-by-month). In all cases, the raw unit of data is a rated chess game from the Lichess public database, but the pipeline aggregates games (and, for time-use, moves) into analysis-ready summaries. Because many variables overlap across tables, the observation structure can be summarized with a small set of representative row types. The examples below demonstrate the core units: (1) player-period summaries (with optional time-use extensions), (2) player opening-level summaries, and (3) population-level metagame summaries for openings.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observation for player-period outcome summaries (tables: \texttt{pm}, \texttt{pm\us tc}).}\label{tab:obs-player-period-outcomes}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & \texttt{u\us 0001cf80083d0517} (stable pseudonymous identifier) \\
\texttt{month} & \texttt{2024-12} (calendar month) \\
\texttt{tc\us bucket} & \texttt{rapid} (present in \texttt{pm\us tc}; omitted in \texttt{pm}) \\
\texttt{games} & 5 (total games in this player-month-(tc) group) \\
\texttt{wins} & 4 \\
\texttt{losses} & 0 \\
\texttt{draws} & 1 \\
\texttt{win\us rate} & 0.80 (wins / games) \\
\texttt{draw\us rate} & 0.20 (draws / games) \\
\texttt{avg\us rating} & 1544.0 (mean pre-game rating across games) \\
\texttt{avg\us opp\us rating} & 1503.2 (mean opponent rating across games) \\
\texttt{avg\us rating\us diff} & 40.8 (mean rating difference: player minus opponent) \\
\texttt{avg\us moves} & 40.4 (mean game length in moves) \\
\texttt{opening\us diversity} & 4 (distinct openings played in the group) \\
\texttt{top\us opening} & \texttt{Ponziani Opening: Jaenisch Counterattack} (most frequent opening) \\
\end{xltabular}


```


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observation for player-period time-use summaries (table: \texttt{ptm\us tc}).}\label{tab:obs-player-timeuse}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\texttt{player} & \texttt{u\us 00004578077e9fa4} (stable pseudonymous identifier) \\
\texttt{month} & \texttt{2024-08} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{move\us rows} & 70 (number of move records contributing to this summary) \\
\texttt{games} & 1 (distinct games represented among the sampled moves) \\
\texttt{avg\us time\us spent} & 1.66 seconds (mean time spent per move) \\
\texttt{median\us time\us spent} & 1.0 seconds (median time spent per move) \\
\texttt{p90\us time\us spent} & 3.1 seconds (90th percentile time spent per move) \\
\texttt{avg\us clock\us after} & 122.5 seconds (mean remaining clock after moves) \\
\texttt{low\us time\us rate} & 0.00 (share of moves flagged as low-time) \\
\texttt{avg\us rating} & 1677.0 (mean pre-move rating across sampled moves) \\
\texttt{avg\us rating\us diff} & 11.0 (mean rating difference across sampled moves) \\
\texttt{avg\us time\us opening} & 0.9 seconds (mean time spent in opening phase) \\
\texttt{avg\us time\us middlegame} & 2.2 seconds (mean time spent in middlegame phase) \\
\texttt{avg\us time\us endgame} & 1.64 seconds (mean time spent in endgame phase; may be missing) \\
\end{xltabular}


```


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} X}
\caption{Sample observations for opening-level summaries: player repertoire entries (\texttt{pom\us tc}) and population metagame cells (\texttt{mom\us tc}).}\label{tab:obs-opening-level}\\
\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endfirsthead

\toprule
\textbf{Field} & \textbf{Example Value/Meaning} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

\multicolumn{2}{l}{\textbf{Player opening summary (table: \texttt{pom\us tc})}} \\
\texttt{player} & \texttt{u\us 0001cf80083d0517} \\
\texttt{month} & \texttt{2024-11} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{eco} & \texttt{B13} \\
\texttt{opening} & \texttt{Caro-Kann Defense: Exchange Variation} \\
\texttt{games} & 4 (games where this player used the opening in this month/tc) \\
\texttt{win\us rate} & 0.50 (opening-conditional win rate for the player) \\
\texttt{draw\us rate} & 0.00 (opening-conditional draw rate for the player) \\
\texttt{avg\us rating} & 1965.3 (mean rating in games using this opening) \\
\texttt{avg\us rating\us diff} & 12.0 (mean rating difference in those games) \\

\midrule
\multicolumn{2}{l}{\textbf{Population opening summary (table: \texttt{mom\us tc})}} \\
\texttt{month} & \texttt{2024-01} \\
\texttt{tc\us bucket} & \texttt{blitz} \\
\texttt{eco} & \texttt{C41} \\
\texttt{opening} & \texttt{Philidor Defense} \\
\texttt{games} & 8{,}825 (sampled games using this opening in the month/tc group) \\
\texttt{share} & 0.0189 (opening share of the month/tc metagame) \\
\texttt{white\us win\us rate} & 0.512 (fraction of games White wins) \\
\texttt{black\us win\us rate} & 0.442 (fraction of games Black wins) \\
\texttt{draw\us rate} & 0.0466 (fraction drawn) \\
\texttt{upset\us rate} & 0.436 (fraction where lower-rated player wins) \\
\texttt{avg\us abs\us rating\us diff} & 50.2 (mean rating gap magnitude) \\
\texttt{avg\us moves} & 35.5 (mean game length) \\
\end{xltabular}


```

Together, these observation types support longitudinal player performance modeling (`pm`, `pm_tc`), repertoire/style measurement from opening choice (`pom_tc`), metagame dynamics via opening adoption and outcomes (`mom_tc`), and bounded-rationality/time-management analyses using move-time features (`ptm_tc`).

<!--
\newpage
-->

### Cleaning

A large amount of cleaning was performed during the pipeline's parsing steps. The filtering and deterministic sampling was already touched upon in #sec-source, but additional standardization and feature construction include:

- `TimeControl` tag parsed into:
  - `initial` (seconds).
  - `increment` (seconds).
  - `tc_bucket` (< 180 = bullet, < 600 = blitz, < 1800 = rapid, else classical),
- Converted PGN result strings into:
  - `winner` (white/black/draw).
  - `score_white` (1/0.5/0).
- Computed upset indicators using ratings:
  - `rating_diff` = `white_rating` - `black rating`.
  - `abs_rating_diff`
  - `is_upset`
- Game-length variables:
  - Computed `piles` and `moves` from move sequences (used to construct final datasets).

The information was then restructured and aggregated into a tidy format.

## Exploratory Data Analysis {#sec-eda}

@fig-coverage, the first and most important of the EDA plots, shows the number of observations the parser sampled per player-month, colored by time control.^[Throughout, 'rows' refer to aggregated observation units (e.g., player-month-time-control or opening-month-time-control records), whereas 'games' refer to the underlying sampled matches contributing to those aggregates. A change in row counts therefore reflects changes in the breadth of observed groups, not necessarily a change in total game volume.] It highlights a fundamental flaw in my parsing method— one that I failed to consider until seeing it visualized. See, while the parser filters the raw data down to a smaller subset of time controls and rating ranges, it does nothing to ensure that there's an equal distribution of games for given time controls or players of given ratings. Rather, it simply samples in a pseudo-random manner. This means that the distribution of certain variables in the cleaned dataset will reflect the true distribution of those variables among all games stored in the Lichess database. In more straightforward terms, popular time controls like blitz are played more often than less-popular ones like classical, and that relationship is mirrored in the data, as evidenced in @fig-coverage.

This finding isn't the end of the world, but it *is* extremely important to control for. It means that some 'missingness' will be structural rather than truly random, and that there will be a general imbalance across time controls, ratings, and months. It also justifies restricting certain analyses to blitz/rapid (or else collapsing them entirely), adding minimum-support thresholds to ensure there's enough data, and treating month as a potential confound only when coverage is stable enough; buckets with adequate support will be emphasized, and low-support buckets should be treated as exploratory only.


```{r Coverage Plot}
#| label: fig-coverage
#| fig-cap: "Proportion of observations per time control is skewed due to the sampling method used when parsing. Any estimate that compares time controls will have unequal precision, especially where classical is concerned. "
#| fig-align: center

cov_pm_tc <- dbGetQuery(con, sprintf("
  SELECT month, tc_bucket, COUNT(*) AS n_rows
  FROM read_parquet(%s)
  GROUP BY month, tc_bucket
  ORDER BY month, tc_bucket
", qpath(paths$pm_tc)))

ggplot(cov_pm_tc, aes(x = month, y = n_rows, group = tc_bucket, color = tc_bucket)) +
  geom_line(linewidth = 1) +
  coord_cartesian(clip = "off") +
  labs(x = "Month", y = "Observations (Player-Month-TC)", title = "Coverage of Player-Month-TC Table", color = "Time Control") +
  scale_color_few(labels = c("Blitz", "Classical", "Rapid")) +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")
```

@fig-reliability and @fig-panel-depth display two different kinds of unreliability inherent to player-month: the former being the number of games played in a given month, and the latter the number of months each player is present for in the data (total; not necessarily consecutive). For @fig-reliability, many player-months contain few games, making outcomes like win rate or average rating noisy at the unit level, justifying a minimum-games filter for player-month analyses, weighting player-month observations by games in aggregated models, and preferring aggregated measures such as bins over raw point clouds; game count should be treated as a precision/weighting signal.



```{r Reliability Plot}
#| label: fig-reliability
#| fig-cap: "Player-month observations have highly unequal reliability. The dataset has a long, right-skewed tail of heavy-activity players, which risk dominating aggregates if not uncontrolled for."
#| fig-align: center

gdist <- dbGetQuery(con, sprintf("
  SELECT games
  FROM read_parquet(%s)
  WHERE games <= 250
", qpath(paths$pm)))

ggplot(gdist, aes(x=games)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "steelblue") +
  coord_cartesian(xlim = c(0, 250)) +
  scale_y_continuous(labels = label_comma()) +
  labs(x = "Games per Player in a Given Month", y = "Rows", title = "Player-Month Reliability Varies Widely") +
  theme_few()
```
For @fig-panel-depth, a majority of players only appear for a few months, which is to be expected for a hobby like chess. There's a notable spike at the 24-month column, but that makes sense— that column reflects long-time players who have likely played for much longer than two-odd years; if this data were being visualized without a two-year window, that spike would reflect a '24+' bin rather than players who have played for exactly 24 months. This attrition justifies the reporting of effective sample size for any next- or k-month analysis, using consecutive-month joins rather than simple lags on sparse panels, and potentially restricting to 'stable' players or months when clean identification is important.

```{r Panel Depth Plot}
#| label: fig-panel-depth
#| fig-cap: "The panel is unbalanced: Many players appear briefly; fewer are observed consistently. Lag-based outcomes are only defined for a subset of observations."
#| fig-align: center

panel_len <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, COUNT(DISTINCT month) AS n_months
    FROM read_parquet(%s)
    GROUP BY player
  )
  SELECT n_months, COUNT(*) AS n_players
  FROM x
  GROUP BY n_months
  ORDER BY n_months
", qpath(paths$pm)))

ggplot(panel_len, aes(x = n_months, y = n_players)) +
  geom_col(fill = "steelblue", color = "steelblue") +
  scale_y_continuous() +
  labs(x = "Months Observed per Player", y = "Players",
       title = "Panel Depth: How Long Players Remain Observable") +
  theme_few()
```

Pictured in @fig-delta-rating is without contest the single most perfect Gaussian distribution I have ever seen in a real-world dataset. This is not a parsing mistake, nor the result of some sort of transformation, but a natural consequence of the Glicko-2 rating system employed by Lichess. Both Glicko-2 and its predecessor Glicko assume that player skill – both for individuals *and* for a given population – follows a normal distribution. Because of that, implementations of Glicko systems almost always result in normally-distributed skill ratings, provided the player count is high enough.^[It's important to note that Glicko-based systems don't 'force' ratings to conform to a normal distribution; it's possible for them to fall into any form, but the way the system is designed tends to converge towards normality in most cases.] Practically, this means that $\Delta$ rating should be treated as a noisy outcome requiring careful aggregation, preferring binned summaries over unconstrained smoothing. It also encourages caution when interpreting statistically-significant-yet-tiny effects, as small, downright impractical or meaningless outcomes can still read as significant when dealing with a distribution like this.

```{r Rating Changes Plot}
#| label: fig-delta-rating
#| fig-cap: "Rating change is centered near zero with roughly symmetric dispersion. Most month-to-month changes are small, so effect sizes will likely be modest."
#| fig-align: center

chg <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT (avg_rating - prev_rating) AS d_rating
  FROM y
  WHERE prev_rating IS NOT NULL AND d_rating >= -300 AND d_rating <= 300
", qpath(paths$pm)))

ggplot(chg, aes(x=d_rating)) +
  geom_histogram(bins=80, fill = "steelblue", color = "steelblue") +
  labs(x = "Month-to-Month Change in Average Rating", y = "Count", 
       title = "Distribution of Within-Player Rating Change") +
  theme_few()
```
Unsurprisingly, @fig-volatility shows that the variance in players' month-to-moth ratings decreases with the number of games played in a given month. Importantly, the figure is log-transformed; though the plot visualizes a linear relationship, the first hundred games or so are comparatively more impactful than the next thousand. This is a relationship that I would expect to see with any rating system – law of large numbers and all – but was something I was expecting from Glicko-2 in particular. On top of rating, Glicko systems also measure uncertainty, which is a Bayesian-style measure of how confident it is in a player's rating. Because uncertainty always reduces as more games are played, it makes sense that this relationship is so clearly present in the data. This figure helps reinforce the need for minimum-games filters (which will reduce heteroskedastic noise) and shows that weighting by games is defensible as precision weighting. It also supports longer-horizon $\Delta$ (over multiple months instead of just one) as a way to increase robustness because it averages noise.

```{r Volatility Plot}
#| label: fig-volatility
#| fig-cap: "Observed rating movement behaves like a measurement-error problem; the number of games played is inversely proportional to rating-change volatility. This is expected behavior for a rating system."
#| fig-align: center

vol <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, games, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      games,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT
    games,
    ABS(avg_rating - prev_rating) AS abs_d_rating
  FROM y
  WHERE prev_rating IS NOT NULL
", qpath(paths$pm)))

ggplot(vol, aes(x = games, y = abs_d_rating)) +
  geom_point(alpha = 0.15) +
  scale_x_continuous(trans = "log10") +
  labs(x = "Games per Month (Log-Transformed)", y = "Absolute Change in Rating (|Δ|)",
       title = "Monthly Rating Volatility Shrinks as Game Volume Rises",
       subtitle = "Log-Transformed for Legibility") +
  theme_few()
```

@fig-low-time shows that low-time rate is massively zero-inflated, to the degree that it's difficult to identify a meaningful relationship when modeled continuously; binning is practically a necessity. Not shown directly in the figure is the fact that rapid and classical time controls don't have enough data to be meaningfully modeled on their own; due to how low-time rate is defined in the parser, a much smaller proportion of moves classify as being performed under time pressure.^[Low-time definition: A move is flagged as 'low time' if the remaining clock time after the move satisfies $\text{clock\_after} \le \max\!\left(10,\; 2 \times \text{increment}\right)$ seconds, where $\text{increment}$ is the per-move increment of the game's time control.] This justifies collapsing across time controls via move-weighted aggregations, near-exclusive use of binned metrics (including a 0% bin) instead of smoothers, and a combination of minimum move-row thresholds and n-per-bin reports for defensibility.

```{r LTR}
#| label: fig-low-time
#| fig-cap: "Low-time rate is zero-inflated and noisy. Time-control-specific curves are underpowered (with the exception of blitz), and pooling is needed for stable inference. Any relationshop between time trouble and improvement is likely both small and nonlinear, which is why loess smoothing wasn't applied to this figure."
#| fig-align: center

min_games_pm   <- 5     # require at least this many games in the month (across all TCs)
min_moves_ptm  <- 25    # require at least this many move rows in the month (across all TCs)

ltr_scatter_all <- dbGetQuery(con, sprintf("
  WITH pm_tc AS (
    SELECT
      player,
      month,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  pm AS (
    -- collapse across time controls using games-weighted avg_rating
    SELECT
      player,
      month,
      SUM(games) AS games_total,
      SUM(avg_rating * games) / SUM(games) AS avg_rating_all
    FROM pm_tc
    GROUP BY player, month
    HAVING SUM(games) >= %d
  ),
  pm_next AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      LEAD(avg_rating_all) OVER (PARTITION BY player ORDER BY month) AS next_rating_all
    FROM pm
  ),
  delta AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      (next_rating_all - avg_rating_all) AS d_rating_next
    FROM pm_next
    WHERE next_rating_all IS NOT NULL
  ),
  ptm_tc AS (
    SELECT
      player,
      month,
      low_time_rate,
      move_rows
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND low_time_rate IS NOT NULL
      AND move_rows IS NOT NULL
  ),
  ptm AS (
    -- collapse across time controls using move_rows-weighted low_time_rate
    SELECT
      player,
      month,
      SUM(move_rows) AS move_rows_total,
      SUM(low_time_rate * move_rows) / SUM(move_rows) AS low_time_rate_all
    FROM ptm_tc
    GROUP BY player, month
    HAVING SUM(move_rows) >= %d
  )
  SELECT
    d.player,
    d.month,
    d.avg_rating_all,
    d.games_total,
    d.d_rating_next,
    p.low_time_rate_all,
    p.move_rows_total
  FROM delta d
  JOIN ptm p USING (player, month)
", qpath(paths$pm_tc), min_games_pm,
   qpath(paths$ptm_tc), min_moves_ptm))

ggplot(ltr_scatter_all, aes(x = low_time_rate_all, y = d_rating_next)) +
  geom_point(alpha = 0.25) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  ylim(-225, 225) +
  labs(
    x = "Low-Time Rate",
    y = "Next-Month Δ Rating",
    title = "Time Trouble vs Next-Month Rating Change",
    subtitle = "Collapsed Across Time Controls"
  ) +
  theme_few()

```

Collectively, these diagnostics show that the dataset is (1) imbalanced across time controls (and ratings, though not visualized), (2) highly variable in within-unit support, and (3) noisy in short-horizon outcomes. Therefore, methods should adopt minimum-support thresholds, use precision weights, and rely on binned or aggregated summaries for stable inference.

### Opening Share

Opening trends over time are one of the main ways this project aims to study metagame development. The results, featured in @fig-opening and @fig-opening-rating, are rather disappointing. In particular, @fig-opening displays an amount of stability that I never would have expected from a dataset spanning as many months as this. It should be noted that similar figures for rapid and classical time controls show similar patterns, but with more variance. I suspect this difference in variance is a result of the fewer number of rows associated with each format, as classical stands out the most. Even if variance differs, however, the pattern remains regardless of time control— there's very little drift in terms of opening popularity, at least so far as the most popular choices are concerned. A better way of exploring this dynamic would have used select months over several years, but doing so would have required a completely different parsing pipeline, so this will have to suffice. 

```{r Opening Trends}
#| label: fig-opening
#| fig-cap: "Opening preference remains remarkably stable in blitz games over the project's two-year period."
#| fig-align: center
#| fig-height: 4

top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz' AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT 8
", qpath(paths$mom_tc)))

open_list <- paste0("('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')")

trend <- dbGetQuery(con, sprintf("
  WITH filtered AS (
    SELECT month, tc_bucket, opening, games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz'
      AND opening IN %s
      AND opening IS NOT NULL
  ),
  opening_level AS (
    SELECT
      month,
      tc_bucket,
      opening,
      SUM(games) AS games
    FROM filtered
    GROUP BY month, tc_bucket, opening
  )
  SELECT
    month,
    tc_bucket,
    opening,
    games,
    CAST(games AS DOUBLE) / SUM(games) OVER (PARTITION BY month, tc_bucket) AS share
  FROM opening_level
  ORDER BY month, opening
", qpath(paths$mom_tc), open_list))


ggplot(trend, aes(x=month, y=share, color=opening, group=opening)) +
  geom_line(linewidth = 1) +
  scale_y_continuous(labels=percent_format(accuracy=0.1)) +
  labs(x="Month", y="Share of Games", color = "", 
  title="Opening Share Over Time (Blitz)", subtitle = "Top Eight Openings in Terms of Game Share") +
  scale_color_few() +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle=60, hjust=1)
  ) + 
  guides(color = guide_legend(nrow = 4, byrow = TRUE))
```

@fig-opening-rating, while still more stable than expected, is more interesting. This figure shows clear changes in preferences as player rating increases, particularly in the cases of the Pirc and Philidor defenses— the former becoming more popular and the latter less. Even where the more stable openings are concerned, there's a more pronounced trend present than anything in @fig-opening. Even if the relationship is minor, this is evidence enough to warrant attempts at modeling opening popularity as a function of player skill.

```{r Opening by Rating}
#| label: fig-opening-rating
#| fig-cap: "Opening trends by rating are more interesting. The Pirc and Philidor defenses show particularly interesting movement as player rating increases."
#| fig-align: center
#| fig-height: 4

# settings
tc_use <- "blitz"
bin_width <- 200
top_k <- 8             # number of openings to plot
min_bin_total <- 2000  # min player-games in a rating bin

# pick top openings within tc_bucket (weighted by games)
top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT %d
", qpath(paths$pom_tc), tc_use, top_k))

open_list <- paste0(
  "('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')"
)

# compute opening share within rating bins
trend_rating <- dbGetQuery(con, sprintf("
  WITH base AS (
    SELECT
      tc_bucket,
      opening,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IN %s
      AND opening IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  binned AS (
    SELECT
      opening,
      CAST(FLOOR(avg_rating / %d) * %d AS INTEGER) AS rating_bin_lo,
      SUM(games) AS player_games
    FROM base
    GROUP BY opening, rating_bin_lo
  ),
  totals AS (
    SELECT
      rating_bin_lo,
      SUM(player_games) AS bin_total
    FROM binned
    GROUP BY rating_bin_lo
  )
  SELECT
    b.rating_bin_lo,
    (b.rating_bin_lo + %d - 1) AS rating_bin_hi,
    b.opening,
    b.player_games,
    t.bin_total,
    CAST(b.player_games AS DOUBLE) / t.bin_total AS share
  FROM binned b
  JOIN totals t USING (rating_bin_lo)
  WHERE t.bin_total >= %d
  ORDER BY b.rating_bin_lo, b.opening
", qpath(paths$pom_tc), tc_use, open_list,
   bin_width, bin_width, bin_width, min_bin_total))


trend_rating <- trend_rating |>
  mutate(
    rating_bin = paste0(rating_bin_lo, "\u2013", rating_bin_hi),
    rating_bin = factor(rating_bin, levels = unique(rating_bin))
  )

ggplot(trend_rating, aes(x = rating_bin, y = share, color = opening, group = opening)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2, alpha = 0.7) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = "Rating",
    y = "Share of Player-Games",
    title = paste0("Opening Share by Rating (Blitz)"),
    color = ""
  ) +
  scale_color_few() +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle=60, hjust=1)
  ) + 
  guides(color = guide_legend(nrow = 4, byrow = TRUE))
```

### Time Usage

@fig-time and @fig-time-allocation are much more in line with what I was expecting than the opening-popularity figures. @fig-time shows a clear difference in thinking time, both between time controls and player ratings. Blitz and classical time controls are mirrors of one another: As players get better, they spend less time per move in blitz formats and more per move in classical formats. Rapid, in contrast, shows an interesting curve, where time usage dips for a little and then increases. Overall, @fig-time is enough on its own to justify time control as a factor when modeling with move time as an outcome.

```{r Time Use}
#| label: fig-time
#| fig-cap: "All three time controls have distinct time-usage patterns as player rating increases."
#| fig-align: center

time_bin <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    floor(avg_rating/100)*100 AS rating_bin,
    AVG(avg_time_spent) AS mean_time_spent,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 20000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

ggplot(time_bin, aes(x=rating_bin, y=mean_time_spent)) +
  geom_line() +
    facet_wrap(~factor(tc_bucket, levels = c("blitz", "rapid", "classical")), scales="free_y", labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))) +
  labs(x="Player Rating", y="Average Time per Move (Seconds)", title="Thinking Time Varies by Skill and Time Control") +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

@fig-time-allocation is in my opinion one of the most interesting. It breaks down the time-per-move from @fig-time into the amount of time allocated in opening, middlegame, and endgame phases.^[Game phases are defined by the Stockfish chess engine rather than a consistent standard.] Regardless of skill or time control, the consistent throughline is that players spend the most time thinking in the middlegame and the least in the opening, with the endgame in the middle. This is about what I'd expect. The opening has the fewest possible permutations, and is often studied and memorized, and while the end- and middlegames are similar, the former tends to be defined when fewer pieces remain on the board, which reduces complexity and thus thinking time. It sugggests factoring by gamestate may also be helpful when modeling move time as an outcome.

```{r Allocation Profile by Rating}
#| label: fig-time-allocation
#| fig-cap: "All three time controls have distinct time-usage patterns when game phases are additionally factored in."
#| fig-align: center

phase <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    FLOOR(avg_rating/200)*200 AS rating_bin,
    AVG(avg_time_opening) AS t_open,
    AVG(avg_time_middlegame) AS t_mid,
    AVG(avg_time_endgame) AS t_end,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 10000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

phase_long <- phase |>
  pivot_longer(cols=c(t_open, t_mid, t_end),
                      names_to="phase", values_to="sec")

ggplot(phase_long, aes(x=rating_bin, y=sec, color=phase, group=phase)) +
  geom_line() +
  facet_wrap(~factor(tc_bucket, levels = c("blitz", "rapid", "classical")), scales="free_y", labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))) +
  labs(x="Rating", y="Seconds per Move", color = "Phase",
       title="Phase Time Allocation by Rating") +
  scale_color_few(labels = c("Endgame", "Middlegame", "Opening")) +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


### Opening Diversity

@fig-strata is the strangest of the figures I've chosen to share here. It's also the most complex, faceting by rating bin and coloring the plot's lines by number of games played. For the most part, players' month-to-month rating seems to remain mostly stable regardless of opening diversity. A few patterns remain consistent, however: (1) At lower ratings, opening diversity has a minor-yet-positive relationship with rating change, indicating slight improvement on average for players with a larger repertoire. This relationship seems to falls off once a player hits a rating of around 1800. And (2) the left tail of each plot tends to skew negative, regardless of where the plot is centered. I'm not entirely sure how useful any of this information will be for modeling, but I felt it was an interesting enough visualization to share.

```{r Div Strata}
#| label: fig-strata
#| fig-cap: "Line color indicates number of games played, while the x-axis show the number of unique openings a player used within a given month."
#| fig-align: center
#| fig-height: 6
#| fig-width: 8

div_strata <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, opening_diversity, games, avg_rating
    FROM read_parquet(%s)
    WHERE opening_diversity IS NOT NULL
      AND games IS NOT NULL
      AND avg_rating IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating
    FROM x
  ),
  z AS (
    SELECT
      CAST(FLOOR(avg_rating/200)*200 AS INTEGER) AS rating_bin,
      CASE
        WHEN games < 10 THEN '<10'
        WHEN games < 25 THEN '10–24'
        WHEN games < 60 THEN '25–59'
        ELSE '60+'
      END AS games_bin,
      opening_diversity,
      (next_rating - avg_rating) AS d_rating_next
    FROM y
    WHERE next_rating IS NOT NULL
  )
  SELECT
    rating_bin,
    games_bin,
    opening_diversity,
    AVG(d_rating_next) AS mean_d,
    COUNT(*) AS n
  FROM z
  GROUP BY rating_bin, games_bin, opening_diversity
  HAVING n >= 5
  ORDER BY rating_bin, games_bin, opening_diversity
", qpath(paths$pm)))

div_strata <- div_strata |>
  mutate(games_bin = factor(games_bin, levels = c("60+", "25–59", "10–24", "<10")))

ggplot(div_strata, aes(x=opening_diversity, y=mean_d, color=games_bin)) +
  geom_line(alpha = 1) +
  facet_wrap(~rating_bin, scales="free_y") +
  labs(x="Opening Diversity (Within-Month)", y="Mean Next-Month Δ Rating",
       color="Games",
       title="Opening Diversity vs Improvement",
       subtitle = "Stratified by Games and Rating") +
  scale_color_few() +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```


<!--
\newpage
-->


## Methodology
<!--
Describe the methods used to analyze the data. Do not discuss the analysis itself. Make sure to discuss any train/test data splitting here.
-->

At the core of this project are three themes, as listed in the @sec-background: (1) descriptive metagame dynamics, (2) how time constraints relate to behavior and performance, and (3) characterization of player types or styles. The analysis is strictly non-causal; models are used for description, association, and prediction under out-of-sample evaluation.

When splitting the data into train and test sets, all predictive modeling uses forward-in-time splits to avoid data leakage. This means that the training set contains earlier months in the sample window, while the validation/test set are a contiguous block of later months that were held out from the training data. This split is applied at the month level so that predictors from any given month are never trained using information from future months. For player progression tasks, where the target is a future rating change, using time-forward splits is essential for keeping evaluation realistic.

When appropriate, models are additionally checked for robustness under time-control and rating-bin stratification, to ensure performance isn't driven by a single skill or time-control band.

### Data Tranformations
<!--
Describe any transformation you did to the data to assist in the analyzing of the data such as creating dummy variables.
-->

Transformations serve to stabilize noisy estimates, make quantities comparable across players and contexts, and construct interpretable features aligned with all three research topics. Different transformations are applied for different purposes.

#### Aggregation-Level Alignment {#sec-aggregation}

Rather than a single, monolithic dataset, this project handles several datasets aligned to differing observational units. Strictly speaking, this alignment occurs during the data pipeline itself rather than a formal, pre-modeling cleaning step but it's important to note nonetheless. Broadly, the observational units for each table can be separated into four categories: metagame (population-level games), player outcomes (player-level games), time usage, and opening repertoire. This, however, is an oversimplification; please refer to @sec-data for more detail.

#### Binning and Standardization

To support nonparametric EDA and stable estimation, rating is often binned for plots and stratified summaries— several of which have already been featured in @sec-eda. For clustering (player type/style discovery), features are standardized using z-scores so that variables measured on different scales (e.g., seconds, proportions, counts) contribute in a comparable manner.

#### Exposure Weighting and Time-Control-Collapsed Quantities

As displayed by @fig-coverage in @sec-eda, in some contexts it's justifiable to collapse observations across time controls, particularly where classical is concerned. When a model needs 'overall' player-month measures rather than those specific to a given time control, overall rating is constructed as a games-weighted average of `avg_rating` across time controls. Similarly, overall time-trouble rate becomes an average of `low_time_rate` across time controls, weighted by move rows. This prevents the volume disparity between time controls from disproportionately influencing aggregated quantities.

#### Reliability Thresholds

Many variables are rates estimated from finite samples. Often, these sample sizes are extremely small, increasing volatility. In order to reduce such noise, minimum thresholds for observation count may be applied; player-month observations may require a minimum number of games per player, time-use observations a number of move-rows, and opening-level estimates a number of games per group. This filtering may still be applied even if the model or models being fit don't explicitly require it to function (or as part of a fundamental model assumption) to ensure that modeled signals are not dominated by small-sample volatility.

#### Derived Features

A set of interpetable style features is constructed from player-opening and time-use tables for use in type discovery and player-progression models. Similarly to @sec-aggregation, some of these features were created during the data pipeline itself, while others are derived directly from them. These include:

- Repertoire breath, measured by `opening_diversity`.
- Repertoire concentration, computed as a share of games in the most-played opening from `player_opening_month`.
- Opening switching, indicated by whether `top_opening` differs from the previous month.
- Repertoire churn, measured jointly as:
  - The distance between consecutive-month opening-share vectors.
  - The share of games in brand new openings relative to the month prior.
- Time-pressure behavior, based on a combination of `low_time_rate` plus summaries like `avg_time_spent`, `p90_time_spent`, and `avg_time_opening`.

For more information on pre-existing features, see @sec-data.

#### Missingness

Finally, phase-specific time metrics are often missing for structural reasons. Almost always, the phase missing is endgame, as many games never reach a computer-defined endgame due to resignation or blundered, early losses. Models that use phase variables either include explicit missingness handling or analyze phases separately with appropriate inclusion rules.

### Types of Models {#sec-types-of-models}
Because this project aims to explore a variety of questions by modeling on multiple datasets, strictly categorizing models by family makes little sense.^[Here, I define "model family" as sets of  models operating under similar principles and fundamental assumptions (e.g., general linear models, tree-based models, etc.), whereas "model category" refers to the five groupings of models I specifically define for this paper.] Were there a single theme and associated dataset, doing so would make it easy to illustrate exactly how each type of model contributes to the project's singular goal. In this case, however, I think it more appropriate to categorize models by their goal; to define them by the questions they intend to answer and the methods they'll use in seeking them.^[It should be noted that this form of model categorization allows for slight overlap between model families; some families may be used by multiple categories of model.] 


#### Metagame Dynamics {#sec-metagame-dynamics}

<!--
Temporal dynamics and environment (time is the key axis)
-->

The goal of metagame models are to quantify how aggregate metagame metrics vary across factors such as time and format. The primary observational unit is `meta_opening_month_tc`, while response variables include `share` and outcome rates (e.g., `white_win_rate`). The model families which fall under this category include generalized additive models (GAM) and weighted linear models (WLM), as well as multinomial models. 

GAM and WLM models are useful because they produce simple, easy-to-read trend estimates. The incorporation of weights let high-support opening-month cells impact the fit more than small, noisy cells, and GAM-style smoothing allows for nonlinear time trends without forcing a specific form. Both are strong at tracking a small set of common openings over time (which, given the EDA, is very relevant #ref) while remaining easy to justify and visualize. Importantly, because opening shares are bounded and compositional, these models can only ever approximate the true structure of opening share, meaning that results will necessarily be biased or even potentially spurious.

Multinomial models exist to solve the problem of compositional variables. By modeling on log-odds and log-ratios, they avoid predicting impossible shares ($\notin 0 < x < 1$); in essence, they're capable of implicitly understanding that if one opening goes up, others must go down. This provides interpretable effects for opening adoption relative to a specific baseline, meaning they're best when relative substitution between openings is of import. Multinomial models tend to work better when modeling a large portion of a distribution (rather than just the top k), but as a consequence are more complicated to both implement and interpret.

Combined, these models attempt to ask questions about the state of the metagame – the Lichess 'ecosystem' –  and its evolution rather than just fitting adoption curves. 

#### Opening Popularity {#sec-opening-popularity}

<!--
Skill-conditioned preferences (rating is the key axis). Update first paragraph now that we're using `pom_tc` instead.
-->

The goal of opening-popularity models is to describe how opening preferences differ by player skill and time control. You'll notice that this goal shares substantial overlap with that of metagame models, and indeed, opening popularity is a subset of metagame dynamics. The core difference between the two is that opening-popularity models exclusively target the adoption rate of openings at the expense of all other metagame descriptors; they care not for the diversity of the opening pool, or how outcome rates, game length, and rating gaps shift, for instance. While both use `meta_opening_month_tc` as an observational unit, popularity models prefer to treat each opening separately (top-k) or as a compositional set focused on `share`. Models in this category include multinomial models, count models, and weighted regression.

Contrary to the metagame category, multinomial models are front and center for opening popularity because the target is explicitly compositional. Metagame models seek to observe large-scale patterns more than anything, so in their case it's possible to get away with using multinomials as a backup for increased accuracy, but when it comes to opening popularity there's no real excuse. Multinomial choice models build player-game as a categorical choice, producing probabilities of choosing each opening as a function of things like rating and time control. They're computationally heavy, but are worth the price if they have access to enough data.

Count models such as poisson and negative binomial are a simpler answer to the same problem, treating observed opening use as counts out of a total number of games. By applying an offset (e.g., taking the log of total games), they can effectively model rates while remaining in count space. Negative binomial can handle overdispersion, which is common in count data. Their main advantage over multinomial models is their ability to produce interpretable, multiplicative effects, or rate ratios, which allows for another angle of analysis.

Finally, weighted regression on `share` takes the 'simple is best' philosophy to an extreme. Just as in the prior section, they don't enforce composition, meaning they're liable to misbehave at the boundaries if used for extrapolation; they can't replace multinomial or count models and only serve as an approximation. They remain for much the same reasons as before: they're easy to implement, easy to explain, and good when the only goal is a half-decent descriptive model focused on top openings. Weighted regression certainly won't be the star of the show, but it's such a simple addition that it costs nothing to include.

Together, these models seek to hone in on opening popularity with a higher degree of specificity than the metagame models, each through varying levels of complexity.

#### Time Use {#sec-time-use}
Time use models attempt to characterize the way time allocation varies across skill and time control. Observational units are `player_time_month_tc` for player-level behavior and `moves_month_tc` for macro behavior. Response variables include time spent, time management, pressure, and phase allocation, while predictors primarily consist of `avg_rating` and `tc_bucket`. Three related pairs of models populate this category: additive- and interaction regression, robust- and quantile regression, and beta- and logistic-style modeling.

Additive regression and interaction regression provide a clean separation between skill effects and constraint effects. Additive regression assumes each predictor contributes its own effect, and that those effects simply add up, best for when predictors are believed to act independently of one another. They also tend to be a substitute for interaction regression when there's not enough data. Interaction regression, as the name implies, allows the effect of one predictor to depend on the level of another predictor. Both are relatively easy to interpret.

Robust regression and quantile regression are useful in situations when ordinary least squares (OLS) is a poor fit. Notably, both solve different problems. Robust regression still estimates the conditional mean like OLS, but makes it less sensitive to outliers, whereas quantile regression breaks the mold by estimating a conditional quantile (e.g., median, 90th percentile, etc.) rather than the mean. As a consequence, they must also be interpreted slightly differently, as they ask how predictors shift a percentile rather than the mean. They're valuable for modeling time use because time-related variables often present heavy tails, which these models are better equipped to handle.

Beta and logistic modeling are along for the ride because of a specific problem. Namely, `low_time_rate` is bounded and almost always piles up at 0. Both beta and logit-style approaches keep predictions bounded in [0, 1], handling proportion structure better than OLS, the core difference between the two being that logistic modeling uses a binary response (or a count of successes out of a total number of trials), while beta modeling uses a continuous proportion between 0 and 1. By binning low-time rate, it's possible to produce stable summaries that avoid smoothing failures, making for a more-easily-defensible model

Together, these models thoroughly explore several different methods of time-use analysis. The reason this section contains three pairs of complimentary models rather than just the 'better' of each pair is that the EDA did not produce conclusive enough evidence to justify jumping straight to a smaller subset of model options; the data is complex enough that a simple visual inspection was insufficient for ruling out one choice or the other, and so the most reliable solution is to test them all.

#### Player Progression {#sec-player-progression}
Player-progression models are straightforward: They wish to relate a player's current-month behavior to future rating change. The purpose of this category is simple, asking if it's possible to isolate certain factors that prove to have a marked effect on a player's growth— positive *or* negative. The baseline observational unit for these models is `player_month`, which can be stratified by time control (`player_month_tc`) or joined to other tables for additional information (e.g., `player_time_month_tc` or `player_opening_month_tc`). The response variables are derived from `avg_rating` and all represent $\Delta rating$ in some form. Horizon spans may differ (i.e., rating change over spans greater than a single month), and constructing a stability model using $|\Delta|$ is also possible. Predictors include the number of games and moves played during a given timespan, style proxies, and time-use features, controlled by baseline rating and other fixed effects. Comprising this category are model families such as regularized linear models, mixed-/fixed effects models, and tree-based regression.

Regularized linear models (ridge, lasso, elastic net) help account for the fact that this category's predictors are correlated. Regularization stabilizes these estimates. Ridge regression is good for prediction under multicollinearity, and lasso, though perhaps less relevant here is good for feature selection. Elastic net is a hybrid of the two. As a whole, they should provide a strong baseline predictive model without sacrificing interpretability.

Fixed effects models and mixed (random) effects models are good at handling longitudinal data, though they differ in how they handle time-invariant predictors. The former controls for time-invariant differences without attempting to predict their effects, while the latter tries to estimate said effects. Fixed effects push towards within-player identification, while mixed effects will pool players to improve stability. Crucially, mixed effects models assume that unit intercepts are uncorrelated to the regressors, which can't be reasonably assumed with this data. Fixed effects models are likely to be the better performing of the two (if non-fixed models even pass muster at all), though the implementation of something like a correlated random effects (Mundlak) model might help bridge the gap.

Tree-based regression both naturally captures nonlinearities and thresholds in the data and automatically handles interactions without explicit specification. They're best for exploratory prediction and detecting complex feature relationships. As with all tree-based models, this comes at the cost of interpretability; they require interpretation tools such as feature importance, partial dependence plots, individual conditional expectation plots, et cetera, and even then struggle to match the interpretability of more-straightforward models.

Together, these models account for the varying facets of longitudinal player-level modeling in order to identify the best angle to attack the problem from.

#### Player Categorization {#sec-player-categorization}
As the name implies, player-categorization models want to categorize players. More specifically, they seek to identify clusters using behavioral features, stratifying by rating and format, and observing for stability. This is by far the most nebulous of the model categories because they're primarily exploratory in nature; rather than beginning with pre-defined player archetypes and asking how their performance differs, these models ask which style clusters (if any) can be identified in the first place. The most straightforward observational unit is a join between `player_month_tc`, `player_time_month_tc`, and repertoire-derived features, and inputs include performance, outcomes, repertoire breadth, time use, and general context. The core model families in this category are dimensionality reduction and clustering, though additional modeling becomes possible after playstyle has been categorized.

Dimensionality reduction methods such as principal component analysis (PCA) and uniform manifold approximation and projection (UMAP) provide low-dimensional representation. PCA yields interpretable axes and helps with noise reduction, whereas UMAP is useful for visual separations and nonlinear manifolds. PCA is a linear method which tries to preserve large-scale structure, while UMAP is a nonlinear method prefers preserving cluster separation. Dimensionality reduction is good at for summarizing multi-feature space *before* clustering or visualization, which should prove helpful in the case of a very broad 'style space'.

Clustering methods (e.g., k-means, hierarchical, GMM) are potentially capable of turning continuous player-behavior into discrete, describable archetypes. K-means clustering is fast, but requires tuning towards an unknown k. Hierarchical clustering is tree-based and doesn't require a known k, but requires a dendrogram for interpretation, and Gaussian mixture models (GMM) creates clusters of varying sizes with associated membership probabilities, which is useful if player archetypes overlap. 

Though perhaps not within the scope of this project, dimensionality reduction and clustering are notable as a potential method of creating discrete types for downstream modeling. Once each player-month has a type, it becomes possible to study stability and movement between types or perform supervised follow-up using type labels (e.g., as predictors of progression).
  
Together, these models explore the bounds of player-archetype/playstyle categorization.

### Model Selection
Because this project does not intend to settle on a single, final, 'best' model, I have not outlined an exact process for model selection. Rather, the sheer quantity and breadth of models tested naturally pushes toward a broader set of more-general rules to follow. The rules I have defined are as follows:

- Start with defensible baselines.
  - Make the simplest possible model (or set of models) in a given category that fulfill reasonable standards for transparency, methodology, and fulfilled model assumptions.
  - Compare increasingly-complex models to their baseline counterparts when assessing their worth.
- Prefer simpler models when performance is comparable.
  - When two models perform similarly, the one with fewer assumptions and clearer interpretation is preferable for reporting.
- Check stability to thresholds and stratification.
  - Key results must be checked for sensitivity to minimum thresholds and to whether analyses are stratified by time control and/or rating bin.
  - If one model performs worse than another but the better-performing model fails to pass for stability, then the 'worse', more-stable model should be preferred.
- For clustering specifically, select solutions by stability and interpretability.
  - Candidate clusterings should be preferred when feature profiles are distinct and assignments are reasonably stable across adjacent months or within rating strata.

Because no single framework is sufficient for selecting between all models tested, a large portion of the model-selection process will instead be covered in @sec-results.^[As a note, I heavily suspect I'll be reworking this section before the final draft of this paper.]

<!--
\newpage
-->

## Modeling {#sec-modeling}

<!--
Professor described this as "a bit of a rough draft of the results section" where he "[expects] completed models with some initial comparisons between them and what models are showing the most usefulness", whereas "the later results section should include this comparison but also include analysis and inference from the final selected model(s)".
-->

This section compares the fitted models introduced in @sec-types-of-models and identifies the specifications retained for the remainder of the analysis. The goal here is not to present substantive inference, but to narrow a larger set of candidate models to a smaller set of primary specifications for each modeling task.

Model retention is based on four considerations: inferential suitability, held-out predictive adequacy, robustness to reasonable specification changes, and interpretability. These criteria do not apply identically to every modeling problem. In some cases, the most useful retained model is the one that supports formal inference, while in others the retained model is better understood as descriptive or predictive only. This distinction is especially important for methods such as clustering, which are useful for summarizing structure in the data but are not intended as a basis for inferential claims.

Because a large number of models were fit during development, the body of this section focuses on the specifications most relevant to downstream analysis. Additional model details, alternative fits, and supplemental comparisons are reported in @sec-appendix.

### Metagame Dynamics {#sec-metagame-dynamics-2}

Metagame dynamics were modeled at two related levels. First, aggregate ecosystem and game-environment variables were modeled at the month-by-time-control level using weighted linear models. These macro outcomes included opening concentration and diversity – measured by the Herfindahl-Hirschman index (HHI) and the effective number of openings – along with draw rate, upset rate, and average moves. Because these series are observed over time, macro inference relies on Newey-West (HAC) standard errors. Second, Opening composition itself was modeled separately by time control using multinomial specifications, with weighted fixed-effects and GAM share models retained only as descriptive comparisons rather than as primary inferential models.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.5cm} >{\raggedright\arraybackslash}p{3.0cm} >{\raggedleft\arraybackslash}X}
\caption{Held-out predictive error for the retained macro metagame models.}\label{tab:metagame-macro-model-comparison}\\
\toprule
\textbf{Outcome} & \textbf{Time Control} & \textbf{Held-Out Error} \\
\midrule
\endfirsthead

\toprule
\textbf{Outcome} & \textbf{Time Control} & \textbf{Held-Out Error} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

HHI Concentration & All Formats & $3.70 \times 10^{-5}$ \\
Effective Number of Openings ($N_{\mathrm{eff}}$) & All Formats & 1.48 \\
Draw Rate & All Formats & 0.00056 \\
Upset Rate & All Formats & 0.00158 \\
Average Moves & All Formats & 0.081 \\
\end{xltabular}
```

Table \ref{tab:metagame-macro-model-comparison} summarizes held-out predictive error for the retained macro models. These weighted linear specifications fit the aggregate metagame series well, with especially small held-out error for concentration and game-environment outcomes. This level of predictive accuracy, together with the interpretability of the resulting time trends, was sufficient to retain the weighted linear models as the primary inferential framework for ecosystem-level metagame claims. 

The retained macro models point most clearly to declining concentration and increasing diversity in opening usage. HHI shows a negative baseline time trend, with additional negative time interactions for classical and rapid, indicating that opening usage became less concentrated over time in all three formats and that the decline was steeper outside blitz. The effective number of openings shows a corresponding positive pattern: The baseline time trend and the classical and rapid interactions are all positive, which indicates broad expansion of the opening ecosystem over time. By contrast, draw rate and average moves show no statistically reliable linear time trends, while upset rate shows a modest positive baseline trend that does not generalize as clearly across time controls. Table \ref{tab:metagame-macro-estimates} reports the estimated macro time effects underlying these retained claims.


<!-- Need to rework tables so that the scientific notation convention is properly followed. -->

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{2.6cm} >{\raggedright\arraybackslash}p{3.5cm} >{\raggedleft\arraybackslash}p{2.1cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}X}
\caption{Estimated time effects from the retained macro metagame models.}\label{tab:metagame-macro-estimates}\\
\toprule
\textbf{Outcome} & \textbf{Effect} & \textbf{Estimate} & \textbf{SE} & \textbf{\textit{p}-Value} \\
\midrule
\endfirsthead

\toprule
\textbf{Outcome} & \textbf{Effect} & \textbf{Estimate} & \textbf{SE} & \textbf{\textit{p}-Value} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Draw Rate & Baseline Time Trend & $-2.322\times10^{-5}$ & $1.849\times10^{-5}$ & 0.215 \\
Draw Rate & Classical $\times$ Time & $2.615\times10^{-5}$ & $1.073\times10^{-4}$ & 0.808 \\
Draw Rate & Rapid $\times$ Time & $9.697\times10^{-6}$ & $1.909\times10^{-5}$ & 0.614 \\

HHI Concentration & Baseline Time Trend & $-2.283\times10^{-6}$ & $6.890\times10^{-7}$ & 0.002 \\
HHI Concentration & Classical $\times$ Time & $-1.694\times10^{-5}$ & $7.154\times10^{-6}$ & 0.022 \\
HHI Concentration & Rapid $\times$ Time & $-7.440\times10^{-6}$ & $1.502\times10^{-6}$ & $< 0.001$ \\

Effective Number of Openings & Baseline Time Trend & $1.021\times10^{-1}$ & $3.096\times10^{-2}$ & 0.002 \\
Effective Number of Openings & Classical $\times$ Time & $5.556\times10^{-1}$ & $2.485\times10^{-1}$ & 0.030 \\
Effective Number of Openings & Rapid $\times$ Time & $2.411\times10^{-1}$ & $5.453\times10^{-2}$ & $< 0.001$ \\

Average Moves & Baseline Time Trend & $-3.957\times10^{-4}$ & $2.461\times10^{-3}$ & 0.873 \\
Average Moves & Classical $\times$ Time & $-1.816\times10^{-2}$ & $1.526\times10^{-2}$ & 0.240 \\
Average Moves & Rapid $\times$ Time & $-3.840\times10^{-3}$ & $3.455\times10^{-3}$ & 0.272 \\

Upset Rate & Baseline Time Trend & $9.540\times10^{-5}$ & $4.099\times10^{-5}$ & 0.024 \\
Upset Rate & Classical $\times$ Time & $1.011\times10^{-4}$ & $2.997\times10^{-4}$ & 0.737 \\
Upset Rate & Rapid $\times$ Time & $7.481\times10^{-5}$ & $6.067\times10^{-5}$ & 0.223 \\
\end{xltabular}
```

Opening composition required a different primary model because opening shares are inherently compositional. Table \ref{tab:metagame-composition-model-comparison} compares the descriptive share models with the multinomial specifications fit separately by time control. The multinomial models were retained as the primary inferential basis for compositional change because they respect the simplex structure of opening shares, whereas the weighted fixed-effects and GAM models treat openings one at a time and therefore cannot fully represent substitution across the opening set. Held-out log loss for the multinomial models was similar across formats at 0.910 in blitz, 0.923 in rapid, and 0.940 in classical, suggesting broadly comparable predictive performance across time controls. At the same time, the descriptive share models produced nearly identical held-out error in blitz and rapid, indicating that they capture much of the same visible trend structure even though they aren't appropriate as the main inferential basis.^[Lorem ipsum.]

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.7cm} >{\raggedright\arraybackslash}p{2.8cm} >{\raggedright\arraybackslash}p{2.3cm} >{\raggedleft\arraybackslash}X}
\caption{Held-out predictive error for opening-composition models.}\label{tab:metagame-composition-model-comparison}\\
\toprule
\textbf{Target} & \textbf{Model} & \textbf{Time Control} & \textbf{Held-Out Error} \\
\midrule
\endfirsthead

\toprule
\textbf{Target} & \textbf{Model} & \textbf{Time Control} & \textbf{Held-Out Error} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Opening Share & FEOLS & Blitz & 0.00243 \\
Opening Share & GAM & Blitz & 0.00243 \\
Opening Share & FEOLS & Rapid & 0.00239 \\
Opening Share & GAM & Rapid & 0.00237 \\
Opening Composition & Multinomial & Blitz & 0.910 \\
Opening Composition & Multinomial & Classical & 0.940 \\
Opening Composition & Multinomial & Rapid & 0.923 \\
\end{xltabular}
```


Robustness checks suggest that the substantive multinomial conclusions are not highly sensitive to the exact top-$k$ threshold used to define the retained opening set. As shown in Table \ref{tab:metagame-multinomial-robustness}, increasing $k$ predictably worsened held-out log loss by making the prediction problem more demanding, but the broad directional pattern among the largest movers remained stable. In blitz, the pooled "Other" category consistently increased, while the main negative mover remained a French Defense variation once the retained opening set was sufficiently large. This suggests that the composition results reflect broad substitution within the opening ecosystem rather than dependence on a single threshold choice.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.4cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}p{2.0cm} X X}
\caption{Robustness of the Blitz multinomial model to the number of retained opening categories.}\label{tab:metagame-multinomial-robustness}\\
\toprule
\textbf{$k$} & \textbf{Categories} & \textbf{Log Loss} & \textbf{Top Increase} & \textbf{Top Decrease} \\
\midrule
\endfirsthead

\toprule
\textbf{$k$} & \textbf{Categories} & \textbf{Log Loss} & \textbf{Top Increase} & \textbf{Top Decrease} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

8 & 9 & 0.674 & \texttt{Other} ($+0.000191$) & \texttt{Philidor Defense} ($-0.000162$) \\
12 & 13 & 0.910 & \texttt{Other} ($+0.000951$) & \texttt{French Defense: Knight Variation} ($-0.000771$) \\
16 & 17 & 1.127 & \texttt{Other} ($+0.001047$) & \texttt{French Defense: Knight Variation} ($-0.000771$) \\
\end{xltabular}
```

Overall, these comparisons support a two-part retained strategy for metagame dynamics: weighted linear models for macro/ecosystem-level claims, and multinomial models for claims about opening substitution and compositional change. Weighted share regressions and GAMs remain useful descriptive checks, but won't be carried forward as the primary inferential basis for this model category.

### Opening Popularity {#sec-opening-popularity-2}

Opening popularity was modeled primarily as a compositional problem. The main inferential candidate was a pooled multinomial model fit to month-by-time-control-by-rating-bin cells, with openings aggregated into the twelve most common training period openings plus an "Other" category after sparse cells were excluded. This specification uses rating bin, time control, month index, and a rating-by-time-control interaction to predict opening composition. Per-opening negative binomial and weighted least squares models were also fit as lower-complexity sensitivity checks, but these models treat openings separately and therefore cannot coherently represent substitution across the opening set.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.6cm} >{\raggedright\arraybackslash}p{3.0cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.2cm}}
\caption{Held-out predictive error for opening-popularity models.}\label{tab:opening-popularity-model-comparison}\\
\toprule
\textbf{Model} & \textbf{Target} & \textbf{Test Rows} & \textbf{Held-Out Error} \\
\midrule
\endfirsthead

\toprule
\textbf{Model} & \textbf{Target} & \textbf{Test Rows} & \textbf{Held-Out Error} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Pooled Multinomial & Opening Composition & 81 & 0.893 \\
Negative Binomial (Per Opening) & Opening-Use Rate & 80--81 & 0.0021--0.0097 \\
Weighted Least Squares (Per Opening) & Opening Share & 80--81 & 0.0024--0.0046 \\
\end{xltabular}
```


Table \ref{tab:opening-popularity-model-comparison} compares held-out predictive error across the candidate model families. The pooled multinomial was retained as the primary model for opening popularity because it directly respects the compositional structure of opening choice while still providing competitive held-out performance. Its average per-game log loss on the held-out test set was 0.893 across 81 test rows. The negative binomial and weighted least squares specifications were useful for describing individual openings, but because they model by one opening at a time, they remain only as supporting checks.

The retained multinomial model produced plausible and structured rating-conditioned opening profiles. Across both blitz and rapid, the "Other" category remained the most likely predicted outcome at every representative rating bin, and its fitted share increased steadily with rating. Among the explicitly modeled openings, lower-rated bins placed relatively more mass on openings such as the Philidor Defense, Scandinavian Defense, and Queen's Pawn Game, whereas higher-rated bins shifted more strongly toward the Caro-Kann Defense and the Accelerated London System. These patterns suggest that the retained model is capturing systematic variation by player strength rather than only fitting noise in sparse cells. Table \ref{tab:opening-popularity-rating-profile} summarizes the representative rating-conditioned predictions, and Table \ref{tab:opening-popularity-time-change} summarizes the largest predicted opening-share changes over time at rating 1600. In both blitz and rapid, the largest increase is absorbed by the residual "Other" category, while the largest named increase is the Accelerated London System and the largest named decrease is the French Defense: Knight Variation, indicating modest but directionally consistent compositional change over time.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.8cm} >{\raggedright\arraybackslash}p{1.8cm} X X X}
\caption{Representative rating-conditioned predictions from the pooled multinomial opening-popularity model.}\label{tab:opening-popularity-rating-profile}\\
\toprule
\textbf{Rating Bin} & \textbf{Time Control} & \textbf{Most Likely Opening} & \textbf{Second Most Likely} & \textbf{Third Most Likely} \\
\midrule
\endfirsthead

\toprule
\textbf{Rating Bin} & \textbf{Time Control} & \textbf{Most Likely Opening} & \textbf{Second Most Likely} & \textbf{Third Most Likely} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

800  & Blitz & \texttt{Other} (0.732) & \texttt{Philidor Defense} (0.041) & \texttt{Queen's Pawn Game} (0.034) \\
1600 & Blitz & \texttt{Other} (0.835) & \texttt{Queen's Pawn Game} (0.023) & \texttt{Queen's Pawn Game: Accelerated London System} (0.017) \\
2400 & Blitz & \texttt{Other} (0.892) & \texttt{Caro-Kann Defense} (0.014) & \texttt{Queen's Pawn Game} (0.014) \\
800  & Rapid & \texttt{Other} (0.768) & \texttt{Scandinavian Defense} (0.034) & \texttt{Philidor Defense} (0.034) \\
1600 & Rapid & \texttt{Other} (0.840) & \texttt{Queen's Pawn Game} (0.022) & \texttt{Philidor Defense} (0.022) \\
2400 & Rapid & \texttt{Other} (0.877) & \texttt{Caro-Kann Defense} (0.018) & \texttt{Queen's Pawn Game: Accelerated London System} (0.016) \\
\end{xltabular}
```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}p{1.6cm} X X}
\caption{Largest predicted opening-share changes over time from the pooled multinomial model.}\label{tab:opening-popularity-time-change}\\
\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Largest Positive Movers} & \textbf{Largest Negative Movers} \\
\midrule
\endfirsthead

\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Largest Positive Movers} & \textbf{Largest Negative Movers} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Blitz & 1600 & \texttt{Other} ($+0.0070$); \texttt{Queen's Pawn Game: Accelerated London System} ($+0.0023$); \texttt{Horwitz Defense} ($+0.0010$) & \texttt{French Defense: Knight Variation} ($-0.0041$); \texttt{Queen's Pawn Game} ($-0.0014$); \texttt{Pirc Defense} ($-0.0013$) \\
Rapid & 1600 & \texttt{Other} ($+0.0066$); \texttt{Queen's Pawn Game: Accelerated London System} ($+0.0029$); \texttt{Horwitz Defense} ($+0.0008$) & \texttt{French Defense: Knight Variation} ($-0.0038$); \texttt{Philidor Defense} ($-0.0014$); \texttt{Queen's Pawn Game} ($-0.0014$) \\
\end{xltabular}
```

Robustness checks generally supported the retained multinomial specification. As shown in Table \ref{tab:opening-popularity-robustness}, allowing rating to vary nonlinearly with a spline slightly improved held-out fit relative to the linear rating form, whereas replacing linear time with a spline slightly worsened fit, so the simpler linear-time specification was retained. Varying the number of explicitly modeled openings produced expected results: Smaller category sets were easier to predict, while expanding from $k = 12$ to $k = 16$ increased held-out log loss. Where one-opening models were available, their estimated time-direction patterns also broadly agreed with the multinomial results for the most common blitz openings, which further increased confidence in the retained specification.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{2.5cm} >{\raggedleft\arraybackslash}p{2.0cm}}
\caption{Robustness checks for the pooled multinomial opening-popularity model.}\label{tab:opening-popularity-robustness}\\
\toprule
\textbf{Check} & \textbf{Specification} & \textbf{Log Loss} \\
\midrule
\endfirsthead

\toprule
\textbf{Check} & \textbf{Specification} & \textbf{Log Loss} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Rating Functional Form & Linear Rating & 0.893 \\
Rating Functional Form & Spline Rating & 0.890 \\
Time Functional Form & Linear Time & 0.893 \\
Time Functional Form & Spline Time & 0.895 \\
Top-$k$ Threshold & $k = 8$ & 0.666 \\
Top-$k$ Threshold & $k = 12$ & 0.893 \\
Top-$k$ Threshold & $k = 16$ & 1.111 \\
\end{xltabular}
```


All in all, these comparisons support the pooled multinomial as the primary model for opening popularity. It respects the compositional structure of opening choice, captures meaningful variation across rating and time control, and remains reasonably stable under alternative functional forms and threshold choices. The negative binomial and weighted share models remain valuable as supporting checks, but are not as carried forward as the main inferential basis for this model category.

### Time Use {#sec-time-use-2}

In contrast to the prior model categories, time-use modeling was organized around two related targets: average time use, and low-time frequency. For the former, candidate specifications were compared at both the macro month level and the player-month level. For the latter, the main comparison was between a fractional quasibinomial specification and a binomial approximation. Across these targets, retained models were chosen by balancing held-out predictive adequacy, interpretability, and robustness to modest specification changes.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{3.0cm} >{\raggedleft\arraybackslash}X}
\caption{Held-out predictive error for mean time-use models.}\label{tab:timeuse-mean-model-comparison}\\
\toprule
\textbf{Level} & \textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endfirsthead

\toprule
\textbf{Level} & \textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Macro Month-Level & Additive & 0.149 \\
Macro Month-Level & Interaction & 0.141 \\
Player-Month & Additive & 1.839 \\
Player-Month & Interaction & 1.832 \\
\end{xltabular}
```

Table \ref{tab:timeuse-mean-model-comparison} compares held-out predictive error for the candidate mean time-use models. At both the macro month level and the player-month level, interaction specifications outperformed additive alternatives, so the interaction forms were retained as the primary models for average time use. The improvement in held-out fit was modest, but it was consistent across levels, which suggests that the relationship between skill and time use is not adequately captured by a single common rating slope across time controls. For this reason, the interaction specifications were carried forward as the main basis for subsequent inference on average time use.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.2cm} >{\raggedleft\arraybackslash}X}
\caption{Held-out predictive error for low-time-rate models.}\label{tab:timeuse-lowtime-model-comparison}\\
\toprule
\textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endfirsthead

\toprule
\textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Fractional Quasibinomial & 0.07445 \\
Binomial Approximation & 0.07445 \\
\end{xltabular}
```

Low-time frequency required a different retained model because the outcome is bounded and concentrated near zero. Table \ref{tab:timeuse-lowtime-model-comparison} shows that the fractional quasibinomial and binomial approximation produced effectively identical held-out error. Accordingly, the fractional quasibinomial was retained as the primary specification because it is better aligned with the scale and distribution of the outcome, while the binomial approximation was retained only as a sensitivity check. In other words, predictive performance did not clearly distinguish the two models, so inferential suitability and outcome structure became the deciding considerations.

Robustness checks did not meaningfully overturn either retention decision. For the retained mean models, Table \ref{tab:timeuse-mean-robustness} shows that replacing linear time with a spline at the macro level worsened held-out fit, which supports keeping the simpler linear-time form, while allowing a spline in rating at the player level produced only a small improvement, suggesting at most mild nonlinearity in the skill gradient. For the retained low-time model, Table \ref{tab:timeuse-lowtime-robustness} shows that replacing the linear rating term with a spline produced essentially no change in held-out error. In short, these checks indicate that the retained time-use models are reasonably stable and are not being driven by a restrictive, functional-form choice.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedright\arraybackslash}p{3.0cm} >{\raggedleft\arraybackslash}X}
\caption{Robustness checks for the retained mean time-use models.}\label{tab:timeuse-mean-robustness}\\
\toprule
\textbf{Level} & \textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endfirsthead

\toprule
\textbf{Level} & \textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Macro Mean Time Use & Linear Time & 0.141 \\
Macro Mean Time Use & Spline Time & 0.154 \\
Player Mean Time Use & Linear Rating & 1.832 \\
Player Mean Time Use & Spline Rating & 1.823 \\
\end{xltabular}
```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.4cm} >{\raggedleft\arraybackslash}X}
\caption{Robustness check for the retained low-time-rate model.}\label{tab:timeuse-lowtime-robustness}\\
\toprule
\textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endfirsthead

\toprule
\textbf{Specification} & \textbf{WRMSE} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Fractional Model, Linear Rating & 0.07445 \\
Fractional Model, Spline Rating & 0.07445 \\
\end{xltabular}
```

Accordingly, interaction-based models were retained as the primary specifications for average time use, and the fractional quasibinomial model was retained as the primary specification for low-time frequency. Robust regression, quantile regression, phase-specific mean models, and the binomial approximation remain useful diagnostics and sensitivity checks, but they are not carried forward as the main inferential basis for this subsection.

### Player Progression {#sec-player-progression-2}

Player-progression models were designed to assess whether month-$t$ player behavior could be used to predict subsequent rating movement under a forward-in-time design. Three targets were retained: 1-month rating change, 3-month rating change, and absolute 1-month rating change. As shown in Table \ref{tab:progression-sample-sizes}, the 3-month target is substantially smaller because it requires observing the same player several months later, whereas the 1-month targets preserved the full, filtered sample. This difference in sample size is important for interpreting later comparisons across targets.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.2cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}p{2.0cm} X}
\caption{Sample sizes for player-progression targets after filtering and forward-month splitting.}\label{tab:progression-sample-sizes}\\
\toprule
\textbf{Target} & \textbf{Train} & \textbf{Test} & \textbf{Definition} \\
\midrule
\endfirsthead

\toprule
\textbf{Target} & \textbf{Train} & \textbf{Test} & \textbf{Definition} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

1-Month Change & 99{,}049 & 18{,}166 & $\Delta \text{Rating}_{t+1} = \text{Rating}_{t+1} - \text{Rating}_t$ \\
3-Month Change & 80{,}517 & 6{,}354 & $\Delta \text{Rating}_{t+3} = \text{Rating}_{t+3} - \text{Rating}_t$ \\
Absolute 1-Month Change & 99{,}049 & 18{,}166 & $|\Delta \text{Rating}_{t+1}|$ \\
\end{xltabular}
```

The main inferential candidate in this subsection was a player fixed-effects model fit on the player-month panel. This choice reflects the goal of isolating within-player associations between current-month behavior and later rating movement rather than attributing progression differences to stable cross-player traits. The baselie feature set combined current rating, volume, and repertoire descriptors: average rating, games played, average moves, opening diversity, top-opening concentration, and whether the player switched their most-played opening. Ridge regression and random forest were also fit as predictive benchmarks. Table \ref{tab:progression-predictive-comparison} compares held-out performance across these model families for all three targets.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.0cm} >{\raggedright\arraybackslash}p{2.6cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}X}
\caption{Held-out predictive performance for player-progression models.}\label{tab:progression-predictive-comparison}\\
\toprule
\textbf{Model} & \textbf{Target} & \textbf{RMSE} & \textbf{MAE} & \textbf{Test Rows} \\
\midrule
\endfirsthead

\toprule
\textbf{Model} & \textbf{Target} & \textbf{RMSE} & \textbf{MAE} & \textbf{Test Rows} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Player Fixed Effects & 1-Month Change & 57.87 & 41.31 & 14{,}878 / 18{,}166 \\
Ridge Regression & 1-Month Change & 61.62 & 41.31 & 18{,}166 \\
Random Forest & 1-Month Change & 61.97 & 42.27 & 18{,}166 \\

Player Fixed Effects & 3-Month Change & 70.76 & 50.83 & 5{,}726 / 6{,}354 \\
Ridge Regression & 3-Month Change & 80.64 & 57.59 & 6{,}354 \\
Random Forest & 3-Month Change & 81.38 & 58.42 & 6{,}354 \\

Player Fixed Effects & Absolute 1-Month Change & 41.74 & 28.56 & 14{,}878 / 18{,}166 \\
Ridge Regression & Absolute 1-Month Change & 45.67 & 30.38 & 18{,}166 \\
Random Forest & Absolute 1-Month Change & 45.70 & 30.66 & 18{,}166 \\
\end{xltabular}
```

Across all three targets, the player fixed-effects models outperformed both ridge regression and random forest on held-out data. The advantage was especially clear for the 3-month and absolute 1-month targets, where the gap in both RMSE and MAE was substantial. This made the player fixed-effects family the strongest overall modeling direction in this subsection, as not only was it the most appropriate model family for inference, but also the best-performing family among the tested benchmarks. Accordingly, player fixed effects were retained as the primary progression model.

Additional sensitivity checks suggest that this retained choice is reasonably stable. Table \ref{tab:progression-fe-robustness} shows that adding collapsed time-use features on the common-support sample did not improve performance relative to the corresponding baseline fit, which indicates that the expanded feature set did not add enough information to offset the associated loss of coverage. The same table also shows that replacing the linear rating term with a spline produced only a very small improvement, suggesting that any nonlinearity in the rating gradient is modest at this stage and not large enough to overturn the simpler retained specification.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.2cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}X}
\caption{Robustness checks for the retained 1-month player fixed-effects model.}\label{tab:progression-fe-robustness}\\
\toprule
\textbf{Specification} & \textbf{Train} & \textbf{Predicted / Total Test} & \textbf{RMSE} & \textbf{MAE} \\
\midrule
\endfirsthead

\toprule
\textbf{Specification} & \textbf{Train} & \textbf{Predicted / Total Test} & \textbf{RMSE} & \textbf{MAE} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Baseline & 99{,}049 & 14{,}878 / 18{,}166 & 57.87 & 41.31 \\
Baseline, Common Support & 2{,}015 & 62 / 405 & 27.86 & 21.23 \\
Baseline + Time Use, Common Support & 2{,}015 & 62 / 405 & 28.51 & 21.96 \\
Linear Rating & 99{,}049 & 14{,}878 / 18{,}166 & 57.87 & 41.31 \\
Spline Rating & 99{,}049 & 14{,}878 / 18{,}166 & 57.72 & 41.25 \\
\end{xltabular}
```

A final structural check asked whether common month shocks should be absorbed directly through added month fixed effects. As shown in Table \ref{tab:progression-fe-structure}, doing so increased within-$R^2$ only slightly. This indicates that, while present, shared month-level variation wasn't strong enough to outweigh the usefulness of the simpler specification.The player-month variant was therefore kept solely as a supporting model.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.2cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}X}
\caption{Fixed-effects structure sensitivity for the 1-month progression model.}\label{tab:progression-fe-structure}\\
\toprule
\textbf{Specification} & \textbf{Train} & \textbf{Within-$R^2$} \\
\midrule
\endfirsthead

\toprule
\textbf{Specification} & \textbf{Train} & \textbf{Within-$R^2$} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Player Fixed Effects & 95{,}428 & 0.2696 \\
Player + Month Fixed Effects & 95{,}428 & 0.2744 \\
\end{xltabular}
```

In summary, these comparisons support player fixed effects as the retained specification for player-progression models. They provide the strongest combination of longitudinal appropriateness, inferential usefulness, and held-out predictive performance, while also remaining reasonably stable under the specification changes examined here. Ridge regression and random forest remain useful predictive benchmarks, and the alternative fixed-effects variants remain informative sensitivities, but none are carried forward as the main inferential basis for this subsection.

### Player Categorization {#sec-player-categorization-2}

The goal in this subsection was not to identify latent player “types” in a strong causal or inferential sense, but to retain a clustering specification that usefully summarizes broad behavioral variation across player-month-time-control observations. The retained clustering sample contained 156,901 observations from 25,547 players across 24 months and all three time-control buckets. As summarized in Table \ref{tab:categorization-spec}, repertoire features were available throughout this sample, whereas stable time-use coverage was sparse in the joined design. For that reason, the retained clustering specification was based on standardized core and repertoire variables only, with time-use treated as a possible downstream overlay rather than as an input to the clustering algorithm.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{4.4cm} >{\raggedleft\arraybackslash}X}
\caption{Sample coverage and retained specification for player categorization.}\label{tab:categorization-spec}\\
\toprule
\textbf{Item} & \textbf{Value} \\
\midrule
\endfirsthead

\toprule
\textbf{Item} & \textbf{Value} \\
\midrule
\endhead

\midrule
\multicolumn{2}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Candidate Observations & 156{,}901 \\
Distinct Players & 25{,}547 \\
Distinct Months & 24 \\
Time-Control Levels & 3 \\
Fraction With Repertoire Features & 1.000 \\
Fraction With Time-Use Overlay & 0.017 \\
Retained Clustering Observations & 156{,}901 \\
Clustering Features & 10 \\
Primary Algorithm & \texttt{K-Means} \\
Chosen $k$ & 6 \\
Feature Set & \texttt{Core Plus Repertoire} \\
\end{xltabular}
```

The retained descriptive specification used $k$-means clustering with $k = 6$ on a 10-variable standardized feature set combining performance, outcome, and repertoire proxies. This choice was intended to produce an interpretable segmentation while avoiding the instability that can arise from very high-dimensional or sparsely observed feature spaces. The resulting solution was reasonably balanced, which makes it more useful for profile-based descriptive interpretation. Table \ref{tab:categorization-cluster-sizes} shows that cluster shares range from 0.112 to 0.217, so no single cluster dominates the retained segmentation, and neither is any cluster so small as to be descriptively unhelpful.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}X}
\caption{Cluster sizes for the retained player-categorization solution.}\label{tab:categorization-cluster-sizes}\\
\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Share} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Share} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

3 & 34{,}057 & 0.217 \\
6 & 31{,}620 & 0.202 \\
2 & 28{,}181 & 0.180 \\
5 & 25{,}390 & 0.162 \\
1 & 20{,}107 & 0.128 \\
4 & 17{,}546 & 0.112 \\
\end{xltabular}
```

The retained solution captures meaningful differences across both general performance characteristics and opening-related behavior. Table @ref(tab:categorization-performance-profiles) shows that the clusters differ in average rating, rating differential, average moves, and win rate, indicating that the segmentation is not simply reproducing a single skill ordering. Table \ref{tab:categorization-repertoire-profiles} shows additional differences in opening diversity, concentration, and opening-switching behavior, which suggests that repertoire structure is an important part of the retained clustering solution. In particular, some clusters are characterized by broad and diverse opening usage, whereas others are defined more by concentrated repertoires or high switching rates. Log-transformed activity measures were also included in the retained clustering design to reduce the influence of heavy-tailed count variables (see @sec-appendix).

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.4cm} >{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}X}
\caption{Performance and outcome means for the retained player-categorization solution.}\label{tab:categorization-performance-profiles}\\
\toprule
\textbf{Cluster} & \textbf{Share} & \textbf{Avg. Rating} & \textbf{Avg. Rating Diff.} & \textbf{Avg. Moves} & \textbf{Win Rate} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{Share} & \textbf{Avg. Rating} & \textbf{Avg. Rating Diff.} & \textbf{Avg. Moves} & \textbf{Win Rate} \\
\midrule
\endhead

\midrule
\multicolumn{6}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

3 & 0.217 & 1518.4 & -3.5 & 32.9 & 0.481 \\
6 & 0.202 & 1506.1 & -2.3 & 33.5 & 0.480 \\
2 & 0.180 & 1214.6 & -51.0 & 27.5 & 0.369 \\
5 & 0.162 & 1669.6 & 42.6 & 33.4 & 0.604 \\
1 & 0.128 & 1760.9 & 7.6 & 35.6 & 0.486 \\
4 & 0.112 & 1649.3 & 2.2 & 37.6 & 0.427 \\
\end{xltabular}
```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.4cm} >{\raggedleft\arraybackslash}p{2.4cm} >{\raggedleft\arraybackslash}X}
\caption{Repertoire means for the retained player-categorization solution.}\label{tab:categorization-repertoire-profiles}\\
\toprule
\textbf{Cluster} & \textbf{Share} & \textbf{Opening Diversity} & \textbf{Top Opening Share} & \textbf{Switched Opening} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{Share} & \textbf{Opening Diversity} & \textbf{Top Opening Share} & \textbf{Switched Opening} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

3 & 0.217 & 48.1 & 0.095 & 1.000 \\
6 & 0.202 & 44.1 & 0.169 & 0.006 \\
2 & 0.180 & 16.6 & 0.166 & 0.619 \\
5 & 0.162 & 16.3 & 0.153 & 0.727 \\
1 & 0.128 & 114.8 & 0.088 & 0.443 \\
4 & 0.112 & 17.8 & 0.154 & 0.712 \\
\end{xltabular}
```

Robustness and stability checks support treating the retained solution as useful, but only in a descriptive sense. As shown in Table \ref{tab:categorization-robustness}, the richer core-plus-repertoire feature set produced a more balanced clustering solution than the core-only alternative. Repeated $k$-means fits under the retained specification converged to the same partition, indicating effectively perfect initialization stability. At the same time, agreement with hierarchical clustering and Gaussian mixture clustering was only moderate, silhouette values were modest, and month-to-month agreement for overlapping players was limited. So, while these diagnostics suggest that the retained segmentation captures real structure in the data, they also suggest that this structure is neither sharply separated nor highly persistent over time.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.8cm} >{\raggedright\arraybackslash}p{3.0cm} >{\raggedleft\arraybackslash}X}
\caption{Robustness and stability diagnostics for the retained player-categorization solution.}\label{tab:categorization-robustness}\\
\toprule
\textbf{Diagnostic} & \textbf{Statistic} & \textbf{Value} \\
\midrule
\endfirsthead

\toprule
\textbf{Diagnostic} & \textbf{Statistic} & \textbf{Value} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Feature-Set Robustness & Core-Only Min. Cluster Share & 0.034 \\
Feature-Set Robustness & Core Plus Repertoire Min. Cluster Share & 0.112 \\
Algorithm Agreement & K-Means vs. Hierarchical ARI & 0.321 \\
Algorithm Agreement & K-Means vs. GMM ARI & 0.298 \\
Silhouette & Mean Silhouette & 0.145 \\
Silhouette & Median Silhouette & 0.151 \\
Initialization Stability & Mean ARI & 1.000 \\
Initialization Stability & Min. ARI & 1.000 \\
Initialization Stability & Max. ARI & 1.000 \\
Month-to-Month Stability & Mean ARI & 0.139 \\
Month-to-Month Stability & Median ARI & 0.143 \\
Month-to-Month Stability & Min. ARI & 0.025 \\
Cluster Balance & Number of Clusters & 6 \\
Cluster Balance & Min. Cluster Share & 0.112 \\
Cluster Balance & Max. Cluster Share & 0.217 \\
\end{xltabular}
```

Dimensionality-reduction methods were used in a supporting role rather than as retained models in their own right. PCA provided a more interpretable low-dimensional summary of the dominant axes of variation in the standardized feature space, while UMAP offered a nonlinear view of local neighborhood structure. Their role here is the same as that of the alternative clustering algorithms—  not to replace the retained clustering specification, but to help assess whether the descriptive segmentation corresponds to visible structure in the player-behavior space.

Ultimately, the retained player-categorization model should be understood as a practical descriptive device. It provides a reproducible way to summarize broad playstyle variation across player-month observations, and the associated diagnostics suggest that the resulting clusters are usable as behavior profiles for downstream description. They are not, however, treated here as definitive latent classes or as a basis for inferential claims about underlying player types.

<!--
\newpage
-->

## Results {#sec-results}

<!-- Describe the results that you received from the methods described earlier. -->

This section presents the substantive findings from the retained models identified in @sec-modeling. The aim is no longer to compare candidate specifications, but to interpret what the retained models suggest about chess behavior and structure across the main domains of the project: metagame dynamics, opening popularity, time use, player progression, and player categorization.

Accordingly, the section emphasizes interpretable summaries of model output, including estimated effects, representative predicted values, and descriptive profile summaries where appropriate. Inferential claims are drawn only from models retained for that purpose, while descriptive models are used more cautiously to summarize structure and variation in the data. Each subsection therefore focuses on what the retained model or models show, how large the relevant patterns are, and how confidently those patterns can be described.

### Metagame Dynamics {#sec-metagame-dynamics-3}

The intent of metagame-dynamics models was to characterize the state and evolution of the metagame across time and format. The model category accomplishes this on two levels. Aggregate metagame metrics (macro models) operate at the ecosystem level, while opening composition models help contextualize those aggregate shifts by clarifying what kinds of opening-level substitutions accompany them. The clearest changes concern concentration and diversity, while broader game-environment outcomes changed less consistently.

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
Y_{tc,t}
&=
\beta_0
+
\beta_1 t
+
\beta_2 \,\mathbf{1}\{\mathrm{tc}=\mathrm{Classical}\}
+
\beta_3 \,\mathbf{1}\{\mathrm{tc}=\mathrm{Rapid}\} \\
&\quad
+
\beta_4 \, t \cdot \mathbf{1}\{\mathrm{tc}=\mathrm{Classical}\}
+
\beta_5 \, t \cdot \mathbf{1}\{\mathrm{tc}=\mathrm{Rapid}\}
+
\varepsilon_{tc,t}.
\end{aligned}
\end{equation*}
\caption{Retained macro metagame model. Here, $Y_{tc,t}$ denotes a month-by-time-control metagame outcome, such as HHI concentration, effective number of openings, draw rate, upset rate, or average moves. Inference used Newey--West heteroskedasticity- and autocorrelation-consistent standard errors.}
\label{fig:eq-metagame-macro}
\end{figure}
```

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
\Pr(Y_t = k)
&=
\frac{\exp(\eta_{k,t})}
{\sum_{j=1}^{K}\exp(\eta_{j,t})}, \\
\eta_{k,t}
&=
\alpha_k + \beta_k t,
\qquad k=1,\dots,K.
\end{aligned}
\end{equation*}
\caption{Retained multinomial metagame-composition model, fit separately by time control. Here, $Y_t$ denotes the opening category at month $t$, and $k$ indexes the retained opening categories.}
\label{fig:eq-metagame-multinomial}
\end{figure}
```


Across formats, opening usage became less concentrated and more diverse (see table \ref{tab:metagame-macro-estimates}). This trend is present for all three time controls, though its magnitude differs. HHI fell over time in the baseline format (blitz), with an estimate of $-2.283 \times 10^{-6}, \ p = 0.002$; this trend was more pronounced in rapid ($-7.440 \times 10^{-6}, \ p < 0.001$) and classical ($-1.694 \times 10^{-5}, \ p = 0.022$). The number of effective openings ($N_\text{eff}$) rose over time in the baseline format, with an estimate of $0.1021, \ p = 0.002$; this trend was also more pronounced in rapid ($0.2411, \ p < 0.001$) and classical ($0.5556, \ p = 0.030$). These trends move in opposite-yet-complementary directions, as declining concentration and increasing diversity are expressing the same core pattern. Overall, the time-control interactions matter substantively, not just statistically.

```{r Metagame Results Figure}
#| label: fig-metagame-results
#| fig-cap: "Predicted model values over time."
#| fig-align: center

p_macro_fit <- readRDS("Paper_files/figure-rds/p_macro_fit.rds")
p_macro_fit
```

Also present in table \ref{tab:metagame-macro-estimates} are outcomes. Draw rate and and average moves present little evidence of a systematic linear trend, the lowest p-value among their estimates being 0.215 for the baseline time trend. Upset rate shows some evidence of a positive baseline time trend ($9.540 \times 10^{-5}, \ p = 0.024$), but it's weak and not clearly generalized across formats; the non-baseline trends for upset rate have p-values more in line with draw rate and average moves. This is important, as it means that ecosystem structure changed more clearly than these broader, game-environment outcomes did.

Whereas the aggregate metrics show that the metagame broadened, the composition model provides a way to describe which opening-share shifts underlie the aggregate broadening of the metagame. The compositional patterns were estimated separately by time control, allowing format-specific movement to be compared. Of note is that the composition findings are based on models with comparable performance across formats; table \ref{tab:metagame-composition-model-comparison} shows blitz, rapid, and classical to have held-out errors of 0.910, 0.923, and 0.940, respectively.

The observed macro diversification reflects broad redistribution across the opening ecosystem rather than a single new opening taking over, and the compositional results reflect that: From table \ref{tab:metagame-multinomial-robustness}, "Other" is the top increase at every $k$, and a named opening decline remains visible once the retained opening set is sufficiently rich. The pattern is stable enough to suggest ecosystem-wide redistribution, as if "Other" gains share constantly, it means that the metagame isn't just rotating among a fixed set of dominant openings. Instead, some of the change is occurring outside the most explicitly-retained openings, which is consistent with diversification.

<!-- Potentially move held-out errors into a new paragraph here and add additional format-specific mover output (additional model stuff) later. -->

The claim of broad redistribution is not merely a modeling artifact of a single cutoff. Table \ref{tab:metagame-multinomial-robustness} shows that "Other" remains the top increase, and the top negative mover stabilizes as French Defense: Knight Variation at $k = 12$ and $k = 16$. While log loss increases from 0.674 at $k = 8$ to 0.910 at $k = 12$ to 1.127 at $k = 16$, this is reflective of increasing $k$ making the prediction mechanically more difficult; the broad directional composition remains stable, even if log loss increases with $k$.

All in all, at the ecosystem level, the metagame became less concentrated and more diverse over time, and those structural changes differed in severity across time controls. The opening-level evidence suggests that this broadening reflects diffuse substitution within the opening ecosystem rather than one simple adoption cycle. The clearest evidence concerns ecosystem structure, and not parallel shifts in draw rate or average moves.

### Opening Popularity {#sec-opening-popularity-3}

The goal of the opening-popularity category of models is simple, because they're really just asking what factors impact opening choice. The retained pooled multinomial model was used so summarize opening popularity as a compositional outcome. This is a very similar approach to the opening-level method applied in @sec-metagame-dynamics-3, the difference being that the prior section was concerned only about explaining and supporting metagame-level drift, whereas this one explores opening choice in much more depth.

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
\Pr(Y_i = k)
&=
\frac{\exp(\eta_{ik})}
{\sum_{j=1}^{K}\exp(\eta_{ij})}, \\
\eta_{ik}
&=
\alpha_k
+
\beta_{1k}\,\mathrm{RatingBin}_i
+
\beta_{2k}\,\mathbf{1}\{\mathrm{tc}_i=\mathrm{Rapid}\} \\
&\quad
+
\beta_{3k}\,t_i
+
\beta_{4k}\,\mathrm{RatingBin}_i
\cdot
\mathbf{1}\{\mathrm{tc}_i=\mathrm{Rapid}\}.
\end{aligned}
\end{equation*}
\caption{Retained pooled multinomial opening-popularity model. Here, $Y_i$ denotes the opening category for month-by-time-control-by-rating-bin cell $i$, $t_i$ is the month index, and $k$ indexes the retained opening categories, including "Other".}
\label{fig:eq-opening-popularity}
\end{figure}
```


Opening popularity is not random across skill levels or formats. This is best evidenced by table \ref{tab:opening-popularity-rating-profile}. As it's an aggregate, the residual "Other" category will almost always be dominant when modeling with a top-$k$ approach (for any reasonable $k$) simply due to the sheer quantity of named openings that see play at every level. Of greater importance are the share of the Other category, the share and presence of the top non-Other openings, and how those numbers change across skill and time control. Among explicitly-modeled openings, lower-rated bins show more mass on openings like the Philidor Defense, Scandinavian Defense, and Queen's Pawn Game, while higher-rated bins shift toward the Caro-Kann Defense and the Accelerated London System^[While the London System technically falls under the umbrella of the Queen's Pawn Game – they begin very similarly – it's often considered to be distinct from the more typical set of Queen's Pawn openings, which is why I'm using it as an example of a shift in opening popularity.] This supports the claim that opening popularity varies systematically by skill level, and that the pattern is visible across time controls.

In both blitz and rapid, the dominant Other category grows with rating, from 0.732/0.768 at an 800 rating, 0.835/0.840 at 1600, and 0.892/0.877 at 2400 for blitz/rapid, respectively. This implies that higher-rated players place more of their opening mass outside the named top openings retained in the model. A rating pattern also presents itself where specific openings are concerned: For both blitz and rapid, the Philidor Defense is popular at 800, the Queen's Pawn Game is popular at 1600, and the Caro-Kann defense is popular at 2400. The identity and order of the top named openings is not identical, however, suggesting that time control shapes popularity within rating strata, even when the broader rating gradient is similar.

```{r Opening-Popularity Results Figure}
#| label: fig-opening-popularity-results
#| fig-cap: "Predicted share for choice openings."
#| fig-align: center

p_opening_rating <- readRDS("Paper_files/figure-rds/p_opening_rating.rds")
p_opening_rating
```

Changes in opening popularity over time are present, but modest in magnitude. As seen in table \ref{tab:opening-popularity-time-change} and elaborated on in @sec-metagame-dynamics-3, the time trend is directionally consistent across formats, the slight positive trend for Other pointing to broad compositional shifts rather than a dramatic takeover by one opening. In both formats, the largest named positive mover is the Accelerated London System, and the largest named negative mover is the French Defense: Knight Variation.

The popularity results are based on the retained pooled multinomial model, which showed reasonable held-out performance on the test set with a log loss of 0.893 (see table \ref{tab:opening-popularity-model-comparison}). The one-opening models provide supporting checks, but the multinomial is the relevant basis for interpreting substitution across openings. Table \ref{tab:opening-popularity-robustness} tested linear and spline fits for both rating and time, as well as different top-$k$ thresholds. Allowing nonlinear rating slightly improved fit (log loss of 0.890 compared to 0.893), whereas allowing nonlinear time slightly worsened fit (0.895 compared to 0.893), though the improvements were incredibly marginal; the simpler linear-time specification is adequate for baseline reporting. As addressed in @sec-metagame-dynamics, an increase in $k$ saw a corresponding increase in log loss (0.666 for $k = 8$ up to 1.111 for $k = 16$), but this increase in modeling difficulty does not overturn the broad pattern of modest, consistent changes. As such, these conclusions don't appear to depend heavily on one narrow functional-form choice.

Ultimately, these models found that opening popularity varies systematically by rating and time control. Higher-rated bins allocate more share outside the explicitly-retained top openings. Over time, the largest changes are modest but directionally consistent across blitz and rapid. The most visible temporal change is a rise in the residual other category alongside smaller named increases and decreases.

### Time Use {#sec-time-use-3}
<!-- I do not yet have coefficients/model estimates, so currently have some placeholders -->

This category of models characterizes how players allocate time and experience time pressure across rating and time control; time use is being treated as a behavioral outcome that may vary by skill and the structure of the format. Two different models were retained for this category: interaction-based mean models for average time use, and a fractional quasibinomial model for low-time frequency. Average time use was modeled separately at the macro and player-month levels, but the results found in this section are taken from the player-level interaction model. Across the retained models, the clearest time-use pattern is a format-specific rating gradient: Stronger players use less time on average in blitz, but more in rapid and especially classical, implying that the relationship between skill and time allocation changes direction across formats. Low-time frequency was found to increase with rating in all formats, especially blitz.

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
\mathrm{AvgTime}_i
&=
\beta_0
+
\beta_1\,\mathrm{Rating}_i
+
\beta_2\,\mathbf{1}\{\mathrm{tc}_i=\mathrm{Classical}\}
+
\beta_3\,\mathbf{1}\{\mathrm{tc}_i=\mathrm{Rapid}\} \\
&\quad
+
\beta_4\,\mathrm{Rating}_i \cdot \mathbf{1}\{\mathrm{tc}_i=\mathrm{Classical}\}
+
\beta_5\,\mathrm{Rating}_i \cdot \mathbf{1}\{\mathrm{tc}_i=\mathrm{Rapid}\}
+
\varepsilon_i.
\end{aligned}
\end{equation*}
\caption{Retained player-level mean time-use model. Here, $\mathrm{AvgTime}_i$ denotes average time spent for observation $i$, and the interaction terms allow the rating slope to vary by time control.}
\label{fig:eq-timeuse-mean}
\end{figure}
```

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
\operatorname{logit}\!\left(\mathbb{E}[L_i \mid X_i]\right)
&=
\beta_0
+
\beta_1\,\mathrm{Rating}_i
+
\beta_2\,\mathbf{1}\{\mathrm{tc}_i=\mathrm{Classical}\}
+
\beta_3\,\mathbf{1}\{\mathrm{tc}_i=\mathrm{Rapid}\},
\end{aligned}
\end{equation*}
\caption{Retained fractional low-time model. Here, $L_i \in [0,1]$ denotes low-time frequency for observation $i$.}
\label{fig:eq-timeuse-lowtime}
\end{figure}
```


Per table \ref{tab:timeuse-mean-model-comparison} interaction specifications were retained at both levels due to slightly superior WRMSE (0.143 vs. 0.149 for macro and 1.832 vs. 1.839 for player-month). Though it's a marginal improvement, it implies that a common rating effect across time controls was slightly inadequate. Therefore, the substantive results should be interpreted in terms of format-specific rating relationships rather than one universal rating slope. As shown in the format difference are much substantial when taking model estimates into account, with blitz at the lowest (4.390 seconds per move at 800, 3.834 at 1600, and 3.279 at 2400), classical at the highest (800: 10.573, 1600: 12.396, 2400: 14.220), and rapid somewhere in between (800: 6.448, 1600: 7.136, 2400: 7.824). Notably, blitz is the only of the three with a negative rating gradient, and classical presents a much wider confidence interval, suggesting greater uncertainty (see table \ref{tab:timeuse-player-mean-profile}).

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{2.6cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.3cm} >{\raggedleft\arraybackslash}X}
\caption{Predicted average time spent from the retained player-level mean time-use model at the reference month.}\label{tab:timeuse-player-mean-profile}\\
\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Predicted Mean} & \textbf{95\% CI} \\
\midrule
\endfirsthead

\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Predicted Mean} & \textbf{95\% CI} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Blitz & 800 & 3.973 & [3.908, 4.038] \\
Blitz & 1600 & 3.718 & [3.677, 3.759] \\
Blitz & 2400 & 3.463 & [3.412, 3.514] \\
Classical & 800 & 11.941 & [9.083, 14.799] \\
Classical & 1600 & 12.778 & [11.150, 14.406] \\
Classical & 2400 & 13.615 & [11.172, 16.058] \\
Rapid & 800 & 6.964 & [6.791, 7.137] \\
Rapid & 1600 & 7.280 & [7.103, 7.456] \\
Rapid & 2400 & 7.595 & [7.286, 7.905] \\
\end{xltabular}
```


Low-time frequency increases with rating in all three formats. As shown in table \ref{tab:timeuse-lowtime-profile} the absolute level of low-time pressure is highest in blitz, lower in rapid, and lowest in classical. For contrast, the lowest predicted low-time rate in blitz (0.0209 or 2.09% in the 800-rating bin) is greater than the highest predicted low-time rate in classical (0.0194 or 1.94% in the 2400-rating bin). Just as with the mean models, classical estimates are more uncertain than those for blitz and rapid. This means that, even though higher-rated players are more likely to enter low-time states in every format, the baseline prevalence of those states differs sharply across formats.


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{2.6cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.4cm} >{\raggedleft\arraybackslash}X}
\caption{Predicted low-time frequency from the retained fractional time-pressure model at the reference month.}\label{tab:timeuse-lowtime-profile}\\
\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Predicted Rate} & \textbf{95\% CI} \\
\midrule
\endfirsthead

\toprule
\textbf{Time Control} & \textbf{Rating} & \textbf{Predicted Rate} & \textbf{95\% CI} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Blitz & 800 & 0.0209 & [0.0183, 0.0239] \\
Blitz & 1600 & 0.0447 & [0.0414, 0.0483] \\
Blitz & 2400 & 0.0931 & [0.0856, 0.1011] \\
Classical & 800 & 0.00014 & [$1.28\times10^{-8}$, 0.611] \\
Classical & 1600 & 0.00167 & [0.000105, 0.0260] \\
Classical & 2400 & 0.0194 & [$6.54\times10^{-5}$, 0.857] \\
Rapid & 800 & 0.00194 & [0.00108, 0.00347] \\
Rapid & 1600 & 0.00965 & [0.00785, 0.0119] \\
Rapid & 2400 & 0.0466 & [0.0333, 0.0649] \\
\end{xltabular}
```


In blitz, higher-rated players use less time on average but experience more low-time pressure. In rapid and classical, higher-rated players use more time on average but still experience more low-time pressure. From this, we can infer that low-time frequency is not merely an inverse of average pacing; it captures a distinct aspect of time-management behavior.

```{r Time-Use Results Figure}
#| label: fig-time-use-results
#| fig-cap: "Predicted time-use values across rating bins."
#| fig-align: center

p_time_panels <- readRDS("Paper_files/figure-rds/p_time_panels.rds")
p_time_panels
```

Though no relevant tables are present for the sake of brevity, it should be noted that interaction specifications were retained for the mean models because a common rating slope was inadequate. The negative baseline rating effect for blitz and positive rapid interaction were directionally consistent across supporting models. The classical interaction, however, was less stable and should therefore be interpreted more cautiously.

On the whole, time-use behavior varies strongly across both rating and format. The relationship between rating and average time-use is format-specific, whereas low-time frequency rises with rating in all formats. Faster formats produce substantially more time pressure overall, which is expected. Therefore, average pace and acute time pressure should be treated as related-yet-distinct aspects of player behavior.

### Player Progression {#sec-player-progression-3}

The retained player fixed-effects models are being used to assess whether within-player changes in current-month behavior are associated with later rating movement, and if so, which behaviors are associated with what kinds of change (i.e., rating gain/loss). Three targets were retained: 1-month rating change, 3-month change, and absolute 1-month change. The signed (non-absolute) targets ask whether a predictor relates to a player moving up or down in rating, with positive values reflecting positive increases and vice versa. In contrast, the absolute target removes direction, and asks whether a predictor relates to moving more or less, regardless of direction—  positive values reflect larger changes, while negative ones imply smaller ones. 

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
Y_{it}^{(h)}
&=
\alpha_i
+
\beta_1\,\mathrm{AvgRating}_{it}
+
\beta_2\,\mathrm{Games}_{it}
+
\beta_3\,\mathrm{AvgMoves}_{it} \\
&\quad
+
\beta_4\,\mathrm{OpeningDiversity}_{it}
+
\beta_5\,\mathrm{TopOpeningShare}_{it}
+
\beta_6\,\mathrm{SwitchedOpening}_{it}
+
\varepsilon_{it},
\end{aligned}
\end{equation*}
\caption{Generic form of the retained player fixed-effects progression model, where $Y_{it}^{(h)} \in \{\Delta \mathrm{Rating}_{i,t+1}, \Delta \mathrm{Rating}_{i,t+3}, |\Delta \mathrm{Rating}_{i,t+1}|\}$.}
\label{fig:eq-progression-fe-generic}
\end{figure}
```


```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.4cm} >{\raggedright\arraybackslash}p{2.3cm} >{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}X}
\caption{Coefficient estimates from the retained player fixed-effects progression models.}\label{tab:progression-fe-results}\\
\toprule
\textbf{Predictor} & \textbf{Target} & \textbf{Estimate} & \textbf{95\% CI} & \textbf{\textit{p}-Value} \\
\midrule
\endfirsthead

\toprule
\textbf{Predictor} & \textbf{Target} & \textbf{Estimate} & \textbf{95\% CI} & \textbf{\textit{p}-Value} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Current Average Rating & 1-Month Change & -0.469 & [-0.511, -0.427] & $< 0.001$ \\
Games Played & 1-Month Change & -0.00295 & [-0.0101, 0.00420] & 0.398 \\
Average Moves & 1-Month Change & 1.403 & [1.213, 1.594] & $< 0.001$ \\
Opening Diversity & 1-Month Change & 0.162 & [0.129, 0.194] & $< 0.001$ \\
Top-Opening Share & 1-Month Change & 6.697 & [-3.782, 17.176] & 0.196 \\
Switched Most-Played Opening & 1-Month Change & 2.722 & [1.035, 4.409] & 0.003 \\

Current Average Rating & 3-Month Change & -0.800 & [-0.859, -0.740] & $< 0.001$ \\
Games Played & 3-Month Change & 0.00345 & [-0.00693, 0.0138] & 0.494 \\
Average Moves & 3-Month Change & 1.112 & [0.870, 1.353] & $< 0.001$ \\
Opening Diversity & 3-Month Change & 0.0614 & [0.0165, 0.106] & 0.010 \\
Top-Opening Share & 3-Month Change & -3.668 & [-17.236, 9.901] & 0.577 \\
Switched Most-Played Opening & 3-Month Change & 2.419 & [0.249, 4.589] & 0.031 \\

Current Average Rating & Absolute 1-Month Change & 0.00027 & [-0.0107, 0.0112] & 0.959 \\
Games Played & Absolute 1-Month Change & 0.00110 & [-0.00469, 0.00689] & 0.694 \\
Average Moves & Absolute 1-Month Change & -0.334 & [-0.503, -0.165] & $< 0.001$ \\
Opening Diversity & Absolute 1-Month Change & -0.0712 & [-0.0918, -0.0505] & $< 0.001$ \\
Top-Opening Share & Absolute 1-Month Change & -5.837 & [-13.635, 1.961] & 0.133 \\
Switched Most-Played Opening & Absolute 1-Month Change & -3.755 & [-4.344, -3.166] & $< 0.001$ \\
\end{xltabular}
```


Table \ref{tab:progression-fe-results} shows that, for signed targets, higher current rating is associated with more negative subsequent rating change. Conversely, average moves, opening diversity, and switching one's most-played opening are positively associated with rating signed rating change. For the absolute-change target, average moves, opening diversity, and opening switching are all negatively associated with month-to-month volatility.

```{r Player-Progression Results Figure}
#| label: fig-player-progression-results
#| fig-cap: "Predicted coefficients for the fixed-effects models"
#| fig-align: center

p_player_coef <- readRDS("Paper_files/figure-rds/p_player_coef.rds")
p_player_coef
```

In concert, the signed and absolute targets paint a very consistent, very interesting picture. Among the predictors that matter, none are linked to greater movement magnitude; the signal is consistently toward smaller movement magnitude, with the weaker ones uncertain, but never clearly positive. This is particularly relevant when it comes to favorable (positive) signed predictors, as it means that every favorable predictor is associated with less absolute movement.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedright\arraybackslash}p{3.7cm} >{\raggedright\arraybackslash}p{2.6cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}X}
\caption{Representative predicted contrasts from the retained player fixed-effects progression models.}\label{tab:progression-fe-contrasts}\\
\toprule
\textbf{Predictor} & \textbf{Target} & \textbf{Predicted Difference} & \textbf{95\% CI} \\
\midrule
\endfirsthead

\toprule
\textbf{Predictor} & \textbf{Target} & \textbf{Predicted Difference} & \textbf{95\% CI} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Current Average Rating (25th vs 75th Pctl.) & 1-Month Change & -247.000 & [-267.706, -226.294] \\
Games Played (25th vs 75th Pctl.) & 1-Month Change & -0.365 & [-1.193, 0.462] \\
Average Moves (25th vs 75th Pctl.) & 1-Month Change & 9.668 & [8.445, 10.891] \\
Opening Diversity (25th vs 75th Pctl.) & 1-Month Change & 7.434 & [6.052, 8.816] \\
Top-Opening Share (25th vs 75th Pctl.) & 1-Month Change & 0.609 & [-0.280, 1.498] \\
Switched Most-Played Opening (0 vs 1) & 1-Month Change & 2.722 & [1.148, 4.296] \\

Current Average Rating (25th vs 75th Pctl.) & 3-Month Change & -412.457 & [-441.158, -383.755] \\
Games Played (25th vs 75th Pctl.) & 3-Month Change & 0.472 & [-0.854, 1.799] \\
Average Moves (25th vs 75th Pctl.) & 3-Month Change & 7.443 & [5.935, 8.950] \\
Opening Diversity (25th vs 75th Pctl.) & 3-Month Change & 2.883 & [0.915, 4.852] \\
Top-Opening Share (25th vs 75th Pctl.) & 3-Month Change & -0.324 & [-1.441, 0.793] \\
Switched Most-Played Opening (0 vs 1) & 3-Month Change & 2.419 & [0.394, 4.443] \\

Current Average Rating (25th vs 75th Pctl.) & Absolute 1-Month Change & 0.143 & [-5.237, 5.523] \\
Games Played (25th vs 75th Pctl.) & Absolute 1-Month Change & 0.136 & [-0.533, 0.806] \\
Average Moves (25th vs 75th Pctl.) & Absolute 1-Month Change & -2.300 & [-3.384, -1.215] \\
Opening Diversity (25th vs 75th Pctl.) & Absolute 1-Month Change & -3.274 & [-4.161, -2.388] \\
Top-Opening Share (25th vs 75th Pctl.) & Absolute 1-Month Change & -0.531 & [-1.192, 0.131] \\
Switched Most-Played Opening (0 vs 1) & Absolute 1-Month Change & -3.755 & [-4.305, -3.205] \\
\end{xltabular}
```

Table \ref{tab:progression-fe-contrasts} helps convert the fixed-effect estimates into practical differences across representative predictor values. It compares continuous predictors are compared at the 25th and 75th percentiles, while the switching indicator is compared from 0 to 1. All magnitude patterns are directionally consistent with the coefficient table. Moving from lower to higher values of average moves corresponds to a meaningful increase in signed rating change. To a degree, this is also true of opening diversity, with the only difference being its modest range at the 3-month target compared to its 1-month counterpart (decreases from [6.052, 8.816] to [0.915, 4.852]). Switching the most-played opening is a smaller but still impactful contrast. All three are associated with smaller absolute movement. 

Of particular note is current average rating, as its signed intervals are orders of magnitude larger than any of the other predictors. This does not imply that its importance as a predictor is equally large, however; rather, it's on a very wide scale, which impacts the size of the interval. As an example, for 1-month change, the 25th percentile rating is ~1272, while the 75th is ~1798, for a difference of ~526 rating points. Multiply that by the coefficient of -0.469 and you get a contrast near $-0.469 \times 526 \approx -247$. Compare this to average moves, where the 25th–75th gap is only about 7 moves, opening diversity where it's ~46 units, or top-opening share, where it's less than 0.1, and you'll see why the contrast band for average rating is so much larger.

Not all retained predictors show equally clear evidence; games played and top-opening share stand out for being weak or ambiguous predictors. Refer back to table \ref{tab:progression-fe-results}. Neither have a p-value below 0.1 for any of the targets, and most are many times that. Their poor evaluation is further reinforced by table \ref{tab:progression-fe-contrasts}, as still neither have a confidence interval that precludes zero. Current average rating is of note for being the only predictor whose statistical significance crosses a meaningful alpha of 0.05 between models, with a p-value of 0.959 for the absolute target. This is expected, however, and actually a good sign that the models are on the right track. The Lichess rating system is designed so that a player's current rating is already a good summary of expected strength, so we actually shouldn't expect that it should predict larger or smaller absolute next-month movement; it would be more worrying if the coefficient weren't centered near 0.

It's important to remember that the fixed-effect design isolates within-player association; these targets are being modeled as within-player associations, not as cross-sectional differences between players. The models are estimated on large within-player panels in order to support stable coefficient estimation. Note that the signed-change models explain substantially more within-player variation than the absolute-change model – within-$R^2$ values of 0.270, 0.453, and 0.004 for 1-month, 3-month, and 1-month absolute, respectively – so the latter is best treated as a supporting magnitude benchmark rather than the main target.

Within players, months characterized by higher move-count games, broader opening diversity, and switching one's most-played opening are associated with more favorable next-month rating movement. Those same predictors are not associated with larger movement magnitude in the absolute model, but are instead most often associated with smaller absolute change. By contrast, games played and top-opening share do not show comparably stable evidence in the retained fixed-effects specification. These results suggest that the behavioral signals most associated with later progression concern game complexity, repertoire breadth, and active repertoire adjustment rather than simple activity volume or repertoire concentration alone.

### Player Categorization {#sec-player-categorization-3}

Though it has been mentioned multiple times in @sec-player-categorization-2, it is important to stress that no player-categorization model is fit for inference; even the retained $k$-means solution should only be treated as a descriptive summary of behavioral variation across player-month observations. The point here is not to treat the clusters as latent causal types, but to describe recurring patterns in performance, repertoire structure, and activity. The results below use standardized cluster centers, time-control distributions, and month-to-month persistence summaries.

```{=latex}
\begin{figure}[htbp]
\centering
\small
\begin{equation*}
\begin{aligned}
\min_{C_1,\dots,C_6}
\sum_{k=1}^{6}
\sum_{i \in C_k}
\left\| x_i - \mu_k \right\|^2,
\end{aligned}
\end{equation*}
\caption{Retained $k$-means clustering objective for player categorization. Here, $x_i \in \mathbb{R}^p$ denotes the standardized feature vector for observation $i$, $C_k$ denotes cluster $k$, and $\mu_k$ denotes the centroid of cluster $k$.}
\label{fig:eq-categorization-kmeans}
\end{figure}
```


As described in tables \ref{tab:categorization-standardized-performance} and \ref{tab:categorization-standardized-repertoire}, clusters differ along multiple axes rather than only average rating. Clusters 1 and 2 are foils: the former contains wide-repertoire, high-activity, and above-average-rated players, whereas the latter has lower-rated, lower-activity, narrow-repertoire players whose games are shorter on average. Cluster 3 is notable for very high opening switching and somewhat lower opening concentration, but is otherwise close to the sample center. Clusters 4 and 5 both contain strong players, but are behaviorally distinct, with cluster 4 grouping players with longer games and high draw rates, and cluster 5 grouping players with high win rates and positive rating differentials. Finally, cluster 6 prefers players with concentrated openings, low switching, and moderate activity. In general, the clusters are especially concerned with diversity, concentration, switching, and game structure.

<!-- The first table in this section needs to be reformatted, as it currently stretches off the page -->

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.2cm} >{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{1.7cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{1.5cm} >{\raggedleft\arraybackslash}X}
\caption{Standardized performance and outcome centers for the retained player-categorization solution. Positive values indicate above-average cluster means on the standardized feature scale, while negative values indicate below-average means.}\label{tab:categorization-standardized-performance}\\
\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Avg. Rating} & \textbf{Rating Diff.} & \textbf{Avg. Moves} & \textbf{Win Rate} & \textbf{Draw Rate} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Avg. Rating} & \textbf{Rating Diff.} & \textbf{Avg. Moves} & \textbf{Win Rate} & \textbf{Draw Rate} \\
\midrule
\endhead

\midrule
\multicolumn{7}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

1 & 20{,}107 & 0.643 & 0.134 & 0.474 & 0.096 & 0.067 \\
2 & 28{,}181 & -0.888 & -0.666 & -1.001 & -0.934 & -0.423 \\
3 & 34{,}057 & -0.037 & -0.016 & -0.017 & 0.055 & -0.110 \\
4 & 17{,}546 & 0.330 & 0.061 & 0.826 & -0.427 & 1.739 \\
5 & 25{,}390 & 0.387 & 0.613 & 0.070 & 1.132 & -0.525 \\
6 & 31{,}620 & -0.071 & -0.000 & 0.095 & 0.041 & -0.090 \\
\end{xltabular}
```

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.4cm} >{\raggedleft\arraybackslash}p{1.8cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}X}
\caption{Standardized repertoire centers for the retained player-categorization solution.}\label{tab:categorization-standardized-repertoire}\\
\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Op. Diversity} & \textbf{Top Share} & \textbf{Switched} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{$n$} & \textbf{Op. Diversity} & \textbf{Top Share} & \textbf{Switched} \\
\midrule
\endhead

\midrule
\multicolumn{5}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

1 & 20{,}107 & 1.994 & -0.623 & -0.286 \\
2 & 28{,}181 & -0.684 & 0.354 & 0.072 \\
3 & 34{,}057 & 0.176 & -0.534 & 0.845 \\
4 & 17{,}546 & -0.651 & 0.202 & 0.261 \\
5 & 25{,}390 & -0.690 & 0.191 & 0.291 \\
6 & 31{,}620 & 0.067 & 0.391 & -1.171 \\
\end{xltabular}
```


Table \ref{tab:categorization-timecontrol-shares} shows that the behavioral clusters aren't distributed uniformly across formats. Blitz is most heavily represented by the switching-heavy and concentrated/stable clusters (3, 6). Classical is dominated by the high-win-rate cluster, the lower-rated/narrower cluster, and the long-game/high-draw cluster (5, 2, 4), with cluster 1 almost absent. Rapid is more evenly distributed, though clusters 2, 3, and 5 are the largest. The format patterns are suggestive rather than definitive, especially in classical, as its counts are much smaller. Overall, the results suggest that format is part of the behavioral structure summarized by the retained clustering solution.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{2.0cm} >{\raggedleft\arraybackslash}X}
\caption{Cluster shares within each time-control bucket. Shares are computed within time control rather than across the full clustering sample.}\label{tab:categorization-timecontrol-shares}\\
\toprule
\textbf{Time Control} & \textbf{Cluster} & \textbf{Share Within Time Control} \\
\midrule
\endfirsthead

\toprule
\textbf{Time Control} & \textbf{Cluster} & \textbf{Share Within Time Control} \\
\midrule
\endhead

\midrule
\multicolumn{3}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

Blitz & 1 & 0.164 \\
Blitz & 2 & 0.158 \\
Blitz & 3 & 0.227 \\
Blitz & 4 & 0.096 \\
Blitz & 5 & 0.140 \\
Blitz & 6 & 0.215 \\

Classical & 1 & 0.013 \\
Classical & 2 & 0.238 \\
Classical & 3 & 0.138 \\
Classical & 4 & 0.222 \\
Classical & 5 & 0.285 \\
Classical & 6 & 0.104 \\

Rapid & 1 & 0.064 \\
Rapid & 2 & 0.218 \\
Rapid & 3 & 0.202 \\
Rapid & 4 & 0.137 \\
Rapid & 5 & 0.199 \\
Rapid & 6 & 0.180 \\
\end{xltabular}
```

The clusters have some continuity over time, but not enough to justify treating them as fixed, latent identities. As seen in table \ref{tab:categorization-persistence}, cluster 1 is the most persistent, cluster 4 is the least persistent, and the others fall somewhere in between. Transition structure is also not random. Cluster 3 is the most common transition destination (i.e., moving from one cluster to another), though cluster 6 serves as a recurring destination from cluster 3. This suggests that cluster 3 might act as a sort of intermediate state within the model.

```{=latex}
\begin{xltabular}{\textwidth}{>{\raggedleft\arraybackslash}p{1.6cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}p{2.2cm} >{\raggedleft\arraybackslash}X}
\caption{Month-to-month persistence and most common transition destination for the retained player-categorization clusters. Persistence is measured as the share of adjacent player-month pairs remaining in the same cluster.}\label{tab:categorization-persistence}\\
\toprule
\textbf{Cluster} & \textbf{Adjacent Pairs} & \textbf{Same-Cluster Rate} & \textbf{Most Common Next Cluster If Switched} \\
\midrule
\endfirsthead

\toprule
\textbf{Cluster} & \textbf{Adjacent Pairs} & \textbf{Same-Cluster Rate} & \textbf{Most Common Next Cluster If Switched} \\
\midrule
\endhead

\midrule
\multicolumn{4}{r}{\small\emph{Continued...}} \\
\endfoot

\bottomrule
\endlastfoot

1 & 18{,}238 & 0.660 & 3 (0.464 Of Switches) \\
2 & 12{,}599 & 0.478 & 3 (0.391 Of Switches) \\
3 & 26{,}988 & 0.469 & 6 (0.329 Of Switches) \\
4 & 8{,}924 & 0.277 & 5 (0.325 Of Switches) \\
5 & 12{,}143 & 0.357 & 3 (0.345 Of Switches) \\
6 & 25{,}661 & 0.510 & 3 (0.430 Of Switches) \\
\end{xltabular}
```

Though unsuitable for inference, the retained $k$-means categorization solution is useful because it captures multiple distinct dimensions of behavior, those profiles are unevenly distributed across format, and that they show only moderate persistence over time. They do not, however, establish sharply-separated latent types or imply stable archetypes. At the end of the day, the clustering solution serves as a descriptive map of player-month behavior, but its moderate persistence and uneven transition structure suggest that these clusters are better understood as recurring behavioral profiles than anything more concrete.

<!--
\newpage
-->

## Discussion

<!--
Original template structure

### Final model interpolation  
Discuss what information the selected model provides that may be new and interesting

### Use of Model
Describe how this model might be used by others.

### Comparison
Compare the things learned from this selected model to the previous information.
-->

### Principal Findings {#sec-discussion}

The retained models collectively point to structured variation in online chess behavior across both ecosystem-level and player-level domains. The metagame became less concentrated and more diverse; opening choice varied systematically by rating and time control; time use reflected multiple behavioral dimensions rather than one simple skill gradient; progression was associated more with complexity, breadth, and adaptation; and player categorization identified recurring behavioral profiles as opposed to fixed player types. The main interpretive question, then, is not whether structure exists, but what the structure suggests about metagame change, player decision-making, and behavioral development.

One of the most consistent patterns present in the results is that they repeatedly argue against simple, one-dimensional explanations: metagame change was not just a story of one opening taking over, time management was not reducible to one general measure of efficiency, and progression was not best explained by activity volume alone. Instead, the findings point toward more diverse outcomes. Diversification is interesting because a simple prior expectation might have been convergence around a smaller number of dominant openings or strategies; it suggests that greater access to theory, engines, and content may be expanding participation in the opening space rather than narrowing it, and implies that player development may benefit more from experimentation and differentiation than optimization towards one best repertoire.

The subsections that follow interpret these findings in greater depth, focusing on what they imply about metagame evolution, opening choice, time-management behavior, progression, and the meaning of the retained behavioral profiles.^[Though no prior expectations were formally established at this project's outset, the subsections may sometimes draw comparisons between the results and hypothetical expectations which might otherwise have been made.]

### Metagame Structure and Opening Choice {#sec-metagame-and-opening-discussion}

<!-- 
Include citation for metagames converging, maybe?
-->

The strongest ecosystem-level finding was the paired decline in concentration and rise in effective diversity across time controls; the metagame broadened over time. This is meaningful because metagames are convergent in nature; they seek to optimize gameplay to its logical extreme, which is an inherently reductive process. While the results cannot support an inferential claim about whether the increase in diversity is a result of an online chess environment compared to chess itself, it isn't implausible to speculate that the massive quantity of diverse people playing online chess may be fostering diversification by pushing back against a strategic collapse toward a few dominant lines. Also of note is the game's immense mathematical complexity, which necessarily makes convergence to an optimal strategy more difficult.

On the opening-choice front, the clearest structure in opening popularity came from rating and format rather than over-time drift. The results found that opening choice was deeply stratified by player strength, with time control modifying the player-strength structure instead of replacing it. Temporal changes are directionally consistent across formats, but modest in magnitude. The opening metagame therefore looks like a layered system of skill-stratified preferences plus moderate temporal drift.

Important is the distinction between diversification as a measure of which openings are most popular and diversification as a measure of total opening share. The former implies that the top openings simply shuffled around, cyclically taking each other's place over time, while the latter implies that the share of the Other category of openings actually increases over time. This dichotomy is important because metagames often move in cyclical patterns, best illustrated by the game of rock-paper-scissors.

Suppose that chess openings can be likened to rock-paper-scissors, where any given opening can map to one of the three options— rock beats scissors, scissors beats paper, and paper beats rock. Now suppose that 50% of chess players favor openings which map to rock, while scissors and paper are each only favored by 25% of players. Assuming completely random matchmaking, scissors-style openings become disproportionately likely to lose, and paper-style openings become conversely likely to win. Canny players will recognize this dynamic, and begin to play paper at a higher rate. Let's say 15% of all scissors players switch to paper. Now, scissors has the disproportionate advantage. That is, until rock players catch on and swap to paper and the cycle repeats itself. In a perfect world, the metagame would converge to a point where each type had an opening share of $1/3$, but in practice, metagames tend to remain cyclical.

Now, it isn't actually reasonable to map chess openings to a rock-paper-scissors structure, but that's not the point of the hypothetical. The point is that even after converging to a point where rock, paper, and scissors each have an equal opening share, there are still only three options in use; if the top openings shift around, but the quantity of unique openings played remains the same, then all that's really happened is a shift among the most popular openings. The opening diversification supported by the results contrast such an eventuality by showing that the Other category of openings is actually increasing in share over time, which suggests that the Lichess playerbase is expanding in breadth.

Overall, these findings complicate at least two simple expectations: First, more theory and shared content might have led to convergence, and second, metagame change might have been expected to appear mostly as visible named-opening turnover. Instead, the results suggest broad diversification, long-tail expansion, and strong rating structuring.

### Time Management as Multi-Dimensional Behavior {#sec-time-discussion}

Average time use and low-time frequency are related, but not interchangeable. In blitz, stronger players use less time on average and low-time frequency increases with rating. In rapid and classical, stronger players use more time on average, but low-time frequency also increases rating. As such, the time-use results suggest that time management has at least two distinct components that shouldn't be treated as a single skill gradient: baseline pacing, and exposure to acute time pressure.

The rating gradient in average time changes sign by format. In other words, using more time does not have the same behavioral meaning in all time controls. In blitz, stronger players may be able to play faster on average because of better pattern recognition, more stable repertoires, or stronger intuition. In slower formats, stronger players may instead choose to invest more time because the format allows deeper calculation and more deliberate planning. With that in mind, faster average play in blitz should not automatically be read as lower engagement or lower complexity, and slower average play in rapid/classical should not automatically be read as 'better' time management in a universal sense.

Low-time frequency rises with rating in all formats. This argues against a plausible prior expectation equating skill with uniformly safer or smoother time use. Instead, stronger players may think more deeply in critical moments, play sharper or more demanding positions, or tolerate more pressure because of greater confidence or comfort. It's reasonable to speculate that, rather than simply playing 'faster' or 'slower' according to the time control, skilled players are more active and selective in how they allocate time, accepting future pressure in exchange for better decisions in critical positions.

### Progression, Adaptation, and Player Development {#sec-player-progression-discussion}

The simplest assumption to make pertaining to improvement in any skill is that improvement is driven primarily by time spent practicing said skill. In the context of chess data, a reasonable prior expectation might be that player activity is the salient predictor of player growth. The results, however, found otherwise; the retained models liked more favorable later-rating movement to higher average moves, greater opening diversity, and switching the most-played opening. In contrast, games played was weak and inconclusive, suggesting that improvement is associated more with how players engage with the game than to how much they play.

Broadening repertoires and changing the most-played opening are behaviors consistent with an adaptive, flexible style, which is meaningful because fixed, narrow repetition is often assumed to be the cleaner developmental path. The results suggest otherwise, supporting the notion that repertoire adjustment and breadth may coincide with favorable rating movement. Ergo, development involve strategic exploration as well as consolidation.

The absolute-change model helps clarify the signed-change results, though due to a low within-$R^2$, it's best read as a supporting benchmark. None of the favorable signed predictors were associated with greater movement magnitude; all had negative coefficients for the absolute-change model. This has interesting implications. For one, it weakens the argument that the strong results for experimentatative behavior is merely a consequence of experimentation's high-variance nature. More importantly, however, it implies that none of the positive-leaning predictors has an extreme or immediate impact on next-month rating. In effect, while the results support experimental play as a potential indicator of further improvement, they failed to find predictors for 'overnight improvement', instead modeling the improvement process as a gradual change occurring over time; there doesn't seem to be 'one quick trick to instantly improve your rating by 100 points'.

To be clear, the progression results are less a claim that playing more does not matter than a claim that raw volume has limited additional explanatory value once game complexity, repertoire breadth, and repertoire adjustment are taken into account. In that sense, the retained models suggest diminishing returns to volume alone and a comparatively stronger role for adaptive engagement. 

Overall, the progression results favor a development story centered on adaptation and complexity, which contrasts with a simpler 'play more and optimize more' account. @sec-metagame-dynamics-3 and @sec-metagame-and-opening-discussion make this pattern plausible, since the ecosystem itself is diverse and players appear to move among behavioral configurations. Improvement may therefore be best understood as an adaptive process rather than a purely accumulative one. 

### Behavioral Profiles as Descriptive States {#sec-categorization-discussion}

The clustering models are in somewhat of a tricky spot, because even though they were largely successful, they're unfit for inference. As such, this section makes no attempts to provide conclusive or inferential claims, and instead intends to ask what it means for the clustering to have succeeded and the potential implications of that success.

The retained clustering solution differentiates player-month among multiple behavioral axes: skill, repertoire breadth, opening switching, game length and draw structure, and activity. For example, cluster 1 contains high-diversity, high-activity, and above-average-rated players, while cluster 2 contrasts this with, with lower-rated, low-activity, narrower-repertoire players. This is salient because it means the model isn't simply reproducing the Lichess skill ranking, and instead identifies distinct configurations of chess behavior.

Were the clusters sharply stable, they might look like enduring player types, but because persistence is moderate and transitions between clusters are common, they're better interpreted as recurring behavioral states or configurations. This descriptive-state reading is consistent with the rest of the results: A broadening opening ecosystem and adaptive progression patterns both suggest behavioral movement, so the clusters may therefore reflect temporary strategic priorities, developmental phases, or context-dependent styles. This makes them useful as summaries of behavior, but not as definitive labels of what a player 'is'.

Ultimately, although the retained clustering solution is not itself inferential, its success suggest that player behavior is structured enough to motivate more explicit state-based analyses in future work. Because the clusters are both interpretable and fluid, a natural next step would be to model transitions among behavioral profiles directly and ask how those movements relate to development, format, or repertoire change. In short, the fact that the models work so well at classifying player behavior is more important than any of their descriptive interpretations.

### Practical Uses and Broader Implications {#sec-implications-discussion}

The models presented here provide a general framework for studying chess behavior across multiple levels. Those levels include metagame structure, opening choice, time-use behavior, player progression, and descriptive behavioral profiling. Because each level showed interpretable structure, the framework appears reusable beyond the bounds of this paper and its accompanying dataset. Future work could apply the same framework to platform comparisons, longer time periods, or more explicit transition modeling.

The results also point toward potential practical uses for coaching and training contexts. The progression models imply that richer and more adaptive engagement may be more informative than game volume alone. The time-use models imply that average pace and low-time pressure should be treated as separate aspects of player behavior. Together, this suggests that player-facing tools could offer more useful diagnostics by tracking breadth and adjustment rather than only game counts or raw performance outcomes.

Beyond direct practical use, the results suggest a broader method of understanding chess behavior, indicating that the multiple behavioral levels modeled are interconnected pieces of a greater whole as opposed to unrelated outcomes. This is important because these domains are often studied independently. The broader value of the retained models is therefore not just that they fit these data, but that they help organize chess-behavior research as a multi-level system.

### Limitations {#sec-limitations-discussion}

The retained model results are laboring under several limitations, two of which are all-encompassing. First, this entire project is observational in nature, and thus cannot make strong causal claims. All inference presented in this work is therefore limited to associations between variables rather than claims that one directly influences the other. For example, opening diversity and switching are associated with rating increase, but the models do not prove that increasing diversity or switching causes improvement; similarly, time-use patterns should not be read as causal mechanisms of strength and instead as correlated behavioral signatures. Even the fixed-effects models, while stronger than purely cross-sectional comparisons, still do not eliminate all time-varying confounding. This is the project's most important limitation: That all retained results are derived from observational data, so even the strongest inferential findings must be interpreted as such.

Second, the data gathered for the project was intentionally limited in size and scope— size for reproducibility purposes, and scope because the project set out to explore several facets of online chess behavior. In a more focused study, the modeling process would have access to a dataset tailored to the project's specific goals (e.g., the inclusion of move data to better model individual games). Greater specificity can also indirectly afford larger datasets, as those built to accommodate multiple different modeling goals effectively 'waste' space by including fields that aren't necessary for all research questions. Admittedly, this is not a limitation of the core modeling framework so much as a consequence of the format this project was built upon and the concern for reproducibility therein, but the reason for a limitations existence does not obviate its effect.

Also related to the scope and scale of the data is granularity. The game of chess is an unfathomably-complex strategic space, to the point that it's functionally impossible to model in full detail; there is simply not a computer on the planet that could even get close to accomplishing such a task. To that end, each model necessarily compresses a much richer behavioral space into something capable of being modeled. Part of this compression happened during the parsing pipeline (see @sec-source), but some models took additional steps to further simplify the data. For instance, the top-$k$ opening models rely on a large residual Other category, which is substantively meaningful but still aggregates many heterogeneous openings; some retained models depend on linear or otherwise simple functional forms; and some estimates, especially in slower formats or sparse settings, are less stable or more uncertain.

Finally, it must be acknowledged that the data used for modeling are both relative to a specific platform in Lichess and dependent on the time period studied. The results should accordingly not be used to describe online chess behavior as a whole, but interpreted in the context of the Lichess population from January 2024 through December 2025. It might be that other platforms like Chess.com show similar results or that the Lichess ecosystem behaved very similarly in years prior to 2024, but it's equally possible that those alternate populations behave in entirely different manners.

<!--
\newpage
-->

### Furture Work {#sec-future-work}

The clearest and most immediate next step is to apply the retained modeling framework to richer and more-targeted datasets that allow greater depth within each substantive domain. This project was intentionally broad and reproducible, which made it possible to study several facets of online chess behaviors at once, but limited how deeply any one domain could be modeled. Future work could accordingly trade breadth for depth by using larger and more specialized datasets to model particular questions with greater specificity. Potential avenues include move-level game data, richer opening annotations, and more detailed and longer-spanning player histories.

The current results also point to several specific questions that deserve more direct study. Included among such questions are: whether metagame broadening reflects experimentation or mere long-tail expansion; whether low-time frequency among stronger players reflects sharper positions or more selective time investment; and whether repertoire adjustment is a genuine precursor to improvement or simply a correlate of broader behavioral change. Follow-up studies could test these mechanisms more directly by building models around the unresolved processes suggested by our results rather than only refining the same descriptive summaries.

A final priority for future work is to evaluate the external generality of the present findings. Our results are tied to Lichess and to a specific time window, making cross-platform and cross-period comparison especially important. Replication on other platforms or different time periods would help identify which patterns are platform-specific and which are more general features of online chess. If the framework proves stable across settings, it could support broader comparative chess research with possible wide-ranging implications.


<!--
\newpage
## References
List the reference used for your paper. This should include anything cited in your background section and the source of your data.
-->

\newpage
## Appendix {#sec-appendix}
### Appendix A Graphs (Use appropriate name)
```{r appendix A, echo=FALSE}
# Use this code to display useful graphs that do not belong in the body of the paper.
# For example use this area for model validation graphs like residual versus fit.
```

#### A.1 (Use appropriate name)
If showing multiple graphs make sure to index them by 0.1 each


\newpage

### Appendix B Tables (Use appropriate name)
```{r appendix B, echo=FALSE}
# Use this area to show results that are in table format
```

#### Appendix B.1 (Use appropriate name)
If showing multiple tables make sure to index them by 0.1 each


### Code
```{r cleaning,echo=FALSE,eval=FALSE}
# Write your cleaning code here
```


```{r ref.label=knitr::all_labels()}
#| echo: true
#| eval: false
```

## Cleanup (Header To-Be Removed for Final Draft)

```{r Cleanup, output = FALSE}
if (exists("con") && DBI::dbIsValid(con)) {
  DBI::dbDisconnect(con, shutdown = TRUE)
}
```





