---
title: "EDA"
format: html
---

```{r Setup}
library(arrow)
library(DBI)
library(duckdb)
library(dplyr)
library(ggplot2)
library(scales)
library(ggthemes)

# Paths
base_dir <- "../data/derived/full_datasets"
paths <- list(
  pm = file.path(base_dir, "player_month.parquet"),
  pm_tc = file.path(base_dir, "player_month_tc.parquet"),
  pom_tc = file.path(base_dir, "player_opening_month_tc.parquet"),
  ptm_tc = file.path(base_dir, "player_time_month_tc.parquet"),
  mom_tc = file.path(base_dir, "meta_opening_month_tc.parquet")
)

con <- dbConnect(duckdb(), dbdir=":memory:")
on.exit(dbDisconnect(con, shutdown=TRUE), add=TRUE)

# Safe read_parquet helper using path quoting
qpath <- function(p) paste0("'", gsub("'", "''", normalizePath(p, winslash="/", mustWork=FALSE)), "'")
```

```{r Coverage}
cov_pm_tc <- dbGetQuery(con, sprintf("
  SELECT month, tc_bucket, COUNT(*) AS n_rows
  FROM read_parquet(%s)
  GROUP BY month, tc_bucket
  ORDER BY month, tc_bucket
", qpath(paths$pm_tc)))

cov_mom_tc <- dbGetQuery(con, sprintf("
  SELECT month, tc_bucket, COUNT(*) AS n_rows
  FROM read_parquet(%s)
  GROUP BY month, tc_bucket
  ORDER BY month, tc_bucket
", qpath(paths$mom_tc)))

ggplot(cov_pm_tc, aes(x = month, y = n_rows, group = tc_bucket, color = tc_bucket)) +
  geom_line(linewidth = 1) +
  coord_cartesian(clip = "off") +
  labs(x = "Month", y = "Games (Player-Month-TC)", title = "Coverage of Player-Month-TC Table", color = "Time Control") +
  scale_color_few(labels = c("Blitz", "Classical", "Rapid")) +
  theme_few() + 
    theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  ) +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

ggplot(cov_mom_tc, aes(x = month, y = n_rows, group = tc_bucket, color = tc_bucket)) +
  geom_line() +
  labs(x = "Month", y = "Rows (Opening-Month-TC)", title = "Coverage of Opening-Level Metagame Table", color = "Time Control") +
  scale_color_discrete(labels = c("Blitz", "Rapid", "Classical")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

This reveals a pretty big oversight that I somehow forgot to account for. That being: The sampling method I implemented did not group by time control before collecting and parsing the data. As a consequence, this means that the number of observations for each time control are proportional to the full dataset; there's an inequality in the quantity of observations for each time control (particularly classical, which is much lower). I must decide whether it's worth my time to re-parse the data to rectify this or if it's okay to abandon the classical TC in some (but hopefully not all) cases.

It's worth noting that the reason the parser doesn't group by time control first is because it does its hash-based sampling *before* unpacking each PGN and reading its header. Doing so massively improved parse time because the pipeline only has to process a small fraction of each month's dataset (compared to wasting a ton of time/compute by opening everything and then throwing upwards of 95% of it away), meaning this might just be something I have to live with. In either case, this should be addressed in the EDA section.

```{r Opening Trends}
top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz' AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT 8
", qpath(paths$mom_tc)))

open_list <- paste0("('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')")

trend <- dbGetQuery(con, sprintf("
  WITH filtered AS (
    SELECT month, tc_bucket, opening, games
    FROM read_parquet(%s)
    WHERE tc_bucket='blitz'
      AND opening IN %s
      AND opening IS NOT NULL
  ),
  opening_level AS (
    SELECT
      month,
      tc_bucket,
      opening,
      SUM(games) AS games
    FROM filtered
    GROUP BY month, tc_bucket, opening
  )
  SELECT
    month,
    tc_bucket,
    opening,
    games,
    CAST(games AS DOUBLE) / SUM(games) OVER (PARTITION BY month, tc_bucket) AS share
  FROM opening_level
  ORDER BY month, opening
", qpath(paths$mom_tc), open_list))


ggplot(trend, aes(x=month, y=share, color=opening, group=opening)) +
  geom_line() +
  scale_y_continuous(labels=percent_format(accuracy=0.1)) +
  labs(x="Month", y="Share of Games", color = "Opening", 
  title="Opening Share Over Time (Blitz)", subtitle = "Top Eight Openings in Terms of Game Share") +
  scale_color_few() +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom",
    axis.text.x = element_text(angle=60, hjust=1)
  )
```

I had a feeling that metagame drift wouldn't be crazy over only a two-month period, but the results surprised me. The code chunk above only produces one plot, but I checked all three and found that opening preferences are actually quite stable. As time controls give more time, the variance increases, from blitz being the most stable and classical the least. Were my only goal to answer this question, I would have sampled the first couple months over, say, a ten-year span rather than two full years, but I parsed and cleaned my data with a wide range of questions in mind, which necessarily weakens my ability to look into each specific question (address in EDA).

- ANOVA?

```{r Popularity vs Success}
cross <- dbGetQuery(con, sprintf("
  SELECT month, tc_bucket, opening, share, white_win_rate, draw_rate, upset_rate, games
  FROM read_parquet(%s)
  WHERE month='2025-06'
    AND opening IS NOT NULL
    AND games >= 200   -- threshold to avoid tiny groups
", qpath(paths$mom_tc)))

ggplot(cross, aes(x=share, y=white_win_rate)) +
  geom_point(alpha=0.4) +
  facet_wrap(~tc_bucket, scales="free_x") +
  scale_x_continuous(labels=percent) +
  labs(x="Opening Share", y="White Win Rate", title="Opening Popularity vs Success (Single-Month Sample)") +
  theme_minimal()
```

- Possible Lichess analysis bias (e.g., high-success lines are traps which only get categorized if sprung, leading to abnormally-high win rates).
- ELO/Glicko almost certainly influences this as well, since it's designed to make win rates converge to 50%.
- Ultimately, this line of questioning isn't very interesting because of these factors; look into opening share by rating instead.

```{r Games per Player Month}
gdist <- dbGetQuery(con, sprintf("
  SELECT games
  FROM read_parquet(%s)
", qpath(paths$pm)))

ggplot(gdist, aes(x=games)) +
  geom_histogram(bins = 400, fill = "steelblue", color = "steelblue") +
  coord_cartesian(xlim = c(0, 250)) +
  scale_y_continuous(labels = label_comma()) +
  labs(x="Games per Player in any Given Month", y="Count", title="Player-Month Reliability Varies Widely") +
  theme_minimal()
```

```{r Winrate by Rating}
wr_bin <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    floor(avg_rating/100)*100 AS rating_bin,
    AVG(win_rate) AS mean_win_rate,
    COUNT(*) AS n_rows,
    SUM(games) AS total_games
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_games >= 5000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$pm_tc)))

ggplot(wr_bin, aes(x=rating_bin, y=mean_win_rate)) +
  geom_line() +
  facet_wrap(~tc_bucket) +
  labs(x="Rating bin", y="Mean win rate", title="Win rate patterns across rating strata and time controls")
```

```{r Rating Changes}
chg <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT (avg_rating - prev_rating) AS d_rating
  FROM y
  WHERE prev_rating IS NOT NULL AND d_rating >= -300 AND d_rating <= 300
", qpath(paths$pm)))

ggplot(chg, aes(x=d_rating)) +
  geom_histogram(bins=80, fill = "steelblue", color = "steelblue") +
  labs(x="Month-to-Month Change in Average Rating", y="Count", title="Distribution of Within-Player Rating Change") +
  theme_few()
```

```{r Time Use}
time_bin <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    floor(avg_rating/100)*100 AS rating_bin,
    AVG(avg_time_spent) AS mean_time_spent,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 20000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

ggplot(time_bin, aes(x=rating_bin, y=mean_time_spent)) +
  geom_line() +
  facet_wrap(~tc_bucket, scales="free_y") +
  labs(x="Player Rating", y="Average Time per Move (Seconds)", title="Thinking Time Varies by Skill and Time Control") +
  theme_minimal()
```

```{r Opening by Rating}
tc_use <- "blitz" # "blitz" / "rapid" / "classical"
bin_width <- 200 
top_k <- 8 # number of openings to plot
min_bin_total <- 2000  # min player-games in a rating bin

# Pick top openings within tc_bucket (weighted by games)
top_openings <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT opening, SUM(games) AS total_games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IS NOT NULL
    GROUP BY opening
  )
  SELECT opening
  FROM x
  ORDER BY total_games DESC
  LIMIT %d
", qpath(paths$pom_tc), tc_use, top_k))

open_list <- paste0(
  "('", paste(gsub("'", "''", top_openings$opening), collapse="','"), "')"
)

# Compute opening share within rating bins
## Rating_bin_lo is based on avg_rating from the player-opening aggregation
## Use SUM(games) as weights, so share is in terms of player-games
trend_rating <- dbGetQuery(con, sprintf("
  WITH base AS (
    SELECT
      tc_bucket,
      opening,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE tc_bucket = '%s'
      AND opening IN %s
      AND opening IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  binned AS (
    SELECT
      opening,
      CAST(FLOOR(avg_rating / %d) * %d AS INTEGER) AS rating_bin_lo,
      SUM(games) AS player_games
    FROM base
    GROUP BY opening, rating_bin_lo
  ),
  totals AS (
    SELECT
      rating_bin_lo,
      SUM(player_games) AS bin_total
    FROM binned
    GROUP BY rating_bin_lo
  )
  SELECT
    b.rating_bin_lo,
    (b.rating_bin_lo + %d - 1) AS rating_bin_hi,
    b.opening,
    b.player_games,
    t.bin_total,
    CAST(b.player_games AS DOUBLE) / t.bin_total AS share
  FROM binned b
  JOIN totals t USING (rating_bin_lo)
  WHERE t.bin_total >= %d
  ORDER BY b.rating_bin_lo, b.opening
", qpath(paths$pom_tc), tc_use, open_list,
   bin_width, bin_width, bin_width, min_bin_total))

# Nicer labels
trend_rating <- trend_rating |>
  mutate(
    rating_bin = paste0(rating_bin_lo, "\u2013", rating_bin_hi),
    rating_bin = factor(rating_bin, levels = unique(rating_bin))
  )


ggplot(trend_rating, aes(x = rating_bin, y = share, color = opening, group = opening)) +
  geom_line() +
  geom_point(size = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = "Rating",
    y = "Share of Player-Games",
    title = paste0("Opening Share by Rating (", tc_use, ")"),
    color = "Opening"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    legend.position = "bottom"
  )
```

```{r Months Observed per Player}
panel_len <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, COUNT(DISTINCT month) AS n_months
    FROM read_parquet(%s)
    GROUP BY player
  )
  SELECT n_months, COUNT(*) AS n_players
  FROM x
  GROUP BY n_months
  ORDER BY n_months
", qpath(paths$pm)))

ggplot(panel_len, aes(x = n_months, y = n_players)) +
  geom_col(fill = "steelblue", color = "steelblue") +
  scale_y_continuous() +
  labs(x = "Months Observed per Player", y = "Players",
       title = "Panel Depth: How Long Players Remain Observable") +
  theme_few()
```

```{r Consecutive Streaks}
streaks <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month
    FROM read_parquet(%s)
    GROUP BY player, month
  ),
  y AS (
    SELECT
      player,
      month,
      STRPTIME(month || '-01', '%%Y-%%m-%%d') AS mdate
    FROM x
  ),
  z AS (
    SELECT
      player,
      mdate,
      mdate - INTERVAL '1 month' * ROW_NUMBER() OVER (PARTITION BY player ORDER BY mdate) AS grp
    FROM y
  ),
  streaks AS (
    SELECT player, COUNT(*) AS streak_len
    FROM z
    GROUP BY player, grp
  )
  SELECT streak_len, COUNT(*) AS n_streaks
  FROM streaks
  GROUP BY streak_len
  ORDER BY streak_len
", qpath(paths$pm)))

ggplot(streaks, aes(x=streak_len, y=n_streaks)) +
  geom_col(fill = "steelblue", color = "steelblue") +
  scale_y_continuous() +
  labs(x="Consecutive-month streak length", y="Streaks",
       title="How often do we observe players in consecutive months?") +
  theme_minimal()

```

```{r Coverage Heatmap}
cov_heat <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    month,
    FLOOR(avg_rating/200)*200 AS rating_bin,
    SUM(games) AS total_games
  FROM read_parquet(%s)
  GROUP BY tc_bucket, month, rating_bin
", qpath(paths$pm_tc)))

ggplot(cov_heat, aes(x=month, y=rating_bin, fill=total_games)) +
  geom_tile() +
  facet_wrap(~tc_bucket, scales="free_x") +
  labs(x="Month", y="Rating bin", title="Coverage heatmap (player-games)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle=60, hjust=1))

```

```{r Games per Month vs Volatility in Delta-Rating}
vol <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, games, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      games,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT
    games,
    ABS(avg_rating - prev_rating) AS abs_d_rating
  FROM y
  WHERE prev_rating IS NOT NULL
", qpath(paths$pm)))

ggplot(vol, aes(x = games, y = abs_d_rating)) +
  geom_point(alpha = 0.15) +
  scale_x_continuous(trans = "log10") +
  labs(x = "Games per Month (Log-Transformed)", y = "Absolute Change in Rating (|Δ|)",
       title = "Monthly Rating Volatility Shrinks as Game Volume Rises",
       subtitle = "Log-Transformed for Legibility") +
  theme_few()
```

```{r Delta-Rating by Rating Bin}
chg_bin <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      avg_rating,
      LAG(avg_rating) OVER (PARTITION BY player ORDER BY month) AS prev_rating
    FROM x
  )
  SELECT
    FLOOR(prev_rating/200)*200 AS rating_bin,
    (avg_rating - prev_rating) AS d_rating
  FROM y
  WHERE prev_rating IS NOT NULL
", qpath(paths$pm)))

ggplot(chg_bin, aes(x=d_rating)) +
  geom_histogram(bins=60) +
  facet_wrap(~rating_bin, scales="free_y") +
  labs(x="Δ rating", y="Count",
       title="Rating change distribution varies by starting skill") +
  theme_minimal()

```

```{r Opening Diversity vs Rating Change}
div_vs_change <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, opening_diversity, games, avg_rating
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      opening_diversity,
      games,
      avg_rating,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating
    FROM x
  )
  SELECT
    opening_diversity,
    games,
    (next_rating - avg_rating) AS d_rating_next
  FROM y
  WHERE next_rating IS NOT NULL
", qpath(paths$pm)))

ggplot(div_vs_change, aes(x=opening_diversity, y=d_rating_next)) +
  geom_point(alpha=0.12) +
  geom_smooth(se=FALSE) +
  labs(x="Opening diversity (month)", y="Next-month Δ rating",
       title="Is repertoire breadth associated with improvement?") +
  theme_minimal()

```

```{r Specialist vs Generalist}
conc <- dbGetQuery(con, sprintf("
  WITH po AS (
    SELECT player, month, tc_bucket, opening, games
    FROM read_parquet(%s)
  ),
  tot AS (
    SELECT player, month, tc_bucket, SUM(games) AS total_games
    FROM po
    GROUP BY player, month, tc_bucket
    HAVING SUM(games) >= 5
  ),
  top AS (
    SELECT
      player, month, tc_bucket,
      MAX(games) AS top_opening_games
    FROM po
    GROUP BY player, month, tc_bucket
  )
  SELECT
    t.player, t.month, t.tc_bucket,
    top_opening_games,
    total_games,
    CAST(top_opening_games AS DOUBLE)/total_games AS top_opening_share
  FROM top
  JOIN tot t USING (player, month, tc_bucket)
", qpath(paths$pom_tc)))

ggplot(conc, aes(x = top_opening_share)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "steelblue") +
  scale_x_continuous(labels=percent_format(accuracy = 1)) +
  labs(x="Share of Games in Most-Played Opening", y = "Player-Months",
       title="Repertoire Concentration") +
  theme_few()

```

```{r Opening Switching vs Improvement}
switch <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, top_opening, avg_rating, games
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player,
      month,
      top_opening,
      avg_rating,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating,
      LAG(top_opening) OVER (PARTITION BY player ORDER BY month) AS prev_opening
    FROM x
  )
  SELECT
    CASE WHEN prev_opening IS NULL THEN NULL
         WHEN top_opening = prev_opening THEN 0 ELSE 1 END AS switched_opening,
    (next_rating - avg_rating) AS d_rating_next
  FROM y
  WHERE next_rating IS NOT NULL AND prev_opening IS NOT NULL
", qpath(paths$pm)))

ggplot(switch, aes(x=factor(switched_opening), y=d_rating_next)) +
  geom_boxplot() +
  labs(x="Changed top opening vs previous month", y="Next-month Δ rating",
       title="Does changing openings relate to improvement?") +
  theme_minimal()

```

```{r Time Use vs Progression}
ltr <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, tc_bucket, low_time_rate, avg_rating, move_rows
    FROM read_parquet(%s)
  ),
  y AS (
    SELECT
      player, month, tc_bucket, low_time_rate, avg_rating, move_rows,
      LEAD(avg_rating) OVER (PARTITION BY player, tc_bucket ORDER BY month) AS next_rating
    FROM x
  )
  SELECT
    tc_bucket,
    low_time_rate,
    move_rows,
    (next_rating - avg_rating) AS d_rating_next
  FROM y
  WHERE next_rating IS NOT NULL AND low_time_rate IS NOT NULL
", qpath(paths$ptm_tc)))

ggplot(ltr, aes(x=low_time_rate, y=d_rating_next)) +
  geom_point(alpha=0.12) +
  geom_smooth(se=FALSE) +
  facet_wrap(~tc_bucket) +
  scale_x_continuous(labels=percent_format(accuracy=1)) +
  labs(x="Low-time rate", y="Next-month Δ rating",
       title="Is time trouble associated with slower improvement?") +
  theme_minimal()

```

```{r Allocation Profile by Rating}
phase <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    FLOOR(avg_rating/200)*200 AS rating_bin,
    AVG(avg_time_opening) AS t_open,
    AVG(avg_time_middlegame) AS t_mid,
    AVG(avg_time_endgame) AS t_end,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 10000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

phase_long <- phase |>
  pivot_longer(cols=c(t_open, t_mid, t_end),
                      names_to="phase", values_to="sec")

ggplot(phase_long, aes(x=rating_bin, y=sec, color=phase, group=phase)) +
  geom_line() +
  facet_wrap(~factor(tc_bucket, levels = c("blitz", "rapid", "classical")), scales="free_y", labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))) +
  labs(x="Rating", y="Seconds per Move", color = "Phase",
       title="Phase Time Allocation Changes With Skill") +
  scale_color_few(labels = c("Endgame", "Middlegame", "Opening")) +
    theme_few() + 
    theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```

```{r Missingness}
miss <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    FLOOR(avg_rating/200)*200 AS rating_bin,
    AVG(CASE WHEN avg_time_endgame IS NULL THEN 1 ELSE 0 END) AS endgame_missing_rate,
    SUM(move_rows) AS total_moves
  FROM read_parquet(%s)
  GROUP BY tc_bucket, rating_bin
  HAVING total_moves >= 20000
  ORDER BY tc_bucket, rating_bin
", qpath(paths$ptm_tc)))

ggplot(miss, aes(x=rating_bin, y=endgame_missing_rate)) +
  geom_line() +
  facet_wrap(~tc_bucket) +
  scale_y_continuous(labels=percent_format(accuracy=1)) +
  labs(x="Rating bin", y="Endgame metric missing rate",
       title="Endgame phase metrics are missing non-randomly") +
  theme_minimal()

```

```{r Distribution Sanity Checks}
dbGetQuery(con, sprintf("
  SELECT COUNT(*) AS n
  FROM read_parquet(%s)
", qpath(paths$ptm_tc)))

# Distribution sanity checks
dbGetQuery(con, sprintf("
  SELECT
    MIN(games) AS min_games,
    APPROX_QUANTILE(games, 0.5) AS med_games,
    APPROX_QUANTILE(games, 0.9) AS p90_games,
    MAX(games) AS max_games,
    MIN(move_rows) AS min_moves,
    APPROX_QUANTILE(move_rows, 0.5) AS med_moves,
    APPROX_QUANTILE(move_rows, 0.9) AS p90_moves,
    MAX(move_rows) AS max_moves,
    AVG(CASE WHEN low_time_rate = 0 THEN 1 ELSE 0 END) AS frac_ltr_zero
  FROM read_parquet(%s)
  WHERE low_time_rate IS NOT NULL
", qpath(paths$ptm_tc)))

# How many rows have a "next" month inside ptm_tc?
dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, tc_bucket, avg_rating
    FROM read_parquet(%s)
    WHERE avg_rating IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating) OVER (PARTITION BY player, tc_bucket ORDER BY month) AS next_rating
    FROM x
  )
  SELECT
    COUNT(*) AS n_total,
    SUM(CASE WHEN next_rating IS NOT NULL THEN 1 ELSE 0 END) AS n_with_next
  FROM y
", qpath(paths$ptm_tc)))

```


```{r LTR Bin}
min_games_pm <- 10  # require at least this many games in the month (across all time controls)
min_moves_ptm <- 45  # require at least this many move rows in the month (across all time controls)

ltr_scatter_all <- dbGetQuery(con, sprintf("
  WITH pm_tc AS (
    SELECT
      player,
      month,
      avg_rating,
      games
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  pm AS (
    -- collapse across time controls using games-weighted avg_rating
    SELECT
      player,
      month,
      SUM(games) AS games_total,
      SUM(avg_rating * games) / SUM(games) AS avg_rating_all
    FROM pm_tc
    GROUP BY player, month
    HAVING SUM(games) >= %d
  ),
  pm_next AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      LEAD(avg_rating_all) OVER (PARTITION BY player ORDER BY month) AS next_rating_all
    FROM pm
  ),
  delta AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      (next_rating_all - avg_rating_all) AS d_rating_next
    FROM pm_next
    WHERE next_rating_all IS NOT NULL
  ),
  ptm_tc AS (
    SELECT
      player,
      month,
      low_time_rate,
      move_rows
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND low_time_rate IS NOT NULL
      AND move_rows IS NOT NULL
  ),
  ptm AS (
    -- collapse across time controls using move_rows-weighted low_time_rate
    SELECT
      player,
      month,
      SUM(move_rows) AS move_rows_total,
      SUM(low_time_rate * move_rows) / SUM(move_rows) AS low_time_rate_all
    FROM ptm_tc
    GROUP BY player, month
    HAVING SUM(move_rows) >= %d
  )
  SELECT
    d.player,
    d.month,
    d.avg_rating_all,
    d.games_total,
    d.d_rating_next,
    p.low_time_rate_all,
    p.move_rows_total
  FROM delta d
  JOIN ptm p USING (player, month)
", qpath(paths$pm_tc), min_games_pm,
   qpath(paths$ptm_tc), min_moves_ptm))

ggplot(ltr_scatter_all, aes(x = low_time_rate_all, y = d_rating_next)) +
  geom_point(alpha = 0.25) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  ylim(-225, 225) +
  labs(
    x = "Low-Time Rate",
    y = "Next-Month Δ Rating",
    title = "Time Trouble vs Next-Month Rating Change",
    subtitle = "Collapsed Across Time Controls"
  ) +
  theme_minimal()

```

```{r Switch 2}
switch2 <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, top_opening, avg_rating, games
    FROM read_parquet(%s)
    WHERE top_opening IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating,
      LEAD(games) OVER (PARTITION BY player ORDER BY month) AS next_games,
      LAG(top_opening) OVER (PARTITION BY player ORDER BY month) AS prev_opening
    FROM x
  )
  SELECT
    CAST(FLOOR(avg_rating/200)*200 AS INTEGER) AS rating_bin,
    CASE WHEN top_opening = prev_opening THEN 0 ELSE 1 END AS switched_opening,
    (next_rating - avg_rating) AS d_rating_next
  FROM y
  WHERE next_rating IS NOT NULL
    AND prev_opening IS NOT NULL
    AND games >= 20
    AND next_games >= 20
", qpath(paths$pm)))

ggplot(switch2, aes(x=factor(switched_opening), y=d_rating_next)) +
  geom_boxplot() +
  facet_wrap(~rating_bin) +
  labs(x="Changed top opening vs previous month", y="Next-month Δ rating",
       title="Opening switching vs improvement (stable months only)") +
  theme_minimal()

```

```{r Div Strata}
div_strata <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, opening_diversity, games, avg_rating
    FROM read_parquet(%s)
    WHERE opening_diversity IS NOT NULL
      AND games IS NOT NULL
      AND avg_rating IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating
    FROM x
  ),
  z AS (
    SELECT
      CAST(FLOOR(avg_rating/200)*200 AS INTEGER) AS rating_bin,
      CASE
        WHEN games < 10 THEN '<10'
        WHEN games < 25 THEN '10–24'
        WHEN games < 60 THEN '25–59'
        ELSE '60+'
      END AS games_bin,
      opening_diversity,
      (next_rating - avg_rating) AS d_rating_next
    FROM y
    WHERE next_rating IS NOT NULL
  )
  SELECT
    rating_bin,
    games_bin,
    opening_diversity,
    AVG(d_rating_next) AS mean_d,
    COUNT(*) AS n
  FROM z
  GROUP BY rating_bin, games_bin, opening_diversity
  HAVING n >= 5
  ORDER BY rating_bin, games_bin, opening_diversity
", qpath(paths$pm)))

div_strata <- div_strata |>
  mutate(games_bin = factor(games_bin, levels = c("60+", "25–59", "10–24", "<10")))

ggplot(div_strata, aes(x=opening_diversity, y=mean_d, color=games_bin)) +
  geom_line(alpha = 1) +
  facet_wrap(~rating_bin, scales="free_y") +
  labs(x="Opening Diversity (Within-Month)", y="Mean Next-Month Δ Rating",
       color="Games",
       title="Opening Diversity vs Improvement",
       subtitle = "Stratified by Games and Rating") +
  scale_color_few() +
  theme_few() + 
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

```

```{r D3}
d3 <- dbGetQuery(con, sprintf("
  WITH x AS (
    SELECT player, month, avg_rating, games, opening_diversity
    FROM read_parquet(%s)
    WHERE avg_rating IS NOT NULL
  ),
  y AS (
    SELECT *,
      LEAD(avg_rating, 3) OVER (PARTITION BY player ORDER BY month) AS rating_3m
    FROM x
  )
  SELECT
    opening_diversity,
    games,
    (rating_3m - avg_rating) AS d_rating_3m
  FROM y
  WHERE rating_3m IS NOT NULL
    AND games >= 15
", qpath(paths$pm)))

ggplot(d3, aes(x=opening_diversity, y=d_rating_3m)) +
  geom_point(alpha=0.08) +
  geom_smooth(se=FALSE) +
  labs(x="Opening diversity", y="3-month Δ rating",
       title="Repertoire breadth vs longer-horizon improvement") +
  theme_minimal()

```

```{r Repertoire Churn vs Improvement}
min_games_pm  <- 20 # require stable months for rating deltas
min_games_pom <- 10 # require enough opening games to define a repertoire
bin_width <- 200 # rating bin width for faceting

churn <- dbGetQuery(con, sprintf("
  WITH pom AS (
    -- collapse across time controls -> player-month-opening counts
    SELECT
      player,
      month,
      opening,
      SUM(games) AS games_opening
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND opening IS NOT NULL
      AND games IS NOT NULL
    GROUP BY player, month, opening
  ),
  pom_tot AS (
    SELECT
      player,
      month,
      SUM(games_opening) AS games_total
    FROM pom
    GROUP BY player, month
    HAVING SUM(games_opening) >= %d
  ),
  pom_share AS (
    SELECT
      p.player,
      p.month,
      p.opening,
      p.games_opening,
      t.games_total,
      CAST(p.games_opening AS DOUBLE) / t.games_total AS share
    FROM pom p
    JOIN pom_tot t USING (player, month)
  ),
  indexed AS (
    SELECT
      player,
      month,
      STRPTIME(month || '-01', '%%Y-%%m-%%d') AS mdate,
      opening,
      share,
      games_total
    FROM pom_share
  ),
  paired AS (
    -- join month t with month t-1 by player and opening
    SELECT
      cur.player,
      cur.month AS month,
      cur.mdate AS mdate,
      cur.opening,
      cur.share AS share_cur,
      prev.share AS share_prev,
      cur.games_total AS games_total
    FROM indexed cur
    LEFT JOIN indexed prev
      ON cur.player = prev.player
     AND cur.opening = prev.opening
     AND prev.mdate = cur.mdate - INTERVAL '1 month'
  ),
  norms AS (
    -- compute vector norms and dot products per player-month
    SELECT
      player,
      month,
      mdate,
      MAX(games_total) AS games_total,
      SUM(share_cur * COALESCE(share_prev, 0.0)) AS dot,
      SQRT(SUM(share_cur * share_cur)) AS norm_cur,
      SQRT(SUM(COALESCE(share_prev, 0.0) * COALESCE(share_prev, 0.0))) AS norm_prev,
      -- share of games in openings not seen last month
      SUM(CASE WHEN share_prev IS NULL THEN share_cur ELSE 0.0 END) AS new_opening_share
    FROM paired
    GROUP BY player, month, mdate
  ),
  churn AS (
    SELECT
      player,
      month,
      games_total,
      CASE
        WHEN norm_cur = 0 OR norm_prev = 0 THEN NULL
        ELSE 1.0 - (dot / (norm_cur * norm_prev))
      END AS churn_cosine,
      new_opening_share
    FROM norms
    WHERE norm_prev > 0  -- require a prior month repertoire exists
  ),
  pm AS (
    SELECT
      player,
      month,
      avg_rating,
      games,
      LEAD(avg_rating) OVER (PARTITION BY player ORDER BY month) AS next_rating,
      LEAD(games)      OVER (PARTITION BY player ORDER BY month) AS next_games
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month  IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
      AND games >= %d
  ),
  pm_delta AS (
    SELECT
      player,
      month,
      avg_rating,
      games,
      (next_rating - avg_rating) AS d_rating_next
    FROM pm
    WHERE next_rating IS NOT NULL
      AND next_games IS NOT NULL
      AND next_games >= %d
  )
  SELECT
    c.player,
    c.month,
    c.games_total AS opening_games,
    d.games      AS pm_games,
    d.avg_rating,
    CAST(FLOOR(d.avg_rating/%d)*%d AS INTEGER) AS rating_bin,
    c.churn_cosine,
    c.new_opening_share,
    d.d_rating_next
  FROM churn c
  JOIN pm_delta d USING (player, month)
  WHERE c.churn_cosine IS NOT NULL
", qpath(paths$pom_tc),
   min_games_pom,
   qpath(paths$pm),
   min_games_pm, min_games_pm,
   bin_width, bin_width
))

# Bin churn for stable curves
churn <- churn |>
  mutate(
    churn_bin = cut(
      churn_cosine,
      breaks = c(-Inf, 0.02, 0.05, 0.10, 0.20, 0.35, 0.50, Inf),
      labels = c("≤2%", "2–5%", "5–10%", "10–20%", "20–35%", "35–50%", ">50%"),
      right = TRUE
    )
  )

churn_summ <- churn |>
  group_by(churn_bin) |>
  summarise(
    n = n(),
    mean_d = mean(d_rating_next, na.rm = TRUE),
    med_d  = median(d_rating_next, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(churn_bin))

ggplot(churn_summ, aes(x = churn_bin, y = mean_d)) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2) +
  labs(
    x = "Repertoire churn (1 - cosine similarity of opening-share vectors)",
    y = "Mean next-month Δ rating",
    title = "Does changing your opening mix relate to next-month improvement?"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

churn_new_summ <- churn |>
  mutate(
    new_bin = cut(
      new_opening_share,
      breaks = c(-Inf, 0.00, 0.05, 0.10, 0.20, 0.35, 0.50, Inf),
      labels = c("0%", "0–5%", "5–10%", "10–20%", "20–35%", "35–50%", ">50%"),
      right = TRUE
    )
  ) |>
  group_by(new_bin) |>
  summarise(
    n = n(),
    mean_d = mean(d_rating_next, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(new_bin))

ggplot(churn_new_summ, aes(x = new_bin, y = mean_d)) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2) +
  labs(
    x = "Share of games in openings not used last month",
    y = "Mean next-month Δ rating",
    title = "Improvement vs experimenting with new openings (month-to-month)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r LTR Binned vs Improvement}
min_games_pm <- 10
min_moves_ptm <- 40

ltr_bins <- dbGetQuery(con, sprintf("
  WITH pm_tc AS (
    SELECT
      player, month,
      avg_rating, games
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month IS NOT NULL
      AND avg_rating IS NOT NULL
      AND games IS NOT NULL
  ),
  pm AS (
    SELECT
      player,
      month,
      SUM(games) AS games_total,
      SUM(avg_rating * games) / SUM(games) AS avg_rating_all
    FROM pm_tc
    GROUP BY player, month
    HAVING SUM(games) >= %d
  ),
  pm_next AS (
    SELECT
      player,
      month,
      games_total,
      avg_rating_all,
      LEAD(avg_rating_all) OVER (PARTITION BY player ORDER BY month) AS next_rating
    FROM pm
  ),
  delta AS (
    SELECT
      player,
      month,
      games_total,
      (next_rating - avg_rating_all) AS d_rating_next
    FROM pm_next
    WHERE next_rating IS NOT NULL
  ),
  ptm_tc AS (
    SELECT
      player, month,
      low_time_rate, move_rows
    FROM read_parquet(%s)
    WHERE player IS NOT NULL
      AND month IS NOT NULL
      AND low_time_rate IS NOT NULL
      AND move_rows IS NOT NULL
  ),
  ptm AS (
    SELECT
      player,
      month,
      SUM(move_rows) AS move_rows_total,
      SUM(low_time_rate * move_rows) / SUM(move_rows) AS low_time_rate_all
    FROM ptm_tc
    GROUP BY player, month
    HAVING SUM(move_rows) >= %d
  )
  SELECT
    d.player,
    d.month,
    d.games_total,
    d.d_rating_next,
    p.move_rows_total,
    p.low_time_rate_all
  FROM delta d
  JOIN ptm p USING (player, month)
", qpath(paths$pm_tc), min_games_pm,
   qpath(paths$ptm_tc), min_moves_ptm
))

# Bin low-time rate to handle zero-mass
ltr_bins <- ltr_bins |>
  mutate(
    ltr_bin = cut(
      low_time_rate_all,
      breaks = c(-Inf, 0, 0.01, 0.02, 0.05, 0.10, 0.20, 0.40, Inf),
      labels = c("0%", "0–1%", "1–2%", "2–5%", "5–10%", "10–20%", "20–40%", ">40%"),
      right = TRUE
    )
  )

ltr_summ <- ltr_bins |>
  group_by(ltr_bin) |>
  summarise(
    n = n(),
    mean_d = mean(d_rating_next, na.rm = TRUE),
    med_d  = median(d_rating_next, na.rm = TRUE),
    mean_games = mean(games_total, na.rm = TRUE),
    mean_moves = mean(move_rows_total, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(!is.na(ltr_bin))

ggplot(ltr_summ, aes(x = ltr_bin, y = mean_d)) +
  geom_line(aes(group = 1)) +
  geom_point(size = 2) +
  labs(
    x = "Low-time rate (collapsed across time controls; move-weighted)",
    y = "Mean next-month Δ rating (games-weighted rating)",
    title = "Time trouble vs improvement (binned summary, no loess)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Show the sample size per bin
ggplot(ltr_summ, aes(x = ltr_bin, y = n)) +
  geom_col() +
  labs(
    x = "Low-time rate bin",
    y = "Player-months",
    title = "Support by low-time-rate bin"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}
bin_width <- 200

support_by_rating2 <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    CAST(FLOOR(avg_rating / %d) * %d AS INTEGER) AS rating_bin,
    COUNT(*) AS n_rows,
    SUM(games) AS total_games
  FROM read_parquet(%s)
  WHERE avg_rating IS NOT NULL
    AND games IS NOT NULL
    AND tc_bucket IS NOT NULL
  GROUP BY tc_bucket, rating_bin
  ORDER BY tc_bucket, rating_bin
", bin_width, bin_width, qpath(paths$pm_tc))) |>
  pivot_longer(
    cols = c(n_rows, total_games),
    names_to = "measure",
    values_to = "support"
  ) |>
  mutate(
    measure = recode(
      measure,
      n_rows = "Row Count",
      total_games = "Summed Games"
    ),
    tc_bucket = factor(tc_bucket, levels = c("blitz", "rapid", "classical"))
  )

ggplot(support_by_rating2, aes(x = rating_bin, y = support)) +
  geom_col(width = bin_width * 0.9, fill = "steelblue", color = "steelblue") +
  facet_grid(
    measure ~ tc_bucket,
    scales = "free_y",
    labeller = labeller(
      tc_bucket = c(blitz = "Blitz", rapid = "Rapid", classical = "Classical")
    )
  ) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = pretty_breaks(n = 7)) +
  labs(
    x = "Rating Bin",
    y = NULL,
    title = "Support Across Rating Bins by Time Control",
    subtitle = "Compares POM_TC Row Counts with Summed Games"
  ) +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  )
```

```{r}
min_moves_ptm <- 40
positive_binwidth <- 0.01

ltr_dist <- dbGetQuery(con, sprintf("
  SELECT
    tc_bucket,
    low_time_rate,
    move_rows
  FROM read_parquet(%s)
  WHERE tc_bucket IS NOT NULL
    AND low_time_rate IS NOT NULL
    AND move_rows IS NOT NULL
    AND move_rows >= %d
", qpath(paths$ptm_tc), min_moves_ptm))

zero_panel <- ltr_dist |>
  mutate(panel = "Zero mass", x_cat = if_else(low_time_rate == 0, "0%", ">0%")) |>
  count(tc_bucket, panel, x_cat, name = "n")

positive_panel <- ltr_dist |>
  filter(low_time_rate > 0) |>
  mutate(panel = "Positive support")

p1 <- ggplot(zero_panel, aes(x = x_cat, y = n)) +
  geom_col(fill = "steelblue", color = "steelblue") +
  facet_wrap(
    ~factor(tc_bucket, levels = c("blitz", "rapid", "classical")),
    labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))
  ) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = NULL,
    y = "Rows",
    title = "Exact Zero Mass"
  ) +
  theme_few()

p2 <- ggplot(positive_panel, aes(x = low_time_rate)) +
  geom_histogram(
    binwidth = positive_binwidth,
    boundary = 0,
    fill = "steelblue",
    color = "steelblue"
  ) +
  facet_wrap(
    ~factor(tc_bucket, levels = c("blitz", "rapid", "classical")),
    scales = "free_y",
    labeller = as_labeller(c(
      blitz = "Blitz",
      rapid = "Rapid",
      classical = "Classical"
    ))
  ) +
  scale_x_continuous(labels = percent_format(accuracy = 1)) +
  scale_y_continuous(labels = label_comma()) +
  labs(
    x = "Low-Time Rate",
    y = "Rows",
    title = "Low-Time Distribution"
  ) +
  theme_few()

p2 / p1
```





