---
title: "Modeling"
format: html
---

#### Setup

```{r Setup, message = FALSE, warning = FALSE}

# Core data + infrastructure
library(yaml)       # read pipeline.yml (paths, parameters)
library(arrow)      # Parquet IO
library(DBI)        # database interface
library(duckdb)     # fast SQL over Parquet

# Data manipulation + dates
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(lubridate)
library(forcats) # factor helpers (bins/labels)

# Visualization (EDA + diagnostics)
library(ggplot2)
library(ggthemes)
library(scales)
library(patchwork) # combine plots
library(kableExtra)

# Modeling utilities + inference
library(broom) # tidy model outputs
library(modelr) # add_predictions / resample helpers
library(sandwich) # robust/clustered SEs
library(lmtest) # coeftest with robust SEs

# Resampling / validation
library(rsample) # time splits, rolling origin
library(yardstick) # performance metrics

# Tidymodels umbrella
library(tidymodels)

# Metagame dynamics: GAM / smoothing
library(mgcv) # GAMs

# Opening popularity: multinomial + count models
library(nnet) # multinomial logistic regression (multinom)
library(VGAM) # flexible categorical models
library(MASS) # glm.nb (NegBin), rlm (robust regression)

# Compositional / proportion modeling
library(betareg) # beta regression for rates in (0,1)
library(DirichletReg) # Dirichlet regression for compositions

# Time-use: robust + quantile regression
library(quantreg) # quantile regression
library(robustbase) # additional robust methods

# Player progression: regularization + panel models
library(glmnet) # ridge/lasso/elastic net
library(fixest) # fixed effects, clustered SEs (panel-friendly)
library(lme4) # mixed effects (random effects)
library(glmmTMB) # mixed models + many families (optional but powerful)

# Tree-based predictive models
library(ranger) # random forest
library(xgboost) # gradient boosting

# Player categorization: dimensionality reduction + clustering
# PCA is in base stats (prcomp), but these help with nonlinear DR + clustering
library(uwot) # UMAP
library(cluster) # silhouette, clustering helpers
library(ClusterR) # Replacement for base-R kmeans
library(mclust) # Gaussian mixture models

# Package conflicts
library(conflicted)
conflicts_prefer(dplyr::slice)
conflicts_prefer(dplyr::filter)
conflicts_prefer(dplyr::lag)
conflicts_prefer(dplyr::select)
conflicts_prefer(dplyr::arrange)
conflicts_prefer(dplyr::mutate)
conflicts_prefer(dplyr::summarise)
conflicts_prefer(dplyr::count)
conflicts_prefer(purrr::map)

pm <- read_parquet("../data/derived/full_datasets/player_month.parquet")
pm_tc <- read_parquet("../data/derived/full_datasets/player_month_tc.parquet")
pom_tc <- read_parquet("../data/derived/full_datasets/player_opening_month_tc.parquet")
ptm_tc <- read_parquet("../data/derived/full_datasets/player_time_month_tc.parquet")
mm_tc <- read_parquet("../data/derived/full_datasets/meta_month_tc.parquet")
mom_tc <- read_parquet("../data/derived/full_datasets/meta_opening_month_tc.parquet")
mvm_tc <- read_parquet("../data/derived/full_datasets/moves_month_tc.parquet")
```

```{r Inference Helpers}

# Tidy coeftest output
tidy_ct <- function(ct) {
  tidy(ct) |>
    rename(
      estimate = estimate,
      std.error = std.error,
      statistic = statistic,
      p.value = p.value
    )
}

# Newey-West / HAC tidy table for lm
tidy_hac <- function(mod, lag = 3, prewhite = FALSE, adjust = TRUE) {
  vc <- NeweyWest(mod, lag = lag, prewhite = prewhite, adjust = adjust)
  ct <- coeftest(mod, vcov. = vc)
  tidy_ct(ct)
}

# Compare classical vs HC1 vs HAC for lm models
compare_lm_ses <- function(mod, hac_lag = 3) {
  cl <- tidy(mod) |>
    select(term, estimate, std.error, p.value) |>
    rename(se_classic = std.error, p_classic = p.value)

  hc <- {
    ct <- coeftest(mod, vcov. = vcovHC(mod, type = "HC1"))
    tidy_ct(ct)
  } |>
    select(term, std.error, p.value) |>
    rename(se_hc1 = std.error, p_hc1 = p.value)

  hac <- tidy_hac(mod, lag = hac_lag) |>
    select(term, std.error, p.value) |>
    rename(se_hac = std.error, p_hac = p.value)

  cl |>
    left_join(hc, by = "term") |>
    left_join(hac, by = "term")
}

# Residual ACF table for quick autocorrelation diagnostics
acf_tbl <- function(mod, lag.max = 12) {
  ac <- acf(resid(mod), plot = FALSE, lag.max = lag.max)
  tibble(
    lag = as.numeric(ac$lag)[-1],
    acf = as.numeric(ac$acf)[-1]
  )
}

plot_resid_acf <- function(mod, title = "", lag.max = 12) {
  acf_df <- acf_tbl(mod, lag.max = lag.max)
  ggplot(acf_df, aes(lag, acf)) +
    geom_col() +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = paste0(title, " — Residual ACF"), x = "Lag", y = "ACF") +
    theme_minimal()
}

# Residual/fitted and QQ plots for lm objects
plot_lm_basic <- function(mod, title = "") {
  d <- tibble(
    fitted = fitted(mod),
    resid  = resid(mod)
  )

  p1 <- ggplot(d, aes(x = fitted, y = resid)) +
    geom_point(alpha = 0.35) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    labs(title = paste0(title, " — Residuals vs Fitted"), x = "Fitted", y = "Residual") +
    theme_minimal()

  p2 <- ggplot(d, aes(sample = resid)) +
    stat_qq(alpha = 0.35) +
    stat_qq_line() +
    labs(title = paste0(title, " — QQ plot"), x = "Theoretical", y = "Residual quantiles") +
    theme_minimal()

  p1 / p2
}

# Compare multiple VCOV specs for a fixest model
compare_fixest_vcovs <- function(mod, specs = list("iid" = "iid")) {
  imap_dfr(specs, function(vs, nm) {
    tidy(mod, vcov = vs) |>
      select(term, estimate, std.error, statistic, p.value) |>
      mutate(vcov = nm, .before = 1)
  })
}
```


```{r Diagnostics Helpers}
wrmse_weighted <- function(truth, pred, w) sqrt(weighted.mean((truth - pred)^2, w = w, na.rm = TRUE))
wmae_weighted <- function(truth, pred, w) weighted.mean(abs(truth - pred), w = w, na.rm = TRUE)

# Residual diagnostics for lm with optional weights
lm_diag_tbl <- function(mod, data = NULL) {
  if (is.null(data)) data <- mod$model
  tibble(
    fitted = fitted(mod),
    resid  = resid(mod),
    w = if (!is.null(mod$weights)) mod$weights else rep(1, nrow(data))
  )
}

# Influence summary for lm
lm_influence_tbl <- function(mod) {
  infl <- influence.measures(mod)
  is_inf <- apply(infl$is.inf, 1, any)
  tibble(
    row = seq_along(is_inf),
    influential = is_inf
  )
}
```


#### Metagame Dynamics

```{r Metagame Prep}
# Clean month and time index
mom_tc2 <- mom_tc |>
  filter(!is.na(opening), !is.na(tc_bucket), !is.na(month)) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date))
  )

mm_tc2 <- mm_tc |>
  filter(!is.na(tc_bucket), !is.na(month)) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date))
  )

# Ordered months and strict forward split
months_ordered <- mom_tc2 |>
  distinct(month, month_date, t) |>
  arrange(month_date)

n_months <- nrow(months_ordered)
cut_idx <- floor(0.80 * n_months)

train_months <- months_ordered |> slice(1:cut_idx) |> pull(month)
test_months <- months_ordered |> slice((cut_idx+1):n_months) |> pull(month)

stopifnot(length(intersect(train_months, test_months)) == 0)

mom_train <- mom_tc2 |> filter(month %in% train_months)
mom_test <- mom_tc2 |> filter(month %in% test_months)

mm_train <- mm_tc2 |> filter(month %in% train_months)
mm_test <- mm_tc2 |> filter(month %in% test_months)

# Sanity
range(mom_train$month_date); range(mom_test$month_date)

# Weighted RMSE helper (used throughout)
eval_wrmse <- function(actual, pred, w) {
  sqrt(sum(w * (actual - pred)^2, na.rm = TRUE) / sum(w, na.rm = TRUE))
}
```
```{r MG Helpers}
show_tc <- "blitz"
plot_tc <- "blitz"
t_name <- "t"

# Compact extractor for key lm terms under HAC
extract_hac_terms <- function(mod, model_name, terms_keep, lag = 3) {
  tidy_hac(mod, lag = lag) |>
    filter(term %in% terms_keep) |>
    mutate(model = model_name, .before = 1)
}

# Simple sign/stability label
sign_label <- function(x, tol = 0) {
  case_when(
    is.na(x) ~ NA_character_,
    x > tol ~ "positive",
    x < -tol ~ "negative",
    TRUE ~ "near_zero"
  )
}
```


```{r Metagame Ecosystem Over Time}
# Aggregate opening shares into ecosystem metrics per month x tc_bucket
eco <- mom_tc2 |>
  group_by(month, month_date, t, tc_bucket) |>
  summarise(
    total_games = sum(games, na.rm = TRUE),
    # concentration / diversity metrics derived from the opening share distribution
    hhi = sum(share^2, na.rm = TRUE),                                        # Herfindahl-Hirschman
    entropy = -sum(if_else(share > 0, share * log(share), 0), na.rm = TRUE), # Shannon entropy
    n_eff = 1 / sum(share^2, na.rm = TRUE),                                  # effective num of openings
    .groups = "drop"
  )

eco_train <- eco |> filter(month %in% train_months)
eco_test  <- eco |> filter(month %in% test_months)

# Weighted linear trends
eco_hhi_wlm <- lm(hhi ~ tc_bucket * t, data = eco_train, weights = total_games)
eco_neff_wlm <- lm(n_eff ~ tc_bucket * t, data = eco_train, weights = total_games)

# Out-of-sample sanity checks (month-forward)
eco_hhi_wrmse  <- eval_wrmse(eco_test$hhi,  predict(eco_hhi_wlm,  eco_test), eco_test$total_games)
eco_neff_wrmse <- eval_wrmse(eco_test$n_eff,predict(eco_neff_wlm, eco_test), eco_test$total_games)

tibble(metric = c("HHI (concentration)", "N_eff (effective openings)"),
       wrmse  = c(eco_hhi_wrmse, eco_neff_wrmse))

ggplot(eco, aes(x = month_date, y = hhi, color = tc_bucket, group = tc_bucket)) +
  geom_line() +
  labs(
    x = "Month",
    y = "HHI concentration (higher = more concentrated opening pool)",
    title = "Metagame ecosystem concentration over time"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")
```

```{r Outcome Environment Over Time}
draw_wlm <- lm(draw_rate ~ tc_bucket * t, data = mm_train, weights = games)
upset_wlm <- lm(upset_rate ~ tc_bucket * t, data = mm_train, weights = games)
moves_wlm <- lm(avg_moves ~ tc_bucket * t, data = mm_train, weights = games)

draw_wrmse <- eval_wrmse(mm_test$draw_rate,  predict(draw_wlm,  mm_test), mm_test$games)
upset_wrmse <- eval_wrmse(mm_test$upset_rate, predict(upset_wlm, mm_test), mm_test$games)
moves_wrmse <- eval_wrmse(mm_test$avg_moves,  predict(moves_wlm, mm_test), mm_test$games)

tibble(outcome = c("draw_rate", "upset_rate", "avg_moves"),
       wrmse = c(draw_wrmse, upset_wrmse, moves_wrmse))

mm_long <- mm_tc2 |>
  select(month_date, tc_bucket, draw_rate, upset_rate) |>
  pivot_longer(cols = c(draw_rate, upset_rate), names_to = "metric", values_to = "value")

ggplot(mm_long, aes(x = month_date, y = value, color = tc_bucket, group = tc_bucket)) +
  geom_line() +
  facet_wrap(~metric, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = "Month",
    y = NULL,
    title = "Metagame outcome environment over time"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")
```

```{r Macro HAC Diagnostics}
macro_mods <- list(
  eco_hhi  = eco_hhi_wlm,
  eco_neff = eco_neff_wlm,
  draw = draw_wlm,
  upset = upset_wlm,
  moves = moves_wlm
)

imap(macro_mods, ~ plot_resid_acf(.x, .y, lag.max = 12))
```
```{r Macro Serial Correlation Tests}
bg_tbl <- imap_dfr(macro_mods, function(mod, nm) {
  bg <- bgtest(mod, order = 3)
  tibble(
    model = nm,
    bg_stat = unname(bg$statistic),
    bg_p = bg$p.value
  )
})
bg_tbl
```

```{r Macro HAC Tables}
macro_hac_compare <- imap(macro_mods, function(mod, nm) {
  compare_lm_ses(mod, hac_lag = 3) |>
    mutate(model = nm, .before = 1)
})

macro_hac_compare$draw
macro_hac_compare$eco_hhi

bind_rows(macro_hac_compare) |>
  arrange(model, term)
```

```{r Macro HAC Lag Sensitivity}
hac_lag_sens <- imap_dfr(macro_mods, function(mod, nm) {
  map_dfr(c(1, 3, 6), function(L) {
    tidy_hac(mod, lag = L) |>
      mutate(model = nm, hac_lag = L, .before = 1)
  })
})
hac_lag_sens
```



```{r Metagame Opening-Share Descriptive Models}
top_k <- 12
min_games_cell <- 200

top_openings_by_tc <- mom_train |>
  group_by(tc_bucket, opening) |>
  summarise(total_games = sum(games, na.rm = TRUE), .groups = "drop") |>
  group_by(tc_bucket) |>
  slice_max(order_by = total_games, n = top_k, with_ties = FALSE) |>
  ungroup()

mom_train_top <- mom_train |>
  inner_join(top_openings_by_tc |> select(tc_bucket, opening), by = c("tc_bucket","opening")) |>
  filter(games >= min_games_cell)

mom_test_top <- mom_test |>
  inner_join(top_openings_by_tc |> select(tc_bucket, opening), by = c("tc_bucket","opening")) |>
  filter(games >= min_games_cell)

# FEOLS gives interpretable opening-by-time trend summaries with clustered VCOV options.
fit_share_feols_tc <- function(tc) {
  dat <- mom_train_top |>
    filter(tc_bucket == tc) |>
    mutate(
      opening = fct_drop(factor(opening)),
      month = factor(month)
    )

  feols(
    share ~ opening * t,
    data = dat,
    weights = ~games
  )
}

fit_gam_tc <- function(tc) {
  dat <- mom_train_top |> filter(tc_bucket == tc) |>
    mutate(opening = fct_drop(factor(opening)))
  gam(
    share ~ opening + s(t, by = opening, k = 8),
    data = dat,
    weights = games,
    method = "REML"
  )
}

tc_levels <- sort(unique(mom_train_top$tc_bucket))
share_feols_models <- set_names(tc_levels) |> map(fit_share_feols_tc)
gam_models <- set_names(tc_levels) |> map(fit_gam_tc)

# Evaluate both on held-out months
share_feols_eval <- imap_dfr(share_feols_models, function(mod, tc) {
  train_levels <- mom_train_top |>
    filter(tc_bucket == tc) |>
    mutate(opening = fct_drop(factor(opening))) |>
    pull(opening) |>
    levels()

  te <- mom_test_top |>
    filter(tc_bucket == tc) |>
    mutate(opening = factor(opening, levels = train_levels)) |>
    filter(!is.na(opening))

  pred <- predict(mod, newdata = te)
  tibble(tc_bucket = tc, model = "FEOLS_share_descriptive", wrmse = eval_wrmse(te$share, pred, te$games))
})

gam_eval <- imap_dfr(gam_models, function(mod, tc) {
  te <- mom_test_top |> filter(tc_bucket == tc) |>
    mutate(opening = factor(opening, levels = levels(mod$model$opening))) |>
    filter(!is.na(opening))
  pred <- predict(mod, newdata = te)
  tibble(tc_bucket = tc, model = "GAM_descriptive", wrmse = eval_wrmse(te$share, pred, te$games))
})

bind_rows(share_feols_eval, gam_eval) |> arrange(tc_bucket, model)

dat_all <- mom_tc2 |>
  inner_join(top_openings_by_tc |> filter(tc_bucket == plot_tc) |> select(tc_bucket, opening),
             by = c("tc_bucket","opening")) |>
  filter(tc_bucket == plot_tc, games >= min_games_cell) |>
  mutate(
    opening = factor(opening, levels = levels(gam_models[[plot_tc]]$model$opening)),
    split = if_else(month %in% train_months, "train", "test")
  ) |>
  filter(!is.na(opening))

dat_all$pred_gam <- predict(gam_models[[plot_tc]], newdata = dat_all)

ggplot(dat_all, aes(x = month_date, y = share, color = opening, group = opening)) +
  geom_point(aes(alpha = split), size = 1) +
  geom_line(aes(y = pred_gam), linewidth = 0.8) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    x = "Month",
    y = "Opening share",
    title = paste0("Supporting: top openings share over time with GAM fit (", plot_tc, ")"),
    alpha = "Split"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")
```
```{r Metagame Multinomial}

# Aggregate counts at opening-month-tc
counts_long_all <- mom_tc2 |>
  filter(!is.na(opening), !is.na(tc_bucket), !is.na(month)) |>
  group_by(tc_bucket, month, month_date, t, opening) |>
  summarise(games = sum(games), .groups = "drop")

# Tag top-k vs Other
top_tbl <- top_openings_by_tc |>
  mutate(is_top = TRUE) |>
  select(tc_bucket, opening, is_top)

counts_long_tagged <- counts_long_all |>
  left_join(top_tbl, by = c("tc_bucket","opening")) |>
  mutate(opening2 = if_else(is_top %in% TRUE, opening, "Other")) |>
  group_by(tc_bucket, month, month_date, t, opening2) |>
  summarise(games = sum(games), .groups = "drop")

# Wide format: one row per (tc_bucket, month), one column per opening2
counts_wide <- counts_long_tagged |>
  pivot_wider(names_from = opening2, values_from = games, values_fill = 0)

mn_train <- counts_wide |> filter(month %in% train_months)
mn_test <- counts_wide |> filter(month %in% test_months)

id_cols <- c("tc_bucket","month","month_date","t")
open_cols <- setdiff(names(mn_train), id_cols)

# Safe names mapping for rename(new = old)
name_map <- tibble(
  opening = open_cols,
  safe = make.names(open_cols, unique = TRUE)
)

rename_vec <- setNames(name_map$opening, name_map$safe)

mn_train2 <- mn_train |> rename(!!!rename_vec)
mn_test2 <- mn_test  |> rename(!!!rename_vec)

safe_open_cols <- name_map$safe

# Sanity
stopifnot(all(safe_open_cols %in% names(mn_train2)))
stopifnot(all(safe_open_cols %in% names(mn_test2)))

# Fit multinomial per tc_bucket
fit_multinom_tc <- function(tc) {
  tr <- mn_train2 |> filter(tc_bucket == tc)

  # response matrix columns (safe names)
  Y <- tr |> select(any_of(safe_open_cols))
  X <- tr |> select(t)

  dat <- bind_cols(X, Y)

  y_part <- paste0("cbind(", paste(names(Y), collapse = ","), ")")
  fml <- as.formula(paste0(y_part, " ~ t"))

  vglm(fml, family = multinomial(), data = dat)
}

mn_models <- mn_train2 |>
  distinct(tc_bucket) |>
  pull(tc_bucket) |>
  set_names() |>
  map(fit_multinom_tc)

# Evaluate with average per-game log loss
mn_logloss <- function(counts, probs, eps = 1e-12) {
  probs <- pmax(probs, eps)
  -sum(counts * log(probs)) / sum(counts)
}

mn_eval <- imap_dfr(mn_models, function(mod, tc) {
  te <- mn_test2 |> filter(tc_bucket == tc)

  Y <- te |> select(any_of(safe_open_cols)) |> as.matrix()
  pred <- predict(mod, newdata = te |> select(t), type = "response")

  # Ensure matrices align by column name (VGAM may drop a baseline column)
  pred <- as.matrix(pred)
  if (!is.null(colnames(pred))) {
    # :eep only overlapping columns
    common <- intersect(colnames(pred), colnames(Y))
    Y_use <- Y[, common, drop = FALSE]
    pred_use <- pred[, common, drop = FALSE]
  } else {
    # Fallback; assume same ordering
    Y_use <- Y
    pred_use <- pred
  }

  probs <- pred_use / rowSums(pred_use)

  row_totals <- rowSums(Y_use)
  row_loss <- map_dbl(seq_len(nrow(Y_use)), \(i) mn_logloss(Y_use[i,], probs[i,]))

  tibble(
    tc_bucket = tc,
    n_test_months = nrow(te),
    avg_logloss_per_game = weighted.mean(row_loss, w = row_totals)
  )
})

mn_eval
```



```{r Metagame Multinomial Predicted Shares}
# Predicted composition at early vs late time for the primary multinomial model

tc_show <- show_tc
mod <- mn_models[[tc_show]]

te_tc <- mn_test2 |>
  filter(tc_bucket == tc_show)

t_min <- min(te_tc$t, na.rm = TRUE)
t_max <- max(te_tc$t, na.rm = TRUE)

newdat <- tibble(t = c(t_min, t_max))

pred <- predict(mod, newdata = newdat, type = "response")
pred <- as.data.frame(pred) |>
  rownames_to_column("row_id") |>
  mutate(period = c("early_test", "late_test")) |>
  select(-row_id) |>
  pivot_longer(-period, names_to = "opening", values_to = "pred_share")

pred |>
  group_by(period) |>
  arrange(period, desc(pred_share)) |>
  slice_head(n = 10) |>
  ungroup()
```

```{r Metagame Multinomial Primary Results}
mn_show <- mn_models[[show_tc]]

cat("Primary inferential model for opening composition\n")
cat("Time control:", show_tc, "\n")

# Coefficient table
coef_tab <- coef(summary(mn_show))

mn_coef <- as.data.frame(coef_tab) |>
  rownames_to_column("term") |>
  rename(
    estimate  = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value   = `Pr(>|z|)`
  ) |>
  mutate(tc_bucket = show_tc)

mn_coef |>
  filter(str_starts(term, "t:")) |>
  arrange(p.value) |>
  head(20)
```

```{r Metagame Multinomial Share Change}
tc_show <- show_tc
mod <- mn_models[[tc_show]]

te_tc <- mn_test2 |>
  filter(tc_bucket == tc_show)

t_min <- min(te_tc$t, na.rm = TRUE)
t_max <- max(te_tc$t, na.rm = TRUE)

newdat <- tibble(t = c(t_min, t_max))
pred <- predict(mod, newdata = newdat, type = "response")
pred <- as.data.frame(pred)

common_cols <- colnames(pred)

share_change <- tibble(
  opening = common_cols,
  pred_early = as.numeric(pred[1, common_cols]),
  pred_late  = as.numeric(pred[2, common_cols])
) |>
  mutate(change = pred_late - pred_early) |>
  arrange(desc(abs(change)))

share_change |> head(15)
```


```{r Metagame Writeup Helpers}

# SE helper for lm objects
tidy_robust <- function(mod, type = "HC1") {
  ct <- coeftest(mod, vcov. = vcovHC(mod, type = type))
  tidy(ct) |>
    rename(estimate = estimate, std.error = std.error, statistic = statistic, p.value = p.value)
}

# Formula-to-string helper
fmt_formula <- function(mod) {
  paste(deparse(formula(mod)), collapse = " ")
}

# Write equation
lm_equation_string <- function(mod, opening_var = "opening", time_var = "t", digits = 4) {
  b <- coef(mod)
  terms <- names(b)

  # Separate intercept, main effects, time, and interactions
  b0 <- b["(Intercept)"]
  b_t <- if (time_var %in% terms) b[time_var] else 0

  # Opening main effects: openingX
  op_main <- b[grepl(paste0("^", opening_var), terms) & !grepl(":", terms)]
  # Interactions: openingX:time
  op_int  <- b[grepl(paste0("^", opening_var), terms) & grepl(paste0(":", time_var, "$"), terms)]

  # Build
  lines <- c()
  lines <- c(lines, sprintf("\\hat{y} = %.4f %s %.4f\\,%s", 
                            round(b0, digits),
                            ifelse(b_t >= 0, "+", "-"),
                            abs(round(b_t, digits)),
                            time_var))

  if (length(op_main) > 0) {
    for (nm in names(op_main)) {
      # nm like openingPhilidor Defense
      op <- sub(paste0("^", opening_var), "", nm)
      val <- op_main[nm]
      lines <- c(lines, sprintf("\\; %s\\; %.4f\\,\\mathbb{1}[%s=%s]", 
                                ifelse(val >= 0, "+", "-"),
                                abs(round(val, digits)),
                                opening_var,
                                gsub("'", "\\\\'", op)))
    }
  }

  if (length(op_int) > 0) {
    for (nm in names(op_int)) {
      # nm like openingPhilidor Defense:t
      op <- sub(paste0("^", opening_var), "", sub(paste0(":", time_var, "$"), "", nm))
      val <- op_int[nm]
      lines <- c(lines, sprintf("\\; %s\\; %.4f\\,%s\\cdot\\mathbb{1}[%s=%s]", 
                                ifelse(val >= 0, "+", "-"),
                                abs(round(val, digits)),
                                time_var,
                                opening_var,
                                gsub("'", "\\\\'", op)))
    }
  }

  paste0("$", paste(lines, collapse = " "), "$")
}
```

```{r Metagame Opening-Share Descriptive Writeup}
wlm_show <- share_feols_models[[show_tc]]

cat("FEOLS descriptive opening-share model — example\n")
cat("Time control:", show_tc, "\n")
cat("Formula:", fmt_formula(wlm_show), "\n\n")

# Standard summary table
wlm_sum <- tidy(wlm_show) |>
  arrange(desc(abs(estimate)))

# Phase 1 primary inference
wlm_cluster_month <- tidy(wlm_show, vcov = ~month) |>
  arrange(desc(abs(estimate)))

wlm_cluster_twoway <- tidy(wlm_show, vcov = ~month + opening) |>
  arrange(desc(abs(estimate)))

wlm_fit_stats <- tibble(
  stat = c("n", "r2", "wr2"),
  value = as.numeric(fitstat(wlm_show, c("n", "r2", "wr2")))
)

wlm_fit_stats
wlm_cluster_month |> head(25)
wlm_cluster_twoway |> head(25)

summary(wlm_show)
```

```{r Metagame Opening-Share GAM Sensitivity}
gam_show <- gam_models[[show_tc]]

cat("GAM sensitivity model for opening-share nonlinearity — example\n")
cat("Time control:", show_tc, "\n")
cat("Formula:", fmt_formula(gam_show), "\n\n")

# Parametric part
gam_param <- tidy(gam_show, parametric = TRUE)

# Smooth part
gam_smooth <- tidy(gam_show, parametric = FALSE)

# Fit stats
gam_fit_stats <- glance(gam_show)

gam_fit_stats
gam_param |> arrange(p.value) |> head(20)
gam_smooth |> arrange(p.value) |> head(20)

summary(gam_show)
```

```{r Metagame Macro Primary Writeup}
macro_show <- draw_wlm # or upset_wlm, moves_wlm

cat("Primary macro inferential model (HAC inference)\n")
cat("Outcome:", all.vars(formula(macro_show))[1], "\n")
cat("Formula:", fmt_formula(macro_show), "\n\n")

macro_hac <- tidy_hac(macro_show, lag = 3)
macro_stats <- glance(macro_show)

macro_stats
macro_hac
```

```{r Metagame Tables All Models}
# Descriptive FEOLS share coefficients for all tc_buckets
share_coef_all_month <- imap_dfr(share_feols_models, function(mod, tc) {
  tidy(mod, vcov = ~month) |>
    mutate(tc_bucket = tc)
})

share_coef_all_twoway <- imap_dfr(share_feols_models, function(mod, tc) {
  tidy(mod, vcov = ~month + opening) |>
    mutate(tc_bucket = tc)
})

# GAM parametric + smooth for all tc_buckets
gam_param_all <- imap_dfr(gam_models, function(mod, tc) {
  tidy(mod, parametric = TRUE) |> mutate(tc_bucket = tc)
})
gam_smooth_all <- imap_dfr(gam_models, function(mod, tc) {
  tidy(mod, parametric = FALSE) |> mutate(tc_bucket = tc)
})

# Primary macro inferential models with HAC
macro_coef_all <- bind_rows(
  tidy_hac(draw_wlm, lag = 3)  |> mutate(outcome = "draw_rate"),
  tidy_hac(upset_wlm, lag = 3) |> mutate(outcome = "upset_rate"),
  tidy_hac(moves_wlm, lag = 3) |> mutate(outcome = "avg_moves")
)

share_coef_all_month |> arrange(tc_bucket, p.value) |> head(15)
share_coef_all_twoway |> arrange(tc_bucket, p.value) |> head(15)
gam_param_all |> arrange(tc_bucket, p.value) |> head(15)
gam_smooth_all |> arrange(tc_bucket, p.value) |> head(15)
macro_coef_all |> arrange(outcome, p.value)

mn_slopes_all <- imap_dfr(mn_models, function(mod, tc) {
  coef_tab <- as.data.frame(coef(summary(mod))) |>
    rownames_to_column("term") |>
    rename(
      estimate  = Estimate,
      std.error = `Std. Error`,
      statistic = `z value`,
      p.value   = `Pr(>|z|)`
    ) |>
    mutate(tc_bucket = tc)

  coef_tab |>
    filter(str_starts(term, "t:"))
})

mn_slopes_all |> arrange(tc_bucket, p.value) |> head(20)
```

```{r Metagame Multinomial Tidy}
mn_show <- mn_models[[show_tc]]
coef_tab <- coef(summary(mn_show))

mn_coef <- as.data.frame(coef_tab) |>
  rownames_to_column("term") |>
  rename(
    estimate  = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value   = `Pr(>|z|)`
  ) |>
  mutate(
    tc_bucket = show_tc
  )

mn_coef |>
  arrange(p.value) |>
  select(tc_bucket, term, estimate, std.error, statistic, p.value) |>
  head(30)
```

```{r Metagame Multinomial Matrices}
mn_show <- mn_models[[show_tc]]

B <- coefvlm(mn_show, matrix = TRUE)  # Coefficients
V <- vcov(mn_show)

# Build SEs from vcov
coef_tab <- coef(summary(mn_show))
se_vec <- coef_tab[, "Std. Error"]
names(se_vec) <- rownames(coef_tab)

# Convert the stacked SEs to a (predictor x category) matrix
make_se_matrix <- function(B, se_vec) {
  out <- B * NA_real_
  for (r in rownames(B)) {
    for (c in colnames(B)) {
      nm <- paste0(r, ":", c)
      if (nm %in% names(se_vec)) out[r, c] <- se_vec[nm]
    }
  }
  out
}

SE <- make_se_matrix(B, se_vec)

B
SE
```
```{r Metagame Multinomial Slopes Only}
mn_show <- mn_models[[show_tc]]

coef_tab <- coef(summary(mn_show))

mn_coef <- as.data.frame(coef_tab) |>
  rownames_to_column("term") |>
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  )

mn_slopes <- mn_coef |>
  filter(str_starts(term, "t:")) |>
  arrange(p.value)

mn_slopes |> head(20)
```

##### Diagnostics

```{r Metagame Macro WLM Diagnostics}
macro_mods <- list(
  eco_hhi  = eco_hhi_wlm,
  eco_neff = eco_neff_wlm,
  draw     = draw_wlm,
  upset    = upset_wlm,
  moves    = moves_wlm
)

# Residual vs fitted, QQ
plot_lm_basic(eco_hhi_wlm,  "eco_hhi_wlm")
plot_lm_basic(draw_wlm,     "draw_wlm")

# Residual autocorrelation checks
plot_resid_acf(eco_hhi_wlm, "eco_hhi_wlm", lag.max = 12)
plot_resid_acf(draw_wlm,    "draw_wlm",    lag.max = 12)

# Heteroskedasticity and serial-correlation diagnostics
bp_tbl <- imap_dfr(macro_mods, function(mod, nm) {
  bp <- bptest(mod)
  tibble(
    model = nm,
    bp_stat = unname(bp$statistic),
    bp_p = bp$p.value
  )
})
bp_tbl

bg_tbl <- imap_dfr(macro_mods, function(mod, nm) {
  bg <- bgtest(mod, order = 3)
  tibble(
    model = nm,
    bg_stat = unname(bg$statistic),
    bg_p = bg$p.value
  )
})
bg_tbl

# Compare classical vs HC1 vs HAC SEs
macro_hac_compare <- imap(macro_mods, function(mod, nm) {
  compare_lm_ses(mod, hac_lag = 3) |>
    mutate(model = nm, .before = 1)
})

bind_rows(macro_hac_compare) |>
  arrange(model, term)

# Influence
infl_tbl <- imap_dfr(macro_mods, function(mod, nm) {
  inf <- lm_influence_tbl(mod)
  tibble(
    model = nm,
    n = nrow(inf),
    n_influential = sum(inf$influential)
  )
})
infl_tbl
```

```{r Metagame Share WLM Diagnostics}
# Diagnostics per tc_bucket for FEOLS share models
share_diag <- imap_dfr(share_feols_models, function(mod, tc) {
  tibble(
    tc_bucket = tc,
    n = nobs(mod),
    r2 = as.numeric(fitstat(mod, "r2"))
  )
})
share_diag

# Visual diagnostics for one bucket
tc_show <- show_tc

share_diag_df <- tibble(
  fitted = fitted(share_feols_models[[tc_show]]),
  resid  = resid(share_feols_models[[tc_show]])
)

ggplot(share_diag_df, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = paste0("FEOLS share (", tc_show, ") — Residuals vs Fitted"),
       x = "Fitted", y = "Residual") +
  theme_minimal()

ggplot(share_diag_df, aes(sample = resid)) +
  stat_qq(alpha = 0.35) +
  stat_qq_line() +
  labs(title = paste0("FEOLS share (", tc_show, ") — QQ plot")) +
  theme_minimal()

# Clustered SEs
compare_fixest_vcovs(
  share_feols_models[[tc_show]],
  specs = list(
    iid = "iid",
    HC1 = "hetero",
    month_cluster = ~month,
    opening_cluster = ~opening,
    two_way = ~month + opening
  )
) |>
  filter(str_detect(term, ":t$") | term == "t") |>
  arrange(term, vcov)
```

```{r Metagame Share GAM Diagnostics}
# GAM diagnostics
tc_show <- show_tc
g <- gam_models[[tc_show]]

# mgcv checks
gam.check(g)

# Concurvity
concurvity(g, full = TRUE)

# Residuals vs fitted, QQ
gdf <- tibble(fitted = fitted(g), resid = residuals(g))
ggplot(gdf, aes(fitted, resid)) +
  geom_point(alpha = 0.35) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = paste0("GAM share (", tc_show, ") — Residuals vs Fitted"),
       x = "Fitted", y = "Residual") +
  theme_minimal()

ggplot(gdf, aes(sample = resid)) +
  stat_qq(alpha = 0.35) +
  stat_qq_line() +
  labs(title = paste0("GAM share (", tc_show, ") — QQ plot")) +
  theme_minimal()
```
```{r Metagame Multinomial Diagnostics}
# Category support (train)
mn_support <- mn_train2 |>
  select(tc_bucket, all_of(safe_open_cols)) |>
  group_by(tc_bucket) |>
  summarise(across(all_of(safe_open_cols), sum), .groups = "drop") |>
  pivot_longer(-tc_bucket, names_to = "cat", values_to = "total") |>
  group_by(tc_bucket) |>
  arrange(total) |>
  slice_head(n = 5) |>
  ungroup()
mn_support

# Calibration-by-month
tc_show <- show_tc
mod <- mn_models[[tc_show]]
te  <- mn_test2 |> filter(tc_bucket == tc_show)

Y <- te |> select(all_of(safe_open_cols)) |> as.matrix()
obs_share <- Y / rowSums(Y)

pred_counts <- predict(mod, newdata = te |> select(t), type = "response")
pred_counts <- as.matrix(pred_counts)

# Align overlapping categories
common <- intersect(colnames(pred_counts), colnames(Y))
obs_use <- obs_share[, common, drop = FALSE]
pred_use <- pred_counts[, common, drop = FALSE]
pred_share <- pred_use / rowSums(pred_use)

# Mean absolute share error per month
month_err <- tibble(
  month = te$month,
  tot = rowSums(Y[, common, drop = FALSE]),
  mae = rowMeans(abs(obs_use - pred_share))
) |>
  group_by(month) |>
  summarise(
    w_mae = weighted.mean(mae, w = tot),
    .groups = "drop"
  )

month_err

ggplot(month_err, aes(x = month, y = w_mae)) +
  geom_line(aes(group = 1)) +
  geom_point() +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = paste0("Metagame multinomial — weighted MAE in share by month (", tc_show, ")"),
    x = "Month (test)",
    y = "Weighted mean |obs - pred| share"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
```
`opening_cluster` result sare not usable for inference; inferential comparison should rely on `month_cluster` and `two_way`.

##### Summary

```{r Metagame Macro Time Shape Check}
fit_macro_gam_time <- function(data, y, w_col) {
  data <- data |>
    mutate(tc_bucket = factor(tc_bucket))

  fml <- as.formula(
    paste0(y, " ~ tc_bucket + s(t, k = 6) + s(t, by = tc_bucket, k = 6)")
  )

  gam(
    fml,
    data = data,
    weights = data[[w_col]],
    method = "REML"
  )
}

macro_primary <- list(
  eco_hhi  = list(train = eco_train, y = "hhi",        w = "total_games", lm = eco_hhi_wlm),
  eco_neff = list(train = eco_train, y = "n_eff",      w = "total_games", lm = eco_neff_wlm),
  draw     = list(train = mm_train,  y = "draw_rate",  w = "games",       lm = draw_wlm),
  upset    = list(train = mm_train,  y = "upset_rate", w = "games",       lm = upset_wlm),
  moves    = list(train = mm_train,  y = "avg_moves",  w = "games",       lm = moves_wlm)
)

macro_time_shape_tbl <- imap_dfr(macro_primary, function(obj, nm) {
  g <- fit_macro_gam_time(obj$train, obj$y, obj$w)

  # cCompare fitted values at first and last observed month by tc_bucket
  t_min <- min(obj$train$t, na.rm = TRUE)
  t_max <- max(obj$train$t, na.rm = TRUE)

  newdat <- crossing(
    t = c(t_min, t_max),
    tc_bucket = sort(unique(obj$train$tc_bucket))
  ) |>
    mutate(tc_bucket = factor(tc_bucket, levels = levels(factor(obj$train$tc_bucket))))

  pred_lm <- predict(obj$lm, newdata = newdat)
  pred_gam <- predict(g, newdata = newdat)

  out <- newdat |>
    mutate(
      model = "linear",
      pred = pred_lm
    ) |>
    bind_rows(
      newdat |>
        mutate(
          model = "flexible",
          pred = pred_gam
        )
    ) |>
    mutate(period = if_else(t == t_min, "early", "late")) |>
    select(-t) |>
    pivot_wider(names_from = period, values_from = pred) |>
    mutate(change = late - early, outcome = nm, .before = 1)

  out
})

macro_time_shape_tbl |>
  arrange(outcome, tc_bucket, model)
```

```{r Metagame Macro Claim Summary}
macro_claims <- bind_rows(
  extract_hac_terms(eco_hhi_wlm,  "eco_hhi",  c("t", "tc_bucketclassical:t", "tc_bucketrapid:t")),
  extract_hac_terms(eco_neff_wlm, "eco_neff", c("t", "tc_bucketclassical:t", "tc_bucketrapid:t")),
  extract_hac_terms(draw_wlm,     "draw",     c("t", "tc_bucketclassical:t", "tc_bucketrapid:t")),
  extract_hac_terms(upset_wlm,    "upset",    c("t", "tc_bucketclassical:t", "tc_bucketrapid:t")),
  extract_hac_terms(moves_wlm,    "moves",    c("t", "tc_bucketclassical:t", "tc_bucketrapid:t"))
) |>
  mutate(
    sign = sign_label(estimate),
    survives_5pct = p.value < 0.05
  ) |>
  arrange(model, term)

macro_claims
```

```{r Metagame Multinomial TopK Robustness}
phase3_top_k <- c(8, 12, 16)
phase3_min_games <- c(100, 200, 500)

fit_mn_spec <- function(top_k_val, min_games_val, tc_show = "blitz") {
  # Top-k openings from train only
  top_tbl_spec <- mom_train |>
    group_by(tc_bucket, opening) |>
    summarise(total_games = sum(games, na.rm = TRUE), .groups = "drop") |>
    group_by(tc_bucket) |>
    slice_max(order_by = total_games, n = top_k_val, with_ties = FALSE) |>
    ungroup()

  # Apply top-k tagging
  counts_long_all_spec <- mom_tc2 |>
    filter(!is.na(opening), !is.na(tc_bucket), !is.na(month)) |>
    group_by(tc_bucket, month, month_date, t, opening) |>
    summarise(games = sum(games), .groups = "drop") |>
    left_join(
      top_tbl_spec |>
        mutate(is_top = TRUE) |>
        select(tc_bucket, opening, is_top),
      by = c("tc_bucket", "opening")
    ) |>
    mutate(opening2 = if_else(is_top %in% TRUE, opening, "Other")) |>
    group_by(tc_bucket, month, month_date, t, opening2) |>
    summarise(games = sum(games), .groups = "drop")

  counts_wide_spec <- counts_long_all_spec |>
    pivot_wider(names_from = opening2, values_from = games, values_fill = 0)

  mn_train_spec <- counts_wide_spec |>
    filter(month %in% train_months, tc_bucket == tc_show)

  mn_test_spec <- counts_wide_spec |>
    filter(month %in% test_months, tc_bucket == tc_show)

  id_cols <- c("tc_bucket", "month", "month_date", "t")
  open_cols <- setdiff(names(mn_train_spec), id_cols)

  if (length(open_cols) < 3 || nrow(mn_train_spec) == 0 || nrow(mn_test_spec) == 0) {
    return(tibble(
      n_categories_train = NA_integer_,
      avg_logloss_per_game = NA_real_,
      top_mover_1 = NA_character_,
      top_mover_1_change = NA_real_,
      top_mover_2 = NA_character_,
      top_mover_2_change = NA_real_
    ))
  }

  # Safe names
  name_map_spec <- tibble(
    opening = open_cols,
    safe = make.names(open_cols, unique = TRUE)
  )
  rename_vec_spec <- setNames(name_map_spec$opening, name_map_spec$safe)

  mn_train2_spec <- mn_train_spec |> rename(!!!rename_vec_spec)
  mn_test2_spec  <- mn_test_spec  |> rename(!!!rename_vec_spec)

  safe_open_cols_spec <- name_map_spec$safe

  # Drop train categories with zero total count
  train_totals <- mn_train2_spec |>
    summarise(across(all_of(safe_open_cols_spec), sum, na.rm = TRUE))

  keep_cols <- names(train_totals)[as.numeric(train_totals[1, ]) > 0]

  if (length(keep_cols) < 2) {
    return(tibble(
      n_categories_train = length(keep_cols),
      avg_logloss_per_game = NA_real_,
      top_mover_1 = NA_character_,
      top_mover_1_change = NA_real_,
      top_mover_2 = NA_character_,
      top_mover_2_change = NA_real_
    ))
  }

  fml <- as.formula(
    paste0("cbind(", paste(keep_cols, collapse = ","), ") ~ t")
  )

  mod <- vglm(fml, family = multinomial(), data = mn_train2_spec)

  pred <- predict(mod, newdata = mn_test2_spec |> select(t), type = "response")
  pred <- as.matrix(pred)

  Y <- mn_test2_spec |> select(any_of(keep_cols)) |> as.matrix()

  common <- intersect(colnames(pred), colnames(Y))
  if (length(common) < 2) {
    return(tibble(
      n_categories_train = length(keep_cols),
      avg_logloss_per_game = NA_real_,
      top_mover_1 = NA_character_,
      top_mover_1_change = NA_real_,
      top_mover_2 = NA_character_,
      top_mover_2_change = NA_real_
    ))
  }

  Y_use <- Y[, common, drop = FALSE]
  pred_use <- pred[, common, drop = FALSE]
  probs <- pred_use / rowSums(pred_use)

  row_loss <- map_dbl(seq_len(nrow(Y_use)), \(i) mn_logloss(Y_use[i, ], probs[i, ]))

  # Early vs late predicted change
  t_min <- min(mn_test2_spec$t, na.rm = TRUE)
  t_max <- max(mn_test2_spec$t, na.rm = TRUE)

  pred_2 <- predict(
    mod,
    newdata = tibble(t = c(t_min, t_max)),
    type = "response"
  ) |>
    as.data.frame()

  share_change <- tibble(
    opening_safe = colnames(pred_2),
    pred_early = as.numeric(pred_2[1, ]),
    pred_late  = as.numeric(pred_2[2, ]),
    change = pred_late - pred_early
  ) |>
    left_join(
      name_map_spec |>
        rename(opening_safe = safe, opening = opening),
      by = "opening_safe"
    ) |>
    mutate(opening = coalesce(opening, opening_safe)) |>
    arrange(desc(abs(change)))

  tibble(
    n_categories_train = length(keep_cols),
    avg_logloss_per_game = weighted.mean(row_loss, w = rowSums(Y_use)),
    top_mover_1 = share_change$opening[1],
    top_mover_1_change = share_change$change[1],
    top_mover_2 = share_change$opening[2],
    top_mover_2_change = share_change$change[2]
  )
}

mn_phase3_thresh <- crossing(
  top_k = phase3_top_k,
  min_games_cell = phase3_min_games
) |>
  mutate(
    out = map2(top_k, min_games_cell, ~ fit_mn_spec(.x, .y, tc_show = "blitz"))
  ) |>
  unnest(out)

mn_phase3_thresh
```

```{r Metagame Composition Agreement Summary}
# Compare multinomial top movers with FEOLS directional slopes for the same tc_bucket
tc_show <- show_tc

mn_change_tbl <- share_change |>
  rename(opening_safe = opening)

safe_to_orig <- name_map |>
  rename(opening_safe = safe, opening = opening)

mn_change_tbl2 <- mn_change_tbl |>
  left_join(safe_to_orig, by = "opening_safe") |>
  mutate(opening = coalesce(opening, opening_safe))

feols_dir_tbl <- tidy(share_feols_models[[tc_show]], vcov = ~month + opening) |>
  filter(str_detect(term, ":t$") | term == "t") |>
  mutate(
    opening = if_else(term == "t", "(baseline)", str_remove(term, "^opening") |> str_remove(":t$")),
    feols_sign = sign_label(estimate)
  ) |>
  select(opening, feols_estimate = estimate, feols_sign, feols_p = p.value)

mn_change_tbl2 |>
  mutate(mn_sign = sign_label(change)) |>
  select(opening, mn_change = change, mn_sign) |>
  left_join(feols_dir_tbl, by = "opening") |>
  arrange(desc(abs(mn_change))) |>
  head(15)
```

```{r Macro WLM Fitted Two-Panel Figure}

show_bands <- TRUE

# Prediction grid over the observed time range
pred_grid <- eco |>
  distinct(month_date, t) |>
  arrange(month_date) |>
  crossing(
    tc_bucket = c("blitz", "rapid", "classical")
  ) |>
  mutate(
    tc_bucket = factor(tc_bucket, levels = c("blitz", "rapid", "classical"))
  )

# Helper to get lm fitted values + pointwise 95% CI
predict_lm_df <- function(mod, newdata, outcome_name) {
  pr <- predict(mod, newdata = newdata, se.fit = TRUE)

  newdata |>
    mutate(
      metric = outcome_name,
      fit = as.numeric(pr$fit),
      se_fit = as.numeric(pr$se.fit),
      lwr = fit - 1.96 * se_fit,
      upr = fit + 1.96 * se_fit
    )
}

# Predictions for both retained macro WLMs
plot_df <- bind_rows(
  predict_lm_df(eco_hhi_wlm,  pred_grid, "HHI"),
  predict_lm_df(eco_neff_wlm, pred_grid, "N_eff")
) |>
  mutate(
    metric = factor(metric, levels = c("HHI", "N_eff")),
    tc_bucket = fct_recode(tc_bucket,
      "Blitz" = "blitz",
      "Rapid" = "rapid",
      "Classical" = "classical"
    ),
    metric_label = factor(
      metric,
      levels = c("HHI", "N_eff"),
      labels = c("HHI", "N_eff")
    )
  )

# Facet labels as plotmath expressions
facet_labs <- labeller(
  metric_label = label_parsed
)

p_macro_fit <- ggplot(
  plot_df,
  aes(x = month_date, y = fit, color = tc_bucket, fill = tc_bucket, group = tc_bucket)
) +
  {
    if (show_bands) {
      geom_ribbon(aes(ymin = lwr, ymax = upr), alpha = 0.20, linewidth = 0, color = NA)
    }
  } +
  geom_line(linewidth = 0.9) +
  facet_wrap(
    ~ metric_label,
    ncol = 1,
    scales = "free_y",
    labeller = facet_labs
  ) +
  labs(
    x = "Month",
    y = NULL,
    color = NULL,
    fill = NULL
  ) +
  scale_y_continuous(
    labels = label_number()
  ) +
  scale_color_few() +
  scale_fill_few() +
  theme_few() +
    theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_text(angle = 60, hjust = 1),
    strip.text = element_text(face = "bold")
  )

# Replace default strip text with parsed math labels
p_macro_fit_final <- p_macro_fit +
  facet_wrap(
    ~ metric,
    ncol = 1,
    scales = "free_y",
    labeller = as_labeller(c(
      HHI   = "Panel~A*':'~Fitted~HHI~Over~Time~by~Time~Control",
      N_eff = "Panel~B*':'~Fitted~N[eff]~Over~Time~by~Time~Control"
    ), label_parsed)
  )
  
p_macro_fit_final

saveRDS(p_macro_fit_final, "../files_for_submission/Paper_files/figure-rds/p_macro_fit.rds")
```

```{r Appendix Macro Observed-vs-Fitted Diagnostics}
macro_diag_df <- bind_rows(
  eco |>
    transmute(
      month,
      month_date,
      tc_bucket,
      observed = hhi,
      fitted = predict(eco_hhi_wlm, newdata = eco),
      outcome = "HHI",
      split = if_else(month %in% train_months, "Train", "Test")
    ),
  eco |>
    transmute(
      month,
      month_date,
      tc_bucket,
      observed = n_eff,
      fitted = predict(eco_neff_wlm, newdata = eco),
      outcome = "N_eff",
      split = if_else(month %in% train_months, "Train", "Test")
    ),
  mm_tc2 |>
    transmute(
      month,
      month_date,
      tc_bucket,
      observed = draw_rate,
      fitted = predict(draw_wlm, newdata = mm_tc2),
      outcome = "Draw Rate",
      split = if_else(month %in% train_months, "Train", "Test")
    ),
  mm_tc2 |>
    transmute(
      month,
      month_date,
      tc_bucket,
      observed = upset_rate,
      fitted = predict(upset_wlm, newdata = mm_tc2),
      outcome = "Upset Rate",
      split = if_else(month %in% train_months, "Train", "Test")
    ),
  mm_tc2 |>
    transmute(
      month,
      month_date,
      tc_bucket,
      observed = avg_moves,
      fitted = predict(moves_wlm, newdata = mm_tc2),
      outcome = "Average Moves",
      split = if_else(month %in% train_months, "Train", "Test")
    )
) |>
  mutate(
    tc_bucket = factor(tc_bucket, levels = c("blitz", "rapid", "classical")),
    split = factor(split, levels = c("Train", "Test")),
    outcome = factor(
      outcome,
      levels = c("HHI", "N_eff", "Draw Rate", "Upset Rate", "Average Moves")
    )
  )

p_macro_diag <- ggplot(
  macro_diag_df,
  aes(
    x = fitted,
    y = observed,
    color = tc_bucket,
    shape = split
  )
) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.55, size = 1.8) +
  facet_wrap(
    ~ outcome,
    scales = "free",
    ncol = 2,
    labeller = as_labeller(c(
      HHI = "HHI",
      N_eff = "N[eff]",
      `Draw Rate` = "Draw~Rate",
      `Upset Rate` = "Upset~Rate",
      `Average Moves` = "Average~Moves"
    ), label_parsed)
  ) +
  labs(
    title = "Observed vs. Fitted Diagnostics for Retained Macro Models",
    x = "Fitted Value",
    y = "Observed Value",
    color = "Time Control",
    shape = "Split"
  ) +
  guides(
    color = guide_legend(override.aes = list(alpha = 1, size = 2.5)),
    shape = guide_legend(override.aes = list(alpha = 1, size = 2.5))
  ) +
  scale_color_few(
    limits = c("blitz", "rapid", "classical"),
    labels = c("Blitz", "Rapid", "Classical")
  ) +
  theme_few() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p_macro_diag

saveRDS(p_macro_diag, "../files_for_submission/Paper_files/figure-rds/p_macro_diag.rds")
```


#### Opening Popularity

```{r Opening Popularity Prep}
top_k <- 12
rating_bin_width <- 200   # Keep consistent with EDA
min_bin_total <- 2000     # Drop sparse rating_bin x month x tc cells

# Base table
pom2 <- pom_tc |>
  filter(
    !is.na(opening), !is.na(tc_bucket), !is.na(month),
    !is.na(avg_rating), !is.na(games)
  ) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date)),
    rating_bin = as.integer(floor(avg_rating / rating_bin_width) * rating_bin_width)
  )

# Month-forward split (80/20)
months_ordered <- pom2 |>
  distinct(month, month_date, t) |>
  arrange(month_date)

cut_idx <- floor(0.80 * nrow(months_ordered))
train_months <- months_ordered |> slice(1:cut_idx) |> pull(month)
test_months  <- months_ordered |> slice((cut_idx + 1):nrow(months_ordered)) |> pull(month)

stopifnot(length(intersect(train_months, test_months)) == 0)

pom_train <- pom2 |> filter(month %in% train_months)
pom_test  <- pom2 |> filter(month %in% test_months)

# Choose Top-K openings pooled from train only
top_openings_pooled <- pom_train |>
  group_by(opening) |>
  summarise(total_games = sum(games, na.rm = TRUE), .groups = "drop") |>
  slice_max(total_games, n = top_k, with_ties = FALSE)

top_tbl <- top_openings_pooled |>
  mutate(is_top = TRUE) |>
  select(opening, is_top)

# Aggregate to adoption counts at (month, tc_bucket, rating_bin, opening2)
counts_long <- pom2 |>
  left_join(top_tbl, by = "opening") |>
  mutate(opening2 = if_else(is_top %in% TRUE, opening, "Other")) |>
  group_by(month, month_date, t, tc_bucket, rating_bin, opening2) |>
  summarise(games = sum(games, na.rm = TRUE), .groups = "drop")

# Total games per (month, tc_bucket, rating_bin)
bin_totals <- counts_long |>
  group_by(month, month_date, t, tc_bucket, rating_bin) |>
  summarise(bin_total = sum(games, na.rm = TRUE), .groups = "drop") |>
  filter(bin_total >= min_bin_total)

counts_long <- counts_long |>
  inner_join(bin_totals, by = c("month","month_date","t","tc_bucket","rating_bin"))

# Wide response for multinomial
counts_wide <- counts_long |>
  pivot_wider(names_from = opening2, values_from = games, values_fill = 0)

train_wide <- counts_wide |> filter(month %in% train_months)
test_wide <- counts_wide |> filter(month %in% test_months)

id_cols <- c("month","month_date","t","tc_bucket","rating_bin","bin_total")
resp_cols <- setdiff(names(train_wide), id_cols)

# Force "Other" to be the baseline by making it the last response column
if ("Other" %in% resp_cols) {
  resp_cols <- c(setdiff(resp_cols, "Other"), "Other")
}
train_wide <- train_wide |> select(all_of(id_cols), all_of(resp_cols))
test_wide <- test_wide  |> select(all_of(id_cols), all_of(resp_cols))

# Make response names syntactic for VGAM
resp_cols_safe <- make.names(resp_cols, unique = TRUE)

train_mn <- train_wide |> rename_with(~ make.names(.x, unique = TRUE), all_of(resp_cols))
test_mn <- test_wide  |> rename_with(~ make.names(.x, unique = TRUE), all_of(resp_cols))

name_map <- tibble(
  opening      = resp_cols,
  opening_safe = resp_cols_safe
)

# Force integer counts explicitly
train_mn <- train_mn |>
  mutate(across(all_of(resp_cols_safe), ~ as.integer(round(.x))))
test_mn <- test_mn  |>
  mutate(across(all_of(resp_cols_safe), ~ as.integer(round(.x))))

# Per-game multinomial log loss
mn_logloss <- function(counts, probs, eps = 1e-12) {
  probs <- pmax(probs, eps)
  -sum(counts * log(probs)) / sum(counts)
}

# Metadata for paper
op_meta <- list(
  top_k = top_k,
  rating_bin_width = rating_bin_width,
  min_bin_total = min_bin_total,
  train_months = train_months,
  test_months = test_months,
  openings_top = top_openings_pooled
)

# Sanity:

# Response categories basically absent in a TC?
train_mn |>
  select(tc_bucket, all_of(resp_cols_safe)) |>
  group_by(tc_bucket) |>
  summarise(across(all_of(resp_cols_safe), sum), .groups = "drop") |>
  pivot_longer(-tc_bucket, names_to = "cat", values_to = "total") |>
  group_by(cat) |>
  summarise(min_total = min(total), .groups = "drop") |>
  arrange(min_total) |>
  head(10)

# "Other" dominating the simplex? 
train_mn |>
  transmute(other_share = .data$Other / rowSums(across(all_of(resp_cols_safe)))) |>
  summarise(
    mean_other = mean(other_share, na.rm = TRUE),
    p90_other = as.numeric(quantile(other_share, 0.9, na.rm = TRUE))
  )
```

```{r Opening Popularity Helper}
mn_time_change_tbl_fullspan <- function(mod, data, rating_ref = NULL) {
  if (is.null(rating_ref)) rating_ref <- median(data$rating_bin, na.rm = TRUE)

  t_early <- min(data$t, na.rm = TRUE)
  t_late  <- max(data$t, na.rm = TRUE)

  newdat <- crossing(
    rating_bin = rating_ref,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = c(t_early, t_late)
  ) |>
    mutate(period = if_else(t == t_early, "early", "late"))

  pred_long <- predict_mn_long(
    mod,
    newdat,
    id_cols = c("rating_bin", "tc_bucket", "t", "period")
  )
  
  pred_long |>
    select(-t) |>
    pivot_wider(
      names_from = period,
      values_from = pred_share
    ) |>
    mutate(change = late - early) |>
    arrange(tc_bucket, desc(abs(change)))
}

clean_opening_name <- function(x) make.names(x, unique = TRUE)

op_all_mn <- train_mn |>
  bind_rows(test_mn)
```

```{r Opening Popularity Prediction Helpers}
make_op_pred_grid <- function(data, rating_ref = NULL, t_ref = NULL) {
  if (is.null(t_ref)) {
    t_ref <- median(data$t, na.rm = TRUE)
  }

  rating_grid <- sort(unique(data$rating_bin))
  rating_grid <- rating_grid[
    seq(1, length(rating_grid), by = max(1, floor(length(rating_grid) / 6)))
  ]

  crossing(
    rating_bin = rating_grid,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = t_ref
  )
}

predict_mn_long <- function(mod, newdata, id_cols = c("rating_bin", "tc_bucket", "t")) {
  pred <- predict(mod, newdata = newdata[, id_cols, drop = FALSE], type = "response") |>
    as.data.frame()

  out <- bind_cols(newdata, pred) |>
    pivot_longer(
      cols = -all_of(id_cols),
      names_to = "opening",
      values_to = "pred_share"
    )

  out
}
```


```{r Opening Popularity Primary Multinomial Fit}
y_part <- paste0("cbind(", paste(resp_cols_safe, collapse = ","), ")")

# Main effects + optional interaction rating_bin:tc_bucket
fml_mn <- as.formula(paste0(
  y_part, " ~ ",
  "scale(rating_bin) + tc_bucket + t + scale(rating_bin):tc_bucket"
))

mn_pooled <- vglm(
  fml_mn,
  family = multinomial(),
  data = train_mn
)

summary(mn_pooled)
```

```{r Opening Popularity Multinomial Eval}
te <- test_mn |> select(t, tc_bucket, rating_bin, bin_total, all_of(resp_cols_safe))
Y <- te |> select(all_of(resp_cols_safe)) |> as.matrix()

pred <- predict(mn_pooled, newdata = te |> select(t, tc_bucket, rating_bin), type = "response")
pred <- as.matrix(pred)

# Align overlapping columns
common <- intersect(colnames(pred), colnames(Y))
Y_use <- Y[, common, drop = FALSE]
pred_use <- pred[, common, drop = FALSE]

probs <- pred_use / rowSums(pred_use)

row_loss <- map_dbl(seq_len(nrow(Y_use)), \(i) mn_logloss(Y_use[i,], probs[i,]))
avg_logloss <- weighted.mean(row_loss, w = rowSums(Y_use))

tibble(
  model = "multinomial pooled",
  n_test_rows = nrow(te),
  avg_logloss_per_game = avg_logloss
)
```

```{r Opening Popularity Multinomial Primary Results}
cat("Primary inferential model for opening popularity/adoption\n")
cat("Formula:\n")
print(fml_mn)

coef_tab <- as.data.frame(coef(summary(mn_pooled))) |>
  rownames_to_column("term") |>
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  )

coef_tab |>
  filter(
    str_detect(term, "^scale\\(rating_bin\\)|^t:|^tc_bucket|^scale\\(rating_bin\\):tc_bucket")
  ) |>
  arrange(p.value) |>
  head(40)
```


```{r Opening Popularity Multinomial Primary Writeup}
coef_tab <- coef(summary(mn_pooled))

mn_coef <- as.data.frame(coef_tab) |>
  rownames_to_column("term") |>
  rename(
    estimate = Estimate,
    std.error = `Std. Error`,
    statistic = `z value`,
    p.value = `Pr(>|z|)`
  )

mn_coef |> arrange(p.value) |> head(40)

# Store the formula text explicitly
fml_text <- deparse(fml_mn)
cat("\nPRIMARY MODEL FORMULA (as fit):\n", fml_text, "\n\n")

# Baseline category
cat("REFERENCE CATEGORY for the primary multinomial model:\n")
cat("  ", tail(resp_cols_safe, 1), "\n\n")

# Coefficient table
coef_tab <- as.data.frame(coef(summary(mn_pooled)))
coef_tab$term <- rownames(coef_tab)
rownames(coef_tab) <- NULL

mn_coef <- coef_tab |>
  rename(
    estimate  = Estimate,
    std_error = `Std. Error`,
    z_value   = `z value`,
    p_value   = `Pr(>|z|)`
  ) |>
  select(term, estimate, std_error, z_value, p_value) |>
  arrange(p_value)

mn_coef |> head(40)
```

```{r Opening Popularity Predicted Shares by Rating}
rating_grid <- sort(unique(test_mn$rating_bin))
rating_grid <- rating_grid[seq(1, length(rating_grid), by = max(1, floor(length(rating_grid) / 6)))]

tc_grid <- sort(unique(test_mn$tc_bucket))
t_ref <- median(test_mn$t, na.rm = TRUE)

newdat <- crossing(
  rating_bin = rating_grid,
  tc_bucket = tc_grid,
  t = t_ref
)

pred <- predict(
  mn_pooled,
  newdata = newdat |> select(t, tc_bucket, rating_bin),
  type = "response"
)

pred_df <- newdat |>
  bind_cols(as.data.frame(pred)) |>
  pivot_longer(
    cols = -c(rating_bin, tc_bucket, t),
    names_to = "opening",
    values_to = "pred_share"
  )

pred_df |>
  group_by(tc_bucket, rating_bin) |>
  arrange(tc_bucket, rating_bin, desc(pred_share)) |>
  slice_head(n = 5) |>
  ungroup()
```


```{r Opening Popularity Predicted Share Change Over Time}
t_early <- min(test_mn$t, na.rm = TRUE)
t_late <- max(test_mn$t, na.rm = TRUE)

rating_ref <- median(test_mn$rating_bin, na.rm = TRUE)
tc_grid <- sort(unique(test_mn$tc_bucket))

newdat <- crossing(
  rating_bin = rating_ref,
  tc_bucket = tc_grid,
  t = c(t_early, t_late)
) |>
  mutate(period = if_else(t == t_early, "early_test", "late_test"))

pred <- predict(
  mn_pooled,
  newdata = newdat |> select(t, tc_bucket, rating_bin),
  type = "response"
)

pred_long <- newdat |>
  bind_cols(as.data.frame(pred)) |>
  pivot_longer(
    cols = -c(rating_bin, tc_bucket, t, period),
    names_to = "opening",
    values_to = "pred_share"
  )

share_change <- pred_long |>
  pivot_wider(names_from = period, values_from = pred_share) |>
  mutate(change = late_test - early_test) |>
  arrange(tc_bucket, desc(abs(change)))

share_change |>
  group_by(tc_bucket) |>
  slice_head(n = 10) |>
  ungroup()
```


```{r Opening Popularity Count Models Prep}
top_openings <- top_openings_pooled |> pull(opening)
openings_to_fit <- top_openings  # length == top_k
stopifnot(length(openings_to_fit) == top_k)

show_opening <- top_openings_pooled |>
  arrange(desc(total_games)) |>
  slice(1) |>
  pull(opening)

# Helpers
wrmse_weighted <- function(truth, pred, w) {
  sqrt(weighted.mean((truth - pred)^2, w = w, na.rm = TRUE))
}

safe_glmnb <- function(df) {
  # NB needs variation and positive exposure
  if (nrow(df) < 20) return(NULL)
  if (all(df$games == 0)) return(NULL)
  if (any(df$bin_total <= 0, na.rm = TRUE)) return(NULL)

  tryCatch(
    glm.nb(
      games ~ scale(rating_bin) + tc_bucket + t + scale(rating_bin):tc_bucket + offset(log(bin_total)),
      data = df
    ),
    error = function(e) NULL
  )
}

safe_lm_wls <- function(df) {
  if (nrow(df) < 20) return(NULL)
  if (all(is.na(df$share))) return(NULL)
  if (all(df$bin_total <= 0, na.rm = TRUE)) return(NULL)

  tryCatch(
    lm(
      share ~ scale(rating_bin) + tc_bucket + t + scale(rating_bin):tc_bucket,
      data = df,
      weights = bin_total
    ),
    error = function(e) NULL
  )
}
```

```{r Opening Popularity NegBin Sensitivity}
counts_train <- counts_long |>
  filter(month %in% train_months) |>
  mutate(rate = games / bin_total)

counts_test <- counts_long |>
  filter(month %in% test_months) |>
  mutate(rate = games / bin_total)

nb_fits <- map(openings_to_fit, function(op) {
  tr <- counts_train |> filter(opening2 == op)
  mod <- safe_glmnb(tr)
  list(opening = op, model = mod)
})
names(nb_fits) <- openings_to_fit

# Eval NegBin
nb_eval <- map_dfr(nb_fits, function(x) {
  op <- x$opening
  mod <- x$model

  te <- counts_test |> filter(opening2 == op)

  if (is.null(mod) || nrow(te) == 0) {
    return(tibble(
      opening = op,
      model_family = "NegBin",
      n_test = nrow(te),
      wrmse_rate = NA_real_
    ))
  }

  pred_counts <- predict(mod, newdata = te, type = "response")
  pred_rate <- pred_counts / te$bin_total

  tibble(
    opening = op,
    model_family = "NegBin",
    n_test = nrow(te),
    wrmse_rate = wrmse_weighted(te$rate, pred_rate, te$bin_total)
  )
})
```

```{r Opening Popularity WLS Sensitivity}
share_tbl <- counts_long |> mutate(share = games / bin_total)

share_train <- share_tbl |>
  filter(month %in% train_months, opening2 != "Other", opening2 %in% openings_to_fit)

share_test <- share_tbl |>
  filter(month %in% test_months, opening2 != "Other", opening2 %in% openings_to_fit)

wls_fits <- map(openings_to_fit, function(op) {
  tr <- share_train |> filter(opening2 == op)
  mod <- safe_lm_wls(tr)
  list(opening = op, model = mod)
})
names(wls_fits) <- openings_to_fit

# Eval WLS
wls_eval <- map_dfr(wls_fits, function(x) {
  op <- x$opening
  mod <- x$model

  te <- share_test |> filter(opening2 == op)

  if (is.null(mod) || nrow(te) == 0) {
    return(tibble(
      opening = op,
      model_family = "WLS",
      n_test = nrow(te),
      wrmse_share = NA_real_
    ))
  }

  pred_share <- predict(mod, newdata = te)
  tibble(
    opening = op,
    model_family = "WLS",
    n_test = nrow(te),
    wrmse_share = wrmse_weighted(te$share, pred_share, te$bin_total)
  )
})
```


```{r Opening Popularity Sensitivity Comparison Tables}
nb_table <- nb_eval |>
  arrange(wrmse_rate)

wls_table <- wls_eval |>
  arrange(wrmse_share)

nb_table
wls_table

# Side-by-side comparison
compare_table <- nb_eval |>
  select(opening, n_test_nb = n_test, wrmse_rate) |>
  left_join(
    wls_eval |> select(opening, n_test_wls = n_test, wrmse_share),
    by = "opening"
  ) |>
  left_join(
    top_openings_pooled |> select(opening, total_games),
    by = "opening"
  ) |>
  arrange(desc(total_games))

compare_table
```

```{r Opening Popularity Sensitivity Examples}
cat("\n\n=== Fully written-out example opening ===\n")
cat("Opening:", show_opening, "\n\n")

cat("NegBin model summary:\n")
if (!is.null(nb_fits[[show_opening]]$model)) {
  print(summary(nb_fits[[show_opening]]$model))
} else {
  cat("NegBin model failed to fit for this opening.\n")
}

cat("\nWLS model summary:\n")
if (!is.null(wls_fits[[show_opening]]$model)) {
  print(summary(wls_fits[[show_opening]]$model))
} else {
  cat("WLS model failed to fit for this opening.\n")
}
```

```{r Opening Popularity Model Roles Summary}
cat("
Opening Popularity — Phase 2 model roles

Primary inferential model:
- mn_pooled (pooled multinomial)

Sensitivity / descriptive models:
- nb_fits  (per-opening NegBin with exposure)
- wls_fits (per-opening weighted share regressions)
")
```


##### Diagnostics

```{r Opening Popularity Multinomial Diagnostics}
# Check baseline category
tail(resp_cols_safe, 1)

# Category support by tc_bucket
cat_support <- train_mn |>
  select(tc_bucket, all_of(resp_cols_safe)) |>
  group_by(tc_bucket) |>
  summarise(across(all_of(resp_cols_safe), sum), .groups = "drop") |>
  pivot_longer(-tc_bucket, names_to = "cat", values_to = "total") |>
  group_by(cat) |>
  summarise(min_total = min(total), .groups = "drop") |>
  arrange(min_total)
cat_support |> head(10)

# Predicted shares in test
te <- test_mn |> select(t, tc_bucket, rating_bin, bin_total, all_of(resp_cols_safe))
pred <- predict(mn_pooled, newdata = te |> select(t, tc_bucket, rating_bin), type = "response")
pred <- as.matrix(pred)
row_sums <- rowSums(pred)
summary(row_sums)

# Convert to shares over the categories returned
pred_share <- pred / rowSums(pred)

# Any negative / NA / >1 entries?
summary(as.vector(pred_share))
sum(is.na(pred_share))
sum(pred_share < -1e-8)
sum(pred_share > 1 + 1e-8)

# Calibration by rating bin
Y <- te |> select(all_of(resp_cols_safe)) |> as.matrix()
obs_share <- Y / rowSums(Y)

common <- intersect(colnames(pred_share), colnames(obs_share))
calib <- tibble(
  rating_bin = te$rating_bin,
  tc_bucket  = te$tc_bucket
) |>
  bind_cols(
    as_tibble(obs_share[, common, drop = FALSE]) |> rename_with(~ paste0("obs_", .x)),
    as_tibble(pred_share[, common, drop = FALSE]) |> rename_with(~ paste0("pred_", .x))
  ) |>
  group_by(tc_bucket, rating_bin) |>
  summarise(
    across(starts_with("obs_"), mean, na.rm = TRUE),
    across(starts_with("pred_"), mean, na.rm = TRUE),
    .groups = "drop"
  )

# Sanity
one_open <- common[1]
calib |>
  select(tc_bucket, rating_bin, paste0("obs_", one_open), paste0("pred_", one_open)) |>
  head(12)
```

```{r Opening Popularity NegBin Diagnostics}
# Convergence / warnings already handled by safe_glmnb
nb_fit_status <- tibble(
  opening = names(nb_fits),
  fit_ok = map_lgl(nb_fits, ~ !is.null(.x$model))
)
nb_fit_status |> count(fit_ok)

# Residual diagnostics
op <- show_opening
nb_mod <- nb_fits[[op]]$model
if (!is.null(nb_mod)) {
  par(mfrow = c(1,2))
  plot(nb_mod, which = 1)  # Residuals vs fitted
  plot(nb_mod, which = 2)  # QQ
  par(mfrow = c(1,1))
}

# Sensitivity to bin_total threshold
thresholds <- c(min_bin_total, 5000, 10000)

nb_sens <- map_dfr(thresholds, function(th) {
  te0 <- counts_test |> filter(opening2 == op, bin_total >= th)
  if (is.null(nb_mod) || nrow(te0) == 0) return(tibble(threshold = th, wrmse_rate = NA_real_, n = nrow(te0)))

  pred_counts <- predict(nb_mod, newdata = te0, type = "response")
  pred_rate <- pred_counts / te0$bin_total
  tibble(
    threshold = th,
    n = nrow(te0),
    wrmse_rate = wrmse_weighted(te0$rate, pred_rate, te0$bin_total)
  )
})
nb_sens
```

```{r Opening Popularity WLS Diagnostics}
op <- show_opening
wls_mod <- wls_fits[[op]]$model

if (!is.null(wls_mod)) {
  plot_lm_basic(wls_mod, paste0("WLS share (", op, ")"))

  bp <- bptest(wls_mod)
  bp

  # Compare classic vs robust SEs
  wls_rob <- tidy_robust(wls_mod, type = "HC1")
  wls_rob |> filter(term %in% c("scale(rating_bin)", "t")) |> arrange(p.value)
}

# Sensitivity to threshold
thresholds <- c(min_bin_total, 5000, 10000)
wls_sens <- map_dfr(thresholds, function(th) {
  te0 <- share_test |> filter(opening2 == op, bin_total >= th)
  if (is.null(wls_mod) || nrow(te0) == 0) return(tibble(threshold = th, wrmse_share = NA_real_, n = nrow(te0)))

  pred <- predict(wls_mod, newdata = te0)
  tibble(
    threshold = th,
    n = nrow(te0),
    wrmse_share = wrmse_weighted(te0$share, pred, te0$bin_total)
  )
})
wls_sens
```

##### Summary

```{r Opening Popularity Rating Robustness}
fml_mn_rating_spline <- as.formula(paste0(
  y_part, " ~ ns(rating_bin, df = 4) + tc_bucket + t + ns(rating_bin, df = 4):tc_bucket"
))

mn_pooled_rating_spline <- vglm(
  fml_mn_rating_spline,
  family = multinomial(),
  data = train_mn
)

# Compare predictive performance
pred_lin <- predict(
  mn_pooled,
  newdata = test_mn |> select(t, tc_bucket, rating_bin),
  type = "response"
) |>
  as.matrix()

pred_spline <- predict(
  mn_pooled_rating_spline,
  newdata = test_mn |> select(t, tc_bucket, rating_bin),
  type = "response"
) |>
  as.matrix()

Y <- test_mn |> select(all_of(resp_cols_safe)) |> as.matrix()

common_lin <- intersect(colnames(pred_lin), colnames(Y))
common_spline <- intersect(colnames(pred_spline), colnames(Y))

Y_lin <- Y[, common_lin, drop = FALSE]
Y_spline <- Y[, common_spline, drop = FALSE]

p_lin <- pred_lin[, common_lin, drop = FALSE] / rowSums(pred_lin[, common_lin, drop = FALSE])
p_spline <- pred_spline[, common_spline, drop = FALSE] / rowSums(pred_spline[, common_spline, drop = FALSE])

row_loss_lin <- map_dbl(seq_len(nrow(Y_lin)), \(i) mn_logloss(Y_lin[i, ], p_lin[i, ]))
row_loss_spline <- map_dbl(seq_len(nrow(Y_spline)), \(i) mn_logloss(Y_spline[i, ], p_spline[i, ]))

op_rating_robust_tbl <- tibble(
  model = c("linear_rating", "spline_rating"),
  avg_logloss_per_game = c(
    weighted.mean(row_loss_lin, w = rowSums(Y_lin)),
    weighted.mean(row_loss_spline, w = rowSums(Y_spline))
  )
)

op_rating_robust_tbl

# Compare top predicted openings across rating bins
pred_grid <- make_op_pred_grid(test_mn)

pred_lin_long <- predict_mn_long(mn_pooled, pred_grid) |>
  mutate(model = "linear_rating")

pred_spline_long <- predict_mn_long(mn_pooled_rating_spline, pred_grid) |>
  mutate(model = "spline_rating")

bind_rows(pred_lin_long, pred_spline_long) |>
  group_by(model, tc_bucket, rating_bin) |>
  arrange(model, tc_bucket, rating_bin, desc(pred_share)) |>
  slice_head(n = 5) |>
  ungroup()
```

```{r Opening Popularity Time Robustness}
fml_mn_time_spline <- as.formula(paste0(
  y_part, " ~ scale(rating_bin) + tc_bucket + ns(t, df = 4) + scale(rating_bin):tc_bucket"
))

mn_pooled_time_spline <- vglm(
  fml_mn_time_spline,
  family = multinomial(),
  data = train_mn
)

# Predictive comparison
pred_lin <- predict(
  mn_pooled,
  newdata = test_mn |> select(t, tc_bucket, rating_bin),
  type = "response"
) |>
  as.matrix()

pred_spline <- predict(
  mn_pooled_time_spline,
  newdata = test_mn |> select(t, tc_bucket, rating_bin),
  type = "response"
) |>
  as.matrix()

Y <- test_mn |> select(all_of(resp_cols_safe)) |> as.matrix()

common_lin <- intersect(colnames(pred_lin), colnames(Y))
common_spline <- intersect(colnames(pred_spline), colnames(Y))

Y_lin <- Y[, common_lin, drop = FALSE]
Y_spline <- Y[, common_spline, drop = FALSE]

p_lin <- pred_lin[, common_lin, drop = FALSE] / rowSums(pred_lin[, common_lin, drop = FALSE])
p_spline <- pred_spline[, common_spline, drop = FALSE] / rowSums(pred_spline[, common_spline, drop = FALSE])

row_loss_lin <- map_dbl(seq_len(nrow(Y_lin)), \(i) mn_logloss(Y_lin[i, ], p_lin[i, ]))
row_loss_spline <- map_dbl(seq_len(nrow(Y_spline)), \(i) mn_logloss(Y_spline[i, ], p_spline[i, ]))

op_time_robust_tbl <- tibble(
  model = c("linear_time", "spline_time"),
  avg_logloss_per_game = c(
    weighted.mean(row_loss_lin, w = rowSums(Y_lin)),
    weighted.mean(row_loss_spline, w = rowSums(Y_spline))
  )
)

op_time_robust_tbl

op_all_mn <- train_mn |> bind_rows(test_mn)

# Compare changes under both specs
time_change_lin <- mn_time_change_tbl_fullspan(mn_pooled, op_all_mn) |>
  mutate(model = "linear_time")

time_change_spline <- mn_time_change_tbl_fullspan(mn_pooled_time_spline, op_all_mn) |>
  mutate(model = "spline_time")

bind_rows(time_change_lin, time_change_spline) |>
  arrange(model, tc_bucket, desc(abs(change))) |>
  group_by(model, tc_bucket) |>
  slice_head(n = 10) |>
  ungroup()
```

```{r Opening Popularity Threshold Robustness}
phase3_top_k <- c(8, 12, 16)
phase3_min_bin_total <- c(1000, 2000, 5000)

fit_op_mn_spec <- function(top_k_val, min_bin_total_val) {
  pom_spec <- pom_tc |>
    filter(
      !is.na(opening), !is.na(tc_bucket), !is.na(month),
      !is.na(avg_rating), !is.na(games)
    ) |>
    mutate(
      month_date = ymd(paste0(month, "-01")),
      t = as.integer(dense_rank(month_date)),
      rating_bin = as.integer(floor(avg_rating / rating_bin_width) * rating_bin_width)
    )

  pom_train_spec <- pom_spec |> filter(month %in% train_months)
  pom_test_spec  <- pom_spec |> filter(month %in% test_months)

  top_open_spec <- pom_train_spec |>
    group_by(opening) |>
    summarise(total_games = sum(games, na.rm = TRUE), .groups = "drop") |>
    slice_max(total_games, n = top_k_val, with_ties = FALSE)

  counts_long_spec <- pom_spec |>
    left_join(
      top_open_spec |> mutate(is_top = TRUE) |> select(opening, is_top),
      by = "opening"
    ) |>
    mutate(opening2 = if_else(is_top %in% TRUE, opening, "Other")) |>
    group_by(month, month_date, t, tc_bucket, rating_bin, opening2) |>
    summarise(games = sum(games, na.rm = TRUE), .groups = "drop")

  bin_totals_spec <- counts_long_spec |>
    group_by(month, month_date, t, tc_bucket, rating_bin) |>
    summarise(bin_total = sum(games, na.rm = TRUE), .groups = "drop") |>
    filter(bin_total >= min_bin_total_val)

  counts_long_spec <- counts_long_spec |>
    inner_join(bin_totals_spec, by = c("month", "month_date", "t", "tc_bucket", "rating_bin"))

  counts_wide_spec <- counts_long_spec |>
    pivot_wider(names_from = opening2, values_from = games, values_fill = 0)

  train_spec <- counts_wide_spec |> filter(month %in% train_months)
  test_spec <- counts_wide_spec |> filter(month %in% test_months)

  id_cols <- c("month", "month_date", "t", "tc_bucket", "rating_bin", "bin_total")
  resp_cols <- setdiff(names(train_spec), id_cols)

  if ("Other" %in% resp_cols) resp_cols <- c(setdiff(resp_cols, "Other"), "Other")

  train_spec <- train_spec |> select(all_of(id_cols), all_of(resp_cols))
  test_spec <- test_spec  |> select(all_of(id_cols), all_of(resp_cols))

  resp_cols_safe <- make.names(resp_cols, unique = TRUE)

  train_mn_spec <- train_spec |> rename_with(~ make.names(.x, unique = TRUE), all_of(resp_cols))
  test_mn_spec <- test_spec  |> rename_with(~ make.names(.x, unique = TRUE), all_of(resp_cols))

  if (nrow(train_mn_spec) == 0 || nrow(test_mn_spec) == 0 || length(resp_cols_safe) < 3) {
    return(tibble(
      n_test_rows = NA_integer_,
      n_categories = NA_integer_,
      avg_logloss_per_game = NA_real_,
      top_mover_1 = NA_character_,
      top_mover_1_change = NA_real_,
      top_mover_2 = NA_character_,
      top_mover_2_change = NA_real_
    ))
  }

  fml_spec <- as.formula(paste0(
    "cbind(", paste(resp_cols_safe, collapse = ","), ") ~ scale(rating_bin) + tc_bucket + t + scale(rating_bin):tc_bucket"
  ))

  mod <- vglm(fml_spec, family = multinomial(), data = train_mn_spec)

  Y <- test_mn_spec |> select(all_of(resp_cols_safe)) |> as.matrix()
  pred <- predict(mod, newdata = test_mn_spec |> select(t, tc_bucket, rating_bin), type = "response") |> as.matrix()

  common <- intersect(colnames(pred), colnames(Y))
  Y_use <- Y[, common, drop = FALSE]
  pred_use <- pred[, common, drop = FALSE]
  probs <- pred_use / rowSums(pred_use)

  row_loss <- map_dbl(seq_len(nrow(Y_use)), \(i) mn_logloss(Y_use[i, ], probs[i, ]))

  all_spec <- train_mn_spec |>
    bind_rows(test_mn_spec)
  
  change_tbl <- mn_time_change_tbl_fullspan(mod, all_spec) |>
    arrange(desc(abs(change)))

  tibble(
    n_test_rows = nrow(test_mn_spec),
    n_categories = length(common),
    avg_logloss_per_game = weighted.mean(row_loss, w = rowSums(Y_use)),
    top_mover_1 = change_tbl$opening[1],
    top_mover_1_change = change_tbl$change[1],
    top_mover_2 = change_tbl$opening[2],
    top_mover_2_change = change_tbl$change[2]
  )
}

op_phase3_thresh <- crossing(
  top_k = phase3_top_k,
  min_bin_total = phase3_min_bin_total
) |>
  mutate(
    out = map2(top_k, min_bin_total, ~ fit_op_mn_spec(.x, .y))
  ) |>
  unnest(out)

op_phase3_thresh
```

```{r Opening Popularity Agreement Summary}
# Compare multinomial time changes to one-opening sensitivity models
mn_change_primary <- mn_time_change_tbl_fullspan(mn_pooled, op_all_mn) |>
  mutate(opening = clean_opening_name(opening)) |>
  select(tc_bucket, opening, mn_change = change) |>
  mutate(mn_sign = sign_label(mn_change))

# NegBin sign for t by opening
nb_sign_tbl <- map_dfr(nb_fits, function(x) {
  mod <- x$model
  op <- clean_opening_name(x$opening)
  if (is.null(mod)) {
    return(tibble(opening = op, nb_t = NA_real_, nb_sign = NA_character_))
  }
  td <- tidy(mod)
  tt <- td |> filter(term == "t")
  tibble(
    opening = op,
    nb_t = if (nrow(tt)) tt$estimate[[1]] else NA_real_,
    nb_sign = sign_label(if (nrow(tt)) tt$estimate[[1]] else NA_real_)
  )
})

# WLS sign for t by opening
wls_sign_tbl <- map_dfr(wls_fits, function(x) {
  mod <- x$model
  op <- clean_opening_name(x$opening)
  if (is.null(mod)) {
    return(tibble(opening = op, wls_t = NA_real_, wls_sign = NA_character_))
  }
  td <- tidy(mod)
  tt <- td |> filter(term == "t")
  tibble(
    opening = op,
    wls_t = if (nrow(tt)) tt$estimate[[1]] else NA_real_,
    wls_sign = sign_label(if (nrow(tt)) tt$estimate[[1]] else NA_real_)
  )
})

mn_change_primary |>
  left_join(nb_sign_tbl, by = "opening") |>
  left_join(wls_sign_tbl, by = "opening") |>
  arrange(tc_bucket, desc(abs(mn_change))) |>
  group_by(tc_bucket) |>
  slice_head(n = 12) |>
  ungroup()
```


```{r Opening Popularity Claim Summary}

# Top predicted opening by rating bin
pred_grid <- make_op_pred_grid(test_mn)

rating_claim_tbl <- predict_mn_long(mn_pooled, pred_grid) |>
  group_by(tc_bucket, rating_bin) |>
  arrange(tc_bucket, rating_bin, desc(pred_share)) |>
  slice_head(n = 3) |>
  ungroup() |>
  mutate(claim_type = "rating_profile", .before = 1)

# Largest movers
time_claim_tbl <- mn_time_change_tbl_fullspan(mn_pooled, op_all_mn) |>
  group_by(tc_bucket) |>
  slice_head(n = 10) |>
  ungroup() |>
  mutate(claim_type = "time_change", .before = 1)

rating_claim_tbl
time_claim_tbl
```

```{r Opening Popularity Figure}
openings_keep <- c(
  "Other",
  "Caro-Kann Defense",
  "Queen's Pawn Game",
  "Philidor Defense",
  "Scandinavian Defense",
  "Accelerated London System"
)

op_all_mn <- bind_rows(train_mn, test_mn)
rating_grid <- sort(unique(op_all_mn$rating_bin))
t_ref <- median(op_all_mn$t, na.rm = TRUE)
tc_keep <- c("blitz", "rapid")

newdat <- crossing(
  rating_bin = rating_grid,
  tc_bucket = tc_keep,
  t = t_ref
)

pred_df <- predict_mn_long(
  mn_pooled,
  newdat,
  id_cols = c("rating_bin", "tc_bucket", "t")
) |>
  mutate(
    opening_raw = opening,
    opening = coalesce(
      name_map$opening[match(opening, name_map$opening_safe)],
      opening
    )
  ) |>
  filter(opening %in% openings_keep) |>
  mutate(
    tc_bucket = factor(tc_bucket, levels = c("blitz", "rapid"),
                       labels = c("Blitz", "Rapid")),
    opening = factor(opening, levels = openings_keep)
  )

# Build two sub-datasets with a shared facet label variable,
# then bind them so facet_grid / facet_wrap can lay them out in a 2x2.
pred_other <- pred_df |>
  filter(opening == "Other") |>
  mutate(panel_row = "Other")

pred_named <- pred_df |>
  filter(opening != "Other") |>
  mutate(panel_row = "Named")

pred_facet <- bind_rows(pred_other, pred_named) |>
  mutate(
    panel_row = factor(panel_row, levels = c("Other", "Named"))
  )

p_opening_rating <- ggplot(
  pred_facet,
  aes(x = rating_bin, y = pred_share, color = opening)
) +
  geom_line(linewidth = 0.95) +
  facet_grid(panel_row ~ tc_bucket, scales = "free_y") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_x_continuous(
    breaks = c(800, 1600, 2400),
    limits = c(800, 2400)
  ) +
  labs(
    x     = "Rating",
    y     = "Predicted Opening Share",
    color = NULL
  ) +
  theme_few() +
  scale_color_few(breaks = c(
    "Caro-Kann Defense",
    "Queen's Pawn Game",
    "Philidor Defense",
    "Scandinavian Defense",
    "Accelerated London System"
  )) +
  guides(
    color = guide_legend(nrow = 2, byrow = TRUE)
  ) +
    theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  ) +
  theme(
    legend.position = "bottom",
    strip.text      = element_text(face = "bold")
  )

p_opening_rating

saveRDS(p_opening_rating, "../files_for_submission/Paper_files/figure-rds/p_opening_rating.rds")
```

```{r Appendix Opening Multinomial Calibration Figure}
te_op <- test_mn |>
  select(month, t, tc_bucket, rating_bin, bin_total, all_of(resp_cols_safe))

pred_op <- predict(
  mn_pooled,
  newdata = te_op |> select(t, tc_bucket, rating_bin),
  type = "response"
) |>
  as.matrix()

obs_counts <- te_op |>
  select(all_of(resp_cols_safe)) |>
  as.matrix()

obs_share <- obs_counts / rowSums(obs_counts)
pred_share <- pred_op / rowSums(pred_op)

common <- intersect(colnames(pred_share), colnames(obs_share))

opening_map <- name_map |>
  transmute(
    opening_safe,
    opening_label = opening
  )

op_calib <- tibble(
  row_id = seq_len(nrow(te_op)),
  tc_bucket = te_op$tc_bucket,
  w = te_op$bin_total
) |>
  bind_cols(
    as_tibble(obs_share[, common, drop = FALSE]) |>
      rename_with(~ paste0("obs__", .x)),
    as_tibble(pred_share[, common, drop = FALSE]) |>
      rename_with(~ paste0("pred__", .x))
  ) |>
  pivot_longer(
    cols = -c(row_id, tc_bucket, w),
    names_to = "name",
    values_to = "share"
  ) |>
  separate(name, into = c("type", "opening_safe"), sep = "__") |>
  pivot_wider(names_from = type, values_from = share) |>
  group_by(tc_bucket, opening_safe) |>
  summarise(
    obs = weighted.mean(obs, w = w, na.rm = TRUE),
    pred = weighted.mean(pred, w = w, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(opening_map, by = "opening_safe") |>
  mutate(
    opening_label = coalesce(opening_label, opening_safe)
  ) |>
  filter(opening_safe != "Other")

p_opening_calib <- ggplot(op_calib, aes(x = pred, y = obs, color = tc_bucket)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_point(alpha = 0.85, size = 2.2) +
  coord_equal() +
  scale_x_continuous(labels = percent_format(accuracy = 0.1)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(
    title = "Observed vs. Predicted Opening Shares for \nRetained Pooled Multinomial Model",
    subtitle = "Test-Set Shares Aggregated Within Time Control; `Other` Category Excluded",
    x = "Predicted Opening Share",
    y = "Observed Opening Share",
    color = "Time Control"
  ) +
  scale_color_few(labels = c("Blitz", "Rapid")) +
  theme_few() +
  theme(
    panel.grid.major.x = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p_opening_calib

saveRDS(p_opening_calib, "../files_for_submission/Paper_files/figure-rds/p_opening_calib.rds")
```


#### Time-Use


```{r Time Use Prep, message = FALSE, warning = FALSE}
# Base tables
ptm2 <- ptm_tc |>
  filter(!is.na(player), !is.na(month), !is.na(tc_bucket),
                !is.na(avg_rating), !is.na(move_rows)) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date))
  )

mvm2 <- mvm_tc |>
  filter(!is.na(month), !is.na(tc_bucket)) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date))
  )

# Reliability thresholds
min_moves_player <- 100    # player-month-tc must have at least this many moves to be included
ptm2 <- ptm2 |> filter(move_rows >= min_moves_player)

# Same forward-split as earlier
months_ordered_tu <- ptm2 |>
  distinct(month, month_date, t) |>
  arrange(month_date)

cut_idx_tu <- floor(0.80 * nrow(months_ordered_tu))
train_months_tu <- months_ordered_tu |> slice(1:cut_idx_tu) |> pull(month)
test_months_tu  <- months_ordered_tu |> slice((cut_idx_tu+1):nrow(months_ordered_tu)) |> pull(month)

ptm_train <- ptm2 |> filter(month %in% train_months_tu)
ptm_test  <- ptm2 |> filter(month %in% test_months_tu)

mvm_train <- mvm2 |> filter(month %in% train_months_tu)
mvm_test  <- mvm2 |> filter(month %in% test_months_tu)

stopifnot(length(intersect(train_months_tu, test_months_tu)) == 0)

# Helpers
wrmse_weighted <- function(truth, pred, w) sqrt(weighted.mean((truth - pred)^2, w = w, na.rm = TRUE))
wmae_weighted  <- function(truth, pred, w) weighted.mean(abs(truth - pred), w = w, na.rm = TRUE)

# Note:
# For player_time_month_tc, move_rows is a natural weight.
# For moves_month_tc, move_rows is already a column in the derived table (called move_rows in the build script).
```

```{r Time Use Summary Helpers}
sign_label <- function(x) {
  case_when(
    is.na(x) ~ NA_character_,
    x > 0 ~ "positive",
    x < 0 ~ "negative",
    TRUE ~ "zero"
  )
}

make_mean_time_profile <- function(
  mod,
  data,
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent",
  predict_type = NULL,
  rating_values = NULL
) {
  if (is.null(rating_values)) {
    rating_grid <- sort(unique(data[[rating_var]]))
    rating_grid <- rating_grid[
      seq(1, length(rating_grid), by = max(1, floor(length(rating_grid) / 6)))
    ]
  } else {
    rating_grid <- rating_values
  }

  newdat <- crossing(
    !!rlang::sym(rating_var) := rating_grid,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = median(data$t, na.rm = TRUE)
  )

  pred <- if (is.null(predict_type)) {
    predict(mod, newdata = newdat)
  } else {
    predict(mod, newdata = newdat, type = predict_type)
  }

  newdat |>
    mutate(
      pred_value = as.numeric(pred),
      outcome = outcome_name
    )
}

make_mean_time_change <- function(
  mod,
  data,
  rating_ref = NULL,
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent",
  predict_type = NULL,
  rating_values = NULL
) {
  t_early <- min(data$t, na.rm = TRUE)
  t_late  <- max(data$t, na.rm = TRUE)

  if (is.null(rating_values)) {
    rating_grid <- if (is.null(rating_ref)) {
      median(data[[rating_var]], na.rm = TRUE)
    } else {
      rating_ref
    }
  } else {
    rating_grid <- rating_values
  }

  newdat_early <- crossing(
    !!rlang::sym(rating_var) := rating_grid,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = t_early
  )

  newdat_late <- crossing(
    !!rlang::sym(rating_var) := rating_grid,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = t_late
  )

  pred_early <- if (is.null(predict_type)) {
    predict(mod, newdata = newdat_early)
  } else {
    predict(mod, newdata = newdat_early, type = predict_type)
  }

  pred_late <- if (is.null(predict_type)) {
    predict(mod, newdata = newdat_late)
  } else {
    predict(mod, newdata = newdat_late, type = predict_type)
  }

  bind_cols(
    newdat_early |>
      rename(t_early_value = t),
    newdat_late |>
      rename(t_late_value = t) |>
      select(t_late_value),
    tibble(
      pred_early = as.numeric(pred_early),
      pred_late  = as.numeric(pred_late),
      change     = as.numeric(pred_late) - as.numeric(pred_early),
      outcome    = outcome_name
    )
  )
}
```


```{r Time Use Primary Macro Mean Models}

# Build a macro table by joining moves_month_tc to meta_month_tc
mvm_env <- mvm2 |>
  left_join(
    mm_tc2 |>
      transmute(
        month, tc_bucket, month_date, t,
        avg_rating_env = (avg_white_rating + avg_black_rating)/2,
        avg_abs_rating_diff = avg_abs_rating_diff,
        draw_rate = draw_rate
      ),
    by = c("month","tc_bucket","month_date","t")
  )

mvm_env_train <- mvm_env |> filter(month %in% train_months_tu)
mvm_env_test  <- mvm_env |> filter(month %in% test_months_tu)

# Interpret tc_bucket as control/stratifier
m_macro_add <- lm(
  avg_time_spent ~ scale(avg_rating_env) + tc_bucket + t,
  data = mvm_env_train,
  weights = move_rows
)

# Allow rating effect to differ by tc_bucket
m_macro_int <- lm(
  avg_time_spent ~ scale(avg_rating_env) * tc_bucket + t,
  data = mvm_env_train,
  weights = move_rows
)

pred_add <- predict(m_macro_add, newdata = mvm_env_test)
pred_int <- predict(m_macro_int, newdata = mvm_env_test)

tibble(
  model = c("macro_additive", "macro_interaction"),
  wrmse = c(
    wrmse_weighted(mvm_env_test$avg_time_spent, pred_add, mvm_env_test$move_rows),
    wrmse_weighted(mvm_env_test$avg_time_spent, pred_int, mvm_env_test$move_rows)
  )
)

summary(m_macro_int)

# Repeat across other macro outcomes
macro_targets <- c("median_time_spent", "p90_time_spent", "avg_clock_after", "low_time_rate")

macro_eval <- map_dfr(macro_targets, function(y) {
  f_add <- as.formula(paste0(y, " ~ scale(avg_rating_env) + tc_bucket + t"))
  f_int <- as.formula(paste0(y, " ~ scale(avg_rating_env) * tc_bucket + t"))

  m_add <- lm(f_add, data = mvm_env_train, weights = move_rows)
  m_int <- lm(f_int, data = mvm_env_train, weights = move_rows)

  p_add <- predict(m_add, newdata = mvm_env_test)
  p_int <- predict(m_int, newdata = mvm_env_test)

  tibble(
    outcome = y,
    model = c("additive", "interaction"),
    wrmse = c(
      wrmse_weighted(mvm_env_test[[y]], p_add, mvm_env_test$move_rows),
      wrmse_weighted(mvm_env_test[[y]], p_int, mvm_env_test$move_rows)
    )
  )
})

macro_eval |> arrange(outcome, model)
```

```{r Time Use Primary Player Mean Models}
# Weight player rows by move_rows
ptm_train2 <- ptm_train |> filter(!is.na(avg_time_spent))
ptm_test2  <- ptm_test  |> filter(!is.na(avg_time_spent))

# Move to feols so clustered inference is easy
m_player_add_fe <- feols(
  avg_time_spent ~ scale(avg_rating) + tc_bucket + t,
  data = ptm_train2,
  weights = ~move_rows
)

m_player_int_fe <- feols(
  avg_time_spent ~ scale(avg_rating) * tc_bucket + t,
  data = ptm_train2,
  weights = ~move_rows
)

p_add <- predict(m_player_add_fe, newdata = ptm_test2)
p_int <- predict(m_player_int_fe, newdata = ptm_test2)

tibble(
  model = c("player_add_fe", "player_int_fe"),
  wrmse = c(
    wrmse_weighted(ptm_test2$avg_time_spent, p_add, ptm_test2$move_rows),
    wrmse_weighted(ptm_test2$avg_time_spent, p_int, ptm_test2$move_rows)
  )
)

summary(m_player_int_fe)
```

```{r Time Use Primary Mean Results}
cat("Primary inferential models for average time-use\n\n")

cat("Macro model (HAC inference)\n")
compare_lm_ses(m_macro_int, hac_lag = 3) |>
  filter(term %in% c("scale(avg_rating_env)", "t") | str_detect(term, "tc_bucket"))

cat("\nPlayer model (clustered inference)\n")
compare_fixest_vcovs(
  m_player_int_fe,
  specs = list(
    player_cluster = ~player,
    two_way = ~player + month
  )
) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  arrange(term, vcov)
```


```{r Time Use Robust Sensitivity}
# Sensitivity model for outlier robustness; not a primary inferential model

ptm_train_r <- ptm_train2 |>
  select(avg_time_spent, avg_rating, tc_bucket, t) |>
  drop_na()

m_rlm <- rlm(
  avg_time_spent ~ scale(avg_rating) * tc_bucket + t,
  data = ptm_train_r
)

summary(m_rlm)
```
```{r Time Use Quantile Sensitivity}
# Sensitivity models for distributional heterogeneity; not primary.

# Median regression (tau = 0.5) and "upper tail" regression (tau = 0.9)
ptm_train_q <- ptm_train2 |>
  select(avg_time_spent, avg_rating, tc_bucket, t) |>
  drop_na()

m_q50 <- rq(
  avg_time_spent ~ scale(avg_rating) * tc_bucket + t,
  tau = 0.5,
  data = ptm_train_q
)

m_q90 <- rq(
  avg_time_spent ~ scale(avg_rating) * tc_bucket + t,
  tau = 0.9,
  data = ptm_train_q
)

summary(m_q50)
summary(m_q90)

# Out-of-sample quantile metric
pinball_loss <- function(y, qhat, tau) mean((tau - (y < qhat)) * (y - qhat))

ptm_test_q <- ptm_test2 |>
  select(avg_time_spent, avg_rating, tc_bucket, t, move_rows) |>
  drop_na()

p50 <- predict(m_q50, newdata = ptm_test_q)
p90 <- predict(m_q90, newdata = ptm_test_q)

tibble(
  model = c("rq_tau_0.5", "rq_tau_0.9"),
  loss = c(
    pinball_loss(ptm_test_q$avg_time_spent, p50, 0.5),
    pinball_loss(ptm_test_q$avg_time_spent, p90, 0.9)
  )
)
```

```{r Time Use Primary Low-Time Model}
ptm_train_ltr <- ptm_train |>
  filter(!is.na(low_time_rate), !is.na(avg_rating), !is.na(tc_bucket)) |>
  mutate(
    # Keep inside (0,1) to avoid boundary issues
    ltr_adj = pmin(pmax(low_time_rate, 1e-6), 1 - 1e-6)
  )

ptm_test_ltr <- ptm_test |>
  filter(!is.na(low_time_rate), !is.na(avg_rating), !is.na(tc_bucket)) |>
  mutate(
    ltr_adj = pmin(pmax(low_time_rate, 1e-6), 1 - 1e-6)
  )

m_frac <- glm(
  ltr_adj ~ scale(avg_rating) * tc_bucket + t,
  family = quasibinomial(),
  data = ptm_train_ltr,
  weights = move_rows
)

summary(m_frac)

# Evaluate
p_frac <- predict(m_frac, newdata = ptm_test_ltr, type = "response")
wrmse_weighted(ptm_test_ltr$low_time_rate, p_frac, ptm_test_ltr$move_rows)
```

```{r Time Use Low-Time Binomial Sensitivity}
ptm_train_bin <- ptm_train |>
  filter(!is.na(low_time_rate), !is.na(move_rows)) |>
  mutate(
    low_time_moves = as.integer(round(low_time_rate * move_rows)),
    low_time_moves = pmin(pmax(low_time_moves, 0L), as.integer(move_rows))
  )

ptm_test_bin <- ptm_test |>
  filter(!is.na(low_time_rate), !is.na(move_rows)) |>
  mutate(
    low_time_moves = as.integer(round(low_time_rate * move_rows)),
    low_time_moves = pmin(pmax(low_time_moves, 0L), as.integer(move_rows))
  )

m_binom <- glm(
  cbind(low_time_moves, move_rows - low_time_moves) ~ scale(avg_rating) * tc_bucket + t,
  family = binomial(),
  data = ptm_train_bin
)

summary(m_binom)

p_bin <- predict(m_binom, newdata = ptm_test_bin, type = "response")
wrmse_weighted(ptm_test_bin$low_time_rate, p_bin, ptm_test_bin$move_rows)
```

```{r Time Use Low-Time Primary Results}
cat("Primary inferential model for low-time frequency\n")

tidy(m_frac) |>
  filter(
    term %in% c("(Intercept)", "scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  arrange(p.value)

cat("\nBinomial approximation sensitivity check\n")
tidy(m_binom) |>
  filter(
    term %in% c("(Intercept)", "scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  arrange(p.value)
```


```{r Time Use Phase Allocation}
phase_vars <- c("avg_time_opening", "avg_time_middlegame", "avg_time_endgame")

phase_eval <- map_dfr(phase_vars, function(y) {
  tr <- ptm_train |> filter(!is.na(.data[[y]]))
  te <- ptm_test  |> filter(!is.na(.data[[y]]))

  if (nrow(tr) < 200 || nrow(te) < 50) {
    return(tibble(phase = y, model = NA_character_, wrmse = NA_real_))
  }

  m_add <- lm(
    reformulate(c("scale(avg_rating)", "tc_bucket", "t"), response = y),
    data = tr,
    weights = move_rows
  )
  m_int <- lm(
    reformulate(c("scale(avg_rating)*tc_bucket", "t"), response = y),
    data = tr,
    weights = move_rows
  )

  p_add <- predict(m_add, newdata = te)
  p_int <- predict(m_int, newdata = te)

  tibble(
    phase = y,
    model = c("additive", "interaction"),
    wrmse = c(
      wrmse_weighted(te[[y]], p_add, te$move_rows),
      wrmse_weighted(te[[y]], p_int, te$move_rows)
    )
  )
})

phase_eval
```

```{r Time Use Model Roles Summary}
cat("
Time-Use — Phase 2 model roles

Primary inferential models:
- Macro average time-use: m_macro_add / m_macro_int (HAC inference)
- Player average time-use: m_player_add_fe / m_player_int_fe (clustered inference)
- Low-time frequency: m_frac
- Low-time companion sensitivity: m_binom

Sensitivity / descriptive models:
- m_rlm
- m_q50, m_q90
- phase allocation models
")
```


##### Diagnostics

```{r Time Use Macro HAC Diagnostics}
compare_lm_ses(m_macro_add, hac_lag = 3)
compare_lm_ses(m_macro_int, hac_lag = 3)

plot_resid_acf(m_macro_int, "Macro time-use interaction", lag.max = 12)

bgtest(m_macro_int, order = 3)
```

`scale(avg_rating):tc_bucket_classical` is not stable, so it shouldn't be treated as an inferential result going forward.


```{r Time Use Primary Mean Model Diagnostics}
# Macro interaction model diagnostics
plot_lm_basic(m_macro_int, "Macro time-use (interaction)")
plot_resid_acf(m_macro_int, "Macro time-use (interaction)", lag.max = 12)

# Player interaction model diagnostics
player_diag_df <- tibble(
  fitted = fitted(m_player_int_fe),
  resid = resid(m_player_int_fe)
)

ggplot(player_diag_df, aes(x = fitted, y = resid)) +
  geom_point(alpha = 0.25) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Player time-use (interaction, FEOLS) — Residuals vs Fitted",
       x = "Fitted", y = "Residual") +
  theme_minimal()

ggplot(player_diag_df, aes(sample = resid)) +
  stat_qq(alpha = 0.25) +
  stat_qq_line() +
  labs(title = "Player time-use (interaction, FEOLS) — QQ plot") +
  theme_minimal()

# BP for macro only
tibble(
  model = "m_macro_int",
  bp_p  = bptest(m_macro_int)$p.value
)

# HAC comparison for macro model
compare_lm_ses(m_macro_int, hac_lag = 3)

# Clustered VCOVs for player model
compare_fixest_vcovs(
  m_player_int_fe,
  specs = list(
    iid = "iid",
    HC1 = "hetero",
    player_cluster = ~player,
    month_cluster = ~month,
    two_way = ~player + month
  )
) |>
  filter(term %in% c("scale(avg_rating)", "t") | str_detect(term, "tc_bucket")) |>
  arrange(term, vcov)
```

```{r Time Use Robust-Quantile Diagnostics}
# Robust regression
if (exists("m_rlm")) {
  rlm_resid <- resid(m_rlm)
  summary(rlm_resid)
  ggplot(tibble(resid = rlm_resid), aes(resid)) +
    geom_histogram(bins = 60) +
    labs(title = "RLM residual distribution") +
    theme_minimal()
}

# Quantile regression
if (exists("m_q50")) {
  ptm_test_q2 <- ptm_test_q |> mutate(pred50 = p50, pred90 = p90)
  ggplot(ptm_test_q2, aes(x = pred50, y = avg_time_spent)) +
    geom_point(alpha = 0.15) +
    labs(title = "Quantile regression (tau=0.5): predicted vs observed", x = "Predicted median", y = "Observed") +
    theme_minimal()

  ggplot(ptm_test_q2, aes(x = pred90, y = avg_time_spent)) +
    geom_point(alpha = 0.15) +
    labs(title = "Quantile regression (tau=0.9): predicted vs observed", x = "Predicted 90th pct", y = "Observed") +
    theme_minimal()
}
```

```{r Time Use Low-Time Diagnostics}
# Check mass at zero
ltr_mass <- ptm2 |>
  summarise(
    n = n(),
    frac_zero = mean(low_time_rate == 0, na.rm = TRUE),
    frac_one  = mean(low_time_rate == 1, na.rm = TRUE)
  )
ltr_mass

# Calibration
ptm_test_ltr2 <- ptm_test_ltr |> mutate(pred = p_frac)

cal_ltr <- ptm_test_ltr2 |>
  mutate(bin = cut(low_time_rate, breaks = c(-Inf, 0, 0.01, 0.02, 0.05, 0.1, 0.2, 0.4, Inf))) |>
  group_by(tc_bucket, bin) |>
  summarise(
    n = n(),
    obs = weighted.mean(low_time_rate, w = move_rows, na.rm = TRUE),
    pred= weighted.mean(pred, w = move_rows, na.rm = TRUE),
    .groups = "drop"
  )
cal_ltr |> head(12)
```


##### Summary

```{r Time Use Mean Robustness}
# Macro robustness
m_macro_int_spline <- lm(
  avg_time_spent ~ scale(avg_rating_env) * tc_bucket + ns(t, df = 4),
  data = mvm_env_train,
  weights = move_rows
)

pred_macro_linear <- predict(m_macro_int, newdata = mvm_env_test)
pred_macro_spline <- predict(m_macro_int_spline, newdata = mvm_env_test)

macro_phase3_tbl <- tibble(
  model = c("macro_linear_time", "macro_spline_time"),
  wrmse = c(
    wrmse_weighted(mvm_env_test$avg_time_spent, pred_macro_linear, mvm_env_test$move_rows),
    wrmse_weighted(mvm_env_test$avg_time_spent, pred_macro_spline, mvm_env_test$move_rows)
  )
)

macro_phase3_tbl

# Player robustness
m_player_int_fe_spline <- feols(
  avg_time_spent ~ ns(avg_rating, df = 4) * tc_bucket + t,
  data = ptm_train2,
  weights = ~move_rows
)

pred_player_linear <- predict(m_player_int_fe, newdata = ptm_test2)
pred_player_spline <- predict(m_player_int_fe_spline, newdata = ptm_test2)

player_phase3_tbl <- tibble(
  model = c("player_linear_rating", "player_spline_rating"),
  wrmse = c(
    wrmse_weighted(ptm_test2$avg_time_spent, pred_player_linear, ptm_test2$move_rows),
    wrmse_weighted(ptm_test2$avg_time_spent, pred_player_spline, ptm_test2$move_rows)
  )
)

player_phase3_tbl

# Interpretable predicted profiles
mean_profile_linear <- make_mean_time_profile(
  m_player_int_fe,
  ptm_test2,
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent"
) |>
  mutate(model = "player_linear_rating")

mean_profile_spline <- make_mean_time_profile(
  m_player_int_fe_spline,
  ptm_test2,
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent"
) |>
  mutate(model = "player_spline_rating")

bind_rows(mean_profile_linear, mean_profile_spline) |>
  arrange(model, tc_bucket, avg_rating)

mean_change_linear <- make_mean_time_change(
  m_player_int_fe,
  ptm_train2 |> bind_rows(ptm_test2),
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent"
) |>
  mutate(model = "player_linear_rating")

mean_change_spline <- make_mean_time_change(
  m_player_int_fe_spline,
  ptm_train2 |> bind_rows(ptm_test2),
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent"
) |>
  mutate(model = "player_spline_rating")

bind_rows(mean_change_linear, mean_change_spline) |>
  arrange(model, desc(abs(change)))
```

```{r Time Use Mean Agreement Summary}
primary_terms <- compare_fixest_vcovs(
  m_player_int_fe,
  specs = list(two_way = ~player + month)
) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    fe_estimate = estimate,
    fe_sign = sign_label(estimate),
    fe_p = p.value
  )

rlm_terms <- tidy(m_rlm) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    rlm_estimate = estimate,
    rlm_sign = sign_label(estimate)
  )

q50_terms <- tidy(m_q50) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    q50_estimate = estimate,
    q50_sign = sign_label(estimate)
  )

q90_terms <- tidy(m_q90) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    q90_estimate = estimate,
    q90_sign = sign_label(estimate)
  )

time_use_mean_agreement <- primary_terms |>
  left_join(rlm_terms, by = "term") |>
  left_join(q50_terms, by = "term") |>
  left_join(q90_terms, by = "term")

time_use_mean_agreement
```

```{r Time Use Low Time Robustness}
# Primary vs companion family
frac_terms <- tidy(m_frac) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    frac_estimate = estimate,
    frac_sign = sign_label(estimate),
    frac_p = p.value
  )

binom_terms <- tidy(m_binom) |>
  filter(
    term %in% c("scale(avg_rating)", "t") |
      str_detect(term, "tc_bucket")
  ) |>
  transmute(
    term,
    binom_estimate = estimate,
    binom_sign = sign_label(estimate),
    binom_p = p.value
  )

low_time_family_agreement <- frac_terms |>
  left_join(binom_terms, by = "term")

low_time_family_agreement

# Flexible rating robustness for low-time
m_frac_spline <- glm(
  ltr_adj ~ ns(avg_rating, df = 4) * tc_bucket + t,
  family = quasibinomial(),
  data = ptm_train_ltr,
  weights = move_rows
)

p_frac_linear <- predict(m_frac, newdata = ptm_test_ltr, type = "response")
p_frac_spline <- predict(m_frac_spline, newdata = ptm_test_ltr, type = "response")

low_time_phase3_tbl <- tibble(
  model = c("frac_linear_rating", "frac_spline_rating"),
  wrmse = c(
    wrmse_weighted(ptm_test_ltr$low_time_rate, p_frac_linear, ptm_test_ltr$move_rows),
    wrmse_weighted(ptm_test_ltr$low_time_rate, p_frac_spline, ptm_test_ltr$move_rows)
  )
)

low_time_phase3_tbl

cat("\nInterpretation note: frac_spline_rating is numerically unstable in the tails, so substantive low-time conclusions should rely on frac_linear_rating plus the binomial companion check.\n")

# Predicted profiles
low_time_profile_linear <- make_mean_time_profile(
  m_frac,
  ptm_test_ltr,
  rating_var = "avg_rating",
  outcome_name = "low_time_rate",
  predict_type = "response"
) |>
  mutate(model = "frac_linear_rating")

low_time_profile_spline <- make_mean_time_profile(
  m_frac_spline,
  ptm_test_ltr,
  rating_var = "avg_rating",
  outcome_name = "low_time_rate",
  predict_type = "response"
) |>
  mutate(model = "frac_spline_rating")

low_time_profile_linear |>
  arrange(tc_bucket, avg_rating)

bind_rows(low_time_profile_linear, low_time_profile_spline) |>
  arrange(model, tc_bucket, avg_rating)
```

```{r Time Use Claim Summary}
clean_ratings <- c(800, 1600, 2400)

mean_claim_tbl <- make_mean_time_change(
  m_player_int_fe,
  ptm_train2 |> bind_rows(ptm_test2),
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent",
  rating_values = clean_ratings
) |>
  mutate(claim_type = "mean_time_change", .before = 1)

mean_profile_claim_tbl <- make_mean_time_profile(
  m_player_int_fe,
  ptm_test2,
  rating_var = "avg_rating",
  outcome_name = "avg_time_spent",
  rating_values = clean_ratings
) |>
  mutate(claim_type = "mean_time_profile", .before = 1)

low_time_profile_claim_tbl <- make_mean_time_profile(
  m_frac,
  ptm_test_ltr,
  rating_var = "avg_rating",
  outcome_name = "low_time_rate",
  predict_type = "response",
  rating_values = clean_ratings
) |>
  mutate(claim_type = "low_time_profile", .before = 1)

bind_rows(
  mean_claim_tbl,
  mean_profile_claim_tbl,
  low_time_profile_claim_tbl
) |>
  arrange(claim_type, tc_bucket, avg_rating)
```

```{r Time Use Uncertainty Tables}
clean_ratings <- c(800, 1600, 2400)
z_crit <- qnorm(0.975)

# Helpers
safe_model_matrix <- function(mod, newdata) {
  tt <- delete.response(terms(mod))
  model.matrix(tt, data = newdata)
}

pred_ci_from_vcov <- function(mod, newdata, vcov_mat,
                              type = c("identity", "response"),
                              level = 0.95) {
  type <- match.arg(type)
  alpha <- 1 - level
  z <- qnorm(1 - alpha/2)

  X <- safe_model_matrix(mod, newdata)
  b <- coef(mod)

  # Align in case vcov/model.matrix names differ in ordering
  keep <- intersect(colnames(X), names(b))
  X <- X[, keep, drop = FALSE]
  b <- b[keep]
  V <- vcov_mat[keep, keep, drop = FALSE]

  eta <- as.numeric(X %*% b)
  se_eta <- sqrt(pmax(0, diag(X %*% V %*% t(X))))

  if (type == "identity") {
    fit <- eta
    lo  <- eta - z * se_eta
    hi  <- eta + z * se_eta
  } else {
    fam <- family(mod)
    fit <- fam$linkinv(eta)
    lo  <- fam$linkinv(eta - z * se_eta)
    hi  <- fam$linkinv(eta + z * se_eta)
  }

  bind_cols(
    newdata,
    tibble(
      estimate = fit,
      std.error = se_eta,
      conf.low = lo,
      conf.high = hi
    )
  )
}

change_ci_from_vcov <- function(mod, newdata_early, newdata_late, vcov_mat,
                                type = c("identity", "response"),
                                level = 0.95) {
  type <- match.arg(type)
  alpha <- 1 - level
  z <- qnorm(1 - alpha/2)

  X0 <- safe_model_matrix(mod, newdata_early)
  X1 <- safe_model_matrix(mod, newdata_late)
  b  <- coef(mod)

  keep <- Reduce(intersect, list(colnames(X0), colnames(X1), names(b)))
  X0 <- X0[, keep, drop = FALSE]
  X1 <- X1[, keep, drop = FALSE]
  b  <- b[keep]
  V  <- vcov_mat[keep, keep, drop = FALSE]

  eta0 <- as.numeric(X0 %*% b)
  eta1 <- as.numeric(X1 %*% b)
  Xd   <- X1 - X0
  se_d <- sqrt(pmax(0, diag(Xd %*% V %*% t(Xd))))

  if (type == "identity") {
    change <- eta1 - eta0
    lo <- change - z * se_d
    hi <- change + z * se_d
    est0 <- eta0
    est1 <- eta1
  } else {
    fam <- family(mod)
    est0 <- fam$linkinv(eta0)
    est1 <- fam$linkinv(eta1)

    # CI is for link-scale contrast, then mapped as a probability-difference approximation
    # via endpoint transformation.
    lo <- fam$linkinv(eta1 - z * se_d) - fam$linkinv(eta0)
    hi <- fam$linkinv(eta1 + z * se_d) - fam$linkinv(eta0)
    change <- est1 - est0
  }

  bind_cols(
    newdata_early |>
      rename(t_early_value = t),
    newdata_late |>
      rename(t_late_value = t) |>
      select(t_late_value),
    tibble(
      estimate_early = est0,
      estimate_late = est1,
      change = change,
      std.error = se_d,
      conf.low = lo,
      conf.high = hi
    )
  )
}

make_profile_grid <- function(data, rating_values, rating_var = "avg_rating") {
  crossing(
    !!rlang::sym(rating_var) := rating_values,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = median(data$t, na.rm = TRUE)
  )
}

make_change_grid <- function(data, rating_values, rating_var = "avg_rating") {
  t_early <- min(data$t, na.rm = TRUE)
  t_late  <- max(data$t, na.rm = TRUE)

  grid_early <- crossing(
    !!rlang::sym(rating_var) := rating_values,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = t_early
  )

  grid_late <- crossing(
    !!rlang::sym(rating_var) := rating_values,
    tc_bucket = sort(unique(data$tc_bucket)),
    t = t_late
  )

  list(early = grid_early, late = grid_late)
}

# VCOV choices aligned to model roles
vcov_macro_hac <- NeweyWest(
  m_macro_int,
  lag = 3,
  prewhite = FALSE,
  adjust = TRUE
)

vcov_player_2way <- vcov(
  m_player_int_fe,
  cluster = ~player + month
)

vcov_frac_quasi <- vcov(m_frac)

# Macro mean time-use uncertainty tables
macro_profile_grid <- make_profile_grid(
  mvm_env_test,
  rating_values = clean_ratings,
  rating_var = "avg_rating_env"
)

macro_change_grid <- make_change_grid(
  bind_rows(mvm_env_train, mvm_env_test),
  rating_values = clean_ratings,
  rating_var = "avg_rating_env"
)

macro_mean_profile_unc <- pred_ci_from_vcov(
  m_macro_int,
  newdata = macro_profile_grid,
  vcov_mat = vcov_macro_hac,
  type = "identity"
) |>
  mutate(
    model = "m_macro_int",
    claim_type = "macro_mean_profile",
    outcome = "avg_time_spent",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating_env)

macro_mean_change_unc <- change_ci_from_vcov(
  m_macro_int,
  newdata_early = macro_change_grid$early,
  newdata_late  = macro_change_grid$late,
  vcov_mat = vcov_macro_hac,
  type = "identity"
) |>
  mutate(
    model = "m_macro_int",
    claim_type = "macro_mean_change",
    outcome = "avg_time_spent",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating_env)

# Player mean time-use uncertainty tables
player_profile_grid <- make_profile_grid(
  ptm_test2,
  rating_values = clean_ratings,
  rating_var = "avg_rating"
)

player_change_grid <- make_change_grid(
  bind_rows(ptm_train2, ptm_test2),
  rating_values = clean_ratings,
  rating_var = "avg_rating"
)

player_mean_profile_unc <- pred_ci_from_vcov(
  m_player_int_fe,
  newdata = player_profile_grid,
  vcov_mat = vcov_player_2way,
  type = "identity"
) |>
  mutate(
    model = "m_player_int_fe",
    claim_type = "player_mean_profile",
    outcome = "avg_time_spent",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating)

player_mean_change_unc <- change_ci_from_vcov(
  m_player_int_fe,
  newdata_early = player_change_grid$early,
  newdata_late  = player_change_grid$late,
  vcov_mat = vcov_player_2way,
  type = "identity"
) |>
  mutate(
    model = "m_player_int_fe",
    claim_type = "player_mean_change",
    outcome = "avg_time_spent",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating)

# Low-time uncertainty tables
low_time_profile_grid <- make_profile_grid(
  ptm_test_ltr,
  rating_values = clean_ratings,
  rating_var = "avg_rating"
)

low_time_change_grid <- make_change_grid(
  bind_rows(ptm_train_ltr, ptm_test_ltr),
  rating_values = clean_ratings,
  rating_var = "avg_rating"
)

low_time_profile_unc <- pred_ci_from_vcov(
  m_frac,
  newdata = low_time_profile_grid,
  vcov_mat = vcov_frac_quasi,
  type = "response"
) |>
  mutate(
    model = "m_frac",
    claim_type = "low_time_profile",
    outcome = "low_time_rate",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating)

low_time_change_unc <- change_ci_from_vcov(
  m_frac,
  newdata_early = low_time_change_grid$early,
  newdata_late  = low_time_change_grid$late,
  vcov_mat = vcov_frac_quasi,
  type = "response"
) |>
  mutate(
    model = "m_frac",
    claim_type = "low_time_change",
    outcome = "low_time_rate",
    .before = 1
  ) |>
  arrange(tc_bucket, avg_rating)

# Combined tables
time_use_uncertainty_tbl <- bind_rows(
  macro_mean_profile_unc |>
    rename(rating_value = avg_rating_env),
  macro_mean_change_unc |>
    rename(rating_value = avg_rating_env),
  player_mean_profile_unc |>
    rename(rating_value = avg_rating),
  player_mean_change_unc |>
    rename(rating_value = avg_rating),
  low_time_profile_unc |>
    rename(rating_value = avg_rating),
  low_time_change_unc |>
    rename(rating_value = avg_rating)
) |>
  select(
    claim_type, model, outcome, tc_bucket, rating_value,
    everything()
  )

time_use_uncertainty_tbl

time_use_uncertainty_pub <- time_use_uncertainty_tbl |>
  mutate(
    estimate_ci = case_when(
      claim_type %in% c("macro_mean_change", "player_mean_change", "low_time_change") ~
        sprintf("%.4f [%.4f, %.4f]", change, conf.low, conf.high),
      TRUE ~
        sprintf("%.4f [%.4f, %.4f]", estimate, conf.low, conf.high)
    )
  ) |>
  select(
    claim_type, model, outcome, tc_bucket, rating_value,
    estimate_ci
  )

time_use_uncertainty_pub
```

```{r Time Use Two-Panel Profile Figure}
rating_vals <- c(800, 1600, 2400)

# Hold time fixed at the retained reference used in profile helpers
t_ref_mean <- median(ptm_test2$t, na.rm = TRUE)
t_ref_low  <- median(ptm_test_ltr$t, na.rm = TRUE)

tc_levels <- sort(unique(ptm_test2$tc_bucket))

# Retained player-level mean interaction model
newdat_mean <- crossing(
  avg_rating = rating_vals,
  tc_bucket = tc_levels,
  t = t_ref_mean
)

pred_mean <- newdat_mean |>
  mutate(
    pred_value = as.numeric(predict(m_player_int_fe, newdata = newdat_mean)),
    panel = "Panel A: Predicted Average Time Spent by Rating and Time Control"
  )


# Retained fractional low-time model
newdat_low <- crossing(
  avg_rating = rating_vals,
  tc_bucket = sort(unique(ptm_test_ltr$tc_bucket)),
  t = t_ref_low
)

pred_low <- newdat_low |>
  mutate(
    pred_value = as.numeric(predict(m_frac, newdata = newdat_low, type = "response")),
    panel = "Panel B: Predicted Low-Time Frequency by Rating and Time Control"
  )

# Combine for faceted plot
plot_df <- bind_rows(pred_mean, pred_low) |>
  mutate(
    tc_bucket = factor(
      tc_bucket,
      levels = c("blitz", "rapid", "classical"),
      labels = c("Blitz", "Rapid", "Classical")
    ),
    panel = factor(
      panel,
      levels = c(
        "Panel A: Predicted Average Time Spent by Rating and Time Control",
        "Panel B: Predicted Low-Time Frequency by Rating and Time Control"
      )
    )
  )

p_time_panels <- ggplot(
  plot_df,
  aes(x = avg_rating, y = pred_value, color = tc_bucket, group = tc_bucket)
) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 2) +
  facet_wrap(~ panel, ncol = 1, scales = "free_y") +
  scale_x_continuous(
    breaks = rating_vals,
    limits = c(min(rating_vals), max(rating_vals))
  ) +
  scale_y_continuous(
    labels = function(x) x
  ) +
  labs(
    x = "Rating",
    y = NULL,
    color = NULL
  ) +
  theme_few() +
  scale_color_few() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold")
  ) +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  )

p_time_panels

saveRDS(p_time_panels, "../files_for_submission/Paper_files/figure-rds/p_time_panels.rds")
```

```{r Appendix Low-Time Calibration Figure}
ptm_ltr_plot <- ptm_test_ltr |>
  mutate(
    pred = p_frac,
    rating_bin = as.character(as.integer(floor(avg_rating / 200) * 200)),
    tc_bucket = factor(tc_bucket, levels = c("blitz", "rapid", "classical"))
  ) |>
  group_by(tc_bucket, rating_bin) |>
  summarise(
    n = n(),
    weight = sum(move_rows, na.rm = TRUE),
    obs = weighted.mean(low_time_rate, w = move_rows, na.rm = TRUE),
    pred = weighted.mean(pred, w = move_rows, na.rm = TRUE),
    .groups = "drop"
  )

p_lowtime_calib <- ggplot(
  ptm_ltr_plot,
  aes(x = pred, y = obs, color = tc_bucket)
) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    color = "grey50"
  ) +
  geom_point(alpha = 0.8, size = 2.2) +
  coord_equal(xlim = c(0, 0.12), ylim = c(0, 0.12)) +
  scale_x_continuous(
    breaks = c(0.00, 0.03, 0.06, 0.09, 0.12),
    labels = percent_format(accuracy = 0.1)
  ) +
  scale_y_continuous(
    breaks = c(0.00, 0.03, 0.06, 0.09, 0.12),
    labels = percent_format(accuracy = 0.1)
  ) +
  scale_color_few(
    name = NULL,
    breaks = c("blitz", "rapid", "classical"),
    labels = c("Blitz", "Rapid", "Classical"),
    drop = FALSE
  ) +
  labs(
    x = "Predicted Low-Time Rate",
    y = "Observed Low-Time Rate"
  ) +
  guides(
    color = guide_legend(
      override.aes = list(alpha = 1, size = 3)
    )
  ) +
  theme_few() +
  theme(
    panel.grid.major.x = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p_lowtime_calib

saveRDS(p_lowtime_calib, "../files_for_submission/Paper_files/figure-rds/p_lowtime_calib.rds")
```


#### Player-Progression Models


```{r Progression Prep}
min_games_pm <- 10   # Reliability filter for player-month
min_moves_ptm <- 40   # If joining time-use, filter for stability
horizons <- c(1, 3)  # Months-ahead targets

# Base player-month table
pm2 <- pm |>
  filter(!is.na(player), !is.na(month), !is.na(avg_rating), !is.na(games)) |>
  mutate(
    month_date = ymd(paste0(month, "-01"))
  ) |>
  arrange(player, month_date) |>
  filter(games >= min_games_pm)

# Build future rating targets
# d_rating_1m = next month's avg_rating - current
# d_rating_3m = 3 months ahead - current
pm_targets <- pm2 |>
  group_by(player) |>
  mutate(
    next_rating_1m = lead(avg_rating, 1),
    next_rating_3m = lead(avg_rating, 3),
    d_rating_1m = next_rating_1m - avg_rating,
    d_rating_3m = next_rating_3m - avg_rating,
    abs_d_rating_1m = abs(d_rating_1m)
  ) |>
  ungroup()

# Month-forward split
months_ordered_pm <- pm_targets |>
  distinct(month, month_date) |>
  arrange(month_date)

cut_idx <- floor(0.80 * nrow(months_ordered_pm))
train_months_pm <- months_ordered_pm |> slice(1:cut_idx) |> pull(month)
test_months_pm  <- months_ordered_pm |> slice((cut_idx+1):nrow(months_ordered_pm)) |> pull(month)

stopifnot(length(intersect(train_months_pm, test_months_pm)) == 0)

pm_train0 <- pm_targets |> filter(month %in% train_months_pm)
pm_test0 <- pm_targets |> filter(month %in% test_months_pm)

range(pm_train0$month_date); range(pm_test0$month_date)
```
```{r Progression Join Time Use}
# Collapse ptm_tc to player-month by move_rows-weighted averages across tc_bucket
ptm_pm <- ptm_tc |>
  filter(!is.na(player), !is.na(month), !is.na(move_rows)) |>
  filter(move_rows >= min_moves_ptm) |>
  group_by(player, month) |>
  summarise(
    move_rows_total = sum(move_rows, na.rm = TRUE),
    avg_time_spent_all = sum(avg_time_spent * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    p90_time_spent_all = sum(p90_time_spent * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    avg_clock_after_all = sum(avg_clock_after * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    low_time_rate_all = sum(low_time_rate * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    avg_time_opening_all = sum(avg_time_opening * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    avg_time_middlegame_all = sum(avg_time_middlegame * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    avg_time_endgame_all = sum(avg_time_endgame * move_rows, na.rm = TRUE) / sum(move_rows, na.rm = TRUE),
    .groups = "drop"
  )

# Join to player-month
pm_with_time <- pm_targets |>
  left_join(ptm_pm, by = c("player","month"))
```

```{r Progression Join Repertoire}
# Collapse pom_tc to player-month opening shares (across tc_bucket)
pom_pm <- pom_tc |>
  filter(!is.na(player), !is.na(month), !is.na(opening), !is.na(games)) |>
  group_by(player, month, opening) |>
  summarise(games_opening = sum(games), .groups = "drop") |>
  group_by(player, month) |>
  mutate(games_total_opening = sum(games_opening)) |>
  ungroup()

# Share in most-played opening (top_opening_share)
top_share <- pom_pm |>
  group_by(player, month) |>
  summarise(
    top_opening_share = max(games_opening / games_total_opening, na.rm = TRUE),
    .groups = "drop"
  )

# Churn proxy
top_opening <- pom_pm |>
  group_by(player, month) |>
  slice_max(order_by = games_opening, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(player, month, top_opening_name = opening)

opening_switch <- top_opening |>
  mutate(month_date = ymd(paste0(month, "-01"))) |>
  arrange(player, month_date) |>
  group_by(player) |>
  mutate(prev_top_opening = lag(top_opening_name),
                switched_opening = if_else(!is.na(prev_top_opening) & top_opening_name != prev_top_opening, 1L, 0L)) |>
  ungroup() |>
  select(player, month, switched_opening)

pm_feat <- pm_with_time |>
  left_join(top_share, by = c("player","month")) |>
  left_join(opening_switch, by = c("player","month"))
```

```{r Progression Model Datasets}
# 1-month improvement target
dat_1m <- pm_feat |>
  filter(!is.na(d_rating_1m)) 

# 3-month improvement target
dat_3m <- pm_feat |>
  filter(!is.na(d_rating_3m))

# Stability target
dat_abs <- pm_feat |>
  filter(!is.na(abs_d_rating_1m))

# Train/test split by month (forward-in-time)
split_train_test <- function(df) {
  list(
    train = df |> filter(month %in% train_months_pm),
    test = df |> filter(month %in% test_months_pm)
  )
}

spl_1m <- split_train_test(dat_1m)
spl_3m <- split_train_test(dat_3m)
spl_abs <- split_train_test(dat_abs)

# Sanity
c(
  n_train_1m = nrow(spl_1m$train), n_test_1m = nrow(spl_1m$test),
  n_train_3m = nrow(spl_3m$train), n_test_3m = nrow(spl_3m$test),
  n_train_abs = nrow(spl_abs$train), n_test_abs = nrow(spl_abs$test)
)
```

```{r Progression Predictive Ridge Benchmark}
# Base feature set
base_feature_cols <- c(
  "avg_rating", "games", "avg_moves", "opening_diversity",
  "top_opening_share", "switched_opening"
)

# Keep only columns that actually exist
base_feature_cols <- intersect(base_feature_cols, names(spl_1m$train))

prep_xy <- function(df, ycol, xcols) {
  df2 <- df |> select(all_of(c(ycol, xcols))) |> drop_na()
  y <- df2[[ycol]]
  X <- model.matrix(reformulate(xcols), data = df2)[, -1, drop = FALSE]
  list(X = X, y = y)
}

fit_glmnet_ridge <- function(spl, ycol, xcols, label) {
  tr <- prep_xy(spl$train, ycol, xcols)
  te <- prep_xy(spl$test,  ycol, xcols)

  cv <- cv.glmnet(tr$X, tr$y, alpha = 0.0)  # Ridge
  pred <- as.numeric(predict(cv, newx = te$X, s = "lambda.1se"))

  metrics <- tibble(
    model = label,
    n_train = nrow(tr$X),
    n_test  = nrow(te$X),
    rmse = rmse_vec(te$y, pred),
    mae  = mae_vec(te$y, pred)
  )

  # Coefficient table
  coef_mat <- coef(cv, s = "lambda.1se")
  coef_df <- as.matrix(coef_mat) |>
    as.data.frame() |>
    rownames_to_column("term") |>
    rename(coef = last_col()) |>
    mutate(abs_coef = abs(coef)) |>
    arrange(desc(abs_coef))

  list(metrics = metrics, coef = coef_df)
}

# Fit for all three targets
ridge_1m <- fit_glmnet_ridge(spl_1m,  "d_rating_1m", base_feature_cols, "glmnet_ridge_1m")
ridge_3m <- fit_glmnet_ridge(spl_3m,  "d_rating_3m", base_feature_cols, "glmnet_ridge_3m")
ridge_abs <- fit_glmnet_ridge(spl_abs, "abs_d_rating_1m", base_feature_cols, "glmnet_ridge_abs1m")

bind_rows(ridge_1m$metrics, ridge_3m$metrics, ridge_abs$metrics) |>
  arrange(rmse)

# Example coefficient table
ridge_1m$coef |> filter(term != "(Intercept)") |> slice_head(n = 25)
```

```{r Progression Primary FE Models}
fit_fe_model <- function(spl, ycol, xcols, label) {
  fe_train <- spl$train |>
    select(player, month, month_date, all_of(c(ycol, xcols))) |>
    drop_na()

  fe_test <- spl$test |>
    select(player, month, month_date, all_of(c(ycol, xcols))) |>
    drop_na()

  fml_fe <- as.formula(paste0(
    ycol, " ~ ",
    paste(xcols, collapse = " + "),
    " | player"
  ))

  # Cluster by player in the inferential fit
  m_fe <- feols(
    fml_fe,
    data = fe_train,
    vcov = ~player
  )

  pred <- predict(m_fe, newdata = fe_test)

  list(
    model = m_fe,
    metrics = tibble(
      model = label,
      n_train = nrow(fe_train),
      n_test_raw = nrow(fe_test),
      n_test_pred = sum(!is.na(pred)),
      rmse = rmse_vec(fe_test[[ycol]][!is.na(pred)], pred[!is.na(pred)]),
      mae = mae_vec(fe_test[[ycol]][!is.na(pred)], pred[!is.na(pred)])
    )
  )
}

fe_1m <- fit_fe_model(spl_1m,  "d_rating_1m",     base_feature_cols, "fixest_player_FE_1m")
fe_3m <- fit_fe_model(spl_3m,  "d_rating_3m",     base_feature_cols, "fixest_player_FE_3m")
fe_abs <- fit_fe_model(spl_abs, "abs_d_rating_1m", base_feature_cols, "fixest_player_FE_abs1m")

bind_rows(fe_1m$metrics, fe_3m$metrics, fe_abs$metrics) |>
  arrange(rmse)
```

```{r Progression Primary FE Results}
cat("Primary inferential models for player progression\n\n")

cat("1-month rating change\n")
compare_fixest_vcovs(
  fe_1m$model,
  specs = list(
    player_cluster = ~player,
    two_way = ~player + month
  )
) |>
  arrange(term, vcov)

cat("\n3-month rating change\n")
compare_fixest_vcovs(
  fe_3m$model,
  specs = list(
    player_cluster = ~player,
    two_way = ~player + month
  )
) |>
  arrange(term, vcov)

cat("\nAbsolute 1-month change\n")
compare_fixest_vcovs(
  fe_abs$model,
  specs = list(
    player_cluster = ~player,
    two_way = ~player + month
  )
) |>
  arrange(term, vcov)
```

```{r Progression FE Primary Writeup}
cat("Primary inferential writeup: 1-month FE model\n")

tidy(fe_1m$model, vcov = ~player + month) |>
  arrange(p.value)

cat("\nModel fit summary\n")
fe_1m$metrics
```


```{r Progression Predictive RF Benchmark}
# Predictive benchmark only; not a primary inferential model

fit_rf <- function(spl, ycol, xcols, label) {
  tr <- spl$train |> select(all_of(c(ycol, xcols))) |> drop_na()
  te <- spl$test  |> select(all_of(c(ycol, xcols))) |> drop_na()

  rf_mod <- ranger(
    reformulate(xcols, response = ycol),
    data = tr,
    num.trees = 500,
    importance = "impurity"
  )

  pred <- predict(rf_mod, data = te)$predictions

  list(
    metrics = tibble(
      model = label,
      n_train = nrow(tr),
      n_test  = nrow(te),
      rmse = rmse_vec(te[[ycol]], pred),
      mae = mae_vec(te[[ycol]], pred)
    ),
    importance = sort(rf_mod$variable.importance, decreasing = TRUE)
  )
}

rf_1m <- fit_rf(spl_1m,  "d_rating_1m", base_feature_cols, "ranger_rf_1m")
rf_3m <- fit_rf(spl_3m,  "d_rating_3m", base_feature_cols, "ranger_rf_3m")
rf_abs <- fit_rf(spl_abs, "abs_d_rating_1m", base_feature_cols, "ranger_rf_abs1m")

bind_rows(rf_1m$metrics, rf_3m$metrics, rf_abs$metrics) |>
  arrange(rmse)

# Example variable importance
head(rf_1m$importance, 20)
```


```{r Progression Predictive Benchmark Comparison}
# Gather metrics from the three families
ridge_tbl <- bind_rows(ridge_1m$metrics, ridge_3m$metrics, ridge_abs$metrics)
fe_tbl <- bind_rows(fe_1m$metrics, fe_3m$metrics, fe_abs$metrics)
rf_tbl <- bind_rows(rf_1m$metrics, rf_3m$metrics, rf_abs$metrics)

bind_rows(ridge_tbl, fe_tbl, rf_tbl) |>
  arrange(model)
```

```{r Progression Model Roles Summary}
cat("
Player Progression — Phase 2 model roles

Primary inferential models:
- fe_1m$model
- fe_3m$model
- fe_abs$model

Predictive-only benchmark models:
- ridge_1m, ridge_3m, ridge_abs
- rf_1m, rf_3m, rf_abs
")
```


##### Diagnostics

```{r Progression Ridge Predictive Diagnostics}
# Compare lambda.min vs lambda.1se performance
glmnet_compare_lambda <- function(fit_obj, spl, ycol, xcols) {
  tr <- prep_xy(spl$train, ycol, xcols)
  te <- prep_xy(spl$test,  ycol, xcols)

  cv <- cv.glmnet(tr$X, tr$y, alpha = 0.0)
  pred_min <- as.numeric(predict(cv, newx = te$X, s = "lambda.min"))
  pred_1se <- as.numeric(predict(cv, newx = te$X, s = "lambda.1se"))

  tibble(
    target = ycol,
    rmse_min = rmse_vec(te$y, pred_min),
    rmse_1se = rmse_vec(te$y, pred_1se),
    mae_min = mae_vec(te$y, pred_min),
    mae_1se = mae_vec(te$y, pred_1se)
  )
}

glmnet_compare_lambda(NULL, spl_1m, "d_rating_1m", base_feature_cols)
glmnet_compare_lambda(NULL, spl_3m, "d_rating_3m", base_feature_cols)
glmnet_compare_lambda(NULL, spl_abs,"abs_d_rating_1m", base_feature_cols)

# Residual distribution for ridge (1m)
tr <- prep_xy(spl_1m$train, "d_rating_1m", base_feature_cols)
te <- prep_xy(spl_1m$test,  "d_rating_1m", base_feature_cols)
cv <- cv.glmnet(tr$X, tr$y, alpha = 0.0)
pred <- as.numeric(predict(cv, newx = te$X, s = "lambda.1se"))
res <- te$y - pred
ggplot(tibble(resid = res), aes(resid)) +
  geom_histogram(bins = 80) +
  labs(title = "GLMNET ridge residual distribution (1m)") +
  theme_minimal()
```

```{r Progression FE Primary Diagnostics}
# Coverage already computed
fe_cov <- fe_1m$metrics |> transmute(share_pred = n_test_pred / n_test_raw)
fe_cov

# Sensitivity
fe_train <- spl_1m$train |>
  select(player, month, d_rating_1m, all_of(base_feature_cols)) |>
  drop_na()

fe_test  <- spl_1m$test |>
  select(player, month, d_rating_1m, all_of(base_feature_cols)) |>
  drop_na()

m_fe_player <- feols(
  as.formula(paste0("d_rating_1m ~ ", paste(base_feature_cols, collapse = " + "), " | player")),
  data = fe_train,
  vcov = ~player
)

m_fe_player_month <- feols(
  as.formula(paste0("d_rating_1m ~ ", paste(base_feature_cols, collapse = " + "), " | player + month")),
  data = fe_train,
  vcov = ~player
)

pred1 <- predict(m_fe_player, newdata = fe_test)
pred2 <- predict(m_fe_player_month, newdata = fe_test)

tibble(
  spec = c("player_FE", "player+month_FE"),
  rmse = c(
    rmse_vec(fe_test$d_rating_1m[!is.na(pred1)], pred1[!is.na(pred1)]),
    rmse_vec(fe_test$d_rating_1m[!is.na(pred2)], pred2[!is.na(pred2)])
  ),
  n_pred = c(sum(!is.na(pred1)), sum(!is.na(pred2)))
)

# Player-clustered vs two-way clustered inference
compare_fixest_vcovs(
  fe_1m$model,
  specs = list(
    player_cluster = ~player,
    two_way = ~player + month
  )
) |>
  arrange(term, vcov)
```

```{r Progression RF Predictive Diagnostics}
# Refit RF with different seeds and compare RMSE + importance rank
rf_stability <- function(seed) {
  set.seed(seed)
  tr <- spl_1m$train |> select(d_rating_1m, all_of(base_feature_cols)) |> drop_na()
  te <- spl_1m$test |> select(d_rating_1m, all_of(base_feature_cols)) |> drop_na()

  mod <- ranger(d_rating_1m ~ ., data = tr, num.trees = 400, importance = "impurity")
  pred <- predict(mod, data = te)$predictions
  imp <- sort(mod$variable.importance, decreasing = TRUE)

  tibble(seed = seed,
                 rmse = rmse_vec(te$d_rating_1m, pred),
                 top1 = names(imp)[1],
                 top2 = names(imp)[2])
}

rf_stab_tbl <- map_dfr(c(1,2,3,4,5), rf_stability)
rf_stab_tbl
```


##### Summary

```{r Progression FE Helpers}

fit_fe_spec <- function(spl, ycol, xcols, fe_rhs = "player", label = NULL) {
  tr <- spl$train |>
    select(player, month, month_date, all_of(c(ycol, xcols))) |>
    drop_na()

  te <- spl$test |>
    select(player, month, month_date, all_of(c(ycol, xcols))) |>
    drop_na()

  fml <- as.formula(
    paste0(ycol, " ~ ", paste(xcols, collapse = " + "), " | ", fe_rhs)
  )

  mod <- feols(
    fml,
    data = tr,
    vcov = ~player
  )

  pred <- predict(mod, newdata = te)

  list(
    model = mod,
    train = tr,
    test = te,
    metrics = tibble(
      model = label %||% paste0(ycol, "_", fe_rhs),
      n_train = nrow(tr),
      n_test_raw = nrow(te),
      n_test_pred = sum(!is.na(pred)),
      rmse = rmse_vec(te[[ycol]][!is.na(pred)], pred[!is.na(pred)]),
      mae  = mae_vec(te[[ycol]][!is.na(pred)], pred[!is.na(pred)])
    )
  )
}

prog_base_features <- base_feature_cols

fe_1m_base_p3 <- fit_fe_spec(
  spl = spl_1m,
  ycol = "d_rating_1m",
  xcols = prog_base_features,
  fe_rhs = "player",
  label = "fe_1m_base"
)
```


```{r Progression Phase3 Feature Sets}
# Time-use features
time_feature_candidates <- c(
  "avg_time_spent_all",
  "p90_time_spent_all",
  "avg_clock_after_all",
  "low_time_rate_all",
  "avg_time_opening_all",
  "avg_time_middlegame_all",
  "avg_time_endgame_all"
)

# Expanded feature set
prog_time_features <- union(
  prog_base_features,
  intersect(time_feature_candidates, names(pm_feat))
)

tibble(
  feature_set = c("base", "base_plus_time"),
  n_features = c(length(prog_base_features), length(prog_time_features))
)
```


```{r Progression Feature Robustness}
common_features <- union(prog_base_features, prog_time_features)

spl_1m_common <- list(
  train = spl_1m$train |>
    select(player, month, month_date, d_rating_1m, all_of(common_features)) |>
    drop_na(),
  test = spl_1m$test |>
    select(player, month, month_date, d_rating_1m, all_of(common_features)) |>
    drop_na()
)

fe_1m_base_common_p3 <- fit_fe_spec(
  spl = spl_1m_common,
  ycol = "d_rating_1m",
  xcols = prog_base_features,
  fe_rhs = "player",
  label = "fe_1m_base_common_support"
)

fe_1m_time_common_p3 <- fit_fe_spec(
  spl = spl_1m_common,
  ycol = "d_rating_1m",
  xcols = prog_time_features,
  fe_rhs = "player",
  label = "fe_1m_base_plus_time_common_support"
)

bind_rows(
  fe_1m_base_common_p3$metrics,
  fe_1m_time_common_p3$metrics
) |>
  arrange(rmse)
```


```{r Progression Phase3 Rating Robustness}
fit_fe_rating_spec <- function(spl, ycol, use_spline = FALSE, fe_rhs = "player", label = NULL) {
  tr <- spl$train |>
    select(
      player, month, month_date, all_of(c(
        ycol, setdiff(prog_base_features, "avg_rating"), "avg_rating"
      ))
    ) |>
    drop_na()

  te <- spl$test |>
    select(
      player, month, month_date, all_of(c(
        ycol, setdiff(prog_base_features, "avg_rating"), "avg_rating"
      ))
    ) |>
    drop_na()

  other_rhs <- setdiff(prog_base_features, "avg_rating")

  rating_term <- if (use_spline) {
    "ns(avg_rating, df = 4)"
  } else {
    "avg_rating"
  }

  rhs <- paste(c(rating_term, other_rhs), collapse = " + ")

  fml <- as.formula(paste0(ycol, " ~ ", rhs, " | ", fe_rhs))

  mod <- feols(
    fml,
    data = tr,
    vcov = ~player
  )

  pred <- predict(mod, newdata = te)

  list(
    model = mod,
    train = tr,
    test = te,
    metrics = tibble(
      model = label,
      n_train = nrow(tr),
      n_test_raw = nrow(te),
      n_test_pred = sum(!is.na(pred)),
      rmse = rmse_vec(te[[ycol]][!is.na(pred)], pred[!is.na(pred)]),
      mae = mae_vec(te[[ycol]][!is.na(pred)], pred[!is.na(pred)])
    )
  )
}

fe_1m_linear_rating_p3 <- fit_fe_rating_spec(
  spl = spl_1m,
  ycol = "d_rating_1m",
  use_spline = FALSE,
  fe_rhs = "player",
  label = "fe_1m_linear_rating"
)

fe_1m_spline_rating_p3 <- fit_fe_rating_spec(
  spl = spl_1m,
  ycol = "d_rating_1m",
  use_spline = TRUE,
  fe_rhs = "player",
  label = "fe_1m_spline_rating"
)

bind_rows(
  fe_1m_linear_rating_p3$metrics,
  fe_1m_spline_rating_p3$metrics
) |>
  arrange(rmse)
```


```{r Progression FE Structure Robustness}
fe_train_1m_p3 <- spl_1m$train |>
  select(player, month, month_date, d_rating_1m, all_of(prog_base_features)) |>
  drop_na()

m_fe_player_p3 <- feols(
  as.formula(paste0("d_rating_1m ~ ", paste(prog_base_features, collapse = " + "), " | player")),
  data = fe_train_1m_p3,
  vcov = ~player
)

m_fe_player_month_p3 <- feols(
  as.formula(paste0("d_rating_1m ~ ", paste(prog_base_features, collapse = " + "), " | player + month")),
  data = fe_train_1m_p3,
  vcov = ~player
)

tibble(
  model = c("fe_1m_player_FE", "fe_1m_player_month_FE"),
  n_train = c(nobs(m_fe_player_p3), nobs(m_fe_player_month_p3)),
  r2_within = c(
    as.numeric(fitstat(m_fe_player_p3, "wr2")),
    as.numeric(fitstat(m_fe_player_month_p3, "wr2"))
  )
)
```

```{r Progression FE Claim Agreement}
# Compare key coefficients across the core FE robustness specs

tidy_key_terms <- function(mod, vcov_spec = ~player + month) {
  tidy(mod, vcov = vcov_spec) |>
    filter(term %in% prog_base_features) |>
    mutate(
      sign = case_when(
        estimate > 0 ~ "positive",
        estimate < 0 ~ "negative",
        TRUE ~ "zero"
      )
    ) |>
    select(term, estimate, p.value, sign)
}

prog_claim_agreement <- list(
  base = tidy_key_terms(fe_1m_base_p3$model),
  base_common_support = tidy_key_terms(fe_1m_base_common_p3$model),
  plus_time_common_support = tidy_key_terms(fe_1m_time_common_p3$model),
  spline_rating = tidy_key_terms(fe_1m_spline_rating_p3$model),
  player_month_fe = tidy_key_terms(m_fe_player_month_p3)
) |>
  imap(~ mutate(.x, spec = .y)) |>
  bind_rows() |>
  pivot_wider(
    names_from = spec,
    values_from = c(estimate, p.value, sign)
  ) |>
  arrange(term)

prog_claim_agreement
```

```{r Progression Predictive Summary}
prog_pred_summary <- bind_rows(
  ridge_1m$metrics, ridge_3m$metrics, ridge_abs$metrics,
  fe_1m$metrics, fe_3m$metrics, fe_abs$metrics,
  rf_1m$metrics, rf_3m$metrics, rf_abs$metrics
) |>
  mutate(
    target = case_when(
      str_detect(model, "1m") & !str_detect(model, "abs") ~ "1m_change",
      str_detect(model, "3m") ~ "3m_change",
      str_detect(model, "abs") ~ "abs_1m_change",
      TRUE ~ NA_character_
    )
  ) |>
  arrange(target, rmse)

prog_pred_summary
```

```{r Progression Predictive Variable Summary}
ridge_1m_top <- ridge_1m$coef |>
  filter(term != "(Intercept)") |>
  slice_head(n = 10) |>
  transmute(
    model = "ridge_1m",
    term,
    importance = abs_coef
  )

rf_1m_top <- tibble(
  term = names(rf_1m$importance),
  importance = as.numeric(rf_1m$importance)
) |>
  slice_head(n = 10) |>
  mutate(model = "rf_1m", .before = 1)

bind_rows(ridge_1m_top, rf_1m_top)
```


```{r Progression Claim Summary}
progression_phase3_claim_summary <- list(
  fe_performance = bind_rows(
    fe_1m_base_p3$metrics,
    fe_1m_base_common_p3$metrics,
    fe_1m_time_common_p3$metrics,
    fe_1m_linear_rating_p3$metrics,
    fe_1m_spline_rating_p3$metrics
  ),
  fe_structure_sensitivity = tibble(
    model = c("fe_1m_player_FE", "fe_1m_player_month_FE"),
    n_train = c(nobs(m_fe_player_p3), nobs(m_fe_player_month_p3)),
    wr2 = c(
      as.numeric(fitstat(m_fe_player_p3, "wr2")),
      as.numeric(fitstat(m_fe_player_month_p3, "wr2"))
    )
  ),
  fe_claim_agreement = prog_claim_agreement,
  predictive_summary = prog_pred_summary,
  predictive_variables_1m = bind_rows(ridge_1m_top, rf_1m_top)
)

progression_phase3_claim_summary$fe_performance
progression_phase3_claim_summary$fe_structure_sensitivity
progression_phase3_claim_summary$fe_claim_agreement
progression_phase3_claim_summary$predictive_summary
progression_phase3_claim_summary$predictive_variables_1m
```

```{r Progression FE Paper Tables}
# Setup
retained_predictors <- base_feature_cols

predictor_labels <- c(
  avg_rating = "current average rating",
  games = "games played",
  avg_moves = "average moves",
  opening_diversity = "opening diversity",
  top_opening_share = "top-opening concentration/share",
  switched_opening = "switched most-played opening"
)

predictor_types <- tribble(
  ~term, ~predictor_class, ~contrast_rule,
  "avg_rating",         "continuous", "p25_vs_p75",
  "games",              "continuous", "p25_vs_p75",
  "avg_moves",          "continuous", "p25_vs_p75",
  "opening_diversity",  "continuous", "p25_vs_p75",
  "top_opening_share",  "continuous", "p25_vs_p75",
  "switched_opening",   "binary",     "0_vs_1"
)

fe_scaling_notes <- tribble(
  ~term, ~label, ~scaling_note,
  "avg_rating", "current average rating", "raw scale; not standardized, logged, or centered in retained FE models",
  "games", "games played", "raw count; not standardized, logged, or centered in retained FE models",
  "avg_moves", "average moves", "raw scale; not standardized, logged, or centered in retained FE models",
  "opening_diversity", "opening diversity", "raw scale; not standardized, logged, or centered in retained FE models",
  "top_opening_share", "top-opening concentration/share", "raw share/proportion scale; not standardized, logged, or centered in retained FE models",
  "switched_opening", "switched most-played opening", "raw binary indicator (0/1); not standardized, logged, or centered in retained FE models"
)

model_map <- list(
  list(
    target = "1-month change",
    ycol = "d_rating_1m",
    model_obj = fe_1m$model,
    metrics_obj = fe_1m$metrics,
    spl = spl_1m
  ),
  list(
    target = "3-month change",
    ycol = "d_rating_3m",
    model_obj = fe_3m$model,
    metrics_obj = fe_3m$metrics,
    spl = spl_3m
  ),
  list(
    target = "absolute 1-month change",
    ycol = "abs_d_rating_1m",
    model_obj = fe_abs$model,
    metrics_obj = fe_abs$metrics,
    spl = spl_abs
  )
)


# Coefficient table
extract_fe_coef_table <- function(mod, target_name, terms_keep) {
  tidy(mod, conf.int = TRUE, vcov = ~player + month) |>
    filter(term %in% terms_keep) |>
    mutate(
      target = target_name,
      label = coalesce(unname(predictor_labels[term]), term)
    ) |>
    select(
      target, term, label,
      estimate,
      std.error,
      conf.low,
      conf.high,
      p.value
    ) |>
    arrange(target, match(term, terms_keep))
}

fe_coef_table <- map_dfr(
  model_map,
  ~ extract_fe_coef_table(.x$model_obj, .x$target, retained_predictors)
)


# Predictor labels actually used
fe_predictor_labels <- tibble(term = retained_predictors) |>
  mutate(
    label = coalesce(unname(predictor_labels[term]), term),
    retained_in_final_fe = TRUE
  )

# Distribution summaries
make_distribution_summary <- function(spl, ycol, terms_keep, target_name) {
  dat <- spl$train |>
    select(player, all_of(c(ycol, terms_keep))) |>
    drop_na()

  map_dfr(terms_keep, function(v) {
    x <- dat[[v]]

    if (all(x %in% c(0, 1), na.rm = TRUE)) {
      tibble(
        target = target_name,
        term = v,
        label = coalesce(unname(predictor_labels[v]), v),
        predictor_class = "binary",
        mean = mean(x, na.rm = TRUE),
        sd = sd(x, na.rm = TRUE),
        p25 = NA_real_,
        p50 = NA_real_,
        p75 = NA_real_,
        prop_1 = mean(x == 1, na.rm = TRUE)
      )
    } else {
      tibble(
        target = target_name,
        term = v,
        label = coalesce(unname(predictor_labels[v]), v),
        predictor_class = "continuous",
        mean = mean(x, na.rm = TRUE),
        sd = sd(x, na.rm = TRUE),
        p25 = as.numeric(quantile(x, 0.25, na.rm = TRUE, names = FALSE)),
        p50 = as.numeric(quantile(x, 0.50, na.rm = TRUE, names = FALSE)),
        p75 = as.numeric(quantile(x, 0.75, na.rm = TRUE, names = FALSE)),
        prop_1 = NA_real_
      )
    }
  })
}

fe_predictor_distribution <- map_dfr(
  model_map,
  ~ make_distribution_summary(.x$spl, .x$ycol, retained_predictors, .x$target)
)


# Reference values for non-focal covariates
make_reference_values <- function(spl, ycol, terms_keep, target_name) {
  dat <- spl$train |>
    select(player, all_of(c(ycol, terms_keep))) |>
    drop_na()

  map_dfr(terms_keep, function(v) {
    x <- dat[[v]]
    tibble(
      target = target_name,
      term = v,
      label = coalesce(unname(predictor_labels[v]), v),
      reference_value = if (all(x %in% c(0,1), na.rm = TRUE)) 0 else mean(x, na.rm = TRUE),
      reference_rule = if (all(x %in% c(0,1), na.rm = TRUE)) {
        "fixed binary reference: 0"
      } else {
        "mean of estimation sample"
      }
    )
  })
}

fe_reference_values <- map_dfr(
  model_map,
  ~ make_reference_values(.x$spl, .x$ycol, retained_predictors, .x$target)
)

# Predicted contrasts from FE models

make_contrast_table <- function(mod, spl, ycol, target_name, terms_keep) {
  dat <- spl$train |>
    select(player, all_of(c(ycol, terms_keep))) |>
    drop_na()

  coef_tbl <- tidy(mod, vcov = ~player + month) |>
    filter(term %in% terms_keep)

  vc <- vcov(mod, vcov = ~player + month)

  map_dfr(terms_keep, function(v) {
    beta <- coef_tbl$estimate[coef_tbl$term == v]
    idx <- which(names(coef(mod)) == v)
    var_beta <- vc[idx, idx]
    se_beta <- sqrt(var_beta)

    x <- dat[[v]]
    is_binary <- all(x %in% c(0,1), na.rm = TRUE)

    if (is_binary) {
      low_val <- 0
      high_val <- 1
      contrast_desc <- "0 vs 1"
      ref_note <- "all non-focal continuous covariates set to estimation-sample means; binary covariates set to 0"
    } else {
      low_val <- as.numeric(quantile(x, 0.25, na.rm = TRUE, names = FALSE))
      high_val <- as.numeric(quantile(x, 0.75, na.rm = TRUE, names = FALSE))
      contrast_desc <- "25th vs 75th percentile"
      ref_note <- "all non-focal continuous covariates set to estimation-sample means; binary covariates set to 0"
    }

    delta_x <- high_val - low_val
    pred_diff <- beta * delta_x
    se_diff <- abs(delta_x) * se_beta
    ci_low <- pred_diff - 1.96 * se_diff
    ci_high <- pred_diff + 1.96 * se_diff

    tibble(
      target = target_name,
      term = v,
      label = coalesce(unname(predictor_labels[v]), v),
      comparison_rule = contrast_desc,
      comparison_low = low_val,
      comparison_high = high_val,
      predicted_difference = pred_diff,
      std.error = se_diff,
      conf.low = ci_low,
      conf.high = ci_high,
      reference_values_note = ref_note
    )
  })
}

fe_contrast_table <- map_dfr(
  model_map,
  ~ make_contrast_table(.x$model_obj, .x$spl, .x$ycol, .x$target, retained_predictors)
)


# FE prediction support
make_support_note <- function(mod, spl, ycol, target_name, metrics_obj, terms_keep) {
  tr <- spl$train |>
    select(player, month, all_of(c(ycol, terms_keep))) |>
    drop_na()

  te <- spl$test |>
    select(player, month, all_of(c(ycol, terms_keep))) |>
    drop_na()

  pred <- predict(mod, newdata = te)

  tibble(
    target = target_name,
    estimation_sample_n = nrow(tr),
    estimation_players = n_distinct(tr$player),
    test_sample_complete_cases = nrow(te),
    supported_test_predictions = sum(!is.na(pred)),
    share_supported_test_predictions = mean(!is.na(pred)),
    coefficient_interpretation_note = "Coefficient interpretation uses the full FE estimation sample after complete-case filtering for the retained predictors.",
    prediction_support_note = "Predicted contrasts are coefficient-based within-player contrasts and are not restricted to supported out-of-sample FE predictions; test-set prediction coverage is reported separately for transparency.",
    interpretability_note = "Out-of-sample FE predictions may be unavailable for player-month observations without supported fixed effects in prediction."
  )
}

fe_prediction_support <- map_dfr(
  model_map,
  ~ make_support_note(.x$model_obj, .x$spl, .x$ycol, .x$target, .x$metrics_obj, retained_predictors)
)


# Compact model summary block
make_model_summary <- function(mod, spl, ycol, target_name) {
  tr <- spl$train |>
    select(player, month, all_of(c(ycol, retained_predictors))) |>
    drop_na()

  tibble(
    target = target_name,
    n_observations = nobs(mod),
    n_players = n_distinct(tr$player),
    within_r2 = as.numeric(fitstat(mod, "wr2"))
  )
}

fe_model_summary <- map_dfr(
  model_map,
  ~ make_model_summary(.x$model_obj, .x$spl, .x$ycol, .x$target)
)

# Standardized-effect block
make_standardized_effects <- function(mod, spl, ycol, target_name, terms_keep) {
  dat <- spl$train |>
    select(player, all_of(c(ycol, terms_keep))) |>
    drop_na()

  coef_tbl <- tidy(mod, vcov = ~player + month) |>
    filter(term %in% terms_keep)

  vc <- vcov(mod, vcov = ~player + month)

  map_dfr(terms_keep, function(v) {
    beta <- coef_tbl$estimate[coef_tbl$term == v]
    idx <- which(names(coef(mod)) == v)
    se_beta <- sqrt(vc[idx, idx])

    x <- dat[[v]]
    is_binary <- all(x %in% c(0,1), na.rm = TRUE)
    scale_amount <- if (is_binary) 1 else sd(x, na.rm = TRUE)

    eff <- beta * scale_amount
    se_eff <- abs(scale_amount) * se_beta

    tibble(
      target = target_name,
      term = v,
      label = coalesce(unname(predictor_labels[v]), v),
      scaling = if (is_binary) "0 to 1 change" else "1 SD increase",
      standardized_effect = eff,
      std.error = se_eff,
      conf.low = eff - 1.96 * se_eff,
      conf.high = eff + 1.96 * se_eff
    )
  })
}

fe_standardized_effects <- map_dfr(
  model_map,
  ~ make_standardized_effects(.x$model_obj, .x$spl, .x$ycol, .x$target, retained_predictors)
)

fe_coef_table
fe_predictor_labels
fe_scaling_notes
fe_contrast_table
fe_reference_values
fe_predictor_distribution
fe_prediction_support
fe_model_summary
fe_standardized_effects
```

```{r Progression FE Coefficient Plot}
# Predictors to keep
coef_terms_keep <- c(
  "avg_rating",
  "avg_moves",
  "opening_diversity",
  "switched_opening"
)

# Display labels
coef_term_labels <- c(
  avg_rating = "Current Average Rating",
  avg_moves = "Average Moves",
  opening_diversity = "Opening Diversity",
  switched_opening = "Switched Most-Played Opening"
)

# Extract retained FE estimates for each target
extract_prog_coef <- function(mod, target_label) {
  tidy(mod, conf.int = TRUE, vcov = ~player + month) |>
    filter(term %in% coef_terms_keep) |>
    mutate(
      target = target_label,
      predictor = recode(term, !!!coef_term_labels)
    ) |>
    select(target, term, predictor, estimate, conf.low, conf.high, std.error, p.value)
}

coef_plot_df <- bind_rows(
  extract_prog_coef(fe_1m$model,  "1-Month Change"),
  extract_prog_coef(fe_3m$model,  "3-Month Change"),
  extract_prog_coef(fe_abs$model, "Absolute 1-Month Change")
) |>
  mutate(
    target = factor(
      target,
      levels = c("1-Month Change", "3-Month Change", "Absolute 1-Month Change")
    ),
    predictor = factor(
      predictor,
      levels = rev(c(
        "Current Average Rating",
        "Average Moves",
        "Opening Diversity",
        "Switched Most-Played Opening"
      ))
    )
  )

p_player_coef <- ggplot(
  coef_plot_df,
  aes(x = estimate, y = predictor, xmin = conf.low, xmax = conf.high, color = target)
) +
  geom_vline(xintercept = 0, linetype = "dashed", linewidth = 0.4, color = "grey50") +
  geom_errorbarh(
    position = position_dodge(width = 0.6),
    height = 0.18,
    linewidth = 0.7
  ) +
  geom_point(
    position = position_dodge(width = 0.6),
    size = 2.4
  ) +
  labs(
    x = "Fixed-Effects Estimate",
    y = NULL,
    color = NULL
  ) +
  theme_few() +
  theme(
    legend.position = "right",
    panel.grid.minor = element_blank()
  ) +
  scale_x_continuous(
    limits = c(-5, 5),
    breaks = seq(-5, 5, by = 2.5)
  ) +
  scale_color_fivethirtyeight() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )

p_player_coef

saveRDS(p_player_coef, "../files_for_submission/Paper_files/figure-rds/p_player_coef.rds")
```

```{r Appendix Progression FE Binned Observed-vs-Predicted}
fe_test_base <- spl_1m$test |>
  select(player, month, d_rating_1m, all_of(base_feature_cols)) |>
  drop_na()

fe_test_plot <- fe_test_base |>
  mutate(
    pred = as.numeric(predict(fe_1m$model, newdata = fe_test_base))
  ) |>
  filter(!is.na(pred), !is.na(d_rating_1m))

fe_binned <- fe_test_plot |>
  mutate(
    pred_bin = ntile(pred, 10)
  ) |>
  group_by(pred_bin) |>
  summarise(
    n = n(),
    pred_lo = min(pred, na.rm = TRUE),
    pred_hi = max(pred, na.rm = TRUE),
    mean_pred = mean(pred, na.rm = TRUE),
    mean_obs  = mean(d_rating_1m, na.rm = TRUE),
    sd_obs    = sd(d_rating_1m, na.rm = TRUE),
    se_obs    = sd_obs / sqrt(n),
    ci_lo     = mean_obs - 1.96 * se_obs,
    ci_hi     = mean_obs + 1.96 * se_obs,
    .groups = "drop"
  )

axis_rng <- range(
  c(fe_binned$pred_lo, fe_binned$pred_hi, fe_binned$ci_lo, fe_binned$ci_hi),
  na.rm = TRUE
)

pad <- 0.04 * diff(axis_rng)
x_rng <- c(axis_rng[1] - pad, axis_rng[2] + pad)
y_rng <- c(axis_rng[1] - pad, axis_rng[2] + pad)

p_prog_fe_binned <- ggplot(fe_binned, aes(x = mean_pred, y = mean_obs)) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "grey50") +
  geom_errorbar(aes(ymin = ci_lo, ymax = ci_hi), width = 0, linewidth = 0.6) +
  geom_point(size = 2.4) +
  coord_equal(xlim = x_rng, ylim = y_rng) +
  labs(
    title = "Observed vs. Predicted Binned Plot for \nRetained FE Progression Model",
    subtitle = "Points are Deciles of Predicted 1-Month Rating Change",
    x = "Mean Predicted Rating Change",
    y = "Mean Observed Rating Change"
  ) +
  theme_few() +
  theme(
    panel.grid.major.x = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.major.y = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.minor = element_blank()
  )

p_prog_fe_binned

saveRDS(p_prog_fe_binned, "../files_for_submission/Paper_files/figure-rds/p_prog_fe_binned.rds")
```


#### Player Categorization

```{r Categorization Prep}
rating_bin_width <- 200 # Keep consistent with earlier sections/EDA
min_games_pm_tc <- 10 # Reliability filter for pm_tc rows
min_open_games <- 10 # Reliability filter for repertoire summaries (per player-month-tc)
min_moves_ptm <- 40 # Used only for overlay profiling, not for clustering inclusion

# Player-month-tc outcomes
pmc <- pm_tc |>
  filter(
    !is.na(player), !is.na(month), !is.na(tc_bucket),
    !is.na(avg_rating), !is.na(games)
  ) |>
  mutate(
    month_date = ymd(paste0(month, "-01")),
    t = as.integer(dense_rank(month_date)),
    rating_bin = as.integer(floor(avg_rating / rating_bin_width) * rating_bin_width)
  ) |>
  filter(games >= min_games_pm_tc)

# Repertoire features from player-opening-month-tc
pom_clean <- pom_tc |>
  filter(
    !is.na(player), !is.na(month), !is.na(tc_bucket),
    !is.na(opening), !is.na(games)
  ) |>
  mutate(
    month_date = ymd(paste0(month, "-01"))
  )

# Total opening games per player-month-tc
pom_tot <- pom_clean |>
  group_by(player, month, tc_bucket) |>
  summarise(open_games_total = sum(games, na.rm = TRUE), .groups = "drop")

# Top opening share per player-month-tc
pom_top_share <- pom_clean |>
  group_by(player, month, tc_bucket) |>
  summarise(
    top_opening_share_tc = max(games, na.rm = TRUE) / sum(games, na.rm = TRUE),
    .groups = "drop"
  )

# Top opening name
pom_top_name <- pom_clean |>
  group_by(player, month, tc_bucket) |>
  slice_max(order_by = games, n = 1, with_ties = FALSE) |>
  ungroup() |>
  select(player, month, tc_bucket, top_opening_name_tc = opening) |>
  mutate(month_date = ymd(paste0(month, "-01"))) |>
  arrange(player, tc_bucket, month_date) |>
  group_by(player, tc_bucket) |>
  mutate(
    prev_top_opening_tc = lag(top_opening_name_tc),
    switched_opening_tc = if_else(
      !is.na(prev_top_opening_tc) & top_opening_name_tc != prev_top_opening_tc,
      1L, 0L
    )
  ) |>
  ungroup() |>
  select(player, month, tc_bucket, switched_opening_tc)

rep_feat <- pom_tot |>
  left_join(pom_top_share, by = c("player","month","tc_bucket")) |>
  left_join(pom_top_name,  by = c("player","month","tc_bucket")) |>
  filter(open_games_total >= min_open_games)

# Core categorization table
cat_tbl <- pmc |>
  left_join(rep_feat, by = c("player","month","tc_bucket"))

# Time-use overlay table
ptm_overlay <- ptm_tc |>
  filter(
    !is.na(player), !is.na(month), !is.na(tc_bucket),
    !is.na(move_rows),
    move_rows >= min_moves_ptm
  ) |>
  select(
    player, month, tc_bucket, move_rows,
    avg_time_spent, median_time_spent, p90_time_spent,
    avg_clock_after, low_time_rate,
    avg_time_opening, avg_time_middlegame, avg_time_endgame
  )

cat_tbl_time_overlay <- cat_tbl |>
  left_join(ptm_overlay, by = c("player","month","tc_bucket"))

# Sanity checks
cat_sanity <- cat_tbl |>
  summarise(
    n = n(),
    n_players = n_distinct(player),
    n_months = n_distinct(month),
    tc_levels = n_distinct(tc_bucket),
    frac_with_rep = mean(!is.na(top_opening_share_tc)),
    .groups = "drop"
  )

cat_sanity

time_cover <- cat_tbl_time_overlay |>
  summarise(
    n = n(),
    frac_with_time = mean(!is.na(avg_time_spent)),
    .groups = "drop"
  )

time_cover
```

```{r Player Categorization Core Features}
core_vars <- c(
  "avg_rating", "avg_rating_diff", "avg_moves",
  "win_rate", "draw_rate",
  "games", "opening_diversity"
)

# Repertoire proxies
rep_vars <- c(
  "top_opening_share_tc", "switched_opening_tc", "open_games_total"
)

use_vars <- intersect(c(core_vars, rep_vars), names(cat_tbl))

# Keep a modeling frame
cat_core <- cat_tbl |>
  select(player, month, month_date, t, tc_bucket, rating_bin, all_of(use_vars)) |>
  drop_na()

# Stabilize heavy-tailed counts
cat_core <- cat_core |>
  mutate(
    log_games = log1p(games),
    log_open_games = if ("open_games_total" %in% names(cat_core)) log1p(open_games_total) else NA_real_
  )

# Update feature list
feature_cols <- setdiff(use_vars, c("games", "open_games_total"))
feature_cols <- c(
  feature_cols,
  "log_games",
  if ("open_games_total" %in% use_vars) "log_open_games" else NULL
)

feature_cols <- intersect(feature_cols, names(cat_core))

# Matrix for modeling
X_raw <- cat_core |>
  select(all_of(feature_cols))

# Standardize
X_scaled <- scale(as.matrix(X_raw))

dim(cat_core)
length(feature_cols)
```

```{r Player Categorization PCA}
pca <- prcomp(X_scaled, center = FALSE, scale. = FALSE)

# Variance explained
pve <- (pca$sdev^2) / sum(pca$sdev^2)

pve_tbl <- tibble(
  pc = paste0("PC", seq_along(pve)),
  pve = pve,
  cum_pve = cumsum(pve)
)

pve_tbl |> slice(1:10)

# Loadings
loadings <- as.data.frame(pca$rotation) |>
  rownames_to_column("feature")

# Top contributors to first few PCs
top_loadings <- loadings |>
  pivot_longer(-feature, names_to = "pc", values_to = "loading") |>
  filter(pc %in% c("PC1","PC2","PC3")) |>
  mutate(abs_loading = abs(loading)) |>
  arrange(pc, desc(abs_loading)) |>
  group_by(pc) |>
  slice_head(n = 10) |>
  ungroup()

top_loadings
```


```{r Player Categorization PCA Scatter}
# Add PCA scores
scores <- as.data.frame(pca$x[, 1:3, drop = FALSE]) |>
  mutate(row_id = row_number())

cat_plot <- cat_core |>
  mutate(row_id = row_number()) |>
  left_join(scores, by = "row_id")

ggplot(cat_plot, aes(x = PC1, y = PC2)) +
  geom_point(alpha = 0.15) +
  facet_wrap(~tc_bucket) +
  labs(title = "PCA projection of player-month-tc behavior (no time-use)",
       subtitle = "Each point is a player-month-tc observation; features standardized",
       x = "PC1", y = "PC2") +
  theme_minimal()
```

```{r Player Categorization UMAP}
set.seed(100)

um <- umap(
  X_scaled,
  n_neighbors = 30,
  min_dist = 0.15,
  metric = "euclidean",
  verbose = TRUE
)

um_df <- tibble(
  UMAP1 = um[,1],
  UMAP2 = um[,2]
) |>
  bind_cols(cat_core |> select(player, month, tc_bucket, rating_bin))

ggplot(um_df, aes(x = UMAP1, y = UMAP2)) +
  geom_point(alpha = 0.12) +
  facet_wrap(~tc_bucket) +
  labs(title = "UMAP projection of player-month-tc behavior (no time-use)") +
  theme_minimal()
```

```{r Player Categorization Primary KMeans}
set.seed(100)

k_grid <- 2:10

# Fit KMeans_rcpp across a grid of k
km_fits <- map(k_grid, ~ KMeans_rcpp(
  X_scaled,
  clusters = .x,
  num_init = 50,
  max_iters = 100,
  initializer = "kmeans++"
))

# Elbow diagnostic
wss <- map_dbl(km_fits, ~ sum(.x$withinss))
wss_tbl <- tibble(k = k_grid, tot_withinss = wss)

ggplot(wss_tbl, aes(x = k, y = tot_withinss)) +
  geom_line() +
  geom_point() +
  labs(title = "Primary clustering spec: k-means elbow curve", x = "k", y = "Total within-cluster SS") +
  theme_minimal()

# Primary descriptive clustering choice
k_choose <- 6
km <- km_fits[[which(k_grid == k_choose)]]

cat_core_km <- cat_core |>
  mutate(cluster_km = factor(km$clusters))

um_km <- um_df |>
  mutate(cluster_km = cat_core_km$cluster_km)

ggplot(um_km, aes(UMAP1, UMAP2, color = cluster_km)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~tc_bucket) +
  labs(
    title = paste0("Primary clustering spec: k-means clusters (k = ", k_choose, ")"),
    subtitle = "Primary descriptive solution used for profiling and robustness checks"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r Player Categorization HClust}
set.seed(100)
n_hc <- 5000
idx_hc <- if (nrow(X_scaled) > n_hc) sample(seq_len(nrow(X_scaled)), n_hc) else seq_len(nrow(X_scaled))

X_h <- X_scaled[idx_hc, , drop = FALSE]

# Distance and linkage
d <- dist(X_h)
hc <- hclust(d, method = "ward.D2")

plot(hc, labels = FALSE, main = "Hierarchical clustering dendrogram (sampled)")

# Cut into k clusters
hc_k <- cutree(hc, k = k_choose)

hc_df <- um_df[idx_hc, ] |>
  mutate(cluster_hc = factor(hc_k))

ggplot(hc_df, aes(UMAP1, UMAP2, color = cluster_hc)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~tc_bucket) +
  labs(title = paste0("UMAP colored by hierarchical clusters (k = ", k_choose, ", sampled)")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r Player Categorization GMM}
set.seed(100)
n_gmm <- 15000
idx_gmm <- if (nrow(X_scaled) > n_gmm) sample(seq_len(nrow(X_scaled)), n_gmm) else seq_len(nrow(X_scaled))

X_g <- X_scaled[idx_gmm, , drop = FALSE]

gmm <- Mclust(X_g, G = 2:10, verbose = FALSE)

summary(gmm)

# BIC plot to choose number of components
plot(gmm, what = "BIC")

gmm_df <- um_df[idx_gmm, ] |>
  mutate(cluster_gmm = factor(gmm$classification))

ggplot(gmm_df, aes(UMAP1, UMAP2, color = cluster_gmm)) +
  geom_point(alpha = 0.25) +
  facet_wrap(~tc_bucket) +
  labs(title = "UMAP colored by GMM clusters (sampled)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r Player Categorization Primary Profiles}
# Primary descriptive cluster profiles
prof_km <- cat_core_km |>
  group_by(cluster_km) |>
  summarise(
    n = n(),
    share = n / nrow(cat_core_km),
    across(all_of(feature_cols), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(desc(n))

prof_km

cluster_mix <- cat_core_km |>
  count(tc_bucket, cluster_km) |>
  group_by(tc_bucket) |>
  mutate(share = n / sum(n)) |>
  ungroup()

ggplot(cluster_mix, aes(x = cluster_km, y = share, fill = tc_bucket)) +
  geom_col(position = "dodge") +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Primary clustering spec: cluster prevalence by time control",
    x = "Cluster",
    y = "Share of observations"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r Player Categorization Stability Jaccard}
# Ensure ordering
clust_assign <- cat_core_km |>
  select(player, tc_bucket, month, month_date, cluster_km) |>
  arrange(player, tc_bucket, month_date)

# Build consecutive-month pairs per player/tc
pairs <- clust_assign |>
  group_by(player, tc_bucket) |>
  mutate(
    next_month      = lead(month),
    next_month_date = lead(month_date),
    next_cluster    = lead(cluster_km),
    # Consecutive month check: difference in months == 1
    month_index     = year(month_date) * 12L + month(month_date),
    next_month_index= year(next_month_date) * 12L + month(next_month_date)
  ) |>
  ungroup() |>
  filter(
    !is.na(next_month_date),
    (next_month_index - month_index) == 1L
  ) |>
  select(-month_index, -next_month_index)

# Month-level stability
stability_month <- pairs |>
  group_by(tc_bucket, month) |>
  summarise(
    n_pairs = n(),
    frac_same_cluster = mean(cluster_km == next_cluster),
    .groups = "drop"
  )

ggplot(stability_month, aes(x = month, y = frac_same_cluster, color = tc_bucket, group = tc_bucket)) +
  geom_line() +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(title = "Month-to-month stability of profile assignments",
     x = "Month",
     y = "Share with same profile next month"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1),
        legend.position = "bottom")

stability_month |>
  arrange(tc_bucket, month) |>
  slice_head(n = 10)
```

##### Diagnostics

```{r Categorization KMeans Diagnostics}
# Silhouette for chosen k
set.seed(100)
n_sil <- 8000
idx_sil <- if (nrow(X_scaled) > n_sil) sample(seq_len(nrow(X_scaled)), n_sil) else seq_len(nrow(X_scaled))
X_s <- X_scaled[idx_sil, , drop = FALSE]
cl_s <- as.integer(cat_core_km$cluster_km[idx_sil])

d <- dist(X_s)
sil <- silhouette(cl_s, d)

summary(sil[, "sil_width"])
plot(sil, main = paste0("Silhouette plot (k = ", k_choose, ", sampled)"), border = NA)
```

```{r Categorization KMeans Init Stability}
# Refit same k multiple times and measure agreement via adjusted Rand index (ARI)
set.seed(100)
n_rep <- 10
km_reps <- map(1:n_rep, ~ KMeans_rcpp(
  X_scaled,
  clusters = k_choose,
  num_init = 10,
  max_iters = 100,
  initializer = "kmeans++"
))

# Compare to first run
base_lab <- km_reps[[1]]$clusters
ari_tbl <- map_dfr(1:n_rep, function(i) {
  lab <- km_reps[[i]]$clusters
  tibble(rep = i, ARI = adjustedRandIndex(base_lab, lab))
})

ari_tbl
ggplot(ari_tbl, aes(x = rep, y = ARI)) + geom_line(aes(group=1)) + geom_point() +
  labs(title = "K-means initialization stability (ARI)", x = "replicate", y = "ARI vs first run") +
  theme_minimal()
```


```{r Categorization Month-to-Month ARI}
assign <- cat_core_km |>
  select(player, tc_bucket, month, month_date, cluster_km) |>
  mutate(month_index = year(month_date) * 12L + month(month_date))

# Build consecutive month pairs by tc_bucket
ari_month <- assign |>
  inner_join(
    assign |> transmute(player, tc_bucket, month_index = month_index - 1L, cluster_prev = cluster_km),
    by = c("player","tc_bucket","month_index")
  ) |>
  group_by(tc_bucket, month_index) |>
  summarise(
    n_overlap = n(),
    ari = adjustedRandIndex(as.integer(cluster_prev), as.integer(cluster_km)),
    .groups = "drop"
  )

ari_month |> head(12)

ggplot(ari_month, aes(x = month_index, y = ari, color = tc_bucket, group = tc_bucket)) +
  geom_line() +
  geom_point(size = 1) +
  labs(title = "Month-to-month clustering stability (ARI on overlapping players)",
       x = "Month index", y = "ARI") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


##### Summary

```{r Categorization Summary Helpers}
make_cat_design <- function(data, vars_core, vars_rep = NULL) {
  use_vars_local <- intersect(c(vars_core, vars_rep), names(data))

  out <- data |>
    select(player, month, month_date, t, tc_bucket, rating_bin, all_of(use_vars_local)) |>
    drop_na()

  out <- out |>
    mutate(
      log_games = if ("games" %in% names(out)) log1p(games) else NA_real_,
      log_open_games = if ("open_games_total" %in% names(out)) log1p(open_games_total) else NA_real_
    )

  feature_cols_local <- setdiff(use_vars_local, c("games", "open_games_total"))
  if ("games" %in% use_vars_local) feature_cols_local <- c(feature_cols_local, "log_games")
  if ("open_games_total" %in% use_vars_local) feature_cols_local <- c(feature_cols_local, "log_open_games")

  feature_cols_local <- intersect(feature_cols_local, names(out))

  X_raw_local <- out |>
    select(all_of(feature_cols_local))

  X_scaled_local <- scale(as.matrix(X_raw_local))

  list(
    data = out,
    feature_cols = feature_cols_local,
    X_scaled = X_scaled_local
  )
}

fit_kmeans_spec <- function(X_scaled, k, seed = 100) {
  set.seed(seed)
  KMeans_rcpp(
    X_scaled,
    clusters = k,
    num_init = 50,
    max_iters = 100,
    initializer = "kmeans++"
  )
}

cluster_profile_tbl <- function(data, cluster_var, feature_cols) {
  data |>
    group_by(.data[[cluster_var]]) |>
    summarise(
      n = n(),
      share = n / nrow(data),
      across(all_of(feature_cols), ~ mean(.x, na.rm = TRUE)),
      .groups = "drop"
    ) |>
    arrange(desc(n))
}
```

```{r Categorization Phase3 Feature-Set Robustness}
core_only_vars <- core_vars
core_rep_vars  <- c(core_vars, rep_vars)

spec_core_only <- make_cat_design(cat_tbl, vars_core = core_only_vars)
spec_core_rep  <- make_cat_design(cat_tbl, vars_core = core_vars, vars_rep = rep_vars)

km_core_only <- fit_kmeans_spec(spec_core_only$X_scaled, k = k_choose, seed = 100)
km_core_rep  <- fit_kmeans_spec(spec_core_rep$X_scaled,  k = k_choose, seed = 100)

cat_core_only_km <- spec_core_only$data |>
  mutate(cluster_km = factor(km_core_only$clusters))

cat_core_rep_km <- spec_core_rep$data |>
  mutate(cluster_km = factor(km_core_rep$clusters))

feature_robust_tbl <- tibble(
  feature_set = c("core_only", "core_plus_repertoire"),
  n_obs = c(nrow(spec_core_only$data), nrow(spec_core_rep$data)),
  n_features = c(length(spec_core_only$feature_cols), length(spec_core_rep$feature_cols)),
  min_cluster_share = c(
    min(prop.table(table(cat_core_only_km$cluster_km))),
    min(prop.table(table(cat_core_rep_km$cluster_km)))
  )
)

feature_robust_tbl

prof_core_only <- cluster_profile_tbl(cat_core_only_km, "cluster_km", spec_core_only$feature_cols) |>
  mutate(feature_set = "core_only", .before = 1)

prof_core_rep <- cluster_profile_tbl(cat_core_rep_km, "cluster_km", spec_core_rep$feature_cols) |>
  mutate(feature_set = "core_plus_repertoire", .before = 1)

prof_core_only
prof_core_rep
```

```{r Categorization Algorithm Robustness}
# k-means assignments
km_assign <- cat_core_km |>
  mutate(row_id = row_number()) |>
  select(row_id, cluster_km)

# Hierarchical assignments
hc_assign <- tibble(
  row_id = idx_hc,
  cluster_hc = factor(hc_k)
)

# GMM assignments
gmm_assign <- tibble(
  row_id = idx_gmm,
  cluster_gmm = factor(gmm$classification)
)

alg_agree_hc <- km_assign |>
  inner_join(hc_assign, by = "row_id") |>
  summarise(
    n_overlap = n(),
    ari = adjustedRandIndex(as.integer(cluster_km), as.integer(cluster_hc))
  ) |>
  mutate(comparison = "kmeans_vs_hclust", .before = 1)

alg_agree_gmm <- km_assign |>
  inner_join(gmm_assign, by = "row_id") |>
  summarise(
    n_overlap = n(),
    ari = adjustedRandIndex(as.integer(cluster_km), as.integer(cluster_gmm))
  ) |>
  mutate(comparison = "kmeans_vs_gmm", .before = 1)

alg_agreement_tbl <- bind_rows(alg_agree_hc, alg_agree_gmm)
alg_agreement_tbl
```

```{r Categorization Stability Summary}
# Consolidate the main descriptive stability checks for the primary clustering solution

sil_summary <- tibble(
  metric = c("mean_silhouette", "median_silhouette"),
  value = c(
    mean(sil[, "sil_width"], na.rm = TRUE),
    median(sil[, "sil_width"], na.rm = TRUE)
  )
)

init_stability_summary <- ari_tbl |>
  summarise(
    mean_init_ari = mean(ARI, na.rm = TRUE),
    min_init_ari = min(ARI, na.rm = TRUE),
    max_init_ari = max(ARI, na.rm = TRUE)
  )

month_stability_summary <- ari_month |>
  summarise(
    mean_month_ari = mean(ari, na.rm = TRUE),
    median_month_ari = median(ari, na.rm = TRUE),
    min_month_ari = min(ari, na.rm = TRUE)
  )

cluster_size_summary <- cat_core_km |>
  count(cluster_km) |>
  mutate(share = n / sum(n)) |>
  summarise(
    n_clusters = n(),
    min_cluster_share = min(share),
    max_cluster_share = max(share)
  )

cat_phase3_stability <- bind_cols(
  init_stability_summary,
  month_stability_summary,
  cluster_size_summary
)

sil_summary
cat_phase3_stability
```


```{r Categorization Claim Summary}
cat_claim_summary <- list(
  primary_spec = tibble(
    item = c("algorithm", "k", "feature_set", "n_obs"),
    value = c(
      "kmeans",
      as.character(k_choose),
      "core_plus_repertoire",
      as.character(nrow(cat_core_km))
    )
  ),

  primary_cluster_sizes = cat_core_km |>
    count(cluster_km) |>
    mutate(share = n / sum(n)) |>
    arrange(desc(n)),

  primary_profiles = prof_km,

  feature_robustness = feature_robust_tbl,

  algorithm_agreement = alg_agreement_tbl,

  stability = cat_phase3_stability
)

cat_claim_summary$primary_spec
cat_claim_summary$primary_cluster_sizes
cat_claim_summary$feature_robustness
cat_claim_summary$algorithm_agreement
cat_claim_summary$stability
```

```{r Categorization Model Roles Summary}
cat("
Player Categorization — Phase 3 roles

Primary descriptive clustering solution:
- cat_core_km
- k-means with k = 6
- standardized core + repertoire features

Phase 3 robustness checks:
- feature-set robustness
- algorithm agreement (k-means vs hierarchical vs GMM)
- initialization stability
- month-to-month stability

Interpretation rule:
- describe only cluster patterns that are reasonably stable across these checks
- treat unstable or tiny clusters as exploratory
")

cat("
Player Categorization — Phase 3 summary

Primary descriptive segmentation:
- k-means on core_plus_repertoire features
- k = 6

Robustness checks:
- feature-set robustness: core_only vs core_plus_repertoire
- algorithm robustness: k-means vs hierarchical vs GMM
- stability: silhouette, initialization ARI, month-to-month ARI

Interpretation:
- descriptive behavioral profiles, not inferential latent classes
")
```
```{r Categorization Paper Output Extraction}
# Standardized cluster centroids for retained features
X_scaled_df <- as.data.frame(X_scaled) |>
  as_tibble() |>
  setNames(feature_cols)

cluster_centroids_std <- bind_cols(
  cat_core_km |> select(cluster_km),
  X_scaled_df
) |>
  group_by(cluster_km) |>
  summarise(
    n = n(),
    across(all_of(feature_cols), ~ mean(.x, na.rm = TRUE)),
    .groups = "drop"
  ) |>
  arrange(cluster_km)

cluster_centroids_std_long <- cluster_centroids_std |>
  pivot_longer(
    cols = all_of(feature_cols),
    names_to = "feature",
    values_to = "standardized_centroid"
  )


# Cluster membership counts by time control
cluster_counts_time_control <- cat_core_km |>
  count(tc_bucket, cluster_km, name = "n") |>
  group_by(tc_bucket) |>
  mutate(share_within_tc = n / sum(n)) |>
  ungroup() |>
  arrange(tc_bucket, cluster_km)

# Cluster membership counts by broad rating bin
cluster_counts_rating_bin <- cat_core_km |>
  count(rating_bin, cluster_km, name = "n") |>
  group_by(rating_bin) |>
  mutate(share_within_rating_bin = n / sum(n)) |>
  ungroup() |>
  arrange(rating_bin, cluster_km)


# Month-to-month same-cluster retention rate by cluster
cluster_retention_by_cluster <- pairs |>
  group_by(cluster_km) |>
  summarise(
    n_pairs = n(),
    same_cluster_rate = mean(cluster_km == next_cluster, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(cluster_km)

cluster_retention_by_cluster_tc <- pairs |>
  group_by(tc_bucket, cluster_km) |>
  summarise(
    n_pairs = n(),
    same_cluster_rate = mean(cluster_km == next_cluster, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(tc_bucket, cluster_km)

# Most common transition destination by cluster

cluster_transition_counts <- pairs |>
  count(cluster_km, next_cluster, name = "n_transitions") |>
  group_by(cluster_km) |>
  mutate(
    total_from_cluster = sum(n_transitions),
    transition_share = n_transitions / total_from_cluster
  ) |>
  ungroup() |>
  arrange(cluster_km, desc(n_transitions))

cluster_top_transition <- pairs |>
  filter(cluster_km != next_cluster) |>
  count(cluster_km, next_cluster, name = "n_transitions") |>
  group_by(cluster_km) |>
  mutate(
    total_switches_from_cluster = sum(n_transitions),
    switch_share = n_transitions / total_switches_from_cluster
  ) |>
  slice_max(order_by = n_transitions, n = 1, with_ties = FALSE) |>
  ungroup() |>
  arrange(cluster_km)

categorization_output_summary <- list(
  standardized_centroids = cluster_centroids_std,
  cluster_counts_time_control = cluster_counts_time_control,
  cluster_counts_rating_bin = cluster_counts_rating_bin,
  cluster_retention_by_cluster = cluster_retention_by_cluster,
  cluster_top_transition = cluster_top_transition
)

cluster_centroids_std
cluster_counts_time_control
cluster_counts_rating_bin
cluster_retention_by_cluster
cluster_top_transition
```

```{r Appendix Clustering Feature Construction Table}
clustering_feature_tbl <- tribble(
  ~final_feature,           ~source_table, ~derived_construction,                                                                 ~exact_transform, ~final_standardization,
  "avg_rating",             "pm_tc",       "Player-month-time-control average rating",                                            "None",           "z-score via scale(as.matrix(X_raw))",
  "avg_rating_diff",        "pm_tc",       "Player-month-time-control average rating differential",                               "None",           "z-score via scale(as.matrix(X_raw))",
  "avg_moves",              "pm_tc",       "Player-month-time-control average moves",                                             "None",           "z-score via scale(as.matrix(X_raw))",
  "win_rate",               "pm_tc",       "Player-month-time-control win rate",                                                  "None",           "z-score via scale(as.matrix(X_raw))",
  "draw_rate",              "pm_tc",       "Player-month-time-control draw rate",                                                 "None",           "z-score via scale(as.matrix(X_raw))",
  "opening_diversity",      "pm_tc",       "Player-month-time-control opening-diversity summary already present in pm_tc",        "None",           "z-score via scale(as.matrix(X_raw))",
  "top_opening_share_tc",   "pom_tc",      "max(games) / sum(games) within player × month × tc_bucket",                          "None",           "z-score via scale(as.matrix(X_raw))",
  "switched_opening_tc",    "pom_tc",      "1 if top opening differs from previous month within player × tc_bucket; else 0",     "None",           "z-score via scale(as.matrix(X_raw))",
  "log_games",              "pm_tc",       "games from pm_tc",                                                                    "log1p(games)",   "z-score via scale(as.matrix(X_raw))",
  "log_open_games",         "pom_tc",      "open_games_total = sum(games) within player × month × tc_bucket",                    "log1p(open_games_total)", "z-score via scale(as.matrix(X_raw))"
)

kable(
  clustering_feature_tbl,
  caption = "Final clustering feature set, source, derivation, transform, and standardization for the retained player-categorization solution."
)
```

```{r Appendix Cluster Transition Heatmap}
transition_pairs <- cat_core_km |>
  select(player, tc_bucket, month_date, cluster_km) |>
  arrange(player, tc_bucket, month_date) |>
  group_by(player, tc_bucket) |>
  mutate(
    next_month_date = lead(month_date),
    next_cluster = lead(cluster_km),
    month_index = year(month_date) * 12L + month(month_date),
    next_month_index = year(next_month_date) * 12L + month(next_month_date)
  ) |>
  ungroup() |>
  filter(
    !is.na(next_cluster),
    (next_month_index - month_index) == 1L
  )

transition_heat <- transition_pairs |>
  count(
    cluster_from = cluster_km,
    cluster_to = next_cluster,
    name = "n"
  ) |>
  group_by(cluster_from) |>
  mutate(
    p = n / sum(n)
  ) |>
  ungroup() |>
  mutate(
    cluster_from = factor(cluster_from, levels = levels(cat_core_km$cluster_km)),
    cluster_to   = factor(cluster_to,   levels = levels(cat_core_km$cluster_km))
  )

p_transition_heat <- ggplot(
  transition_heat,
  aes(x = cluster_to, y = cluster_from, fill = p)
) +
  geom_tile(color = "white", linewidth = 0.4) +
  geom_text(aes(label = percent(p, accuracy = 0.1),
                color = if_else(p >= 0.35, "grey95", "grey20")), size = 3) +
  scale_color_identity() +
  scale_fill_distiller(
    palette = "Blues",
    direction = 1,
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Month-to-Month Transition Matrix for Retained Clusters",
    subtitle = "Rows Sum to 100%; Transitions Computed Within Player and Time Control for Consecutive Months",
    x = "Next-Month Cluster",
    y = "Current-Month Cluster",
    fill = "Row Share"
  ) +
  theme_minimal() +
  theme(
    panel.grid = element_blank()
  )

p_transition_heat

saveRDS(p_transition_heat, "../files_for_submission/Paper_files/figure-rds/p_transition_heat.rds")
```


```{r Appendix PCA-UMAP Cluster Structure}
set.seed(100)

n_plot <- min(25000, nrow(cat_core_km))
idx_plot <- sample(seq_len(nrow(cat_core_km)), n_plot)

pca_cluster_df <- tibble(
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2],
  cluster_km = cat_core_km$cluster_km
) |>
  slice(idx_plot)

umap_cluster_df <- um_df |>
  mutate(cluster_km = cat_core_km$cluster_km) |>
  slice(idx_plot)

p_pca_clusters <- ggplot(
  pca_cluster_df,
  aes(x = PC1, y = PC2, color = cluster_km)
) +
  geom_point(alpha = 0.22, size = 0.7) +
  labs(
    title = "PCA",
    x = "PC1",
    y = "PC2",
    color = "Cluster"
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  theme_few() +
  theme(legend.position = "bottom")

p_umap_clusters <- ggplot(
  umap_cluster_df,
  aes(x = UMAP1, y = UMAP2, color = cluster_km)
) +
  geom_point(alpha = 0.22, size = 0.7) +
  labs(
    title = "UMAP",
    x = "UMAP1",
    y = "UMAP2",
    color = "Cluster"
  ) +
  guides(color = guide_legend(override.aes = list(alpha = 1, size = 2.5))) +
  theme_few() +
  theme(legend.position = "bottom")

p_cluster_structure <- (p_pca_clusters + p_umap_clusters) +
  plot_layout(guides = "collect") +
  plot_annotation(
    theme = theme(legend.position = "bottom")
  )

p_cluster_structure

saveRDS(p_cluster_structure, "../files_for_submission/Paper_files/figure-rds/p_cluster_structure.rds")
```



















